### 2023-09-19 ms/Methods.Rmd
reconstituting most of the repo

---
title: "Methods"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_knit$set(root.dir = "..")
```

```{r}
# source header
source("r/header.R", keep.source = TRUE)

# source superseded
source("r/superseded.R", keep.source = TRUE)
```

# Custom Functions

Below are the custom functions to deal with 1.) single surface data, and 2.) Amphistomatous data. The general work flow is to simulate random and dispersed stomatal grids based on a Gamma distribution of the number of stomata in the real data for use in testing the null hypothesis that stomata are randomly dispersed on the leaf surface using the Average Nearest Neighbor test. For amphistomatous data, the work flow is to first pair associated surfaces, then build kernel densities (or distance squared) of the stomata on each surface and perform a pearson pairwise raster correlation on the paired kernel density rasters. And use simulated data to test the null hypothesis that stomata on the abaxial and adaxial surface are independently distributed.

## Single Surface Data

### Density to Distance

```{r Density to Distance}
D2U
```

Explanation: `D2U` takes a stomatal density value ($D_S$) in units # of stomata / $pixels^2$ and returns a distance value ($U$) which describes the distance between stomata with density ($D_S$) assuming perfect uniform distribution. It works by calculating the area of a regular hexagon within which each pixel is closest to the stomate at the center of the hexagon. Once the area of the hexagon is known, the function works to calculate the distance between stomata ($U$) assuming regular packing of hexagons of area ($A_\text{hex}$). This is done using known equations of the geometry of regular hexagons.

### Distance to Density

```{r Distance to Density}
U2D
```

Explanation: `U2D` takes a distance measurement ($U$) in units pixels and returns a stomatal density value ($D_S$). This function is the reciprocal of `D2U` and works using the same geometric concepts.

### Building Dispersed (Equal) Stomatal Grids

```{r Building Dispersed (Equal) Stomatal Grids 1}
make_equal_stomata_grid
```

Explanation: The `make_equal_stomata_grid` function takes arguments: distance between stomata ($U$), # of horizontal pixels ($\text{pixels}_x$), # of vertical pixels ($\text{pixels}_y$), a logical (`jitter`), and a logical (`rotate`) and outputs a tibble of two columns (`x`, `y`) with $x, y$ coordinates of dispersed stomata. It works via the geometry of dispersed objects which are always arrayed in an equilateral triangle grid such that the area occupied (closest to) by each stomate is a hexagon of ($A_\text{hex}$) from the previous functions. The function builds a sequence of $x$ values starting at `start_end` separated by $U$ until it reaches $(\text{pixels}_x - \text{start_end})$; then the same sequence is repeated adding $U/2$. This way, two "rows" of $x$ values are calculated together. That whole sequence is then repeated $\text{x_rep}$ number of times depending on the grid size and stomatal density. $y$ is built in a similar manner by sequencing by the proper distance $((\sqrt{3}) / 2) U$ until reaching $2 * \text{x_rep}$ number of repetitions. Then, sorting at the end to match the appropriate $x$ values in vector `x`. Jitter is added (if `jitter = TRUE`) by performing vector addition of a uniform random number between 0 and $U$ for $x$ and between 0 and $\sqrt{3} / 2 U$ for $y$. Rotation about a central point is added if (`rotate == TRUE`) based on a random angle between 0 and 45 degrees (the pattern repeats after 45 degrees of rotation). The function returns a tibble of two columns: `x` and `y`, which represent the theoretical positions of stomata in a perfectly dispersed array of size $\text{pixels}_x$ by $\text{pixels}_y$ with distance between stomata ($U$).

```{r Building Dispersed (Equal) Stomatal Grids 2}
make_equal_stomata_grids
```

Explanation: The function `make_equal_stomata_grids` reruns the previous function `n_grid` # of times; returning a list with `n_grid` # of tibbles.

### Building Random Stomatal Grids

```{r Building Random Stomatal Grids 1}
make_random_stomata_grid
```

Explanation: The function `make_random_stomata_grid` takes arguments: number of stomata (`n`), $\text{pixels}_x$, and $\text{pixels}_y$ and returns a tibble of two columns: `x` and `y`, which represent coordinates of a hypothetical array of randomly distributed stomata within an area ($\text{pixels}_x \times \text{pixels}_y$). It works by converting distance between stomata ($U$) to a density ($D$) and then to an integer equal to # of stomata ($n$); then utilizing the function `runif()` which returns a vector of length $n$ of uniformly distributed random numbers between 0 and $\text{pixels}_x$ or $\text{pixels}_y$ Those two vectors are placed in a `tibble` and the `tibble` is returned as the output of the function.

```{r Building Random Stomatal Grids 2}
make_random_stomata_grids
```

Explanation: The function `make_random_stomata_grids` reruns the previous function `n_grid` # of times; returning `n_grid` # of tibbles.

### Plotting Grids

```{r Plotting Grids, include = T}
plot_stomatal_grid
```

Explanation: The function `plot_stomatal_grid()` is a general stomatal plotting function. It takes arguments: tibble of $x, y$ coordinates of stomata in `data.frame` (`df_stomata`), and # of horizontal pixels ($\text{pixels}_x$) and # of vertical pixels ($\text{pixels}_y$) and returns a plot with a border of the proper dimensions ($\text{pixels}_x \times \text{pixels}_y$).

## Simulate Stomatal Grids Based on Real Data

```{r Simulate Stomatal Grids Based on Real Data, include = T}
simulate_grids_from_data
```

Explanation: The function `simulate_grids_from_data()` takes arguments: tibble of $x, y$ coordinates of stomata (`data`), number of horizontal pixels ($\text{pixels}_x$), number of vertical pixels ($\text{pixels}_y$), and number of simulations (`n_grids`) and returns a nested list of 1.) random stomatal grids (`random_grids`) and 2.) dispersed stomatal grids (`equal_grids`). It works by building a gamma distribution (`lambdas`) around number of stomata (n) assuming a scale of 1. Then simulating random and equal grids based on that vector of lambdas.

## Average Nearest Neighbor

```{r Average Nearest Neighbor Function, include = T}
stomata_ann_test
```

Explanation: The function `stomata_ann_test()` takes the arguments: tibble of $x, y$ coordinates of stomata (`df_stomata`), number of horizontal pixels ($\text{pixels}_x$), number of vertical pixels ($\text{pixels}_y$) and returns a list of values returned by an Average Nearest Neighbor Test via the package `spatialEco`. The average nearest neighbor (ANN) is a commonly used spatial statistic tool to determine if objects are 1) clustered, 2) random, or 3) regularly dispersed in geographic space. It works by calculating the euclidean distance between each combination of points (stomata) on a plane (leaf surface), then determining each point's nearest neighbor (by finding which stomata pair gives the smallest distance value). All the nearest neighbor distances are averaged, giving the average nearest neighbor value (observed mean distance). This distance is compared to an expected mean distance ($\text{expected mean distance} = 0.5 * sqrt(A / # of stomata)$) for a hypothetical set of points (same # of points) randomly distributed points across the same area (A). The nearest neighbor index (`nni`) is the ratio of observed mean distance to the expected mean distance. NNI values range from 0 to around 3. Values close to 0 are more likely to be clustered, values around 1, random, and values greater than 1 are more likely to be regularly dispersed. A p-value can be assigned to the NNI value to test for a non-random distribution (clumped or dispersed); however, in this study, we will calculate a p-value based on simulations rather than a the traditional normal distribution assumed by the package `spatialEco` as the small sample size and study area violate the assumptions of the traditional ANN test as edge effects are not trivial.

```{r Tests on Simulations, include = T}
stomata_ann_tests
```

Explanation: The function `stomata_ann_tests()` takes a nested list of simulations (`list`), number of horizontal pixels ($\text{pixels}_x$), number of vertical pixels ($\text{pixels}_y$), and returns a list of ANN test results. It works by applying the previous function `stomata_ann_test()` to each element in a list via an `lapply` function.

```{r Building a Dataframe, include = T}
list_to_frame
```

Explanation: The function `list_to_frame()` takes the result of `stomata_ann_tests()` as an argument and builds a `data.frame` for more convenient analysis.

```{r p-value and Final Result Generation, include = T}
real_test
```

Explanation: The function `real_test()` takes a tibble of real stomatal $x, y$ coordinates (plus metadata) (`data`), number of horizontal pixels ($\text{pixels}_x$), number of vertical pixels ($\text{pixels}_y$), and number of simulations to perform (`n_grids`) and returns a list of values: p-value based on simulations (`p`), number of stomata (`n_stom`), dispersion index (`dispersed`), treatment (`treatment`), leaf number (`leaf_number`), image number (`image_number`), and surface (`surface`). It works by running simulations based on the `tibble` provided, then performing an Average Nearest Neighbor test on the real data and simulated data. It then calculates a p-value based on the proportion of simulated random grids that received a higher NNI than the real data. A dispersion index (`dispersed`) is calculated by determining the how close the real NNI is to the simulated equal grids. Values closer to 1 are more dispersed, while values close to 0 are randomly dispersed, and negative values are more clumped and would tend to be associated with high p-values.

```{r Histogram of Simulated Average Nearest Neighbor Tests, include = T}
stom_hist
```

Explanation: The function stom_hist() takes outputs from list_to_frame() and real_test() to build a histogram which visualizes the result from a single picture.

## Performing Average Nearest Neighbor For Each Surface

```{r Performing Average Nearest Neighbor For Each Surface 1, include = T}
real_tests
```

Explanation: The function `real_tests()` applies `real_test()` across every item in a list. It takes a list of `tibbles` of real stomata $x, y$ coordinates (`list`), number of horizontal pixels ($\text{pixels}_x$), number of vertical pixels ($\text{pixels}_y$), and number of simulations to perform (`n_grids`) and returns a list of outputs from `real_test()`.

```{r Performing Average Nearest Neighbor For Each Surface 2, include = T}
list_to_frame_final
```

Explanation: The function `list_to_frame_final()` takes the list output from `real_tests()` and modifies it into a `data.frame` for data analysis convenience.

# Amphistomatic Data

These functions are curated for the analysis of two associated leaf surfaces: abaxial and adaxial.

## Kernel Density Raster Generation

```{r Kernel Density Raster Generation 1, include = T}
kerneldensity
```

Explanation: The function `kerneldensity()` takes arguments: `tibble` of $x, y$ coordinates of stomata, number of horizontal pixels ($\text{pixels}_x$), and number of vertical pixels ($\text{pixels}_y$) and returns a `raster` of dimensions ($\text{pixels}_x \times \text{pixels}_y$) with values between 0 and 1 based on the kernel density estimate. It works by utilizing the `sp.kde()` function from the `spatialEco` package, which builds kernel density estimate `rasters` from `SpatialPointDataFrames`.

```{r Kernel Density Raster Generation 2, include = T}
kerneldensities
```

Explanation: The function `kerneldensities()` takes arguments list of stomatal grids (`list`), number of horizontal pixels ($\text{pixels}_x$), and number of vertical pixels ($\text{pixels}_y$) and returns a `list` of kernel density estimate `rasters`.

## Pairing Abaxial and Adaxial Surfaces

```{r Pairing Abaxial and Adaxial Surfaces, include = T}
df_pair
```

Explanation: The function `df_pair()` takes a `list` of `tibbles` and pairs them such that the associated leaf surfaces match by ordering the list (all abaxial surfaces ordered first, followed by all adaxial surfaces, with each sample corresponding in position), splitting the `list` in half, renaming the elements, and pairing them using the `map` function of the `purrr` package. It returns a `list` of lists where each `list` is of length 2; one abaxial surface, and one adaxial surface.

## Simulate Stomatal Grids from Paired Data

```{r Simulate Stomatal Grids from Paired Data, include = T}
simulate_grids_from_pairs
```

Explanation: The function `simulate_grids_from_pairs()` takes a `list` of length 2 of stomatal grids $x, y$, number of horizontal pixels ($\text{pixels}_x$), number of vertical pixels ($\text{pixels}_y$), and number of simulations desired (`n_grids`) and returns a `list` of lists of paired, simulated stomatal grids. It works by building a gamma distribution (`lambdas`) around the number of stomata (`n_stom`) and simulating random grids using the `make_random_stomata_grid` function.

## Pairwise Correlation of Paired Kernel Densities

```{r Pairwise Correlation of Paired Kernel Densities 1, include = T}
corr_pair
```

Explanation: The function `corr_pair()` takes arguments list of associated stomatal grids $x, y$, number of horizontal pixels ($\text{pixels}_x$), and number of vertical pixels ($\text{pixels}_y$) and returns a `list` of correlation coefficients and surface average values. It works by building kernel density estimate `rasters` using the `kerneldensities()` function and stacking the rasters in a `raster::rasterstack` object for input into the `layerStats` function of the `raster` package.

```{r Pairwise Correlation of Paired Kernel Densities 2, include = T}
corr_pairs
```

Explanation: The function `corr_pairs()` takes arugments `list` of `lists` of paired stomatal grids $x, y$, number of horizontal pixels ($\text{pixels}_x$), and number of vertical pixels ($\text{pixels}_y$) and returns a `list` of correlation coefficients and raster averages for each pair of stomatal grids. It works by using the `lapply` function over the `corr_pair` function.

```{r Pairwise Correlation of Paired Kernel Densities 3, include = T}
corr_to_frame
```

Explanation: The function `corr_to_frame()` takes the output from `corr_pairs()` function and returns a `tibble` for easier data analysis.

## Pairwise Correlation of Paired Distance Squared Rasters

```{r Pairwise Correlation of Paired Distance Squared Rasters 1, include = T}
create_raster
```

Explanation: The function `create_raster` takes arguments $\text{pixels}_x$ (pixels_x) and $\text{pixels}_y$ (pixels_y) and creates a blank raster for input into the `dist_sq` function. It utilizes the `raster` function of the `raster` package.

```{r Pairwise Correlation of Paired Distance Squared Rasters 2, include = T}
dist_sq
```

Explanation: The function `dist_sq()` takes arguments `data` (a `tibble` of stomatal positions $x, y$) and `leaf` (an empty `raster` object created by the function `create_raster`) and returns a `raster` of distance squared values (the distance of each cell in the raster to the nearest stomate). It does this by utilizing the `distanceFromPoints` function of the `raster` package and then squaring the result for biological relevance.

```{r Pairwise Correlation of Paired Distance Squared Rasters 3, include = T}
dist_sqs
```

Explanation: The function `dist_sqs` applies the function `dist_sq` over each element in a `list`. It utilizes the `lapply` function.

```{r Pairwise Correlation of Paired Distance Squared Rasters 4, include = T}
dist_sq_corr
```

Explanation: The function `dist_sq_corr` takes paired `dist_sqs` outputs (`list`) (paired by `df_pair` function) and `create_raster` function (`leaf`). It works by stacking rasters and running a pearson pairwise correlation across each cell in the raster using `LayerStats` function from the `raster` package.

```{r Pairwise Correlation of Paired Distance Squared Rasters 5, include = T}
dist_sq_corrs
```

Explanation: The function `dist_sq_corrs` performs `dist_sq_corr` over each element in a `list` using the `lapply` function.

## Hypothesis Testing with Simulated Data

```{r Hypothesis Testing with Simulated Data 1, include = T}
real_corr_test
```

Explanation: The function `real_corr_test()` takes arguments list of associated stomatal grids $x, y$, number of horizontal pixels ($\text{pixels}_x$), number of vertical pixels ($\text{pixels}_y$), number of simulations to run (`n_grids`), and raster type (`dist_or_kern`) and returns a `list` of relevant information including a p-value (`p`), the real correlation coefficient (`corr_coef`), number of stomata on each surface (`n_stom`), and associated metadata. It works by using the `simulate_grids_from_pairs()` function to simulate grids, running `corr_pairs()` function (if `dist_or_kern` == FALSE) or `dist_sq_corrs` function (if `dist_or_kern` == TRUE) on simulation data and `corr_pair()` (or `dist_sq_corr`) on real data, and comparing the results by calculating what proportion of simulated correlation coefficients are greater than the real correlation coefficient (p-value).

```{r Hypothesis Testing with Simulated Data 2, include = T}
real_corr_tests
```

Explanation: The function `real_corr_tests()` takes arguments: `list` of `lists` of associated stomatal grids $x, y$, number of horizontal pixels ($\text{pixels}_x$), number of vertical pixels ($\text{pixels}_y$), number of simulations to run (`n_grids`), distance squared or kernel logical (`dist_or_kern`), and number of cores to use (`num_cores`) and returns a `list` of `lists` of test results. It works by running the `mclapply` function of the `parallel` package over the `real_corr_test()` function.

```{r Hypothesis Testing with Simulated Data 3, include = T}
corr_to_frame_final
```

Explanation: The function `corr_to_frame_final()` takes the output from `real_corr_tests()` and returns a `tibble` with extra columns: `light` for light treatment and `stom_ratio` for stomatal ratio.

## Plot a Histogram of Result

```{r Plot a Histogram of Result, include = T}
amphi_stom_hist
```

Explanation: The function `amphi_stom_hist()` takes the output from the simulations and outputs from the tests to build a histogram and visualize the result.

# Test Analysis of a Single Sample

## Load the Data

```{r Load the Data, include = T}

# set working directory
dir = getwd()
setwd(dir)

# load the raw data
stomata = read.csv("stomata_positionR.csv")
stomata = stomata[1:4720, 1:7]
stomata = na.omit(stomata)
# make factors
stomata$treatment = as.factor(stomata$treatment)
stomata$leaf_number = as.factor(stomata$leaf_number)
stomata$image_number = as.factor(stomata$image_number)
stomata$surface = as.factor(stomata$surface)
# rename columns
stomata = stomata %>% rename(
  x = X,
  y = Y
)
# make numeric
stomata$x = as.numeric(stomata$x)
stomata$y = as.numeric(stomata$y)
# display data
str(stomata)

# split data into a list of tibbles for each surface
factors = list(stomata$treatment, stomata$leaf_number,
               stomata$image_number, stomata$surface)
densities = stomata %>%
  split(factors)
# delete dataframes with less than 1 row
densities = densities[sapply(densities, nrow) > 1]

# show random test data
str(densities[[17]])
str(densities[[83]])

# make into a tibble
tib_density = lapply(densities, as_tibble)

# remove duplicate stomata
tib_density = lapply(tib_density, function(x) x[!duplicated(x[c("x", "y")]), ])

```

Explanation: Load in the raw data and split into a list of tibbles.

## Apply Functions for Single Surface Analysis

```{r Apply Functions for Single Surface Analysis, include = T}

# set seed
set.seed(10)

# simulate grids from data
simulations = simulate_grids_from_data(data = tib_density[[17]], 512, 512, n_grids = 1000)

# perform Average Nearest Neighbor Tests
ann_tests = stomata_ann_tests(list = simulations, 512, 512)

# build a dataframe
simulation = list_to_frame(ann_tests)

# calculate p-value and dispersion value
p = real_test(tib_density[[17]], 512, 512, 1000)

# return p
p

# plot
plot = stom_hist(simulation, real_NNI = p[[2]], p_value = p[[1]], dispersed = p[[4]])

# return plot
plot

```

## Apply Functions for Amphistomatic Analysis

```{r Apply Functions for Amphistomatic Analysis, include = T}

# pair stomatal grids
pairs = df_pair(tib_density)

# set seed
set.seed(100)

# run simulations
simulation = simulate_grids_from_pairs(pairs[[17]], 512, 512, n_grids = 100)
simulation = corr_pairs(simulation, 512, 512)
simulation = corr_to_frame(simulation)

# perform test
test = real_corr_test(pairs[[17]], 512, 512, 100, FALSE)

# change to a dataframe
test = corr_to_frame_final(test)

# return result
test

# build histogram
histogram = amphi_stom_hist(test, simulation)

# return histogram
histogram

```

# Test Analysis of All Samples

In this test, the number of simulations (100) to obtain the p-values are smaller than it will be during the final analysis (1000+).

## Single Surface Test Analysis

```{r Single Surface Test Analysis - All Samples, include = T}

# load data
ss_test_df = read.csv("single_surface_prelim.csv")

str(ss_test_df)

ss_test_df$p = as.numeric(ss_test_df$p)
ss_test_df$real_NNI = as.numeric(ss_test_df$real_NNI)
ss_test_df$num_stom = as.numeric(ss_test_df$num_stom)
ss_test_df$dispersed = as.numeric(ss_test_df$dispersed)
str(ss_test_df)

# p-values above 0.95 and below 0.5
sum(ss_test_df$p <= 0.05) / nrow(ss_test_df)
sum(ss_test_df$p >= 0.95) / nrow(ss_test_df)
sum(ss_test_df$p >= 0.95)
sum(ss_test_df$p <= 0.05)

# p-values above 0.975 and below 0.025

sum(ss_test_df$p <= 0.025) / nrow(ss_test_df)
sum(ss_test_df$p >= 0.975) / nrow(ss_test_df)
sum(ss_test_df$p >= 0.975)
sum(ss_test_df$p <= 0.025)

# separate sides
abaxial = ss_test_df[as.factor(ss_test_df$surface) == "abaxial",]
adaxial = ss_test_df[as.factor(ss_test_df$surface) == "adaxial",]

# p-values for abaxial and adaxial surfaces
sum(abaxial$p >= 0.975)
sum(abaxial$p <= 0.025)
sum(adaxial$p >= 0.975)
sum(adaxial$p <= 0.025)

# summary statistics
summstats = summary(ss_test_df, digits = 2)
summstats

# light and dispersed index
# boxplot()
ggplot(ss_test_df, aes(x = light, y = dispersed)) + geom_boxplot()

ssaov1 = aov(dispersed ~ light, data = ss_test_df)
summary(ssaov1)

sstukey1 = TukeyHSD(ssaov1)
sstukey1

# dispersed index and surface
ggplot(ss_test_df, aes(x = surface, y = dispersed)) + geom_boxplot()

ssaov2 = aov(dispersed ~ surface, data = ss_test_df)
summary(ssaov2)

```

## Amphistomatic Test Analysis

```{r Amphistomatic Test Analysis - using Kernel Density, include = T}

# load data
test_all_df = read.csv("dual_surface_kern_prelim.csv")

# preliminary analysis

names(test_all_df)
str(test_all_df)

test_all_df$p = as.numeric(test_all_df$p)
test_all_df$corr_coef = as.numeric(test_all_df$corr_coef)
test_all_df$n_stom_ab = as.numeric(test_all_df$n_stom_ab)
test_all_df$n_stom_ad = as.numeric(test_all_df$n_stom_ad)
test_all_df$stom_ratio = as.numeric(test_all_df$stom_ratio)
str(test_all_df)

# p-values above 0.95 and below 0.5
sum(test_all_df$p <= 0.05) / nrow(test_all_df)
sum(test_all_df$p >= 0.95) / nrow(test_all_df)
sum(test_all_df$p >= 0.95)
sum(test_all_df$p <= 0.05)

binom1 = prop.test(sum(test_all_df$p >= 0.95), nrow(test_all_df), 
                    p = 0.05, alternative = "two.sided")
binom1

binom2 = prop.test(sum(test_all_df$p <= 0.05), nrow(test_all_df), 
                    p = 0.05, alternative = "two.sided")
binom2

# p-values above 0.975 and below 0.025

sum(test_all_df$p <= 0.025) / nrow(test_all_df)
sum(test_all_df$p >= 0.975) / nrow(test_all_df)
sum(test_all_df$p >= 0.975)
sum(test_all_df$p <= 0.025)

binom3 = prop.test(sum(test_all_df$p >= 0.975), nrow(test_all_df), 
                    p = 0.025, alternative = "two.sided")
binom3

binom4 = prop.test(sum(test_all_df$p >= 0.025), nrow(test_all_df), 
                    p = 0.025, alternative = "two.sided")
binom4

# summary statistics

sumstats = summary(test_all_df, digits = 2)
sumstats

# plot light and corr_coef
ggplot(test_all_df, aes(x = light, y = corr_coef)) + geom_boxplot()

# run anova
aov1 = aov(corr_coef ~ light, data = test_all_df)
summary(aov1)

# build histograms
histogram1 <-
  ggplot(test_all_df, aes(x = corr_coef)) +
  geom_histogram(color = "#e9ecef", alpha=0.6, position = 'identity',
                 binwidth = 0.05) +
  theme_pubr() +
  labs(fill="")
histogram1

histogram2 <-
  ggplot(test_all_df, aes(x = stom_ratio)) +
  geom_histogram(color = "#e9ecef", alpha=0.6, position = 'identity',
                 binwidth = 0.05) +
  theme_pubr() +
  labs(fill="")
histogram2

# plot a scatter plot
scatter1 =
  ggplot(test_all_df, aes(x = stom_ratio, y = corr_coef)) +
          geom_point() +
  theme_pubr()
scatter1

# stomatal ratio and corr_coef
lm1 = lm(corr_coef ~ stom_ratio, data = test_all_df)

summary(lm1)

# stomatal ratio and light
lm = lm(stom_ratio ~ light, data = test_all_df)
summary(lm)

ggplot(test_all_df, aes(x = light, y = stom_ratio)) + geom_boxplot()

# number of stomata and light
# abaxial surface
aov = aov(n_stom_ab ~ light, data = test_all_df)
summary(aov)

ggplot(test_all_df, aes(x = light, y = n_stom_ab)) + geom_boxplot()

# adaxial surface
aov = aov(n_stom_ad ~ light, data = test_all_df)
summary(aov)

ggplot(test_all_df, aes(x = light, y = n_stom_ad)) + geom_boxplot()

# manova

manova1 = manova(cbind(test_all_df$n_stom_ab, test_all_df$n_stom_ad) ~ as.factor(test_all_df$light))
summary(manova1)

# paired t.test
t.test1 = t.test(test_all_df$n_stom_ab, test_all_df$n_stom_ad, paired = T)
t.test1

```

## Compare Distance Squared Raster Calc to Kernel Density

```{r Compare Distance Squared Raster Calc to Kernel Density, include = TRUE}

# load data
test_all_dist_df = read.csv("dual_surface_dist_sq_prelim.csv")

names(test_all_dist_df)
str(test_all_dist_df)

test_all_dist_df$p = as.numeric(test_all_dist_df$p)
test_all_dist_df$corr_coef = as.numeric(test_all_dist_df$corr_coef)
test_all_dist_df$n_stom_ab = as.numeric(test_all_dist_df$n_stom_ab)
test_all_dist_df$n_stom_ad = as.numeric(test_all_dist_df$n_stom_ad)
test_all_dist_df$stom_ratio = as.numeric(test_all_dist_df$stom_ratio)
str(test_all_dist_df)

# p-values above 0.95 and below 0.5
sum(test_all_dist_df$p <= 0.05) / nrow(test_all_dist_df)
sum(test_all_dist_df$p >= 0.95) / nrow(test_all_dist_df)
sum(test_all_dist_df$p >= 0.95)
sum(test_all_dist_df$p <= 0.05)

# binom test
binom1 = prop.test(sum(test_all_dist_df$p >= 0.95), nrow(test_all_df), 
                    p = 0.05, alternative = "two.sided")
binom1

binom2 = prop.test(sum(test_all_dist_df$p <= 0.05), nrow(test_all_df), 
                    p = 0.05, alternative = "two.sided")
binom2

# p-values above 0.975 and below 0.025

sum(test_all_dist_df$p <= 0.025) / nrow(test_all_dist_df)
sum(test_all_dist_df$p >= 0.975) / nrow(test_all_dist_df)
sum(test_all_dist_df$p >= 0.975)
sum(test_all_dist_df$p <= 0.025)

binom3 = prop.test(sum(test_all_dist_df$p >= 0.975), nrow(test_all_df), 
                    p = 0.025, alternative = "two.sided")
binom3

binom4 = prop.test(sum(test_all_dist_df$p >= 0.025), nrow(test_all_df), 
                    p = 0.025, alternative = "two.sided")
binom4

# summary statistics

sumstats = summary(test_all_dist_df, digits = 2)
sumstats

# plot light and corr_coef
ggplot(test_all_dist_df, aes(x = light, y = corr_coef)) + geom_boxplot()

# run anova
aov1 = aov(corr_coef ~ light, data = test_all_dist_df)
summary(aov1)

# build histograms
histogram1 <-
  ggplot(test_all_dist_df, aes(x = corr_coef)) +
  geom_histogram(color = "#e9ecef", alpha=0.6, position = 'identity',
                 binwidth = 0.05) +
  theme_pubr() +
  labs(fill="")
histogram1

histogram2 <-
  ggplot(test_all_dist_df, aes(x = stom_ratio)) +
  geom_histogram(color = "#e9ecef", alpha=0.6, position = 'identity',
                 binwidth = 0.05) +
  theme_pubr() +
  labs(fill="")
histogram2

# plot a scatter plot
scatter1 =
  ggplot(test_all_dist_df, aes(x = stom_ratio, y = corr_coef)) +
          geom_point() +
  theme_pubr()
scatter1

# stomatal ratio and corr_coef
lm1 = lm(corr_coef ~ stom_ratio, data = test_all_dist_df)

summary(lm1)

# stomatal ratio and light
lm = lm(stom_ratio ~ light, data = test_all_dist_df)
summary(lm)

ggplot(test_all_dist_df, aes(x = light, y = stom_ratio)) + geom_boxplot()

# number of stomata and light
# abaxial surface
aov = aov(n_stom_ab ~ light, data = test_all_dist_df)
summary(aov)

ggplot(test_all_dist_df, aes(x = light, y = n_stom_ab)) + geom_boxplot()

# adaxial surface
aov = aov(n_stom_ad ~ light, data = test_all_dist_df)
summary(aov)

ggplot(test_all_dist_df, aes(x = light, y = n_stom_ad)) + geom_boxplot()

# manova

manova1 = manova(cbind(test_all_dist_df$n_stom_ab, test_all_dist_df$n_stom_ad) ~ as.factor(test_all_dist_df$light))
summary(manova1)

# paired t.test
t.test1 = t.test(test_all_dist_df$n_stom_ab, test_all_dist_df$n_stom_ad, paired = T)
t.test1

# final t.test
t.test_final = t.test(test_all_df$p, test_all_dist_df$p)
t.test_final

```

# Ideal Abaxial, Adaxial Stomatal Patterning

```{r Make Ideal Amphistomatal Grid, include = TRUE}
make_amphi_grid
```

Explanation: The function `make_amphi_grid()` utilizes `make_equal_stomata_grid()` to build an idealized amphistomatous patterning. It takes inputs U, interstomatal distance, pixels_x, pixels_y, jitter, and rotate. works by adding $U * \sqrt3 / 3$ to the y position of each abaxial stomata to build the adaxial stomatal grid. This distance is calculated by finding the center of mass of an equilateral triangle with side lengths $U$.

```{r Plot Ideal Amphistomatal Grid, include = TRUE}
plot_amphi_grid
```

Explanation: The function `plot_amphi_grid()` plots the output from `make_amphi_grid()` to represent an idealized stomatal grid in the manuscript.

# NNI - New

```{r Perform NNI Test, include = TRUE}
get_nni
```

Explanation: The function `get_nni()` takes inputs df_stomata, pixels_x, and pixels_y, and a logical `progress` and returns nni values. It works by utilizing the nni function from the `spatialEco` package.

# Correlate Abaxial and Adaxial Surfaces

```{r Correlate Surfaces, include = TRUE}
amphi_corr
```

Explanation: The function `amphi_corr()` takes a grouped dataframe of raw data (`data`) and an empty raster (`leaf`) and builds a distance squared raster and performs a pearson raster correlation to determine if two leaf surfaces are independent.

# Tessellation Area Calculation

```{r Tessellation and Area Calculation, include = T}
tessellate_stomata
```

### 2023-09-19 ms/ms.Rmd
reconstituting most of the repo

---
# This template is largely functionally equivalent to that for elsevier
# The two fields that are not in the elsevier format are:
#      `corresponding_author` and `acknowledgements`
title: Spatial Coordination Of Stomatal Patterning Between Leaf Surfaces In Amphistomatous *Arabidopsis thaliana* Incurrs No Photosynthetic Advantage
#date: "`r Sys.Date()`"
author:
  - name: Jacob L. Watts
    email: jwatts@colgate.edu
    corresponding_author: yes # Footnote using corresponding_author
    affiliation: Colgate University
  - name: Graham Dow
    email: graham.dow@usys.ethz.ch
    affiliation: ETH
  - name: Thomas N. Buckley
    email: tnbuckley@ucdavis.edu
    affiliation: University of California Davis
    footnote: 1 # Footnote using reference
  - name: Christopher D. Muir
    email: cdmuir@hawaii.edu
    affiliation: University of Hawaii at Manoa
    footnote: 1 # Footnote using reference
address:
  - code: Colgate University
    address: 13 Oak Drive, Hamilton, NY 13346
  - code: University of Hawaii at Manoa
    address: 2500 Campus Rd, Honolulu, HI 96822
footnote:
  - code: 1
    text: "Equal contribution"
  - code: 2
    text: "Current email address: \\href{mailto:cat@example.com}{cat@example.com}"
abstract: |
  This is the abstract.

  It consists of two paragraphs.
acknowledgements: |
  This is an acknowledgement.

  It consists of two paragraphs.
keywords:
  - stomatal patterning
  - Aribidopsis thaliana
  - amphistomatous leaves
  - photosynthesis modeling
  - optimizing photosynthesis
  - CO$_2$ diffusion
fontsize: 12pt
#spacing: halfline # could also be oneline
#classoptions:
#  - endnotes
bibliography: 8_17.bib
#csl: mycitationstyle.csl
link-citations: yes
output: rticles::oup_article
#urlcolor: orange
#linkcolor: green
#citecolor: red
header-includes:
 - \usepackage[nomarkers,tablesfirst]{endfloat} # For figures and tables at end
 - \usepackage{lineno} # For line numbering
 - \linenumbers # For line numbering
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, message = FALSE) 
knitr::opts_chunk$set(fig.pos = 'p') # Places figures on pages separate from text
knitr::opts_chunk$set(out.width = '100%', dpi = 300) # Figure resolution and size
knitr::opts_chunk$set(fig.env = "figure") # Latex figure environment
knitr::opts_knit$set(root.dir = "..")
```

```{r}
source("r/header.R")
```

# Introduction

Stomatal anatomy (e.g. size, density, distribution, and patterning) and physiological responses to the environment regulate gas exchange during photosynthesis; namely CO$_2$ assimilation and water loss through transpiration. Since waxy cuticles are mostly impermeable to CO$_2$ and $H_2O$, stomata are discrete points through which gas exchange occurs and account only for a small percentage of the leaf area [@lange_responses_1971]. They consist of two guard cells which open and close upon changes in turgor pressure or hormonal cues [@mcadam_linking_2016]. The stomatal pore leads to an internal space known as the substomatal cavity where gases (and pathogens) contact the mesophyll [@muir_stomatal_2020]. Once in the mesophyll, CO$_2$ diffuses throughout a network of intercellular air space (IAS) and into mesophyll cells where CO$_2$ assimilation ($A$) occurs within the chloroplasts [@lee_diffusion_1964]. Maximum conductance ($g_\text{max}$) and transpiration are determined by numerous environmental and anatomical parameters such as: vapor pressure deficit (VPD), irradiance, temperature, wind speed, leaf water potential, IAS geometry, mesophyll cell anatomy, and stomatal anatomy.

Natural selection should optimize the anatomy and physiology of stomata to maximize CO$_2$ gain per unit of water loss, thereby maximizing assimilation rates for a given set of hydraulic constraints and light environments [@cowan_stomatal_1977; @buckley_optimal_2017; @sperry_predicting_2017]. However, optimal stomatal anatomy and physiology depends on developmental and environmental constraints and other physiological co-variates [@croxdale_stomatal_2000; @harrison_influence_2020; MUIR IN PRESS]. There is a wide variety of stomatal traits in nature, but certain common features as well. For example, almost all stomata follow the one cell spacing rule to maintain proper stomatal functioning [@geisler_oriented_2000; @dow_physiological_2014]; however some species (notably in *Begonia*) appear to benefit from overlapping vapor shells caused by stomatal clustering [@yi_gan_stomatal_2010; @lehmann_effects_2015; @papanatsiou_stomatal_2017]. Stomatal density positively co-varies with irradiance during leaf development and negatively co-varies with CO$_2$ concentration [@gay_influence_1975; @schoch_dependence_1980; @woodward_stomatal_1987; @royer_stomatal_2001]. Stomatal size is jointly controlled by genome size, light, and stomatal density [@jordan_environmental_2015]. Size positively co-varies with genome size [@roddy_scaling_2020] and negatively co-varies with stomatal density [@camargo_density_2011]. Total stomatal area (size * density) is optimized for operational conductance ($gs_\text{op}$) rather than maximum conductance ($gs_\text{max}$) such that stomatal apertures are most responsive to changes in the environment at their operational aperture [@franks_physiological_2012; @liu_scaling_2021]. Stomatal aperture can compensate for maladaptive stomatal densities to an extent [@bussis_stomatal_2006], but stomatal density and size ultimately determine a leaf's theoretical $gs_\text{max}$ [@sack_developmental_2016]. Additionally, low stomatal densities lead to irregular and insufficient CO$_2$ supply and reduced photosynthetic efficiency in areas far from stomata [@pieruschka_lateral_2006; @morison_lateral_2005], while high stomatal densities can reduce water use efficiency (WUE) [@bussis_stomatal_2006] and incur excessive metabolic costs [@deans_optimization_2020]. In most species, stomata occur on the abaxial (usually lower) leaf surface; but amphistomy, the occurrence of stomata on both abaxial and adaxial leaf surfaces, is also prevalent in high light environments with constant or intermittent access to sufficient water [@mott_adaptive_1982; @jordan_using_2014; @muir_light_2018; @drake_two_2019; @muir_is_2019]. Amphistomy effectively halves CO$_2$ diffusion path length and boundary layer resistance by doubling boundary layer area [@parkhurst_adaptive_1978; @harrison_influence_2020; @mott_amphistomy_1991]. Historically, stomatal patterning in dicot angiosperms was thought to be random with an exclusionary distance surrounding each stomate [@sachs_developmental_1974]; however, the developmental controls of stomatal patterning are poorly understood and likely more complex than random development along the leaf surface. @croxdale_stomatal_2000] reviews three developmental theories which attempt to explain stomatal patterning in angiosperms: inhibition, cell lineage, and cell cycle, ultimately arguing for a cell cycle based control of stomatal patterning.

The patterning and spacing of stomata on the leaf affects photosynthesis in $C_3$ leaves by altering the CO$_2$ diffusion path length from stomata to sites of carboxylation in the mesophyll. Maximum photosynthetic rate ($A_\text{max}$) in $C_3$ plants is generally co-limited by biochemistry and diffusion, but limitations depend largely on light availability [@parkhurst_intercellular_1990; @manter_ci_2004; @carriqui_diffusional_2015]. In low light environments, assimilation rates are low; consequently, CO$_2$ flux is low and the internal CO$_2$ concentration ($C_\text{i}$) remains high; thus $A_\text{max}$ is constrained by biochemical limitations in low light [@kaiser_metabolic_2016]. For well hydrated leaves in high light, photosynthesis is often limited by CO$_2$ supply as resistances from the boundary layer, stomatal pore, and mesophyll can result in insufficient $C_\text{i}$ to supply efficient photosynthesis at the chloroplast [@farquhar_biochemical_1980; lehmeier_cell_2017]. In this study, we focus primarily on how stomatal patterning affects diffusion.

To maximize CO$_2$ supply from the stomatal pore to chloroplasts, stomata should be uniformly distributed in an equilateral triangular grid on the leaf surface so as to minimize stomatal number and CO$_2$ diffusion path length. As the diffusion rate of CO$_2$ though liquid is approximately 10,000 times slower than CO$_2$ diffusion through air, mesophyll resistance is generally thought to be primarily limited by liquid diffusion [@aalto_three-dimensional_2002], but diffusion through the IAS has also been shown to be a rate limiting process because the tortuous, disjunct nature of the IAS can greatly increase diffusion path lengths [@harwood_understanding_2021]. Additionally, tortuosity is higher in horizontal directions (parallel to leaf surface) than vertical directions (perpendicular to leaf surface) because of the cylindrical shape and vertical arrangement of pallisade mesophyll cells [@earles_beyond_2018; @harwood_understanding_2021]; however, the ratio of lateral to vertical diffusion rate is still largely unknown [@morison_lateral_2005; @pieruschka_lateral_2005; @pieruschka_lateral_2006]. Depending on the thickness of the leaf, porosity of the leaf mesophyll, tortuosity of the IAS, and lateral to vertical diffusion rate ratio, minimizing diffusion path length for CO$_2$ via optimally distributed stomata may yield significant increases in CO$_2$ supply for photosynthesis and higher $A_\text{max}$.

We hypothesized that natural selection will favor stomatal patterning and distribution to minimize the diffusion path length. In amphistomatous leaves, this would be accomplished by 1) a uniform distribution of stomata on both abaxial and adaxial leaf surfaces and 2) coordinated stomatal spacing on each surface that offsets the position of stomata (Fig. \ref{fig:amphi_grid}). Coordination between leaf surfaces is defined in this study as the occurrence of stomata in areas farthest from stomata on the opposite leaf surface. Additionally, because CO$_2$ is more limiting for photosynthesis under high light, we hypothesize that in high light 3) there should be more stomata, and 4) stomata should be more uniformly distributed than in low light. Finally, as stomatal densities are selected for optimal operational aperture, we hypothesize that 5) stomatal length will be positively correlated with the area of the leaf surface to which it is closest (stomatal zone area: the area a stomate occupies and supplies with CO$_2$). This way, each stomate can be optimally sized for its relative leaf area and associated CO$_2$ demand.

To test these hypotheses, we grew *A. thaliana* in high, medium, and low light and measured stomatal density, size, and patterning on both leaf surfaces, and spatial coordination between them. We use voronoi tessellation techniques to calculate the stomatal leaf area. We also use CO$_2$ diffusion modeling to ask: what environmental and physiological conditions maximize the photosynthetic advantage ($A\text{gain}$) of abaxial-adaxial stomatal coordination? Specifically, we predicted that traits which affect diffusion path length (leaf thickness, stomatal density, leaf porosity, lateral-vertical diffusion rate ratio), diffusion rate (temperature, pressure), and CO$_2$ demand (Rubisco concentration, light) would modulate the advantage of optimal stomatal arrangement following the relationships outlined in Table \ref{tab:hypotheses}. Here, we integrate over reasonable parameter space to determine the ecophysiological context most likely to favor stomatal spatial coordination in amphistomatous leaves.

```{r hypotheses, results="asis"}

df = tibble::tribble(
  ~trait, ~relationship,
  "leaf thickness", "+",
  "stomatal density", "-",
  "leaf porosity", "-", 
  "lat.-vert. diffusion ratio", "-", 
  "temperature", "-",
  "pressure", "-",
  "Rubisco concentration", "+",
  "light", "+"
)

# Creates tables that follow OUP guidelines using xtable
print(xtable::xtable(df, caption = "A summary of the hypothesized relationships between leaf traits and environmental conditions and photosynthetic advantage of stomatal spatial coordination in amphistomatous leaves.", label = "tab:hypotheses"),
      comment = FALSE)
```

# Materials and methods

## Data Preparation

[CDM: Graham Dow provided these images. We'll need to add him as a co-author and ask him to write methods on image acquisition.]

*Arabidopsis thaliana* plants were grown in three different light environments: low light (50 PAR), medium light (100 PAR), and high light (200 PAR) following the light standards from (CITE). Once leaves were mature, we captured images of the abaxial and adaxial leaf surfaces using xxx bifocal microscope. This microscope allows the capture of two focal lengths that are spatially correlated such that pixel (1,1) of the abaxial surface image is directly above pixel (1,1) of the adaxial surface image, and so on... Images were 512 by 512 pixels and covered X leaf surface area. We captured 132 images in total, making 66 abaxial-adaxial image pairs. The position of all stomata were recorded using ImageJ [@schneider_nih_2012].

## Single Surface Analysis

We tested whether stomata are non-randomly distributed by comparing the observed stomatal patterning to a random uniform pattern. For each leaf surface image with $n$ stomata we generated $10^3$ synthetic surfaces with $n$ stomata uniformly randomly distributed on the surface. For each sample image, we compared the observed Nearest Neighbor Index ($\mathrm{NNI}$) to the null distribution of $\mathrm{NNI}$ values calculated from the synthetic data set. The observed stomatal distribution is dispersed relative to a uniform random distribution if the observed $NNI$ is greater than 95% of the synthetic $\mathrm{NNI}$ values (one-tailed test).

$\mathrm{NNI}$ is the ratio of observed mean distance ($\overline{D}_O$) to the expected mean distance ($\overline{D}_E$) where $\overline{D}_E$ is:

\begin{equation}\label{eq:emd}
  \mathrm{\overline{D}_E} = \frac{0.5}{\sqrt{A_l/n_\text{stomata}}}
\end{equation}

$A_l$ is leaf area and $n_\text{stomata}$ the number of stomata. $\overline{D}_E$ is the theoretical average distance to the nearest neighbor of each stomate if stomata were randomly distributed (CITE; Clark and Evans, 1954). And $\overline{D}_O$ calculated for each synthetic data set is:

\begin{equation}\label{eq:omd}
  \mathrm{\overline{D}_O} = \frac{\sum_{i=1}^{n_\text{stomata}}d_i}{n_\text{stomata}}
\end{equation}

$d_i$ is the distance between $\text{stomate}_i$ and its nearest neighbor. We calculated $\mathrm{NNI}$ using the R package **spatialEco** version `r packageVersion("spatialEco")`.

For each sample image, we also simulated $10^3$ synthetic data with $n$ stomata ideally dispersed in an equilateral triangular grid. For these grids, we integrated over plausible stomatal densities and then conditioned on stomatal grids with exactly $n$ stomata. The simulated stomatal count was drawn from a Poisson distribution with the mean parameter $\lambda$ drawn from a Gamma distribution with shape $n$ and scale 1 $\lambda \sim \Gamma(n, 1)$. $\Gamma(n, 1)$ is the posterior distribution of $\lambda$ with a flat prior distribution. This allows us to integrate over uncertainty in the stomatal density from the sample image.

We calculated the dispersion index $\mathrm{DI}$) which varies from zero to one. Zero is uniformly random and one is ideally dispersed:

\begin{equation}\label{eq:disp}
  \mathrm{DI} = \frac{\mathrm{NNI} - \text{median}(\mathrm{NNI_{random}})}{\text{median}(\mathrm{NNI_{uniform}}) - \text{median}(\mathrm{NNI_{random}})}
\end{equation}

Uniformly distributed stomata on a leaf with stomatal density ($D_S$) maximize $\overline{D}_O$ in an equilateral triangle pattern with side lengths ($s$) equal to two times the incircle radius or apothem, $a$, of a regular hexagon with area ($A_\text{hex}$) equal to the area of the leaf ($A_l$) divided by the number of stomata ($n_\text{stomata}$) on the leaf according to Eq. \ref{eq:eq2}:

\begin{equation}\label{eq:eq2}
  s = \frac{\sqrt{2}} {3^{1/4} \sqrt{n_\text{stomata} / A_l}}
\end{equation}

As all sides of an equilateral triangle have the same length, the average nearest neighbor distance for a given $D_S$ is maximized and each stomate occupies the same hexagonal area ($A_\text{hex}$). The closer to one $DI$ is, the more uniform an area each of its stomata supply with CO$_2$ during photosynthesis. We tested whether light treatment affects $\mathrm{DI}$ and $D_S$ using analysis of variance (ANOVA).

Finally, to test our hypothesis that stomatal length is modulated by the area of the leaf to which it supplies CO$_2$, we examined the relationship between stomatal zone area and stomatal length using a Bayesian generalized non-linear multilevel model with the R package **brms** version `r packageVersion("brms")`. Stomatal zone area was calculated using voronoi tessellation (e.g. Fig. \ref{fig:example}). The stomatal zone area, $S_\text{area}$, is the region of the leaf surface whose distance to stomate, $S$, is less than the distance to any other stomate, $S$. Stomatal length was measured in `ImageJ` [@schneider_nih_2012].

## Paired Abaxial and Adaxial Surface Analysis

To test whether the position of ab- and adaxial stomata are coordinated we compared the observed distribution to a null distribution where the positions on each surface are random. For each pair of surfaces (observed or synthetic) we calculated the distance squared between each to the nearest stomatal centroid with the R package **raster** version `r packageVersion("raster")`. Then we calculated the cell-wise Pearson correlation coefficient. If stomatal positions on each surface are coordinated to minimize the distance between mesophyll and the nearest stomate, then we expect a negative correlation. A cell that is far from a stomate on one surface should be near a stomate on the other surface (Fig. \ref{fig:amphi}). We generated a null distribution of the correlation coefficient by simulating $10^3$ synthetic data sets for each observed pair. For each synthetic data set, we simulated stomatal position using a random uniform distribution, as described above, matching the number of stomata on abaxial and adaxial leaf surfaces. Stomatal positions on each surface are coordinated if the correlation coefficient is greater than 95% of the synthetic correlation values (one-tailed test).

## Modeling Photosynthesis

We modeled photosynthesis CO$_2$ assimilation rate using a spatially-explicit two-dimensional reaction diffusion model using a porous medium approximation (Parkhurst 1994). Consider a two-dimensional leaf where stomata occur on each surface in a regular sequence with interstomatal distance $U$. The main outcome we assessed is the advantage of offsetting the position of stomata on each surface compared to have stomata on the same $x$ position on each surface. With these assumptions, by symmetry, we only need to model two stomata, one abaxial and one adaxial, from $x = 0$ to $x = U/2$ and from the adaxial surface at $y = 0$ to the abaxial surface at $y = L$, the leaf thickness. We arbitrarily set the adaxial stomate at $x = 0$ and toggled the abaxial stomata position between $x = U/2$ (offset) or $x = 0$ (below adaxial stomate). The advantage of offsetting stomatal position on each surface is the photosynthetic rate of the leaf with offset stomata compared to that with stomata aligned in the same $x$ position:

\begin{equation} \label{eq:coordination_advantage}
  \text{coordination advantage} = \frac{A_\text{offset}}{A_\text{aligned}}
\end{equation}

We modelled the coordination advantage over a range of leaf thicknesses, stomatal densities, photosynthetic capacities, and light environments to understand when offsetting stomatal position on each surface might deliver a significant photosynthetic advantage.

### Light propogation model

[NOTE: we should probably adjust light attenuation to be proportional to the chlorophyll concentration which is one of the parameters in the biochemical models.]

Irradiance at depth $y$ in a leaf with thickness $L$ is modeled following Lloyd *et al.* (1992):

\begin{equation}\label{eq:light_propogation}
  I(y) = 1.1 I_0 e ^ {-2.4 y / L}
\end{equation}

where $I_0$ is photosynthetically active irradiance incident on the adaxial leaf surface.

### Biochemical model

All parameter symbols, units, descriptions, and values are described in Table X below. Following Gutschick (1984), we modeled photosynthetic rate per unit chlorophyll $A_\text{chl}$ then calculated the volumetric photosynthetic rate $A_\text{volume}$ by multiplying by the chlorophyll concentration $Q_\text{chl}$:

$$A_\text{volume} = A_\text{chl} Q_\text{chl}$$
A description of the model is given on page 553-556 of Gutschick (1984). The R code below is how we implemented the model.

* `C_m` is a vector CO$_2$ concentrations in $\text{mmol m}^{-3}$ at different positions within the leaf
* `I_m` is a vector of irradiances in $\mu \text{mol m}^{-2} \text{s}^{-1}$ at different positions within the leaf (same order as `C_m`).
* `pars` is a list of parameters

```{r, echo = TRUE, eval = FALSE}

pars = list(

    P = set_units(101.325, kPa), # air pressure at sea level
    temp = set_units(298.15, K), # assume constant temperature
    R_gas = set_units(8.314, J/K/mol), # ideal gas constant

    # Calculations for C_a and O ued below
    # 21% O2
    # set_units(0.21 * P / (R_gas * temp), mol/m^3)
    # 415 ppm
    # set_units((415/1e6) * P / (R_gas * temp), mmol/m^3)

    # Environmental
    C_a = set_units(16.96367, mmol/m^3),
    O = set_units(8.584027, mol/m^3),
    PAR = set_units(1000, umol/m^2/s),

    # Biochemical
    E_t = set_units(0.01, mol/mol),
    eta_t = set_units(0.59, 1),
    k_c = set_units(20, mol/mol/s),
    k_o = set_units(4.2, mol/mol/s),
    K_c = set_units(0.0184, mol/m^3),
    K_o = set_units(13.2, mol/m^3),
    Q_chl = set_units(3.3, mol/m^3),
    R_p = set_units(0.273, mol/mol),

    # Diffusivity of CO2 in leaf airspace. From Gutschick 1984. Should be updated and include temperature dependence, effect of cell packing, tortuosity, etc.
    D_mc = set_units(7e-7, m^2/s)

)

# strip units to speed up calculation
upars = purrr::map(pars, ~ {if(inherits(.x, "units")) {drop_units(.x)}})

# multiply/divide by 1e3 because C_m is mmol and parameters are in mol
k_vc = upars[["k_c"]] / (1 + (1e3 * upars[["K_c"]] / C_m) *
                           (1 + upars[["O"]] / upars[["K_o"]])) # [1/s]
# Given light_propagation assumptions, k = 2.4/L assuming no scattering
k_i = 2.4 / upars[["leaf_thickness"]] # [1/um] # I'm not sure sure this is right
b_x = k_i * I_m # [mol/m^3/s]
a_x = b_x / upars[["Q_chl"]] # [1/s]
j = 0.5 * a_x * upars[["eta_t"]] # [1/s] # eqn 9
phi = upars[["k_o"]] / upars[["k_c"]] * (upars[["O"]] / upars[["K_o"]]) /
  (1e-3 * C_m / upars[["K_c"]]) # [1]
v_cj = j / (4 + 4 * phi) # [1/s]
v_cp = k_vc * upars[["R_p"]] # [1/s]
v_cr = apply(cbind(v_cj, v_cp), 1, min) # [1/s]
v_c = apply(cbind(k_vc * upars[["E_t"]], v_cr), 1, min) # [1/s]
v_o = phi * v_c # [1/s]
A_chl = v_c - 0.5 * v_o # [1/s] # i.e mol CO2 / mol Chl / s
A_volume = A_chl * upars[["Q_chl"]] # [mol/m^3/s]

```

To model photosynthesis and CO$_2$ transport within a two-dimensional cross section of the leaf, we built a grid of nodes of dimensions leaf thickness by half the interstomatal distance. Nodes represent the leaf mesophyll where CO$_2$ diffusion and assimilation occur.

**Table X** Glossary of mathematical symbols. The columns indicate the mathematical Symbol used in the paper, the associated symbol used in R scripts, scientific Units, and a verbal Description.

 Symbol              | Value(s) | Units            | Description
---------|--------|--------------------------|----------------------------------
Biochemical Parameters |||
$A_\text{volume}$    | NA    | $mol CO_2/m^3/s$    | volumetric assimilation rate
$A_\text{chl}$       | NA    | $mol CO2/mol Chl/s$ | assimilation rate per mol chlorophyll
$E_t$                | 0.01  | $mol / mol Chl$     | rubisco octamer concentration per chlorophyll  
$eta_t$              | 0.59  | unitless            | quantum efficiency of photoexcitation transfer to reaction centers
$k_c$                | 20    | $mol CO2/mol Rubisco/s$ at 25 C | maximal carboxylation velocity of Rubisco
$k_o$                | 4.2   | $mol O2/mol Rubisco/s$ at 25 C | maximal oxygenation velocity of Rubisco
$K_c$                | 0.0184 | $mol/m^3$ at 25 C  | Michaelis constant for CO2
$K_o$                | 13.2  | $mol/m^3$ at 25 C   | Michaelis constant for O2
$J_\text{max}$       | 0.253 | $mol e\text{-}/mol Chl/s$ | maximal electron transport rate per mol Chl
$Q_\text{chl}$       | 3.3   | $mol/m^3$           | Chl volume concentration
$v_c$                | NA    | $mol CO2/mol Chl/s$ | carboxylation rate
$v_o$                | NA    | $mol O2/mol Chl/s$  | oxygenation rate
$k_\text{vc}$        | NA    | $mol CO2/mol Rubisco/mol Chl$ | carboxylation velocity per Rubisco octamer at ambient CO2
$R_p$                | 0.273 | $mol RuBP/mol Chl$  | RuBP pool size
$k_i$                | 2.4   | $1/m^2$ or $1/L$    | light attenuation coefficient
$D_\text{mc}$        | 7e-7  | $m^2/s$             | diffusivity of CO$_2$ in mesophyll airspace
Environmental Parameters |||
$P$                  | 101.325 | $kPa$             | air pressure at sea level
$temp$               | 298.15 | $K$                | temperature
$R_\text{gas}$       | 8.314 | $J/K/mol$           | ideal gas constant
$C_a$                | 16.96367 | $mmol CO_2/m^3$  | atmospheric CO$_2$ concentration
$O$                  | 8.584027 | $mol O_2/m^3$    | atmospheric $O_2$ concentration
$PAR$                | 1000  | $umol photons/m^2/s$ | photosynthetically active radiation
Variables |||
$U_s$                | .05, .10, .15... | $mm$     | interstomatal distance
$T_l$                | .1, .2, .3... | $mm$        | leaf thickness
$g_\text{sc}$        | 0.2, 0.3, 0.4... | $mol/m^2/s$ | stomatal conductance

## Two-dimensional model

FYI - something is wrong with the model. I think the units are off somewhere because the result is very sensitive to grid size and the values often aren't reasonable. I'll need to troubleshoot more, but maybe something will be obvious to you.

We used the `steady.2D()` function in the R package **rootSolve**. The function needs a function to calculate the time differential for node $ij$, $\frac{dC_{m,ij}}{dt}$. The R function below uses the light propagation, biochemical, and diffusional model to calculate a matrix of $\frac{dC_{m,ij}}{dt}$ values:

```{r, echo = TRUE, eval=FALSE}

diffusion2D <- function (time, state, pars, stomata_offset) {

  n_row = pars[["n_row"]]
  n_col = pars[["n_col"]]

  n_node = n_row * n_col

  node_length_m = node_length * 1e-6 # node_length in [m]

  # matrix of C_m values
  C_m_mat = matrix(nrow = n_row, ncol = n_col, state)

  # Photosynthetic demand
  ## Light matrix
  I_mat = seq(0, by = pars[["node_length"]], length.out = n_row) |>
    light_propogation(pars[["leaf_thickness"]], pars[["PAR"]]) |>
    matrix(n_row, ncol = n_col)

  ## Biochemical model
  # Units have to be correct for this to work, but unitless = TRUE should be faster
  A = biochemical_model(C_m_mat, I_mat, pars = pars, unitless = TRUE)

  ## Photosynthesis matrix
  ## multiply by 1000 to convert from mol / m^3 / s to mmol / m^3 / s
  A_mat = matrix(nrow = n_row, ncol = n_col, 1e3 * A)

  # CO2 diffusion
  C_a = drop_units(pars[["C_a"]])
  flux_mat = matrix(nrow = n_row, ncol = n_col, 0)
  flux_mat[1, 1] = pars[["g_sc"]] * (C_a - C_m_mat[1, 1])

  ## 1. Flux through stomata
  if (stomata_offset) {
    flux_mat[n_row, n_col] = pars[["g_sc"]] * (C_a - C_m_mat[n_row, n_col])
  } else {
    flux_mat[n_row, 1] = pars[["g_sc"]] * (C_a - C_m_mat[n_row, 1])
  }

  zero_x <- rep(0, n_row)
  zero_y <- rep(0, n_col)

  ## 2. Mesophyll flux; zero fluxes near boundaries
  flux_above = rbind(rep(0, n_col), C_m_mat[1:(n_row - 1), ] - C_m_mat[2:n_row, ])
  flux_below = rbind(C_m_mat[2:n_row, ] - C_m_mat[1:(n_row - 1), ], rep(0, n_col))
  flux_left  = cbind(rep(0, n_col), C_m_mat[, 1:(n_col - 1)] - C_m_mat[, 2:n_col])
  flux_right = cbind(C_m_mat[, 2:n_col] - C_m_mat[, 1:(n_col - 1)], rep(0, n_col))

  # g_mc [m/s] = D_mc [m^2/s] / node_length [m]
  flux_mat = flux_mat + pars[["g_mc"]] * (flux_above + flux_below + flux_left + flux_right)

    return(list(c(as.vector(flux_mat / node_length_m - A_mat))))

}

```

## Global Results

Stomatal density of *Arabidopsis thaliana* the 132 leaves measured ranged from 12 to 93 (units) with high light leaves ranging from 93 to 55 (units), medium light from 15 to 35 (units), and low light from 12 to 42 (units). Leaves were amphistomatous with a mean stomatal ratio of 0.45.

## Single Surface Results

If our hypotheses that natural selection will act to reduce the diffusion path length of CO$_2$ while minimizing stomatal number are correct, then we would expect to find leaf surfaces with uniformly distributed stomata. However, to the contrary, we find that though 57 of the 132 (43.1%) leaf surfaces were significantly more uniformly dispersed than uniform random synthetic stomatal grids ($\alpha$ = 0.05), none of the leaf surfaces exhibited perfectly uniform stomatal patterning (dispersion index = 1) (Fig. \ref{fig:dispersion}). Additionally, we hypothesized that as CO$_2$ is more limiting to photosynthesis under high light, stomata would be more uniformly dispersed in plants grown in high light than plants grown in low and medium light. The data also fail to support this hypothesis as there is no strong, discernible trend between light and stomatal patterning. Interestingly, adaxial leaf surfaces were more uniformly dispersed than associated abaxial leaf surfaces across all light treatments (F$_\text{1,126}$ = 28.8; p-value < 0.001). Rather than regulate stomatal patterning in response to light regimes, plants respond by increasing stomatal density (Fig. \ref{fig:density}). Stomatal density drastically increased in plants grown under high light, validating the long held hypothesis that light strongly influences stomatal density (F$_\text{2,126}$ = 680.7; p-value < 0.001).

Across all light treatments and leaf surfaces, stomatal length and stomatal area were weakly positively correlated, indicating some support for our hypothesis that stomata are selected for an optimal operational aperture (Fig. \ref{fig:length-area}).

## Dual Surface Analysis

As evidence against our hypothesis that natural selection should favor spatial coordination in the placement of stomata between abaxial and adaxial leaf surfaces, we found no correlation between paired abaxial and adaxial leaf surfaces (Fig. \ref{fig:dual-surface}). Light treatment had no effect on correlation between surfaces (F$_\text{2,63}$ = 2.28; p-value = 0.11). All but one abaxial, adaxial surface pairs were independent ($\alpha$ = 0.05).

### Example output

We wrapped the above functions to adjust environmental and anatomical variables. We're going to simulate over a grid of parameters, but here is what the output looks like plotted:

### DRAFT TEXT

During a model run, CO$_2$ flux between neighboring nodes was determined according to the mesophyll CO$_2$ concentration ($C_m$) in each node, distance between cells, temperature, pressure, and leaf porosity. Assimilation rate ($A$) in each node was determined by $A_\text{max}$, $c_i$, irradiance, temperature, chlorophyll volume concentration, Rubisco octamer concentration per mol chlorophyll, and maximal oxygenation and carboxylation rates of Rubisco. Stomata were placed either 1) directly on top of one another to represent anticoordination or 2) offset from one another to represent coordination between leaf surfaces. $c_i$ and CO$_2$ assimilation rate (A) were set to 415ppm and 0 respectively and the model ran until $c_i$ and A reached equilibrium in each cell. Total A and average $c_i$ were then calculated for each parameter combination.

A was calculated in terms of $A_\text{max}$ and $c_i$ following Eq. \ref{eq:eq4}

\begin{equation}\label{eq:eq4}
A = A_\text{max} * c_i / (c_i + K_c)
\end{equation}

where $K_c$ is the Michaelis constant for CO$_2$: 0.0184 mol $m^\text{-3}$.

$A_\text{max}$ was calculated in terms of irradiance (i), light saturated maximum A ($A_\text{mm}$), and quantum yield ($\varphi$) following Eq. \ref{eq:eq5}

\begin{equation}\label{eq:eq5}
A_\text{max} = A_\text{mm} * \varphi * i / (A_\text{mm} + \varphi * i)
\end{equation}

# Results

## Single surface analysis

\begin{figure}[ht]
\includegraphics[width=\textwidth]{figures/single-surface.pdf}
\caption{Stomata are more dispersed than expected under the null model of uniform random position (dispersion index = 0) but far from a distribution that maximizes distance between stomata (dispersion index = 1). Significant differences between light treatments are indicated by asterisks according to analysis of variance followed by a post-hoc tukey honest significant difference test ($\alpha$ = 0.05).}
\label{fig:dispersion}
\end{figure}

\begin{figure}[ht]
\includegraphics[width = \textwidth]{figures/density.pdf}
\caption{Stomatal density is higher in plants grown under high light conditions. Significant differences between light treatments are indicated by asterisks according to analysis of variance followed by a post-hoc tukey honest significant difference test ($\alpha$ = 0.05).}
\label{fig:density}
\end{figure}

\begin{figure}[ht]
\includegraphics[width = \textwidth]{figures/tessellation-example.pdf}
\caption{Examples of synthetic and real leaf surfaces.  A) Uniform random synthetic leaf surface; B) Example of real leaf surface; C) Regularly distributed synthetic leaf surface. The zone defined by each stomate was calculated with voronoi tessellation and correlated with stomatal length in real leaves.}
\label{fig:example}
\end{figure}

\begin{figure}[ht]
\includegraphics[width = \textwidth]{figures/length-area.pdf}
\caption{Stomatal length and stomatal zone area. Lines of best fit computed using a bayesian generalized non-linear multilevel model. Each light level and leaf surface exhibits a unique, weakly positive relationship between stomatal zone area and length.}
\label{fig:length-area}
\end{figure}

## Dual surface analysis

\begin{figure}[ht]
\includegraphics[width = \textwidth]{figures/ideal-amphi-grid.pdf}
\caption{Idealized amphistomatous stomatal grid with uniform stomatal patterning and perfect abaxial-adaxial coordination.}
\label{fig:amphi}
\end{figure}

\begin{figure}[ht]
\includegraphics[width = \textwidth]{figures/dual-surface.pdf}
\caption{Correlation between paired abaxial and adaxial leaf surfaces. Dashed line indicates no correlation. Weak positive correlations are not significantly different from null simulations. No differences in abaxial-adaxial correlation were observed between light levels according to an analysis of variance ($\alpha$ = 0.05).}
\label{fig:dual-surface}
\end{figure}

# Discussion

Stomata are expensive. A theoretical, optimized plant would minimize stomatal density while also allowing competitive gas exchange rates for its environment so as to maximize C assimilation per unit investment in stomata. Natural selection operates within developmental and physical constraints to drive each plant species toward its theoretical optimum. This study provides evidence that stomata in *Arbidopsis thaliana*, the model angiosperm, are non-randomly distributed, favoring dispersion over clustering (Fig. \ref{fig:dispersion}). However, stomata are not ideally dispersed in an equilateral triangular grid as would be optimal to minimize CO$_2$ diffusion path length and standardize the area supplied by each stomate (Fig. \ref{fig:example}). Additionally, when grown in high light environments, *A. thaliana* exhibited increased stomatal density rather than increased stomatal dispersion (Fig. \ref{fig:density}), which suggests that natural selection has acted more strongly on developmental pathways that modulate stomatal density than those that control stomatal dispersion. In other words, plants optimize gas exchange by adding more stomata rather than dispersing them more evenly across the leaf surface. This study also demonstrates that stomata that supply larger leaf areas with CO$_2$ tend to be larger (Fig. \ref{fig:length-area}). These results could suggest that 1) the added energetic and hydraulic cost of non-ideally dispersed stomata is negligible and therefore not acted on by natural selection; 2) no developmental pathway exists to ensure the ideal placement of stomata on the leaf; or 3) the regulation of stomatal size limits the cost incurred by non-ideal stomatal dispersion.

In high light environments, amphistomy is favorable as high light photosynthesis is limited by CO$_2$ and amphistomy halves diffusion path length and boundary layer resistance, thereby reducing CO$_2$ limitation - increasing theoretical $A_\text{max}$. An optimal amphistomatous leaf has offset stomata such that stomata are more likely to appear on one leaf surface if there is not a stomata directly opposite it on the other surface as shown in figure \ref{fig:amphi}. However, our results show that leaf surfaces are not coordinated but are independent, regardless of light (Fig. \ref{fig:dual-surface}). Additionally, gas exchange models show little photosynthetic efficiency gain from abaxial-adaxial stomatal coordination compared to anticoordination (INSERT FIG FROM MODELING). We posit that this marginal gain is not sufficient to be acted upon strongly by natural selection. Thus, amphistomatous plants do not exhibit abaxial-adaxial stomatal coordination for there is little selective advantage of it.

Our study corroborates previous studies which demonstrate that stomata are non-randomly distributed along the leaf surface as a result of developmental mechanisms such as spatially biased arrest of stomatal initials [@boetsch_arrest_1995], oriented asymmetric cell division [@geisler_oriented_2000], and cell cycle controls [@croxdale_stomatal_2000]. We do not investigate the potential developmental pathways that influence stomatal dispersion in this study; however, they are important to consider as these pathways could limit plants from reaching the theoretical peak in the adaptive landscape: uniform stomatal dispersion. Instead, as this study suggests, plants may simply compensate with higher stomatal density and by fitting stomatal size to the area that they supply with CO$_2$. To understand why stomata are not ideally dispersed, more modelling should be done to estimate the fitness gain of stomatal dispersion. Additionally, genetic manipulation studies should attempt to create mutants with clustered and ideally dispersed stomata for a comparison of their photosynthetic traits. This could have extremely important implications for maximum assimilation rates in crops as most crop species are grown in high light where CO$_2$ is often limiting. In drought-prone environments, increased stomatal dispersion may increase water use efficiency by reducing the number of stomata needed to achieve the same internal CO$_2$ concentration, $C_\text{i}$.

Beyond disperson on a single surface, gas exchange can be optimized via stomatal coordination of abaxial and adaxial surfaces in amphistomatous leaves. Given that leaf thicknesses are generally multiple times greater than interstomatal distance (GIVE DISTANCES HERE). As a result, abaxial-adaxial stomatal coordination reduces CO$_2$ diffusion path length far less than single surface dispersion, so we hypothesize this strategy to afford less photosynthetic advantage to the leaf. Our modelling results demonstrate that, even in ideal conditions, i.e. thick leaf, low stomatal densities, high light, low leaf porosity, high rubisco concentration, etc., the photosynthetic advantage of coordination is minimal. We are not surprised by these results, but still highlight them here as we are the first to report this finding.

Amphistomy is a unique and important adaptation found around the world across many plant lineages [@muir_light_2018], yet much of the dynamics of amphistomy remain poorly understood. Here, we show that in *Arabidopsis thaliana* 1) stomata in are non-randomly dispersed, but not ideally dispersed; 2) stomatal size and density are modulated by light; 3) stomatal size is positively correlated with the area to which it supplies CO$_2$; and 4) abaxial-adaxial stomatal coordination is not exhibited and is not shown to provide a strong photosynthetic advantage using CO$_2$ diffusion models. Interestingly, these findings did not validate many of our hypotheses which were based on first principles, suggesting that there may be limits on plants' ability to control stomatal placement. Future studies which elucidate these limitations may have important implications for agriculturaly productivity in a rapidly changing world.

# References

### 2023-09-19 ms/test.Rmd
reconstituting most of the repo

---
title: "test"
output:
  pdf_document: default
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r dispersion, fig.cap = "Stomata are more dispersed than expected under the null model of uniform random position (dispersion index = 0) but far from a distribution that maximizes distance between stomata (dispersion index = 1)", echo = TRUE}

knitr::include_graphics("figures/dispersion.pdf")

```

### 2023-09-19 r/00_process-data.R

@@ -1,0 +1,54 @@
+source("r/header.R")
+
+stomata_position = readr::read_csv("raw-data/stomata_position.csv") |>
+
+  # add light column
+  dplyr::mutate(
+    light = dplyr::case_when(
+      str_detect(treatment, "HL") ~ "high",
+      str_detect(treatment, "LL") ~ "low"
+    ),
+    light = tidyr::replace_na(light, "medium")
+  ) |>
+
+  # modify leaf_number
+  tidyr::unite(trt_leaf_number, treatment, leaf_number) |>
+  dplyr::group_by(trt_leaf_number, image_number) |>
+  dplyr::mutate(n_stomata = dplyr::n())
+
+# add image_number1
+stomata_position$image_number1 = as.character(gsub("^.{0,3}", "", stomata_position$image_number))
+
+# Check for exact duplicates
+stomata_position |>
+  dplyr::group_by(trt_leaf_number, image_number, surface) |>
+  dplyr::summarize(is_dup = duplicated(tibble(x, y))) |>
+  dplyr::filter(is_dup)
+
+# identify near duplicates
+stomata_position |>
+  # group by
+  dplyr::group_by(trt_leaf_number, image_number, surface) |>
+  #first, arrange by theta
+  arrange(x) |>
+  #mark rows with theta <=5 of next row
+  mutate(x_mark = if_else(abs(x - lag(x)) <= 7, 1, 0)) |>
+  #mark rows with rho <=15 of next row
+  mutate(y_mark = if_else(abs(y - lag(y)) <= 7, 1, 0)) |>
+  #replace NA's with 0
+  tidyr::replace_na(list(x_mark = 0, y_mark = 0)) |>
+  #filter all unwanted rows
+  filter(x_mark == 1 & y_mark == 1) |>
+  #drop unwanted columns
+  select(-x_mark, -y_mark)
+
+# add tessellation info
+stomata_tess = stomata_position |>
+  dplyr::group_modify(~ tessellate_stomata(.x, pixels_x = 512, pixels_y = 512)) |>
+  dplyr::full_join(stomata_position,
+                   by = c("trt_leaf_number", "image_number", "x", "y")) |>
+  mutate(area = ifelse(test = reject, NA, area))
+
+readr::write_rds(stomata_position, "objects/stomata_position.rds")
+readr::write_rds(stomata_tess, "objects/stomata_position_length_area.rds")
+

### 2023-09-19 r/01_simulate-single-surface.R

@@ -1,0 +1,19 @@
+source("r/header.R")
+
+stomata = readr::read_csv("raw-data/stomata_position.csv") |>
+  group_by(treatment, leaf_number, image_number, surface)
+
+pb <- progress_bar$new(
+  format = "[:bar] :current/:total (:percent) elapsed :elapsed eta :eta",
+  total = nrow(attr(stomata, "groups"))
+)
+
+set.seed(197789263)
+synthetic_data <- stomata |>
+  purrrlyr::by_slice(
+    simulate_grids_from_data,
+    pixels_x = 512, pixels_y = 512, n_grid = 1e3, progress = TRUE,
+    .collate = "rows"
+  )
+
+write_rds(synthetic_data, "objects/synthetic_data.rds")

### 2023-09-19 r/02_get-nni-single-surface.R

@@ -1,0 +1,22 @@
+source("r/header.R")
+
+stomata = readr::read_csv("raw-data/stomata_position.csv") |>
+  dplyr::group_by(treatment, leaf_number, image_number, surface)
+
+synthetic_data = readr::read_rds("objects/synthetic_data.rds") |>
+  dplyr::group_by(treatment, leaf_number, image_number, surface, sim, grid)
+
+nni_observed <- stomata |>
+  purrrlyr::by_slice(get_nni, pixels_x = 512, pixels_y = 512, .collate = "rows")
+
+pb <- progress_bar$new(
+  format = "[:bar] :current/:total (:percent) elapsed :elapsed eta :eta",
+  total = nrow(attr(synthetic_data, "groups"))
+)
+
+nni_synthetic <- synthetic_data |>
+  purrrlyr::by_slice(get_nni, pixels_x = 512, pixels_y = 512, progress = TRUE,
+                     .to = "NNI_synthetic")
+
+readr::write_rds(nni_observed, "objects/nni_observed.rds")
+readr::write_rds(nni_synthetic, "objects/nni_synthetic.rds")

### 2023-09-19 r/03_test-single-surface.R

@@ -1,0 +1,72 @@
+source("r/header.R")
+
+nni_observed = readr::read_rds("objects/nni_observed.rds") |>
+  dplyr::select(treatment, leaf_number, image_number, surface, n_stomata,
+                NNI_observed = NNI)
+
+nni_synthetic = readr::read_rds("objects/nni_synthetic.rds") |>
+  dplyr::select(treatment, leaf_number, image_number, surface, #n_stomata,
+                sim, grid, NNI_synthetic)
+
+stomata_area_summ = readr::read_rds("objects/area_summ.rds")
+
+
+single_surface_results <- nni_observed |>
+  dplyr::rowwise() |>
+  purrr::pmap_dfr(~ {
+
+    sim_ran = nni_synthetic |>
+      dplyr::filter(
+        treatment == ..1,
+        leaf_number == ..2,
+        image_number == ..3,
+        surface == ..4,
+        grid == "random"
+      ) |>
+      dplyr::pull(NNI_synthetic) |>
+      purrr::map(magrittr::extract, "NNI") |>
+      purrr::map_dbl(unlist)
+
+    sim_equ = nni_synthetic |>
+      dplyr::filter(
+        treatment == ..1,
+        leaf_number == ..2,
+        image_number == ..3,
+        surface == ..4,
+        grid == "equal"
+      ) |>
+      dplyr::pull(NNI_synthetic) |>
+      purrr::map(magrittr::extract, "NNI") |>
+      purrr::map_dbl(unlist)
+
+    ret = tibble(
+      treatment = ..1,
+      leaf_number = ..2,
+      image_number = ..3,
+      surface = ..4,
+      n_stomata = ..5,
+      NNI_observed = ..6,
+      p_value = length(which(sim_ran >= ..6)) / length(sim_ran),
+      dispersion = (..6 - median(sim_ran)) / (median(sim_equ - median(sim_ran)))
+    ) #|>
+      #full_join(as.tibble(stomata_area_summ), by = c("treatment", "leaf_number", "image_number", "surface"))
+      #cbind(stomata_area_summ)
+
+    # I'm trying to merge these datasets using cbind and full_join. but I really
+    # don't know why it's not working.
+
+    return(ret)
+
+  }) |>
+  # add light column
+  dplyr::mutate(
+    light = dplyr::case_when(
+      str_detect(treatment, "HL") ~ "high",
+      str_detect(treatment, "LL") ~ "low"
+    ),
[…]

### 2023-09-19 r/04_plot-single-surface.R

@@ -1,0 +1,72 @@
+source("r/header.R")
+
+single_surface_results = readr::read_rds("objects/single_surface_results.rds") |>
+  mutate(light_reorder = fct_relevel(light, "low", "medium", "high"))
+
+# anova
+single_surface_anova = aov(dispersion ~ light * surface, data = single_surface_results)
+summary(single_surface_anova)
+
+# some preliminary ideas how to make a table for this
+tmp = single_surface_anova |>
+  broom::tidy()
+  #mutate(p = formatC(tmp$p.value, format = "e", digits = 2))
+tmp
+tmp$p.value = format(tmp$p.value, format = "e", digits = 2)
+tmp
+  mutate(x = if.else(is.na(p.value), NA, scientize(p.value)))
+  rename(SS = sumsq, MS = meansq, `$F$` = statistic, `$P$` = p.value) |>
+
+  knitr::kable()
+
+stat.test <- aov(dispersion ~ light * surface, data = single_surface_results) |>
+  tukey_hsd()
+stat.test = stat.test[1:3,]
+
+# anova
+density_anova = aov(n_stomata ~ light * surface, data = single_surface_results)
+summary(density_anova)
+
+stat.test.density <- aov(n_stomata ~ light * surface, data = single_surface_results) |>
+  tukey_hsd()
+stat.test.density = stat.test.density[1:3,]
+
+# 1. plot dispersion index ----
+plot = ggplot(
+    single_surface_results,
+    aes(light_reorder, dispersion, fill = surface)
+  ) +
+  stat_boxplot() +
+  xlab("Light treatment") +
+  ylab("Dispersion Index\n0 = random; 1 = maximum") +
+  scale_fill_grey(start = 0.3) +
+  geom_hline(yintercept = 0, linetype = "dashed") +
+  geom_hline(yintercept = 1, linetype = "dashed") +
+  stat_pvalue_manual(
+    stat.test, label = "p.adj.signif",
+    y.position = c(.5, .575, .65),
+    inherit.aes = FALSE
+  ) +
+  theme_pubr()
+
+ggsave("single-surface.pdf", plot = plot, path = "ms/figures",
+       width = 4, height = 4)
+
+# 2. plot stomatal density ----
+density_plot = ggplot(
+    single_surface_results,
+    aes(light_reorder, n_stomata, fill = surface)
+  ) +
+  stat_boxplot() +
+  xlab("Light treatment") +
[…]

### 2023-09-19 r/05_dual-surface-correlation.R

@@ -1,0 +1,35 @@
+source("r/header.R")
+
+# load processed data
+stomata = readr::read_rds("objects/stomata_position.rds") |>
+  dplyr::group_by(trt_leaf_number, image_number1)
+
+# load synthetic data and group by
+synthetic_data = readr::read_rds("objects/synthetic_data.rds") |>
+  # add image_number1 to synthetic data for abaxial, adaxial pairing
+  dplyr::mutate(image_number1 = stringr::str_extract(image_number, "[0-9]{1}$")) |>
+  # group synthetic data
+  dplyr::group_by(treatment, leaf_number, image_number1, sim) |>
+  # exclude equal grids - only concerned with random grids
+  dplyr::filter(grid == "random")
+
+# run correlation on real data
+leaf = create_raster(pixels_x = 512, pixels_y = 512)
+corr_observed = stomata |>
+  purrrlyr::by_slice(amphi_corr, leaf = leaf, .collate = "rows",
+                     .to = "corr_coefficient")
+
+# run correlation on synthetic data
+pb <- progress_bar$new(
+  format = "[:bar] :current/:total (:percent) elapsed :elapsed eta :eta",
+  total = nrow(attr(synthetic_data, "groups"))
+)
+
+corr_synthetic = synthetic_data |>
+  purrrlyr::by_slice(amphi_corr, leaf = leaf, progress = TRUE,
+                     .collate = "rows", .to = "corr_coefficient")
+
+# write rds files
+readr::write_rds(corr_observed, "objects/corr_observed.rds")
+readr::write_rds(corr_synthetic, "objects/corr_synthetic.rds")
+

### 2023-09-19 r/06_test-dual-surface.R

@@ -1,0 +1,35 @@
+source("r/header.R")
+
+# load data
+corr_observed = readr::read_rds("objects/corr_observed.rds") |>
+  # separate columns
+  tidyr::separate(trt_leaf_number, c("treatment", "leaf_number"), sep = "_",
+                  remove = TRUE) |>
+  dplyr::select(treatment, leaf_number, image_number1, corr = corr_coefficient)
+
+# synthetic data
+corr_synthetic = readr::read_rds("objects/corr_synthetic.rds") |>
+  dplyr::select(treatment, leaf_number, image_number1, sim,
+                corr_sim = corr_coefficient) |>
+  dplyr::mutate(leaf_number = as.character(leaf_number))
+
+# make dataframe of results
+
+dual_surface_results <- corr_observed |>
+  dplyr::full_join(
+    corr_synthetic,
+    by = c("treatment", "leaf_number", "image_number1")
+  ) |>
+  dplyr::group_by(treatment, leaf_number, image_number1) |>
+  # two-sided test
+  dplyr::mutate(corrsim_ge_corr = abs(corr_sim) >= abs(corr)) |>
+  dplyr::summarise(
+    corr = dplyr::first(corr),
+    mu_corr_sim = mean(corr_sim),
+    n_corr_ge_corrsim = sum(corrsim_ge_corr),
+    n = n(),
+    p_value = n_corr_ge_corrsim / n
+  )
+
+readr::write_rds(dual_surface_results, "objects/dual_surface_results.rds")
+

### 2023-09-19 r/07_plot-dual-surface.R

@@ -1,0 +1,34 @@
+##### Plot Dual Surface Results #####
+source("r/header.R")
+
+dual_surface_results = readr::read_rds("objects/dual_surface_results.rds") |>
+  # add light column
+  dplyr::mutate(
+    light = dplyr::case_when(
+      str_detect(treatment, "HL") ~ "high",
+      str_detect(treatment, "LL") ~ "low"
+    ),
+    light = tidyr::replace_na(light, "medium")) |>
+  mutate(corr_diff = corr - mu_corr_sim) |>
+  ungroup() |>
+  mutate(light_reorder = fct_relevel(light, "low", "medium", "high"))
+
+# anova
+correlation_anova = aov(corr ~ light, data = dual_surface_results)
+summary(correlation_anova)
+
+stat.test.correlation <- aov(corr ~ light, data = dual_surface_results) |>
+  tukey_hsd()
+stat.test.correlation
+
+plot = ggplot(dual_surface_results, aes(light_reorder, corr)) +
+  stat_boxplot() +
+  geom_hline(yintercept = 0, linetype = "dashed") +
+  xlab("Light treatment") +
+  ylab(expression(paste("Correlation between NS", D^2))) +
+  theme_pubr()
+
+
+ggsave("dual-surface.pdf", plot = plot, path = "ms/figures",
+       width = 3, height = 3)
+

### 2023-09-19 r/08_fit-length-area.R

@@ -1,0 +1,15 @@
+source("r/header.R")
+
+stomata = readr::read_rds("objects/stomata_position_length_area.rds") |>
+  dplyr::ungroup() |>
+  filter(length > 5, !reject, !is.na(area)) |>
+  mutate(sqrt_area = sqrt(area))
+
+fit <- brm(bf(
+  sqrt_area ~ surface + length:light + (1 + surface:length|trt_leaf_number),
+  sigma ~ light
+),
+  data = stomata, backend = "cmdstanr", chains = 1L, seed = 226140189
+)
+
+write_rds(fit, "objects/fit.rds")

### 2023-09-19 r/09_plot-length-area.R

@@ -1,0 +1,48 @@
+source("r/header.R")
+
+fit <- read_rds("objects/fit.rds")
+
+summary(fit)
+
+df_new <- fit$data |>
+  dplyr::group_by(light, surface) |>
+  summarise(length_min = min(length), length_max = max(length)) |>
+  rowwise() |>
+  pmap_dfr(~{
+    tibble(light = ..1, surface = ..2, length = seq(..3, ..4, length.out = 1e2))
+  })
+
+df <- posterior_epred(fit, newdata = df_new, re_formula = NA)
+
+df_med <- df |>
+  apply(2, median) |>
+  as_tibble(.name_repair = "minimal") |>
+  rename(sqrt_area = value)
+
+df_qi <- df |>
+  apply(2, ggdist::qi) |>
+  t() |>
+  as_tibble(.name_repair = "minimal") |>
+  magrittr::set_colnames(c("lower", "upper"))
+
+df_pred <- df_new |>
+  bind_cols(df_med, df_qi) |>
+  arrange(light)
+
+plot = ggplot(
+  fit$data,
+  aes(length, sqrt_area, color = light, fill = light, linetype = surface)
+) +
+  geom_ribbon(data = df_pred, mapping = aes(ymin = lower, ymax = upper),
+              alpha = 0.35) +
+  geom_line(data = df_pred) +
+  geom_point(alpha = 0.2) +
+  scale_color_viridis_d() +
+  scale_fill_viridis_d() +
+  xlab("Stomata length [UNITS]") +
+  ylab(expression(paste(sqrt(Tessellation~zone), " [UNITS]"))) +
+  theme_pubr() +
+  theme(legend.position = "right", panel.background = element_rect(fill = "grey"))
+
+ggsave("length-area.pdf", plot = plot, path = "ms/figures",
+       width = 5, height = 4)

### 2023-09-19 r/functions.R

@@ -1,0 +1,1450 @@
+# Custom functions
+
+scientize <- function(x, threshold = -1L, digits = 2L) {
+
+  purrr::map_chr(x, function(.x, threshold, digits) {
+
+    oom <- log10(abs(.x))
+    if (oom < threshold) {
+      x1 <- .x |>
+        magrittr::multiply_by(10 ^ -floor(oom)) |>
+        round(digits)
+
+      x2 <- sprintf(glue::glue("%.{digits}f"), x1) |>
+        stringr::str_c(" \\times 10^{", floor(oom), "}")
+
+      return(x2)
+    } else {
+      x1 <- .x |>
+        signif(digits + 1L) |>
+        as.character()
+      return(x1)
+    }
+
+  }, threshold = threshold, digits = digits)
+
+}
+## Milestone 1: Simulate and plot stomatal grids ----
+
+### Functions for converting from D to U ----
+# D = stomatal density in 1 / pixel^2
+# U = interstomatal distance
+D2U <- function(D) {
+
+  checkmate::assert_numeric(D, lower = 0, finite = TRUE, any.missing = FALSE,
+                            min.len = 1L)
+  # D must have units of 1 / pixel ^ 2
+  # U = 2 * Rhex where Rhex is major radius of a regular hexagon
+  # Ahex = 3 * sqrt(3) / 2 * Rhex ^ 2
+
+  # Ahex is the area of the hexagon that each stomata would occupy in a
+  # perfectly dispersed array of stomata with density (D).
+  Ahex <- 1 / D
+
+  # Rhex is the outcircle radius Ahex calculated using equations which describe
+  # the radius of a regular hexagon
+  Rhex <- sqrt(2 * Ahex / (3 * sqrt(3)))
+
+  # a is the apothem, or incircle radius of a regular hexagon.
+  a <- (sqrt(3) / 2) * Rhex
+
+  # U is the distance between stomata assuming they are separated by two a.
+  U <- 2 * a # in pixels
+
+  # return U
+  return(U)
+
+}
+
+# a function that converts distance to density
+
+U2D <- function(U) {
+
+  # checks
+  checkmate::assert_numeric(U, lower = 0, finite = TRUE, any.missing = FALSE,
+                            min.len = 1L)
+  # U must have units of pixel
+  # a is the apothem or incircle radius of a regular hexagon
+  a <- U / 2
+
[…]

### 2023-09-19 r/header.R

@@ -1,0 +1,103 @@
+# Clear workspace
+rm(list = ls())
+graphics.off()
+
+# hello world
+# Source custom functions
+source("r/functions.R")
+
+# Load libraries
+
+library(arulesViz)
+library(brms)
+library(caret)
+library(cmdstanr)
+library(checkmate)
+library(corrplot)
+library(cowplot)
+library(deSolve)
+library(dplyr)
+library(forcats)
+library(ggdist)
+library(ggforce)
+library(ggimage)
+library(ggplot2)
+library(ggpubr)
+library(ggsignif)
+library(ggvoronoi)
+library(Hmisc)
+library(magrittr)
+library(maptools)
+library(metR)
+library(mgcv)
+library(parallel)
+library(progress)
+library(purrr)
+library(purrrlyr)
+library(RANN)
+library(raster)
+library(readr)
+library(rgdal)
+library(rlist)
+library(rootSolve)
+library(rstatix)
+library(sp)
+library(spatialEco)
+library(spatstat)
+library(stringr)
+library(stats)
+library(terra)
+library(tibble)
+library(tripack)
+library(units)
+
+# Fit Rubisco model
+x_depth = c(145, 830.5)
+y_depth = c(0, 700)
+x_rubisco = c(558, 56.5)
+y_rubisco = c(0, 100)
+nishio_carbon_1993_fig5 = read_csv("raw-data/nishio_carbon_1993_fig5.csv") |>
+  mutate(
+    depth = raw_depth * (diff(y_depth) / diff(x_depth)) - x_depth[1],
+    rubisco = raw_rubisco * (diff(y_rubisco) / diff(x_rubisco)) -
+      x_rubisco[1] * (diff(y_rubisco) / diff(x_rubisco)),
+    # Reversing so that 0 is at abaxial, 1 is at adaxial, to match Earles et al.
+    rel_depth = rev(depth / (min(depth) + max(depth)))
+  )
+
+fit_rubisco = gam(rubisco ~ s(rel_depth), data = nishio_carbon_1993_fig5)
+
+# Resolve conflicts
+# library(conflicted)
+# conflict_prefer("abbreviate", "arules")
+# conflict_prefer("write", "arules")
+# conflict_prefer("intersect", "dplyr")
+# conflict_prefer("recode", "dplyr")
+# conflict_prefer("setdiff", "dplyr")
+# conflict_prefer("setequal", "dplyr")
+# conflict_prefer("union", "dplyr")
[…]

### 2023-09-21 r/photo_2d_pm.R
starting to add photo_2d_pm() functions

@@ -1,0 +1,377 @@
+# Model photosynthesis using a 2D porous medium approximation
+#
+# The concept is that this file will contain all functions to run 2D porous
+# medium model and can be transitioned to a package script once it's ready
+
+# Note that n_z = 0 is bottom (abaxial) side of leaf
+
+# 2D photosynthesis model using a porous medium approximation
+photo_2d_pm = function(...) {
+
+  # Set parameters
+  params = set_2d_pm_param(...)
+
+  # Find solution
+  # Structure output
+
+}
+
+# Set parameters for photo_2d_pm()
+set_2d_pm_param = function(...) {
+
+}
+
+# Get set of parameters for photo_2d_pm()
+get_2d_pm_default_param = function(...) {
+
+  # Stomta spaced 100 um apart, 200 um thick leaf
+  # With hexagonal grid, this is stomatal density of:
+  # 2 / sqrt(3) / 100 ^ 2 = 0.0001154701 um^-2 = 115.4701 mm^-2
+  list(
+    n_x = 100,
+    n_z = 200,
+    t_elem = 1e-6,    # thickness of element [m]
+    Dz = 1.54e-5,     # Diffusivity of CO2 in air [m^2 / s]
+    por_int = 0.2,    # Porosity at interface [m^3 air / m^3 leaf]
+    por_spg = 0.3,    # Porosity of the spongy mesophyll [m^3 air / m^3 leaf]
+    por_pal = 0.1,    # Porosity of the palisade mesophyll [m^3 air / m^3 leaf]
+    tort = 1.55,      # Tortuosity of the palisade and spongy mesophyll [m / m]
+
+    # Note that 400 ppm = 0.01764706 is in mol / m^3, according to Earles et al. (2017)
+    # Implies 44.11765 mol air / m^3 - find eq for this based on temp and P
+    #
+    # THERE IS SOME ISSUE WITH THE WAY I AM CODING THE BOUNDARY LAYER. I SET C_S = 0.21
+    # AS A HACK TO GET RESULTS SIMILAR TO EARLES ET AL. THEY ASSUMED 0.015 (PG. 1093)
+    C_s = 0.015,      # CO2 concentration at stomate [mol / m^3]
+
[…]

### 2023-09-21 sync.sh

@@ -1,0 +1,2 @@
+# Backup on UW ResearchDrive
+rsync -r /Users/cdmuir/Documents/research/02_active/stomata-spacing /Volumes/cdmuir/research/02_active

### 2023-09-21 r/photo_2d_pm.R
working on r/hoto_2d_pm.R

@@ -31,7 +31,7 @@ get_2d_pm_default_param = function(...) {
     n_x = 100,
     n_z = 200,
     t_elem = 1e-6,    # thickness of element [m]
-    Dz = 1.54e-5,     # Diffusivity of CO2 in air [m^2 / s]
+    D_c = 1.54e-5,     # Diffusivity of CO2 in air [m^2 / s]
     por_int = 0.2,    # Porosity at interface [m^3 air / m^3 leaf]
     por_spg = 0.3,    # Porosity of the spongy mesophyll [m^3 air / m^3 leaf]
     por_pal = 0.1,    # Porosity of the palisade mesophyll [m^3 air / m^3 leaf]
@@ -49,7 +49,7 @@ get_2d_pm_default_param = function(...) {
     K_m = 18.7e-3,    # Rubisco effective Michaelis-Menten constant [mol / m^3]
     Gamma = 1.75e-3,  # CO2 compensation point [mol / m^3]
     Theta = 1,        # Curvature factor in FvCB model [1]
-    Rd = 0.066,       # Dark respiratory rate [mol / m^3 / s]
+    R_d = 0.066,       # Dark respiratory rate [mol / m^3 / s]
     Jmax = 275e-6,    # Maximum electron transport rate [mol / m^2 / s]
     Beta = 0.44,      # Fraction of light absorbed by PSII [mol / mol]
     Alpha = 0.72,     # Leaf level absorption [mol / mol]
@@ -244,11 +244,11 @@ bnd_z_top[param[["x_adaxial_stomata"]]] = param[["C_s"]]
 bnd_x_left = C_ias_mat[, 1]
 bnd_x_right = C_ias_mat[, param[["n_x"]]]
 
-Dx = 1 # placeholder for actual coef
+D_e = param[["D_c"]] * param[["por_mat"]] / param[["tort"]]
 
 # diffusion in Z-direction; boundaries = imposed concentration
 ## flux from bottom to top
-flux_z = -Dx * rbind(
+flux_z = -rbind(
   C_ias_mat[1, ] - bnd_z_bottom,
   (C_ias_mat[2:param[["n_z"]], ] - C_ias_mat[1:(param[["n_z"]] - 1), ]),
[…]

### 2023-09-22 r/photo_2d_pm.R
working on r/photo_2d_pm.R

@@ -8,21 +8,21 @@
 # 2D photosynthesis model using a porous medium approximation
 photo_2d_pm = function(...) {
 
-  # Set parameters
-  params = set_2d_pm_param(...)
+  # Set parmseters
+  parmss = set_2d_pm_parms(...)
 
   # Find solution
   # Structure output
 
 }
 
-# Set parameters for photo_2d_pm()
-set_2d_pm_param = function(...) {
+# Set parmseters for photo_2d_pm()
+set_2d_pm_parms = function(...) {
 
 }
 
-# Get set of parameters for photo_2d_pm()
-get_2d_pm_default_param = function(...) {
+# Get set of parmseters for photo_2d_pm()
+get_2d_pm_default_parms = function(...) {
 
   # Stomta spaced 100 um apart, 200 um thick leaf
   # With hexagonal grid, this is stomatal density of:
@@ -49,7 +49,7 @@ get_2d_pm_default_param = function(...) {
     K_m = 18.7e-3,    # Rubisco effective Michaelis-Menten constant [mol / m^3]
     Gamma = 1.75e-3,  # CO2 compensation point [mol / m^3]
     Theta = 1,        # Curvature factor in FvCB model [1]
-    R_d = 0.066,       # Dark respiratory rate [mol / m^3 / s]
+    r_d = 0.066,       # Dark respiratory rate [mol / m^3 / s]
     Jmax = 275e-6,    # Maximum electron transport rate [mol / m^2 / s]
     Beta = 0.44,      # Fraction of light absorbed by PSII [mol / mol]
     Alpha = 0.72,     # Leaf level absorption [mol / mol]
@@ -64,8 +64,8 @@ get_2d_pm_default_param = function(...) {
     # Quadratic parameters describing relative chlorophyll content as a function
     # of relative leaf depth (Johnson et al. [2005]; Borsuk and Brodersen [2019])
     b0_chl = 67.52,
-    b1_chl = 100 * 0.4149, # converts parameters to 0-1 rather than 0-100 scale
-    b2_chl = 100 ^ 2 * -0.0029, # converts parameters to 0-1 rather than 0-100 scale
[…]

### 2023-09-22 r/scratch-photo_2d_pm.R

@@ -1,0 +1,96 @@
+source("r/header.R")
+
+diffusion2D <- function(t, Y, par)   {
+
+  y    <- matrix(nrow = n_z, ncol = n_x, data = Y)  # vector to 2-D matrix
+
+  # hard-coding parameters for now
+  S_m = 20
+  Vstrom = 1.74e-6
+  k_c = 3
+  X_c = 2.5
+  K_m = 18.7e-3
+  Gamma = 1.75e-3
+
+  r_c = (k_c * X_c * y) / (K_m + y)        # carboxylation
+  r_d = 0.066
+  r_p = r_c * Gamma / y
+
+  # I'M NOT SURE THIS IS THE RIGHT WAY TO SCALE TO STROMA VOLUME
+  # 2e4 assumes 200 um thick leaf because 1 m^2 has volume of 2e-4 m^3
+  # S_m [m^2 meso / m^2 leaf] * 2e4 [m^2 leaf / m^3 leaf]
+  dY   <- -(r_c + r_p + r_d) * (S_m * 2e4) * Vstrom
+
+  bound_bottom <- y[1,] # boundary concentration
+  bound_top <- y[n_z,]   # boundary concentration
+  bound_left = y[,1]
+  bound_right = y[, n_x]
+  bound_top[1] = 0.015
+  bound_bottom[n_x] = 0.015
+
+  #diffusion in Z-direction; boundaries=imposed concentration
+  Flux <- -Dz * rbind(
+    y[1, ] - bound_bottom,
+    (y[2:n_z, ] - y[1:(n_z - 1), ]),
+    bound_top - y[n_z,]
+  ) / dz
+  dY   <- dY - (Flux[2:(n_z+1),]-Flux[1:n_z,])/dz
+
+  #diffusion in X-direction
+  Flux <- -Dx * cbind(
+    y[, 1] - bound_left,
+    (y[, 2:n_x] - y[, 1:(n_x - 1)]),
+    bound_right - y[, n_x]
+  ) / dx
+  dY    <- dY - (Flux[,2:(n_x+1)]-Flux[,1:n_x])/dx
+
+  return(list(as.vector(dY)))
+}
+
+# parameters
+n_x = 100
+n_z = 200
+dz    <- dx <- 1e-6   # grid size
+Dz    <- Dx <- 1.54e-5   # diffusion coeff, X- and Y-direction
+
+C_ias_mat  <- matrix(nrow = n_z, ncol = n_x, 0.1)
+
[…]

### 2023-09-22 r/scratch-photo_2d_pm.R
debuggin photo_2d_pm()

@@ -1,92 +1,108 @@
 source("r/header.R")
+library(tidyr)
 
 diffusion2D <- function(t, Y, par)   {
 
-  y    <- matrix(nrow = n_z, ncol = n_x, data = Y)  # vector to 2-D matrix
+  n = n_z * n_x
+  C_ias = matrix(nrow = n_z, ncol = n_x, data = Y[1:n])
+  C_liq = matrix(nrow = n_z, ncol = n_x, data = Y[(n + 1):(2 * n)])
+  dC_ias = dC_liq = numeric(length = n) # empty vectors
 
   # hard-coding parameters for now
+  D_c = 1.54e-5
+  por = 0.2
+  tort = 1.55
   S_m = 20
-  Vstrom = 1.74e-6
+  V_strom = 1.74e-6
   k_c = 3
   X_c = 2.5
   K_m = 18.7e-3
   Gamma = 1.75e-3
 
-  r_c = (k_c * X_c * y) / (K_m + y)        # carboxylation
+  g_liq = 0.25e-3
+  r_c = (k_c * X_c * C_liq) / (K_m + C_liq)        # carboxylation
   r_d = 0.066
-  r_p = r_c * Gamma / y
+  r_p = r_c * Gamma / C_liq
 
-  # I'M NOT SURE THIS IS THE RIGHT WAY TO SCALE TO STROMA VOLUME
-  # 2e4 assumes 200 um thick leaf because 1 m^2 has volume of 2e-4 m^3
-  # S_m [m^2 meso / m^2 leaf] * 2e4 [m^2 leaf / m^3 leaf]
-  dY   <- -(r_c + r_p + r_d) * (S_m * 2e4) * Vstrom
+  D_e = D_c * por / tort
 
-  bound_bottom <- y[1,] # boundary concentration
-  bound_top <- y[n_z,]   # boundary concentration
-  bound_left = y[,1]
-  bound_right = y[, n_x]
+  # FLUX ----
+  bound_bottom <- C_ias[1,] # boundary concentration
+  bound_top <- C_ias[n_z,]   # boundary concentration
+  bound_left = C_ias[,1]
+  bound_right = C_ias[, n_x]
   bound_top[1] = 0.015
   bound_bottom[n_x] = 0.015
 
-  #diffusion in Z-direction; boundaries=imposed concentration
-  Flux <- -Dz * rbind(
[…]

### 2023-09-23 r/header.R
some progress

@@ -48,6 +48,7 @@ library(stringr)
 library(stats)
 library(terra)
 library(tibble)
+library(tidyr)
 library(tripack)
 library(units)
 

### 2023-09-23 r/photo_2d_pm.R

@@ -293,3 +293,4 @@ rd_2p_pm = function(t, y, parms, ...) {
   return(list(c(dC_ias, dC_liq))) # return output from function
 
 }
+

### 2023-09-23 r/scratch-photo_2d_pm.R

@@ -1,5 +1,15 @@
+# I think I did this before, but calculating g_liq from Evans et al. 2009
+# Measured range of r_liq = 25-300 m^2 chloroplast s bar / mol
+# 1 s / m  = 0.025 m^2 chloroplast s bar / mol
+# SO, 1000 - 12000 s / m
+# SO g_liq = 0.001 - 8.33e-5
 source("r/header.R")
-library(tidyr)
+source("r/photo_2d_pm.R")
+
+parms = get_2d_pm_default_parms() |>
+  derive_2d_pm_parms(fit = fit_rubisco)
+
+parms$X_c_z
 
 diffusion2D <- function(t, Y, par)   {
 
@@ -18,9 +28,13 @@ diffusion2D <- function(t, Y, par)   {
   X_c = 2.5
   K_m = 18.7e-3
   Gamma = 1.75e-3
-
   g_liq = 0.25e-3
-  r_c = (k_c * X_c * C_liq) / (K_m + C_liq)        # carboxylation
+  J_max = 0.000275
+  j_max = J_max / (S_m * V_strom)
+
+  w_c = (k_c * X_c * C_liq) / (K_m + C_liq)        # carboxylation
+  w_j = C_liq * j_max / (4 * C_liq + 8 * Gamma)
+  r_c = pmin(w_c, w_j)
   r_d = 0.066
   r_p = r_c * Gamma / C_liq
 
@@ -50,22 +64,23 @@ diffusion2D <- function(t, Y, par)   {
   ) / dx
   dC_ias = dC_ias - (Flux[, 2:(n_x + 1)] - Flux[, 1:n_x]) / dx
 
-  dC_ias = dC_ias + g_liq * (C_liq - C_ias) / dx
+  dC_ias = dC_ias + g_liq * (C_liq - C_ias) / V_strom
 
   # CARBOXYLATION ----
   # I'M NOT SURE THIS IS THE RIGHT WAY TO SCALE TO STROMA VOLUME
   # 2e4 assumes 200 um thick leaf because 1 m^2 has volume of 2e-4 m^3
   # S_m [m^2 meso / m^2 leaf] * 2e4 [m^2 leaf / m^3 leaf]
-  dC_liq = dC_liq + g_liq * (C_ias - C_liq) / dx + (-r_c + r_p + r_d) * (S_m * 2e4) * V_strom
[…]

### 2023-09-25 ms/si.qmd
starting model description in SI

---
title: "2-D Photosynthesis Model"
format: docx
editor: visual
bibliography: ../references.bib
---

We model leaf photosynthesis using a two-dimensional porous medium approximation. The model is solved using a finite element method (FEM) in the *R* package **deSolve** [@soetaert2010]. Table 1 is a glossary model terms and symbols. Here we describe the model and associated *R* code.

## Leaf anatomy

We assume that the leaf is a homogenous 2-D medium. In the final version, we will incorporate differences in spongy and palisade porosity, gradients in light absorption, electron transport capacity, and Rubisco concentration. The mesophyll is $T_\text{leaf}$ thick and the stomata are regularly spaced apart by distance $U$ on both ab- and adaxial surfaces. In this scenario, we assume that the stomata on each surface are precisely offset from each other by distance $U/2$. This minimizes the average distance between any point in the mesophyll and its nearest stomate. Because of the regular spacing, we only need to model the region between a stomate on surface and the next stomate on the other surface (fig. 1). The rest of the mesophyll will be the same because of symmetry.

```{r, message = FALSE, fig.cap='Figure 1: Example leaf anatomy analyzed by the 2-D FEM.', echo = FALSE, fig.height=6, fig.width=3}

library(dplyr)
library(ggforce)
library(ggplot2)
library(tidyr)

# Example leaf where leaf thickness equals interstomatal distance
 xz_ratio = 0.5

  df_stomata = crossing(
    nesting(x = c(0, 1), y = c(0, 1) / xz_ratio),
    r = 0.04,
    x_offset = c(-1.5, 1.5)
  ) |>
    mutate(x0 = x + x_offset * r)

  df_epidermis = df_stomata |>
    summarise(stomata_min = min(x0 - r), stomata_max = max(x0 + r), .by = "y") |>
    mutate(
      xmin = ifelse(stomata_min < 0, stomata_max, 0),
      xmax = ifelse(stomata_max > 1, stomata_min, 1)
    )

  ggplot() +
    geom_circle(data = df_stomata, mapping = aes(x0 = x0, y0 = y, r = r)) +
    geom_segment(data = df_epidermis,
                 mapping = aes(
                   x = xmin,
                   xend = xmax,
                   y = y,
                   yend = y
                 ), linewidth = 1.5) +
    geom_text(
      data = df_stomata |>
        select(x, y) |>
        distinct(),
      mapping = aes(x, y, label = "stomate"),
      position = position_nudge(y = c(-0.05, 0.05) / xz_ratio)
    ) +
    geom_text(
      mapping = aes(x = 0.5, y = c(-0.1, 0.5, 1.1) / xz_ratio, label = c("abaxial", "mesophyll", "adaxial"))
    ) +
    geom_segment(mapping = aes(x = 0, y = -0.15 / xz_ratio, xend = 1, yend = -0.15 / xz_ratio),
                 arrow = arrow(angle = 90, length = unit(0.0125 / xz_ratio, "npc"), ends = "both") ) +
    geom_segment(mapping = aes(x = -0.15, y = 0 / xz_ratio, xend = -0.15, yend = 1 / xz_ratio),
                 arrow = arrow(angle = 90, length = unit(0.0125 / xz_ratio, "npc"), ends = "both") ) +
    geom_text(mapping = aes(x = 0.5, y = -0.2 / xz_ratio, label = "paste(italic(U) / 2, ', half the interstomatal distance [', mu, 'm]')"),
              parse = TRUE) +
    geom_text(mapping = aes(x = -0.2, y = 0.5 / xz_ratio, label = "paste(italic(T)[leaf], ', leaf thickness [', mu, 'm]')"),
              parse = TRUE, angle = 90) +
    coord_equal(clip = "off") +
    theme_void()

```

### 2023-09-25 r/functions.R

@@ -1448,3 +1448,59 @@ check_parameters = function(co2d_pars) {
   # no checks yet
   co2d_pars
 }
+
+
+
+# TESTING OUT FUNCTION TO DRAW LEAF FOR MODEL EXPLANATION
+
+plot_photo_2d_pm = function(parms, soln = NULL, ...) {
+
+  # scaling ratio of U / 2 to T_leaf
+  # assumes that dx == dz
+  xz_ratio = parms[["n_x"]] / parms[["n_z"]]
+
+  df_stomata = crossing(
+    nesting(x = c(0, 1), y = c(0, 1) / xz_ratio),
+    r = 0.04,
+    x_offset = c(-1.5, 1.5)
+  ) |>
+    mutate(x0 = x + x_offset * r)
+
+  df_epidermis = df_stomata |>
+    summarise(stomata_min = min(x0 - r), stomata_max = max(x0 + r), .by = "y") |>
+    mutate(
+      xmin = ifelse(stomata_min < 0, stomata_max, 0),
+      xmax = ifelse(stomata_max > 1, stomata_min, 1)
+    )
+
+  ggplot() +
+    geom_circle(data = df_stomata, mapping = aes(x0 = x0, y0 = y, r = r)) +
+    geom_segment(data = df_epidermis,
+                 mapping = aes(
+                   x = xmin,
+                   xend = xmax,
+                   y = y,
+                   yend = y
+                 ), linewidth = 1.5) +
+    geom_text(
+      data = df_stomata |>
+        select(x, y) |>
+        distinct(),
+      mapping = aes(x, y, label = "stomate"),
+      position = position_nudge(y = c(-0.05, 0.05) / xz_ratio)
+    ) +
+    geom_text(
+      mapping = aes(x = 0.5, y = c(-0.1, 0.5, 1.1) / xz_ratio, label = c("abaxial", "mesophyll", "adaxial"))
+    ) +
+    geom_segment(mapping = aes(x = 0, y = -0.15 / xz_ratio, xend = 1, yend = -0.15 / xz_ratio),
+                 arrow = arrow(angle = 90, length = unit(0.025 / xz_ratio, "npc"), ends = "both") ) +
+    geom_segment(mapping = aes(x = -0.15, y = 0 / xz_ratio, xend = -0.15, yend = 1 / xz_ratio),
+                 arrow = arrow(angle = 90, length = unit(0.025 / xz_ratio, "npc"), ends = "both") ) +
[…]

### 2023-09-25 r/photo_2d_pm.R

@@ -32,9 +32,9 @@ get_2d_pm_default_parms = function(...) {
     n_z = 200,
     t_elem = 1e-6,    # thickness of element [m]
     D_c = 1.54e-5,     # Diffusivity of CO2 in air [m^2 / s]
-    por_int = 0.2,    # Porosity at interface [m^3 air / m^3 leaf]
-    por_spg = 0.3,    # Porosity of the spongy mesophyll [m^3 air / m^3 leaf]
-    por_pal = 0.1,    # Porosity of the palisade mesophyll [m^3 air / m^3 leaf]
+    phi_int = 0.2,    # Porosity at interface [m^3 air / m^3 leaf]
+    phi_spg = 0.3,    # Porosity of the spongy mesophyll [m^3 air / m^3 leaf]
+    phi_pal = 0.1,    # Porosity of the palisade mesophyll [m^3 air / m^3 leaf]
     tort = 1.55,      # Tortuosity of the palisade and spongy mesophyll [m / m]
 
     # Note that 400 ppm = 0.01764706 is in mol / m^3, according to Earles et al. (2017)
@@ -95,9 +95,9 @@ derive_2d_pm_parms = function(parms, fit_rubisco, ...) {
   parms$Sm_z = rep(c(parms[["Sm_spg"]], parms[["Sm_pal"]]),
                  c(parms[["n_z_spg"]], parms[["n_z_pal"]]))
 
-  parms$por_z = rep(c(parms[["por_spg"]], parms[["por_pal"]]),
+  parms$phi_z = rep(c(parms[["phi_spg"]], parms[["phi_pal"]]),
                   c(parms[["n_z_spg"]], parms[["n_z_pal"]]))
-  parms$por_z[parms[["n_z_spg"]]] = parms$por_int
+  parms$phi_z[parms[["n_z_spg"]]] = parms$phi_int
 
   parms %<>%
     derive_2d_pm_rubisco(fit = fit_rubisco, ...)# |>
@@ -202,7 +202,7 @@ derive_2d_pm_j = function(parms, ...) {
 # Make matrices of porosity, Sm, X_c, j_e for photo_2d_pm()
 make_2d_pm_mat = function(parms, ...) {
 
-  parms$por_mat = matrix(parms[["por_z"]], nrow = parms[["n_z"]], ncol = parms[["n_x"]])
[…]

### 2023-09-25 r/scratch-photo_2d_pm.R

@@ -20,8 +20,8 @@ diffusion2D <- function(t, Y, par)   {
 
   # hard-coding parameters for now
   D_c = 1.54e-5
-  por = 0.2
-  tort = 1.55
+  phi = 0.2
+  tau = 1.55
   S_m = 20
   V_strom = 1.74e-6
   k_c = 3
@@ -38,7 +38,7 @@ diffusion2D <- function(t, Y, par)   {
   r_d = 0.066
   r_p = r_c * Gamma / C_liq
 
-  D_e = D_c * por / tort
+  D_e = D_c * phi / tau
 
   # FLUX ----
   bound_bottom <- C_ias[1,] # boundary concentration
@@ -46,7 +46,8 @@ diffusion2D <- function(t, Y, par)   {
   bound_left = C_ias[,1]
   bound_right = C_ias[, n_x]
   bound_top[1] = 0.015
-  bound_bottom[n_x] = 0.015
+  # bound_bottom[n_x] = 0.015
+  bound_bottom[1] = 0.015
 
   # diffusion in Z-direction; boundaries=imposed concentration
   Flux <- -D_e * rbind(
@@ -67,11 +68,18 @@ diffusion2D <- function(t, Y, par)   {
   dC_ias = dC_ias + g_liq * (C_liq - C_ias) / V_strom
 
   # CARBOXYLATION ----
+  # I *THINK* V_strom is the right thing to divide here because g_liq is
+  #  conductance per m^2 chloroplast area and that needs to be converted to
+  #  volume. Specifically, how diffusion from IAS to stroma changes
+  #  concentration
+  #
   # I'M NOT SURE THIS IS THE RIGHT WAY TO SCALE TO STROMA VOLUME
+  # r_c, r_p, and r_d are mol CO2 per stroma volume. Need to convert to per leaf volume.
+  # (mol CO2 / s / m^3 stroma) * (m^3 stroma / m^2 mesophyll) * (m^2 mesophyll * m^2 leaf) * (m^2 leaf / m^3 leaf)
   # 2e4 assumes 200 um thick leaf because 1 m^2 has volume of 2e-4 m^3
-  # S_m [m^2 meso / m^2 leaf] * 2e4 [m^2 leaf / m^3 leaf]
[…]

### 2023-09-25 r/photo_2d_pm.R
table of 2d-pm parameters

@@ -47,7 +47,7 @@ get_2d_pm_default_parms = function(...) {
     k_c = 3,          # Rubisco turnover rate [1 / s]
     X_c = 2.5,        # Rubisco concentration [mol / m^3]
     K_m = 18.7e-3,    # Rubisco effective Michaelis-Menten constant [mol / m^3]
-    Gamma = 1.75e-3,  # CO2 compensation point [mol / m^3]
+    gamma_star = 1.35e-3,  # CO2 compensation point [mol / m^3]
     Theta = 1,        # Curvature factor in FvCB model [1]
     r_d = 0.066,       # Dark respiratory rate [mol / m^3 / s]
     Jmax = 275e-6,    # Maximum electron transport rate [mol / m^2 / s]
@@ -214,7 +214,7 @@ make_2d_pm_mat = function(parms, ...) {
 calc_2d_pm_rc = function(C_liq_mat, parms, ...) {
 
   w_c = (parms[["k_c"]] * parms[["X_c_mat"]] * C_liq_mat) / (parms[["K_m"]] + C_liq_mat)
-  w_j = C_liq_mat * parms[["j_e_mat"]] / (4 * C_liq_mat + 8 * parms[["Gamma"]])
+  w_j = C_liq_mat * parms[["j_e_mat"]] / (4 * C_liq_mat + 8 * parms[["gamma_star"]])
 
   r_c = pmin(w_c, w_j)
 
@@ -279,7 +279,7 @@ rd_2p_pm = function(t, y, parms, ...) {
 
   r_c = calc_2d_pm_rc(C_liq_mat, parms, ...)
 
-  r_p = r_c * parms[["Gamma"]] / C_liq_mat
+  r_p = r_c * parms[["gamma_star"]] / C_liq_mat
 
   dC_ias = dC_liq = flux
   # dC_ias = (flux + parms[["g_liq"]] * (C_liq_mat - C_ias_mat) / parms[["t_elem"]]) /

### 2023-09-25 r/scratch-photo_2d_pm.R

@@ -27,16 +27,16 @@ diffusion2D <- function(t, Y, par)   {
   k_c = 3
   X_c = 2.5
   K_m = 18.7e-3
-  Gamma = 1.75e-3
+  gamma_star = 1.75e-3
   g_liq = 0.25e-3
   J_max = 0.000275
   j_max = J_max / (S_m * V_strom)
 
   w_c = (k_c * X_c * C_liq) / (K_m + C_liq)        # carboxylation
-  w_j = C_liq * j_max / (4 * C_liq + 8 * Gamma)
+  w_j = C_liq * j_max / (4 * C_liq + 8 * gamma_star)
   r_c = pmin(w_c, w_j)
   r_d = 0.066
-  r_p = r_c * Gamma / C_liq
+  r_p = r_c * gamma_star / C_liq
 
   D_e = D_c * phi / tau
 
@@ -119,16 +119,16 @@ V_strom = 1.74e-6
 k_c = 3
 X_c = 2.5
 K_m = 18.7e-3
-Gamma = 1.75e-3
+gamma_star = 1.35e-3
 
 J_max = 0.000275
 j_max = J_max / (S_m * V_strom)
 
 w_c = (k_c * X_c * df_C$C_liq) / (K_m + df_C$C_liq)        # carboxylation
-w_j = df_C$C_liq * j_max / (4 * df_C$C_liq + 8 * Gamma)
+w_j = df_C$C_liq * j_max / (4 * df_C$C_liq + 8 * gamma_star)
 r_c = pmin(w_c, w_j)
 r_d = 0.066
-r_p = r_c * Gamma / df_C$C_liq
+r_p = r_c * gamma_star / df_C$C_liq
 
 t_leaf = n_z * dz # [m]
 a_n = (r_c - r_p - r_d) * (S_m * 1 / t_leaf) * V_strom

### 2023-09-25 r/photo_2d_pm.R
more with default 2d-pm parameter values and functions

@@ -24,62 +24,68 @@ set_2d_pm_parms = function(...) {
 # Get set of parmseters for photo_2d_pm()
 get_2d_pm_default_parms = function(...) {
 
+  # OLD: migrate information to raw-data/2d-pm-parameteters.csv
   # Stomta spaced 100 um apart, 200 um thick leaf
   # With hexagonal grid, this is stomatal density of:
   # 2 / sqrt(3) / 100 ^ 2 = 0.0001154701 um^-2 = 115.4701 mm^-2
-  list(
-    n_x = 100,
-    n_z = 200,
-    t_elem = 1e-6,    # thickness of element [m]
-    D_c = 1.54e-5,     # Diffusivity of CO2 in air [m^2 / s]
-    phi_int = 0.2,    # Porosity at interface [m^3 air / m^3 leaf]
-    phi_spg = 0.3,    # Porosity of the spongy mesophyll [m^3 air / m^3 leaf]
-    phi_pal = 0.1,    # Porosity of the palisade mesophyll [m^3 air / m^3 leaf]
-    tort = 1.55,      # Tortuosity of the palisade and spongy mesophyll [m / m]
+  # list(
+  #   n_x = 100,
+  #   n_z = 200,
+  #   t_elem = 1e-6,    # thickness of element [m]
+  #   D_c = 1.54e-5,     # Diffusivity of CO2 in air [m^2 / s]
+  #   phi_int = 0.2,    # Porosity at interface [m^3 air / m^3 leaf]
+  #   phi_spg = 0.3,    # Porosity of the spongy mesophyll [m^3 air / m^3 leaf]
+  #   phi_pal = 0.1,    # Porosity of the palisade mesophyll [m^3 air / m^3 leaf]
+  #   tort = 1.55,      # Tortuosity of the palisade and spongy mesophyll [m / m]
+  #
+  #   # Note that 400 ppm = 0.01764706 is in mol / m^3, according to Earles et al. (2017)
+  #   # Implies 44.11765 mol air / m^3 - find eq for this based on temp and P
+  #   #
+  #   # THERE IS SOME ISSUE WITH THE WAY I AM CODING THE BOUNDARY LAYER. I SET C_S = 0.21
+  #   # AS A HACK TO GET RESULTS SIMILAR TO EARLES ET AL. THEY ASSUMED 0.015 (PG. 1093)
+  #   C_s = 0.015,      # CO2 concentration at stomate [mol / m^3]
+  #
+  #   k_c = 3,          # Rubisco turnover rate [1 / s]
[…]

### 2023-09-25 ms/ms.Rmd
updated bibliography

@@
 #  - endnotes
-bibliography: 8_17.bib
+bibliography: stomata-spacing.bib
 #csl: mycitationstyle.csl

### 2023-09-25 ms/si.qmd
updated bibliography

@@
 editor: visual
-bibliography: ../references.bib
+bibliography: ../ms/stomata-spacing.bib
 ---
@@
 
-```{r, message = FALSE, fig.cap='Figure 1: Example leaf anatomy analyzed by the 2-D FEM.', echo = FALSE, fig.height=6, fig.width=3}
+## Code
+
+I've copied the *R* code to set up the model and solve it if you want to copy and paste on your own machine. First, here are the packages you'll need.
+
+```{r, packages, message=FALSE}
 
@@
 library(ggplot2)
+library(readr)
 library(tidyr)
+
+```
+
+```{r, functions, echo=FALSE, message=FALSE}
+
+source("../r/functions.R")
+
+```
+
+```{r, message = FALSE, fig.cap='Figure 1: Example leaf anatomy analyzed by the 2-D FEM.', echo = FALSE, fig.height=6, fig.width=3}
 
@@
 ```
+
+```{r, glossary, echo = FALSE, message = FALSE}
+
+readr::read_csv("../raw-data/2d-pm-parameters.csv") |>
+  filter(!calculated) |>
+  mutate(Value = glue::glue("${x}$", x = scientize(default_value))) |>
+  select(Name = name, Symbol = symbol, Value, Units = units, Notes = notes) |>
+  knitr::kable()
+```
+
+## References

### 2023-09-26 ms/si.qmd
editing model description for feedback from Tom

@@
 We model leaf photosynthesis using a two-dimensional porous medium approximation.
-The model is solved using a finite element method (FEM) in the *R* package **deSolve** [@soetaert2010].
+The model is solved using a finite element method (FEM) in the *R* package **deSolve** [@soetaert_solving_2010].
 Table 1 is a glossary model terms and symbols.
@@
 
-## Code
-
-I've copied the *R* code to set up the model and solve it if you want to copy and paste on your own machine. First, here are the packages you'll need.
-
-```{r, packages, message=FALSE}
+```{r, packages1, echo = FALSE, message=FALSE}
 
@@
 library(ggplot2)
+library(magrittr)
 library(readr)
+library(rootSolve)
 library(tidyr)
@@
 
-```{r, message = FALSE, fig.cap='Figure 1: Example leaf anatomy analyzed by the 2-D FEM.', echo = FALSE, fig.height=6, fig.width=3}
+```{r, fig1, message = FALSE, fig.cap='Figure 1: Example leaf anatomy analyzed by the 2-D FEM.', echo = FALSE, fig.height=6, fig.width=3}
 
@@
 readr::read_csv("../raw-data/2d-pm-parameters.csv") |>
-  filter(!calculated) |>
+  # filter(!calculated) |>
   mutate(Value = glue::glue("${x}$", x = scientize(default_value))) |>
   select(Name = name, Symbol = symbol, Value, Units = units, Notes = notes) |>
-  knitr::kable()
+  knitr::kable(caption = "Table 1: glossary of model terms and mathematical symbols.")
+```
+
+## Solution
+
+Below I have copied the section from @earles_excess_2017 (pg.
+1094) on which I based my model.
+After that, I'll explain my modification for 2-D.
+I also think @earles_excess_2017 made one error that I will describe below.
+
+"We developed a FEM to solve a set of partial differential equations that describe CO$_2$ diffusion, photosynthesis, and respiration throughout the 1-D leaf geometry.
+The partial differential equations were solved for steady state using the R library **deSolve** [@soetaert_solving_2010].
+Specifically, the diffusive flux of CO$_2$ through the stomatal boundaries, intercellular airspace, and mesophyll cells was described by:
+
+$$D_\text{e} \frac{\partial^2 C_\text{ias}}{\partial x^2} = -f_\text{liq}$$ {#eq-1d_pm_flux}
+
+$$f_\text{liq} = r_\text{d} + r_\text{d} - r_\text{c}$$ {#eq-1d_pm_fliq}
+
+where
+
+$$
+D_\text{e} = \frac{\phi}{\tau} D_\text{c}
+$$ {#eq-De}
+
+is the effective diffusivity of a porous medium composed of a porous intercellular airspace with a given porosity ($\phi$; m$^3$ m$^{-3}$) and tortuosity ($\tau$; m m$^{-1}$), $D_\text{c}$\] is the diffusion coefficient (m s$^{-1}$) for CO$_2$ in the intercellular airspace, $C_\text{ias}$ is the \[CO$_2$\] (mol m$^{-3}$) at a depth $z$ in the intercellular airspace, $f_\text{liq}$ is the volumetric rate of CO$_2$ diffusion from the intercellular airspace into the chloroplast stroma (mol m$^{-3}$ s$^{-1}$), $r_\text{c}$ is the volumetric rate of ribulose 1,5-bisphosphate (RuBP) carboxylation (mol m$^{-3}$ s$^{-1}$), $r_\text{d}$ is the volumetric respiration rate (mol m$^{-3}$ s$^{-1}$), and $r_\text{p}$ is the volumetric photorespiration rate by Rubisco (mol m$^{-3}$ s$^{-1}$).
+
+The volumetric rate of CO$_2$ diffusion from the intercellular airspace into the chloroplast stroma, $f_\text{liq}$, is defined as:
+
+$$f_\text{liq} = \frac{g_\text{liq} (C_\text{liq} - C_\text{ias})}{l_z}$$ {#eq-fliq1}
+
+where $g_\text{liq}$ is the CO$_2$ conductance from the intercellular airspace into the chloroplast stroma (m s$^{-1}$), $C_\text{liq}$ (mol m$^{-3}$) is the \[CO$_2$\] in the stroma, and $l_z$ is the finite element length through which diffusion occurs (m)." (n.b.
+$l_z$ is the same as $t_\text{elem}$ in my Table 1)
+
+In the 2-D model, we extend the flux equation to $x$ (length) and $z$ (depth) dimensions:
+
+$$D_\text{e} \nabla^2 C_\text{ias} = D_\text{e} \bigg(\frac{\partial^2 C_\text{ias}}{\partial x^2} + \frac{\partial^2 C_\text{ias}}{\partial z^2}\bigg) = -f_\text{liq}$$ {#eq-2d_pm_flux}
+
+As I worked through @earles_excess_2017 model, it did not seem like dividing $g_\text{liq}$ by $l_z$ in equation @eq-fliq1 made sense.
+The FEM is a discretization of a continuous model.
+Intuitively, it does make sense why a coarser grid (greater $l_z$) would lead to a larger \[CO$_2$\] drawdown for a given assimilation rate.
+Noting that $g_\text{liq}$ is conductance per m$^2$ of stroma, I think the correct way is to divide by $V_\text{strom}$, which gives the mesophyll surface area per stroma volume available for liquid-phase diffusion.
+Here I assume that the chloroplasts line the entire inner wall of the mesophyll, so 1 [m$^2$] mesophyll = 1 [m$^2$] stroma, but this could be modified to the correct value.
+My equation is therefore:
+
+$$f_\text{liq} = \frac{g_\text{liq} (C_\text{liq} - C_\text{ias})}{V_\text{strom}}$$ {#eq-fliq2}
+
+Here's how I work out the units for equation @eq-fliq2.
+
+$$\frac{\text{mol}}{\text{m}^3~\text{s}} = \bigg(\frac{\text{m}^3}{\text{m}^2~\text{stroma}} \times \frac{1~\text{m}^2~\text{stroma}}{1~\text{m}^2~\text{meso}} \bigg) \times \bigg(\frac{\text{mol}}{\text{m}^3} \bigg) \times \bigg( \frac{\text{m}^2~\text{meso}}{\text{m}^3} \bigg)$$
+
+This still doesn't seem quite right to me because it seems like the change in airspace concentration should be affected by the porosity since a smaller volume of air will experience a greater drop in concentration for the same amount of assimilation, but maybe I am thinking about this incorrectly.
+
+Coincidentally, this had little effect on @earles_excess_2017 model because their element size was $\approx 1~\mu$m, which is quantitatively similar to $V_\text{strom} = 1.74 \times 10^{-6}$m.
+However, when you change the element size on their model, the solution is quite different; when you change the element size on my model, the solution is only slightly affected but the discretization is an approximation of the true solution.
+
+I calculated assimilation and respiration the same way as @earles_excess_2017 using the standard C\$\_3\$ biochemical model.
+
+The boundary conditions are that the CO$_2$ concentration in the substomatal cavity is constant at $C_\text{stom}$. The fluxes on the left and right sides are 0 because of symmetry.
+
+## R code
+
+I've copied the *R* code to set up the model and solve it if you want to copy and paste on your own machine. I annotated the *R* code you would need to run the model and include an example result.
+
+```{r, model}
+
+library(dplyr)
+library(ggforce)
+library(ggplot2)
+library(magrittr)
+library(readr)
+library(rootSolve)
+library(tidyr)
+
+source("../r/photo_2d_pm.R")
+
+# Define function for 2D FEM
+photo_2d_pm = function(t, Y, parms) {
+
+  # Arguments:
+  # * t, time step used by steady.2d() to find steady-state solution
+  # * Y, vector of length 2 * n_x * n_z. The first half are the elements
+  # corresponding to C_ias[i,j]; the second half are correspoding elements for
+  # C_liq[i,j].
+  # * parms, list of model parameters. Use `get_2d_pm_default_parms()` for
+  #  default parameter values.
+
+  # Set up:
+  # The leaf is divided to an area n_x elements wide, n_z units deep
+  n = parms[["n_z"]] * parms[["n_x"]]
+
+  # Create empty matrices for computation
+  C_ias = matrix(nrow = parms[["n_z"]], ncol = parms[["n_x"]],
+                 data = Y[1:n])
+  C_liq = matrix(nrow = parms[["n_z"]], ncol = parms[["n_x"]],
+                 data = Y[(n + 1):(2 * n)])
+  dC_ias = dC_liq = numeric(length = n) # empty vectors
+
+  # CARBOXYLATION AND RESPIRATION ----
+  # Calculate volumetric j_max from area-based J_max. Here, I assume a single
+  # j_max for every part of the leaf, but in the final model I will have a
+  # gradient of j_max following Earles et al. (2017).
+  j_max = parms[["J_max"]] / (parms[["S_m"]] * parms[["V_strom"]])
+
+  # Carboxylation (n.b. Rubisco concentration is assumed constant throughout
+  # the leaf, but in the final model I will have a garudent of X_c following
+  # Earles et al. (2017))
+  w_c = (parms[["k_c"]] * parms[["X_c"]] * C_liq) / (parms[["K_m"]] + C_liq)
+  w_j = C_liq * j_max / (4 * C_liq + 8 * parms[["gamma_star"]])
+  r_c = pmin(w_c, w_j)
+  r_d = parms[["r_d"]]
+  r_p = r_c * parms[["gamma_star"]] / C_liq
+
+  # FLUX ----
+  D_e = parms[["D_c"]] * parms[["phi"]] / parms[["tau"]]
+
+  # Boundary conditions
+  bound_bottom = C_ias[1,]
+  bound_top = C_ias[parms[["n_z"]],]
+  bound_left = C_ias[,1]
+  bound_right = C_ias[, parms[["n_x"]]]
+  bound_top[1] = parms[["C_stom"]]
+  bound_bottom[parms[["n_x"]]] = parms[["C_stom"]]
+  # bound_bottom[1] = parms[["C_stom"]]
+
+  # diffusion in Z-direction
+  Flux = -D_e / parms[["t_elem"]] * rbind(
+    C_ias[1, ] - bound_bottom,
+    (C_ias[2:parms[["n_z"]], ] - C_ias[1:(parms[["n_z"]] - 1), ]),
+    bound_top - C_ias[parms[["n_z"]], ]
+  )
+  dC_ias = dC_ias -
+    (Flux[2:(parms[["n_z"]] + 1), ] - Flux[1:parms[["n_z"]], ]) / parms[["t_elem"]]
+
+  # diffusion in X-direction
+  Flux = -D_e / parms[["t_elem"]] * cbind(
+    C_ias[, 1] - bound_left,
+    (C_ias[, 2:parms[["n_x"]]] - C_ias[, 1:(parms[["n_x"]] - 1)]),
+    bound_right - C_ias[, parms[["n_x"]]]
+  )
+  dC_ias = dC_ias -
+    (Flux[, 2:(parms[["n_x"]] + 1)] - Flux[, 1:parms[["n_x"]]]) /
+    parms[["t_elem"]]
+
+  dC_ias = dC_ias + parms[["g_liq"]] * (C_liq - C_ias) / parms[["V_strom"]]
+
+  # CARBOXYLATION ----
+  # Leaf thickness - needed to scale from stroma volume per leaf volume
+  T_leaf = parms[["n_z"]] * parms[["t_elem"]] # [m]
+  dC_liq = dC_liq + parms[["g_liq"]] * (C_ias - C_liq) / parms[["V_strom"]] +
+    (-r_c + r_p + r_d) * (parms[["S_m"]] * 1 / T_leaf) * parms[["V_strom"]]
+
+  return(list(c(dC_ias, dC_liq)))
+
+}
+
+# Model parameters
+parms = get_2d_pm_default_parms() |>
+  derive_2d_pm_parms()
+
+# Initial values
+C_ias_mat = matrix(nrow = parms[["n_z"]], ncol = parms[["n_x"]], parms[["C_stom"]])
+C_liq_mat = matrix(nrow = parms[["n_z"]], ncol = parms[["n_x"]], parms[["C_stom"]])
+
+# Solve for C_ias and C_liq
+soln = steady.2D(c(C_ias_mat, C_liq_mat), func = photo_2d_pm, parms = parms,
+                 pos = FALSE, dimens = c(parms[["n_z"]], parms[["n_x"]]),
+                 nspec = 2, lrw = 1e8, atol = 1e-10, rtol = 1e-10, ctol = 1e-10)
+
+# Plot results
+df_C = expand.grid(
+  z = seq_len(parms[["n_z"]]),
+  x = seq_len(parms[["n_x"]]),
+  name = c("C_ias", "C_liq")
+) |>
+  mutate(value = soln$y)
+
+ggplot(df_C, aes(x, z, z = value)) +
+  facet_grid(~ name) +
+  geom_contour_filled() +
+  coord_equal()
+
+# Calculate area-based net photosynthesis
+C_liq = df_C |>
+  filter(name == "C_liq") |>
+  pull(value)
+
+j_max = parms[["J_max"]] / (parms[["S_m"]] * parms[["V_strom"]])
+
+w_c = (parms[["k_c"]] * parms[["X_c"]] * C_liq) / (parms[["K_m"]] + C_liq)
+w_j = C_liq * j_max / (4 * C_liq + 8 * parms[["gamma_star"]])
+r_c = pmin(w_c, w_j)
+r_d = parms[["r_d"]]
+r_p = r_c * parms[["gamma_star"]] / C_liq
+
+T_leaf = parms[["n_z"]] * parms[["t_elem"]]
+a_n = (r_c - r_p - r_d) * (parms[["S_m"]] * 1 / T_leaf) * parms[["V_strom"]]
+
+# 1 m^2 of 200 um thick leaf is 2e-04 m^3
+mean(a_n) * T_leaf * 1e6
+
 ```

### 2023-09-26 r/functions.R

@@ -4,6 +4,7 @@ scientize <- function(x, threshold = -1L, digits = 2L) {
 
   purrr::map_chr(x, function(.x, threshold, digits) {
 
+    if (is.na(.x)) {return(NA)}
     oom <- log10(abs(.x))
     if (oom < threshold) {
       x1 <- .x |>

### 2023-09-26 r/photo_2d_pm.R

@@ -16,7 +16,7 @@ photo_2d_pm = function(...) {
 
 }
 
-# Set parmseters for photo_2d_pm()
+# Set parameters for photo_2d_pm()
 set_2d_pm_parms = function(...) {
 
 }
@@ -82,7 +82,7 @@ get_2d_pm_default_parms = function(...) {
   #
   # )
 
-  read_csv("raw-data/2d-pm-parameters.csv", col_types = "cccdccl") |>
+  read_csv("../raw-data/2d-pm-parameters.csv", col_types = "cccdccl") |>
     filter(!calculated) |>
     select(r, default_value) |>
     with(split(default_value, r))
@@ -92,21 +92,22 @@ get_2d_pm_default_parms = function(...) {
 # Calculate derived parmseters for photo_2d_pm()
 derive_2d_pm_parms = function(parms, fit_rubisco, ...) {
 
-  # parms = get_2d_pm_default_parms()
+  # COMMENTING OUT PARTS I DON'T NEED FOR SIMPLER MODEL
+  #
   # derived quantities
-  parms$t_leaf = parms[["n_z"]] * parms[["t_elem"]] # leaf thickness [m]
-  parms$frac_spg = 1 - parms[["frac_pal"]] # Fraction spongy mesophyll
-  parms$n_z_pal = round(parms[["n_z"]] * parms[["frac_pal"]])
-  parms$n_z_spg = parms[["n_z"]] - parms[["n_z_pal"]]
-  parms$Sm_z = rep(c(parms[["Sm_spg"]], parms[["Sm_pal"]]),
-                 c(parms[["n_z_spg"]], parms[["n_z_pal"]]))
+  parms$T_leaf = parms[["n_z"]] * parms[["t_elem"]] # leaf thickness [m]
+  # parms$frac_spg = 1 - parms[["frac_pal"]] # Fraction spongy mesophyll
+  # parms$n_z_pal = round(parms[["n_z"]] * parms[["frac_pal"]])
+  # parms$n_z_spg = parms[["n_z"]] - parms[["n_z_pal"]]
+  # parms$Sm_z = rep(c(parms[["Sm_spg"]], parms[["Sm_pal"]]),
+  #                c(parms[["n_z_spg"]], parms[["n_z_pal"]]))
+  #
+  # parms$phi_z = rep(c(parms[["phi_spg"]], parms[["phi_pal"]]),
[…]

### 2023-09-26 r/scratch-photo_2d_pm.R

@@ -5,67 +5,89 @@
 # SO g_liq = 0.001 - 8.33e-5
 source("r/header.R")
 source("r/photo_2d_pm.R")
+library(rootSolve)
 
-parms = get_2d_pm_default_parms() |>
-  derive_2d_pm_parms(fit = fit_rubisco)
+photo_2d_pm = function(t, Y, parms) {
 
-parms$X_c_z
+  # Arguments:
+  # * t, time step used by steady.2d() to find steady-state solution
+  # * Y, vector of length 2 * n_x * n_z. The first half are the elements
+  # corresponding to C_ias[i,j]; the second half are correspoding elements for
+  # C_liq[i,j].
+  # * parms, list of model parameters. Use `get_2d_pm_default_parms()` for
+  #  default parameter values.
 
-diffusion2D <- function(t, Y, par)   {
+  # Set up:
+  # The leaf is divided to an area n_x elements wide, n_z units deep
+  n = parms[["n_z"]] * parms[["n_x"]]
 
-  n = n_z * n_x
-  C_ias = matrix(nrow = n_z, ncol = n_x, data = Y[1:n])
-  C_liq = matrix(nrow = n_z, ncol = n_x, data = Y[(n + 1):(2 * n)])
+  # Create empty matrices for computation
+  C_ias = matrix(nrow = parms[["n_z"]], ncol = parms[["n_x"]],
+                 data = Y[1:n])
+  C_liq = matrix(nrow = parms[["n_z"]], ncol = parms[["n_x"]],
+                 data = Y[(n + 1):(2 * n)])
   dC_ias = dC_liq = numeric(length = n) # empty vectors
 
+  # CARBOXYLATION AND RESPIRATION ----
+  # Calculate volumetric j_max from area-based J_max. Here, I assume a single
+  # j_max for every part of the leaf, but in the final model I will have a
+  # gradient of j_max following Earles et al. (2017).
+  #
   # hard-coding parameters for now
-  D_c = 1.54e-5
-  phi = 0.2
-  tau = 1.55
-  S_m = 20
-  V_strom = 1.74e-6
-  k_c = 3
[…]

### 2023-09-26 ms/ms.Rmd
add new bibliography

@@
 #      `corresponding_author` and `acknowledgements`
-title: Spatial Coordination Of Stomatal Patterning Between Leaf Surfaces In Amphistomatous *Arabidopsis thaliana* Incurrs No Photosynthetic Advantage
+title: Spatial Coordination Of Stomatal Patterning Between Leaf Surfaces In Amphistomatous *Arabidopsis thaliana* Incurs No Photosynthetic Advantage
 #date: "`r Sys.Date()`"

### 2023-09-26 ms/si.qmd
corrected f_liq calculation following Tom's email on 2023-09-26

@@
 Intuitively, it does make sense why a coarser grid (greater $l_z$) would lead to a larger \[CO$_2$\] drawdown for a given assimilation rate.
-Noting that $g_\text{liq}$ is conductance per m$^2$ of stroma, I think the correct way is to divide by $V_\text{strom}$, which gives the mesophyll surface area per stroma volume available for liquid-phase diffusion.
-Here I assume that the chloroplasts line the entire inner wall of the mesophyll, so 1 [m$^2$] mesophyll = 1 [m$^2$] stroma, but this could be modified to the correct value.
-My equation is therefore:
+Noting that $g_\text{liq}$ is conductance per m$^2$ of stroma, this means the length scale to divide by should be 1/(stroma area per unit bulk leaf volume).
+This is equivalent to 1/\[(stroma area per leaf area)x(leaf area per bulk leaf volume)\] = 1/\[Sc x (1/Tleaf)\] = Tleaf/Sc (for now I am assuming Sm = Sc, but we can correct this later).
+Our equation is therefore:
 
-$$f_\text{liq} = \frac{g_\text{liq} (C_\text{liq} - C_\text{ias})}{V_\text{strom}}$$ {#eq-fliq2}
-
-Here's how I work out the units for equation @eq-fliq2.
-
-$$\frac{\text{mol}}{\text{m}^3~\text{s}} = \bigg(\frac{\text{m}^3}{\text{m}^2~\text{stroma}} \times \frac{1~\text{m}^2~\text{stroma}}{1~\text{m}^2~\text{meso}} \bigg) \times \bigg(\frac{\text{mol}}{\text{m}^3} \bigg) \times \bigg( \frac{\text{m}^2~\text{meso}}{\text{m}^3} \bigg)$$
-
-This still doesn't seem quite right to me because it seems like the change in airspace concentration should be affected by the porosity since a smaller volume of air will experience a greater drop in concentration for the same amount of assimilation, but maybe I am thinking about this incorrectly.
-
-Coincidentally, this had little effect on @earles_excess_2017 model because their element size was $\approx 1~\mu$m, which is quantitatively similar to $V_\text{strom} = 1.74 \times 10^{-6}$m.
-However, when you change the element size on their model, the solution is quite different; when you change the element size on my model, the solution is only slightly affected but the discretization is an approximation of the true solution.
+$$f_\text{liq} = \frac{g_\text{liq} (C_\text{liq} - C_\text{ias})}{T_\text{leaf} / S_\text{m}}$$ {#eq-fliq2}
 
@@
 
-  # CARBOXYLATION AND RESPIRATION ----
-  # Calculate volumetric j_max from area-based J_max. Here, I assume a single
-  # j_max for every part of the leaf, but in the final model I will have a
-  # gradient of j_max following Earles et al. (2017).
-  j_max = parms[["J_max"]] / (parms[["S_m"]] * parms[["V_strom"]])
-
-  # Carboxylation (n.b. Rubisco concentration is assumed constant throughout
-  # the leaf, but in the final model I will have a garudent of X_c following
-  # Earles et al. (2017))
-  w_c = (parms[["k_c"]] * parms[["X_c"]] * C_liq) / (parms[["K_m"]] + C_liq)
-  w_j = C_liq * j_max / (4 * C_liq + 8 * parms[["gamma_star"]])
-  r_c = pmin(w_c, w_j)
-  r_d = parms[["r_d"]]
-  r_p = r_c * parms[["gamma_star"]] / C_liq
-
     # FLUX ----
@@
 
-  # diffusion in Z-direction
+  # diffusion in Z-direction; boundaries=imposed concentration
   Flux = -D_e / parms[["t_elem"]] * rbind(
@@
 
-  dC_ias = dC_ias + parms[["g_liq"]] * (C_liq - C_ias) / parms[["V_strom"]]
+  dC_ias = dC_ias + parms[["g_liq"]] * (C_liq - C_ias) /
+    (parms[["T_leaf"]] / parms[["S_m"]])
 
-  # CARBOXYLATION ----
-  # Leaf thickness - needed to scale from stroma volume per leaf volume
-  T_leaf = parms[["n_z"]] * parms[["t_elem"]] # [m]
-  dC_liq = dC_liq + parms[["g_liq"]] * (C_ias - C_liq) / parms[["V_strom"]] +
-    (-r_c + r_p + r_d) * (parms[["S_m"]] * 1 / T_leaf) * parms[["V_strom"]]
+  # CARBOXYLATION AND RESPIRATION ----
+  # Calculate volumetric j_max from area-based J_max. Here, I assume a single
+  # j_max for every part of the leaf, but in the final model I will have a
+  # gradient of j_max following Earles et al. (2017).
+  j_max = parms[["J_max"]] / (parms[["S_m"]] * parms[["V_strom"]])
+
+  # Carboxylation (n.b. Rubisco concentration is assumed constant throughout
+  # the leaf, but in the final model I will have a garudent of X_c following
+  # Earles et al. (2017))
+  w_c = (parms[["k_c"]] * parms[["X_c"]] * C_liq) / (parms[["K_m"]] + C_liq)
+  w_j = C_liq * j_max / (4 * C_liq + 8 * parms[["gamma_star"]])
+  r_c = pmin(w_c, w_j)
+  r_d = parms[["r_d"]]
+  r_p = r_c * parms[["gamma_star"]] / C_liq
+
+  dC_liq = dC_liq +
+    parms[["g_liq"]] * (C_ias - C_liq) / (parms[["T_leaf"]] / parms[["S_m"]]) +
+    (-r_c + r_p + r_d) * (parms[["S_m"]] / parms[["T_leaf"]]) * parms[["V_strom"]]
 

### 2023-09-26 r/photo_2d_pm.R

@@ -82,7 +82,8 @@ get_2d_pm_default_parms = function(...) {
   #
   # )
 
-  read_csv("../raw-data/2d-pm-parameters.csv", col_types = "cccdccl") |>
+  # read_csv("raw-data/2d-pm-parameters.csv", col_types = "cccdccl") |>
+    read_csv("../raw-data/2d-pm-parameters.csv", col_types = "cccdccl") |>
     filter(!calculated) |>
     select(r, default_value) |>
     with(split(default_value, r))

### 2023-09-26 r/scratch-photo_2d_pm.R

@@ -3,6 +3,7 @@
 # 1 s / m  = 0.025 m^2 chloroplast s bar / mol
 # SO, 1000 - 12000 s / m
 # SO g_liq = 0.001 - 8.33e-5
+
 source("r/header.R")
 source("r/photo_2d_pm.R")
 library(rootSolve)
@@ -27,34 +28,6 @@ photo_2d_pm = function(t, Y, parms) {
   C_liq = matrix(nrow = parms[["n_z"]], ncol = parms[["n_x"]],
                  data = Y[(n + 1):(2 * n)])
   dC_ias = dC_liq = numeric(length = n) # empty vectors
-
-  # CARBOXYLATION AND RESPIRATION ----
-  # Calculate volumetric j_max from area-based J_max. Here, I assume a single
-  # j_max for every part of the leaf, but in the final model I will have a
-  # gradient of j_max following Earles et al. (2017).
-  #
-  # hard-coding parameters for now
-  # D_c = 1.54e-5
-  # phi = 0.2
-  # tau = 1.55
-  # S_m = 20
-  # V_strom = 1.74e-6
-  # k_c = 3
-  # X_c = 2.5
-  # K_m = 18.7e-3
-  # gamma_star = 1.75e-3
-  # g_liq = 0.25e-3
-  # J_max = 0.000275
-  j_max = parms[["J_max"]] / (parms[["S_m"]] * parms[["V_strom"]])
-
-  # Carboxylation (n.b. Rubisco concentration is assumed constant throughout
-  # the leaf, but in the final model I will have a garudent of X_c following
-  # Earles et al. (2017))
-  w_c = (parms[["k_c"]] * parms[["X_c"]] * C_liq) / (parms[["K_m"]] + C_liq)
-  w_j = C_liq * j_max / (4 * C_liq + 8 * parms[["gamma_star"]])
-  r_c = pmin(w_c, w_j)
-  r_d = parms[["r_d"]]
-  r_p = r_c * parms[["gamma_star"]] / C_liq
 
   # FLUX ----
   D_e = parms[["D_c"]] * parms[["phi"]] / parms[["tau"]]
@@ -87,23 +60,33 @@ photo_2d_pm = function(t, Y, parms) {
     (Flux[, 2:(parms[["n_x"]] + 1)] - Flux[, 1:parms[["n_x"]]]) /
     parms[["t_elem"]]
 
[…]

### 2023-09-27 ms/ms.Rmd
changed latex engine to xelatex

@@
 link-citations: yes
-output: rticles::oup_article
+output: 
+  rticles::oup_article:
+    latex_engine: xelatex
 #urlcolor: orange

### 2023-09-28 ms/ms.Rmd
edited introduction

--- 2023-09-26 ms/ms.Rmd
@@
 keywords:
-  - stomatal patterning
-  - Aribidopsis thaliana
-  - amphistomatous leaves
-  - photosynthesis modeling
-  - optimizing photosynthesis
+  - amphistomy
+  - Arabidopsis thaliana
   - CO$_2$ diffusion
+  - finite element method
+  - optimality
+  - photosynthesis
+  - stomata
 fontsize: 12pt
@@
 Stomatal anatomy (e.g.
-size, density, distribution, and patterning) and physiological responses to the environment regulate gas exchange during photosynthesis; namely CO$_2$ assimilation and water loss through transpiration.
-Since waxy cuticles are mostly impermeable to CO$_2$ and $H_2O$, stomata are discrete points through which gas exchange occurs and account only for a small percentage of the leaf area [@lange_responses_1971].
-They consist of two guard cells which open and close upon changes in turgor pressure or hormonal cues [@mcadam_linking_2016].
-The stomatal pore leads to an internal space known as the substomatal cavity where gases (and pathogens) contact the mesophyll [@muir_stomatal_2020].
+size, density, distribution, and patterning) and movement regulate gas exchange during photosynthesis, namely CO$_2$ assimilation and water loss through transpiration.
+Since waxy cuticles are mostly impermeable to CO$_2$ and H$_2$O, stomata are the primary entry points through which gas exchange occurs despite making up a small percentage of the leaf area [@lange_responses_1971].
+Stomata consist of two guard cells which open and close upon changes in turgor pressure or hormonal cues [@mcadam_linking_2016].
+The stomatal pore leads to an internal space known as the substomatal cavity where gases contact the mesophyll.
 Once in the mesophyll, CO$_2$ diffuses throughout a network of intercellular air space (IAS) and into mesophyll cells where CO$_2$ assimilation ($A$) occurs within the chloroplasts [@lee_diffusion_1964].
-Maximum conductance ($g_\text{max}$) and transpiration are determined by numerous environmental and anatomical parameters such as: vapor pressure deficit (VPD), irradiance, temperature, wind speed, leaf water potential, IAS geometry, mesophyll cell anatomy, and stomatal anatomy.
+Stomatal conductance and transpiration are determined by numerous environmental and anatomical parameters such as vapor pressure deficit (VPD), irradiance, temperature, wind speed, leaf water potential, IAS geometry, mesophyll cell anatomy, and stomatal anatomy.
 
-Natural selection should optimize the anatomy and physiology of stomata to maximize CO$_2$ gain per unit of water loss, thereby maximizing assimilation rates for a given set of hydraulic constraints and light environments [@cowan_stomatal_1977; @buckley_optimal_2017; @sperry_predicting_2017].
-However, optimal stomatal anatomy and physiology depends on developmental and environmental constraints and other physiological co-variates [@croxdale_stomatal_2000; @harrison_influence_2020; MUIR IN PRESS].
-There is a wide variety of stomatal traits in nature, but certain common features as well.
+Many successful predictions about stomata and other leaf traits can be made by hypothesizing that natural selection should optimizes CO$_2$ gain per unit of water loss [@cowan_stomatal_1977; @buckley_optimal_2017; @sperry_predicting_2017].
+However, stomatal anatomy may be partially constrained by physical and developments limits on phenotypic expression [@croxdale_stomatal_2000; @harrison_influence_2020; @muir_how_2023].
+Sometimes optimization leads to similar phenotypes across many disparate species.
 For example, almost all stomata follow the one cell spacing rule to maintain proper stomatal functioning [@geisler_oriented_2000; @dow_physiological_2014]; however some species (notably in *Begonia*) appear to benefit from overlapping vapor shells caused by stomatal clustering [@yi_gan_stomatal_2010; @lehmann_effects_2015; @papanatsiou_stomatal_2017].
-Stomatal density positively co-varies with irradiance during leaf development and negatively co-varies with CO$_2$ concentration [@gay_influence_1975; @schoch_dependence_1980; @woodward_stomatal_1987; @royer_stomatal_2001].
+Stomatal traits also vary adaptively in different environments.
+Stomatal density positively co-varies with irradiance during leaf development and negatively co-varies with CO$_2$ concentration [@gay_influence_1975; @schoch_dependence_1980; @woodward_stomatal_1987; @royer_stomatal_2001], consistent with optimality predictions.
 Stomatal size is jointly controlled by genome size, light, and stomatal density [@jordan_environmental_2015].
 Size positively co-varies with genome size [@roddy_scaling_2020] and negatively co-varies with stomatal density [@camargo_density_2011].
-Total stomatal area (size * density) is optimized for operational conductance ($gs_\text{op}$) rather than maximum conductance ($gs_\text{max}$) such that stomatal apertures are most responsive to changes in the environment at their operational aperture [@franks_physiological_2012; @liu_scaling_2021].
-Stomatal aperture can compensate for maladaptive stomatal densities to an extent [@bussis_stomatal_2006], but stomatal density and size ultimately determine a leaf's theoretical $gs_\text{max}$ [@sack_developmental_2016].
+Total stomatal area (size $\times$ density) is optimized for operational conductance ($gs_\text{s,op}$) rather than maximum conductance ($gs_\text{s,max}$) such that stomatal apertures are most responsive to changes in the environment at their operational aperture [@franks_physiological_2012; @liu_scaling_2021].
+Stomatal aperture can compensate for maladaptive stomatal densities to an extent [@bussis_stomatal_2006], but stomatal density and size ultimately determine a leaf's theoretical $gs_\text{s,max}$ [@sack_developmental_2016], which is proportional to $gs_\text{s,op}$ [@murray_consistent_2020].
 Additionally, low stomatal densities lead to irregular and insufficient CO$_2$ supply and reduced photosynthetic efficiency in areas far from stomata [@pieruschka_lateral_2006; @morison_lateral_2005], while high stomatal densities can reduce water use efficiency (WUE) [@bussis_stomatal_2006] and incur excessive metabolic costs [@deans_optimization_2020].
 In most species, stomata occur on the abaxial (usually lower) leaf surface; but amphistomy, the occurrence of stomata on both abaxial and adaxial leaf surfaces, is also prevalent in high light environments with constant or intermittent access to sufficient water [@mott_adaptive_1982; @jordan_using_2014; @muir_light_2018; @drake_two_2019; @muir_is_2019].
-Amphistomy effectively halves CO$_2$ diffusion path length and boundary layer resistance by doubling boundary layer area [@parkhurst_adaptive_1978; @harrison_influence_2020; @mott_amphistomy_1991].
+Amphistomy effectively halves the CO$_2$ diffusion path length and boundary layer resistance by doubling boundary layer conductance [@parkhurst_adaptive_1978; @harrison_influence_2020; @mott_amphistomy_1991].
 Historically, stomatal patterning in dicot angiosperms was thought to be random with an exclusionary distance surrounding each stomate [@sachs_developmental_1974]; however, the developmental controls of stomatal patterning are poorly understood and likely more complex than random development along the leaf surface.
@@
 The patterning and spacing of stomata on the leaf affects photosynthesis in $C_3$ leaves by altering the CO$_2$ diffusion path length from stomata to sites of carboxylation in the mesophyll.
-Maximum photosynthetic rate ($A_\text{max}$) in $C_3$ plants is generally co-limited by biochemistry and diffusion, but limitations depend largely on light availability [@parkhurst_intercellular_1990; @manter_ci_2004; @carriqui_diffusional_2015].
-In low light environments, assimilation rates are low; consequently, CO$_2$ flux is low and the internal CO$_2$ concentration ($C_\text{i}$) remains high; thus $A_\text{max}$ is constrained by biochemical limitations in low light [@kaiser_metabolic_2016].
-For well hydrated leaves in high light, photosynthesis is often limited by CO$_2$ supply as resistances from the boundary layer, stomatal pore, and mesophyll can result in insufficient $C_\text{i}$ to supply efficient photosynthesis at the chloroplast [@farquhar_biochemical_1980; lehmeier_cell_2017].
-In this study, we focus primarily on how stomatal patterning affects diffusion.
+Maximum photosynthetic rate ($A_\text{max}$) in $C_3$ plants is generally co-limited by biochemistry and diffusion, but modulated by light availability [@parkhurst_intercellular_1990; @manter_ci_2004; @carriqui_diffusional_2015].
+Low light decreasing CO$_2$ demand by limiting electron transport rate, leading to relatively high internal CO$_2$ concentration ($C_\text{i}$) and low $A_\text{max}$ [@kaiser_metabolic_2016].
+In contrast, well hydrated leaves with open stomata in high light, photosynthesis is often limited by CO$_2$ supply as resistances from the boundary layer, stomatal pore, and mesophyll can result in insufficient CO$C_2$ supply at the chloroplast to maxmimize photosynthesis [@farquhar_biochemical_1980; @lehmeier_cell_2017].
+In this study, we focus primarily on how stomatal patterning affects diffusion, ignoring boundary layer and mesophyll resistances.
 
-To maximize CO$_2$ supply from the stomatal pore to chloroplasts, stomata should be uniformly distributed in an equilateral triangular grid on the leaf surface so as to minimize stomatal number and CO$_2$ diffusion path length.
-As the diffusion rate of CO$_2$ though liquid is approximately 10,000 times slower than CO$_2$ diffusion through air, mesophyll resistance is generally thought to be primarily limited by liquid diffusion [@aalto_three-dimensional_2002], but diffusion through the IAS has also been shown to be a rate limiting process because the tortuous, disjunct nature of the IAS can greatly increase diffusion path lengths [@harwood_understanding_2021].
-Additionally, tortuosity is higher in horizontal directions (parallel to leaf surface) than vertical directions (perpendicular to leaf surface) because of the cylindrical shape and vertical arrangement of pallisade mesophyll cells [@earles_beyond_2018; @harwood_understanding_2021]; however, the ratio of lateral to vertical diffusion rate is still largely unknown [@morison_lateral_2005; @pieruschka_lateral_2005; @pieruschka_lateral_2006].
+To maximize CO$_2$ supply from the stomatal pore to chloroplasts, stomata should be uniformly distributed in an equilateral triangular grid on the leaf surface so as to minimize stomatal number and CO$_2$ diffusion path length [@parkhurst_diffusion_1994].
+As the diffusion rate of CO$_2$ though liquid is approximately $10^4\times$ slower than CO$_2$ diffusion through air, mesophyll resistance is generally thought to be primarily limited by liquid diffusion [@aalto_three-dimensional_2002; @evans_resistances_2009], but diffusion through the IAS has also been shown to be a rate limiting process because the tortuous, disjunct nature of the IAS can greatly increase diffusion path lengths [@harwood_understanding_2021].
+Additionally, tortuosity is higher in horizontal directions (parallel to leaf surface) than vertical directions (perpendicular to leaf surface) because of the cylindrical shape and vertical arrangement of pallisade mesophyll cells [@earles_beyond_2018; @harwood_understanding_2021].
+However, the ratio of lateral to vertical diffusion rate is still largely unknown [@morison_lateral_2005; @pieruschka_lateral_2005; @pieruschka_lateral_2006].
 Depending on the thickness of the leaf, porosity of the leaf mesophyll, tortuosity of the IAS, and lateral to vertical diffusion rate ratio, minimizing diffusion path length for CO$_2$ via optimally distributed stomata may yield significant increases in CO$_2$ supply for photosynthesis and higher $A_\text{max}$.
@@
 Additionally, because CO$_2$ is more limiting for photosynthesis under high light, we hypothesize that in high light 3) there should be more stomata, and 4) stomata should be more uniformly distributed than in low light.
-Finally, as stomatal densities are selected for optimal operational aperture, we hypothesize that 5) stomatal length will be positively correlated with the area of the leaf surface to which it is closest (stomatal zone area: the area a stomate occupies and supplies with CO$_2$).
-This way, each stomate can be optimally sized for its relative leaf area and associated CO$_2$ demand.
+Finally, as stomatal densities are selected for optimal operational aperture, we hypothesize that 5) stomatal length will be positively correlated with the area of the leaf surface to which it is closest.
+We refer to this as the 'stomatal zone', the leaf area surrounding a focal stomate closest to that stomate and therefore the zone it supplies with CO$_2$).
+This way, each stomate can be optimally sized relative to the mesophyll volume it supplies.
 
-To test these hypotheses, we grew *A.
-thaliana* in high, medium, and low light and measured stomatal density, size, and patterning on both leaf surfaces, and spatial coordination between them.
-We use voronoi tessellation techniques to calculate the stomatal leaf area.
-We also use CO$_2$ diffusion modeling to ask: what environmental and physiological conditions maximize the photosynthetic advantage ($A\text{gain}$) of abaxial-adaxial stomatal coordination?
+To test these hypotheses, we grew the model plant *Arabidopsis thaliana* in high, medium, and low light and measured stomatal density, size, and patterning on both leaf surfaces, and spatial coordination between them.
+We use Voronoi tessellation techniques to calculate stomatal zones.
+We also used a 2-D porous medium approximation of CO$_2$ diffusion and photosynthesis to predict the photosynthetic advantage of optimal versus suboptimal coordination in stomatal coordination between surfaces.
 Specifically, we predicted that traits which affect diffusion path length (leaf thickness, stomatal density, leaf porosity, lateral-vertical diffusion rate ratio), diffusion rate (temperature, pressure), and CO$_2$ demand (Rubisco concentration, light) would modulate the advantage of optimal stomatal arrangement following the relationships outlined in Table \ref{tab:hypotheses}.

### 2023-09-28 ms/ms.Rmd
edited single surface analysis methods

@@
 
-*Arabidopsis thaliana* plants were grown in three different light environments: low light (50 PAR), medium light (100 PAR), and high light (200 PAR) following the light standards from (CITE).
+*Arabidopsis thaliana* plants were grown in three different light environments: low light (50 PAR), medium light (100 PAR), and high light (200 PAR)...
 Once leaves were mature, we captured images of the abaxial and adaxial leaf surfaces using xxx bifocal microscope.
@@
 
-## Single Surface Analysis
+## Single surface analyses
 
@@
 \begin{equation}\label{eq:emd}
-  \mathrm{\overline{D}_E} = \frac{0.5}{\sqrt{A_l/n_\text{stomata}}}
+  \overline{D}_E = \frac{0.5}{\sqrt{A_\text{leaf} / n_\text{stomata}}}
 \end{equation}
 
-$A_l$ is leaf area and $n_\text{stomata}$ the number of stomata.
-$\overline{D}_E$ is the theoretical average distance to the nearest neighbor of each stomate if stomata were randomly distributed (CITE; Clark and Evans, 1954).
+\noindent $A_\text{leaf}$ is leaf area visible in the sampled field and $n_\text{stomata}$ the number of stomata.
+$\overline{D}_E$ is the theoretical average distance to the nearest neighbor of each stomate if stomata were uniformly randomly distributed [@clark_distance_1954].
 And $\overline{D}_O$ calculated for each synthetic data set is:
@@
 \begin{equation}\label{eq:omd}
-  \mathrm{\overline{D}_O} = \frac{\sum_{i=1}^{n_\text{stomata}}d_i}{n_\text{stomata}}
+  \overline{D}_O = \frac{\sum_{i=1}^{n_\text{stomata}}d_i}{n_\text{stomata}}
 \end{equation}
 
-$d_i$ is the distance between $\text{stomate}_i$ and its nearest neighbor. We calculated $\mathrm{NNI}$ using the R package **spatialEco** version `r packageVersion("spatialEco")`.
+$d_i$ is the distance between $\text{stomate}_i$ and its nearest neighbor.
+We calculated $\mathrm{NNI}$ using the *R* package **spatialEco** version `r packageVersion("spatialEco")` [@evans_spatialeco_2023].
 
@@
 
-We calculated the dispersion index $\mathrm{DI}$) which varies from zero to one. Zero is uniformly random and one is ideally dispersed:
+We developed a dispersion index $\mathrm{DI}$ to quantify how close observed stomatal distributions are to random uniform versus maximally dispered in an equilateral triangular grid.
+$\mathrm{DI}$ varies from zero to one, where zero is uniformly random and one is ideally dispersed:
 
@@
 
-Uniformly distributed stomata on a leaf with stomatal density ($D_S$) maximize $\overline{D}_O$ in an equilateral triangle pattern with side lengths ($s$) equal to two times the incircle radius or apothem, $a$, of a regular hexagon with area ($A_\text{hex}$) equal to the area of the leaf ($A_l$) divided by the number of stomata ($n_\text{stomata}$) on the leaf according to Eq.
+\noindent $\mathrm{NNI}$ is calculated for each sample image as described above; $\text{median}(\mathrm{NNI_{random}})$ and $\text{median}(\mathrm{NNI_{uniform}})$ are calculated from the dry og synthetic data specific to each sample image as described above.
+We tested whether light treatment affects $\mathrm{DI}$ and $D_S$ using analysis of variance (ANOVA).
+
+<!-- CDM: I am not sure this part is needed
+Uniformly distributed stomata on a leaf with stomatal density ($D_S$) maximize $\overline{D}_O$ in an equilateral triangle pattern with side lengths ($s$) equal to two times the incircle radius or apothem, $a$, of a regular hexagon with area ($A_\text{hex}$) equal to $A_\text{leaf}$ divided by the number of stomata ($n_\text{stomata}$) on the leaf according to Eq.
 \ref{eq:eq2}:
@@
 \begin{equation}\label{eq:eq2}
-  s = \frac{\sqrt{2}} {3^{1/4} \sqrt{n_\text{stomata} / A_l}}
+  s = \frac{\sqrt{2}} {3^{1/4} \sqrt{n_\text{stomata} / A_\text{leaf}}}
 \end{equation}
@@
 As all sides of an equilateral triangle have the same length, the average nearest neighbor distance for a given $D_S$ is maximized and each stomate occupies the same hexagonal area ($A_\text{hex}$).
-The closer to one $DI$ is, the more uniform an area each of its stomata supply with CO$_2$ during photosynthesis.
-We tested whether light treatment affects $\mathrm{DI}$ and $D_S$ using analysis of variance (ANOVA).
+As $DI$ approaches 1, the more uniform an area each of its stomata supply with CO$_2$ during photosynthesis.
+ -->
 
-Finally, to test our hypothesis that stomatal length is modulated by the area of the leaf to which it supplies CO$_2$, we examined the relationship between stomatal zone area and stomatal length using a Bayesian generalized non-linear multilevel model with the R package **brms** version `r packageVersion("brms")`.
-Stomatal zone area was calculated using voronoi tessellation (e.g.
+Finally, we examined the relationship between stomatal zone area and stomatal length using a Bayesian generalized non-linear multilevel model with the R package **brms** version `r packageVersion("brms")` [@burkner_brms_2017; @burkner_advanced_2018].
+Stomatal zone area was calculated using Voronoi tessellation (e.g.
 Fig.

### 2023-09-28 r/01_simulate-single-surface.R

@@ -3,13 +3,13 @@ source("r/header.R")
 stomata = readr::read_csv("raw-data/stomata_position.csv") |>
   group_by(treatment, leaf_number, image_number, surface)
 
-pb <- progress_bar$new(
+pb = progress_bar$new(
   format = "[:bar] :current/:total (:percent) elapsed :elapsed eta :eta",
   total = nrow(attr(stomata, "groups"))
 )
 
 set.seed(197789263)
-synthetic_data <- stomata |>
+synthetic_data = stomata |>
   purrrlyr::by_slice(
     simulate_grids_from_data,
     pixels_x = 512, pixels_y = 512, n_grid = 1e3, progress = TRUE,

### 2023-09-28 r/functions.R

@@ -373,11 +373,11 @@ simulate_grids_from_data <- function(
   while (i < n_grid) {
     lambda = rgamma(1, shape = n_stomata, scale = 1)
     U = D2U(lambda / (pixels_x * pixels_y))
-    g <- make_equal_stomata_grid(U, pixels_x = pixels_x, pixels_y = pixels_y,
+    g = make_equal_stomata_grid(U, pixels_x = pixels_x, pixels_y = pixels_y,
                                  jitter = TRUE, rotate = TRUE)
     if (g$n_stomata[1] == n_stomata) {
-      i <- i + 1
-      equal_grids[[i]] <- g
+      i = i + 1
+      equal_grids[[i]] = g
     }
   }
 
@@ -480,16 +480,12 @@ kerneldensity = function(
 
   # kernel density function
   # define extent of analysis
-  e <- extent(0, pixels_x, 0, pixels_y)
+  e = extent(0, pixels_x, 0, pixels_y)
   # run sp.kde which builds a kernel density estimate
-  kernelden <- sp.kde(x = stomata_sp, bw = 120,
+  kernelden = sp.kde(x = stomata_sp, bw = 120,
                       nr = pixels_y, nc = pixels_x,
                       newdata = e, standardize = TRUE,
-                      mask = F)
-
-  # return kernel density raster
-  return(kernelden)
-
+                      mask = FALSE)
 
   # return result
   return(kernelden)

### 2023-09-28 r/header.R
working version of photo_2d_pm()

@@ -15,7 +15,6 @@ library(cmdstanr)
 library(checkmate)
 library(corrplot)
 library(cowplot)
-library(deSolve)
 library(dplyr)
 library(forcats)
 library(ggdist)

### 2023-09-28 r/photo_2d_pm.R

@@ -6,111 +6,81 @@
 # Note that n_z = 0 is bottom (abaxial) side of leaf
 
 # 2D photosynthesis model using a porous medium approximation
-photo_2d_pm = function(...) {
+photo_2d_pm = function(stomatal_arrangment, replace = NULL, ...) {
 
-  # Set parmseters
-  parmss = set_2d_pm_parms(...)
+  # Set parameters
+  parms = set_2d_pm_parms(replace) |>
+    derive_2d_pm_parms(fit_rubisco = fit_rubisco) |>
+    make_2d_pm_mat() |>
+    add_stomatal_position(stomatal_arrangment)
 
   # Find solution
+  C_init = rep(parms[["C_stom"]], 2 * parms[["n_z"]] * parms[["n_x"]])
+  fit = steady.2D(C_init, func = rd_2p_pm, parms = parms,
+                  pos = FALSE, dimens = c(parms[["n_z"]], parms[["n_x"]]),
+                  nspec = 2, lrw = 1e8, atol = 1e-10, rtol = 1e-10, ctol = 1e-10)
+
   # Structure output
+  list(fit = fit, parms = parms)
+
+}
+
+# Function to add stomatal position to parms.
+# For this project, it makes an amphistomatous leaf with stomata aligned or
+# offset, but could be expanded for other purposes
+add_stomatal_position = function(parms, stomatal_arrangment) {
+
+  parms[["x_abaxial_stomata"]] = 1
+  parms[["x_adaxial_stomata"]] = switch(
+    stomatal_arrangment,
+    aligned = 1,
+    offset = parms[["n_x"]]
+  )
+  parms
 
 }
 
 # Set parameters for photo_2d_pm()
-set_2d_pm_parms = function(...) {
+set_2d_pm_parms = function(replace, ...) {
+
+  parms = get_2d_pm_default_parms()
+  if (!is.null(replace)) {
+    parms[names(replace)] = replace[names(replace)]
+  }
+  parms
 
 }
 
-# Get set of parmseters for photo_2d_pm()
+# Get set of parameters for photo_2d_pm()
 get_2d_pm_default_parms = function(...) {
 
[…]

### 2023-09-28 r/scratch-photo_2d_pm.R

@@ -6,8 +6,50 @@
 
 source("r/header.R")
 source("r/photo_2d_pm.R")
-library(rootSolve)
 
+ph2d = photo_2d_pm("aligned", replace = list(I_0 = 0.002, n_z = 300))
+
+# Plot results
+df_C = expand.grid(
+  z = seq_len(ph2d$parms[["n_z"]]),
+  x = seq_len(ph2d$parms[["n_x"]]),
+  name = c("C_ias", "C_liq")
+) |>
+  mutate(value = ph2d$fit$y)
+
+ggplot(df_C, aes(x, z, z = value)) +
+  facet_grid(~ name) +
+  geom_contour_filled() +
+  coord_equal()
+
+# Calculate area-based A_n
+C_liq_mat = matrix(
+  ph2d$fit$y[(ph2d$parms$n_x * ph2d$parms$n_z + 1):(2 * ph2d$parms$n_x * ph2d$parms$n_z)],
+  nrow = ph2d$parms$n_z, ncol = ph2d$parms$n_x
+)
+r_c = calc_2d_pm_rc(C_liq_mat, ph2d$parms)
+r_d = ph2d$parms[["r_d"]]
+r_p = r_c * ph2d$parms[["gamma_star"]] / C_liq_mat
+
+a_n = (r_c - r_p - r_d) * (ph2d$parms[["S_m_mat"]] / ph2d$parms[["T_leaf"]]) * ph2d$parms[["V_strom"]]
+
+# 1 m^2 of 200 um thick leaf is 2e-04 m^3
+mean(a_n) * ph2d$parms[["T_leaf"]] * 1e6 # 24.7955
+
+df_A = expand.grid(
+  z = seq_len(parms[["n_z"]]),
+  x = seq_len(parms[["n_x"]])
+) |>
+  mutate(a = c(a_n))
+
+ggplot(df_A, aes(x, z, z = a)) +
+  geom_contour_filled() +
+  coord_equal()
+
+
+
+
+# SIMPLE VERSION I SENT TO TOM
 photo_2d_pm = function(t, Y, parms) {
 
   # Arguments:

### 2023-09-29 r/photo_2d_pm.R
working parallel version of photo_2d_pm()

@@ -142,7 +142,7 @@ derive_2d_pm_jmax = function(parms, ...) {
   jscale = parms[["J_max"]] / mu_j_max_z1
   parms$j_max_z = jscale * j_max_z1
   # Check
-  checkmate::assert_true(parms[["J_max"]] == mean(parms$j_max_z * parms[["S_m_z"]] * parms[["V_strom"]]))
+  checkmate::assert_true(abs(parms[["J_max"]] - mean(parms$j_max_z * parms[["S_m_z"]] * parms[["V_strom"]])) < 1e-6)
 
   parms
 
@@ -158,7 +158,7 @@ derive_2d_pm_jinf = function(parms, ...) {
   jscale = J_inf / mu_j_inf_z1
   parms$j_inf_z = jscale * j_inf_z1
   # Check
-  checkmate::assert_true(J_inf == mean(parms$j_inf_z * parms[["S_m_z"]] * parms[["V_strom"]]))
+  checkmate::assert_true(abs(J_inf - mean(parms$j_inf_z * parms[["S_m_z"]] * parms[["V_strom"]])) < 1e-6)
 
   parms
 
@@ -286,3 +286,27 @@ rd_2p_pm = function(t, y, parms, ...) {
 
 }
 
+# Calculate A_n per area from output
+calc_2d_pm_An = function(ph2d) {
+
+  # ph2d is object returned from photo_2d_pm()
+  df_C = expand.grid(
+    z = seq_len(ph2d$parms[["n_z"]]),
+    x = seq_len(ph2d$parms[["n_x"]]),
+    name = c("C_ias", "C_liq")
+  ) |>
+    mutate(value = ph2d$fit$y)
+
+  C_liq_mat = matrix(
+    ph2d$fit$y[(ph2d$parms$n_x * ph2d$parms$n_z + 1):(2 * ph2d$parms$n_x * ph2d$parms$n_z)],
+    nrow = ph2d$parms$n_z, ncol = ph2d$parms$n_x
+  )
+  r_c = calc_2d_pm_rc(C_liq_mat, ph2d$parms)
+  r_d = ph2d$parms[["r_d"]]
+  r_p = r_c * ph2d$parms[["gamma_star"]] / C_liq_mat
+
+  a_n = (r_c - r_p - r_d) * (ph2d$parms[["S_m_mat"]] / ph2d$parms[["T_leaf"]]) * ph2d$parms[["V_strom"]]
+
+  mean(a_n) * ph2d$parms[["T_leaf"]] * 1e6
+
+}

### 2023-09-29 r/scratch-photo_2d_pm.R

@@ -7,7 +7,53 @@
 source("r/header.R")
 source("r/photo_2d_pm.R")
 
-ph2d = photo_2d_pm("aligned", replace = list(I_0 = 0.002, n_z = 300))
+library(furrr)
+plan(multisession, workers = 10)
+
+library(progressr)
+handlers(global = TRUE)
+handlers("progress", "beepr")
+
+input = crossing(
+  n_x = c(50, 100),
+  n_z = c(250, 500),
+  stomatal_arrangement = c("aligned", "offset"),
+  I_0 = round(seq(0.0, 0.001, 0.0001), 4)
+) %>%
+  split(~ seq_len(nrow(.)))
+
+run_photo_2d_pm = function(input, fit_rubisco) {
+
+  p = progressor(along = input)
+  out = future_map(input, \(.x, fit_rubisco) {
+
+    r = with(.x, list(I_0 = I_0, n_x = n_x, n_z = n_z))
+    ph2d = photo_2d_pm(.x$stomatal_arrangement, fit_rubisco = fit_rubisco, replace = r)
+    ph2d$parms$stomatal_arrangement = .x$stomatal_arrangement
+    ph2d$parms$A_n = calc_2d_pm_An(ph2d)
+    p()
+    ph2d
+
+  }, fit_rubisco = fit_rubisco)
+
+  out
+
+}
+
+tmp = run_photo_2d_pm(input, fit_rubisco)
+
+tmp |>
+  map("parms") |>
+  map_dfr(\(x) as_tibble(x[c("stomatal_arrangement", "n_z", "n_x", "I_0", "A_n")])) |>
+  ggplot(aes(I_0, A_n, color = stomatal_arrangement)) +
+  facet_grid(n_z ~ n_x) +
+  geom_point()
+
+
+r = list(I_0 = 0.0003, n_x = 50, n_z = 500)
+ph2d = photo_2d_pm("aligned", replace = r)
+
+ph2d = photo_2d_pm("aligned", replace = list(I_0 = 0.001))
 
 # Plot results
 df_C = expand.grid(
@@ -23,18 +69,6 @@ ggplot(df_C, aes(x, z, z = value)) +
   coord_equal()
 
 # Calculate area-based A_n
-C_liq_mat = matrix(
[…]

### 2023-09-29 r/10_run-model.R
added 2d porous medium model output

@@ -1,0 +1,47 @@
+source("r/header.R")
+source("r/photo_2d_pm.R")
+
+plan(multisession, workers = 10)
+
+handlers(global = TRUE)
+handlers("cli")
+
+# U / 2 %in% c(169, 53, 17) corresponds to approximately 10, 100, 1000 mm^2
+# Stomatal density (number per mm^2)
+c(10, 100, 1000) |>
+  # convert to per um^2
+  multiply_by(1e-6) |>
+  D2U() |>
+  divide_by(2)
+
+# Palisade porosity based on Lundgren et al. (2019), fig 5I
+input = crossing(
+  n_x = c(17, 53, 169),
+  n_z = c(100, 300, 500),
+  stomatal_arrangement = c("aligned", "offset"),
+  I_0 = round(c(0.00005, 0.00025, 0.001), 4),
+  phi_pal = c(0.1, 0.2, 0.3)
+) %>%
+  split(~ seq_len(nrow(.)))
+
+run_photo_2d_pm = function(input, fit_rubisco) {
+
+  p = progressor(along = input)
+  out = future_map(input, \(.x, fit_rubisco) {
+
+    r = with(.x, list(I_0 = I_0, n_x = n_x, n_z = n_z, phi_pal = phi_pal))
+    ph2d = photo_2d_pm(.x$stomatal_arrangement, fit_rubisco = fit_rubisco, replace = r)
+    ph2d$parms$stomatal_arrangement = .x$stomatal_arrangement
+    ph2d$parms$A_n = calc_2d_pm_An(ph2d)
+    p()
+    ph2d
+
+  }, fit_rubisco = fit_rubisco)
+
+  out
+
+}
+
+ph2d = run_photo_2d_pm(input, fit_rubisco)
+
+write_rds(ph2d, "objects/model_output.rds")

### 2023-09-29 r/header.R

@@ -17,6 +17,7 @@ library(corrplot)
 library(cowplot)
 library(dplyr)
 library(forcats)
+library(furrr)
 library(ggdist)
 library(ggforce)
 library(ggimage)
@@ -31,6 +32,7 @@ library(metR)
 library(mgcv)
 library(parallel)
 library(progress)
+library(progressr)
 library(purrr)
 library(purrrlyr)
 library(RANN)
@@ -56,7 +58,8 @@ x_depth = c(145, 830.5)
 y_depth = c(0, 700)
 x_rubisco = c(558, 56.5)
 y_rubisco = c(0, 100)
-nishio_carbon_1993_fig5 = read_csv("raw-data/nishio_carbon_1993_fig5.csv") |>
+nishio_carbon_1993_fig5 = read_csv("raw-data/nishio_carbon_1993_fig5.csv",
+                                   show_col_types = FALSE) |>
   mutate(
     depth = raw_depth * (diff(y_depth) / diff(x_depth)) - x_depth[1],
     rubisco = raw_rubisco * (diff(y_rubisco) / diff(x_rubisco)) -

### 2023-09-29 r/10_run-model.R
initial plot of model output

@@ -15,11 +15,12 @@ c(10, 100, 1000) |>
   divide_by(2)
 
 # Palisade porosity based on Lundgren et al. (2019), fig 5I
+
 input = crossing(
   n_x = c(17, 53, 169),
-  n_z = c(100, 300, 500),
+  n_z = c(101, 301, 501),
   stomatal_arrangement = c("aligned", "offset"),
-  I_0 = round(c(0.00005, 0.00025, 0.001), 4),
+  I_0 = round(c(0.00005, 0.00025, 0.001), 5),
   phi_pal = c(0.1, 0.2, 0.3)
 ) %>%
   split(~ seq_len(nrow(.)))

### 2023-09-29 r/11_plot-model.R

@@ -1,0 +1,22 @@
+source("r/header.R")
+source("r/photo_2d_pm.R")
+
+ph2d = read_rds("objects/model_output.rds")
+
+# STUFF for processing
+vars = c("stomatal_arrangement", "n_z", "n_x", "I_0", "A_n", "phi_pal")
+
+ph2d_vars = ph2d |>
+  map("parms") |>
+  map_dfr(\(.x) as_tibble(.x[vars]))
+
+ph2d_vars |>
+  pivot_wider(names_from = "stomatal_arrangement", values_from = "A_n") |>
+  # coordination advantage
+  mutate(ca = log(offset / aligned)) |>
+  ggplot(aes(I_0, ca, color = phi_pal)) +
+  facet_grid(n_z ~ n_x) +
+  geom_point()
+
+ph2d_vars |>
+  filter(n_z == 101, n_x == 169)

### 2023-09-29 r/header.R

@@ -53,6 +53,8 @@ library(tidyr)
 library(tripack)
 library(units)
 
+theme_set(theme_cowplot())
+
 # Fit Rubisco model
 x_depth = c(145, 830.5)
 y_depth = c(0, 700)

### 2023-10-16 ms/ms.Rmd
edits to ms and model table

@@
 source("r/header.R")
+fit = readr::read_rds("objects/fit.rds")
 ```
@@
 
-*Arabidopsis thaliana* plants were grown in three different light environments: low light (50 PAR), medium light (100 PAR), and high light (200 PAR)...
-Once leaves were mature, we captured images of the abaxial and adaxial leaf surfaces using xxx bifocal microscope.
-This microscope allows the capture of two focal lengths that are spatially correlated such that pixel (1,1) of the abaxial surface image is directly above pixel (1,1) of the adaxial surface image, and so on...
-Images were 512 by 512 pixels and covered X leaf surface area.
+*Arabidopsis thaliana* plants were grown in three different light environments: low light (PAR = 50 $\mu \text{mol}~\text{m}^{-2}~\text{s}^{-1}$), medium light (100 $\mu \text{mol}~\text{m}^{-2}~\text{s}^{-1}$), and high light (200 $\mu \text{mol}~\text{m}^{-2}~\text{s}^{-1}$).
+PAR stands for photosynthetically active radiation.
+Once leaves were mature, we captured images of the abaxial and adaxial leaf surfaces using XXX.
 We captured 132 images in total, making 66 abaxial-adaxial image pairs.
-The position of all stomata were recorded using ImageJ [@schneider_nih_2012].
+We measured stomatal position and size using ImageJ [@schneider_nih_2012].
 
@@
 For each sample image, we compared the observed Nearest Neighbor Index ($\mathrm{NNI}$) to the null distribution of $\mathrm{NNI}$ values calculated from the synthetic data set.
-The observed stomatal distribution is dispersed relative to a uniform random distribution if the observed $NNI$ is greater than 95% of the synthetic $\mathrm{NNI}$ values (one-tailed test).
-
 $\mathrm{NNI}$ is the ratio of observed mean distance ($\overline{D}_O$) to the expected mean distance ($\overline{D}_E$) where $\overline{D}_E$ is:
@@
 \begin{equation}\label{eq:emd}
-  \overline{D}_E = \frac{0.5}{\sqrt{A_\text{leaf} / n_\text{stomata}}}
+  \overline{D}_E = \frac{0.5}{\sqrt{A_\text{leaf} / n_\text{stomata}}}.
 \end{equation}
@@
 $\overline{D}_E$ is the theoretical average distance to the nearest neighbor of each stomate if stomata were uniformly randomly distributed [@clark_distance_1954].
-And $\overline{D}_O$ calculated for each synthetic data set is:
+$\overline{D}_O$ calculated for each synthetic data set is:
 
 \begin{equation}\label{eq:omd}
-  \overline{D}_O = \frac{\sum_{i=1}^{n_\text{stomata}}d_i}{n_\text{stomata}}
+  \overline{D}_O = \frac{\sum_{i=1}^{n_\text{stomata}}d_i}{n_\text{stomata}},
 \end{equation}
 
-$d_i$ is the distance between $\text{stomate}_i$ and its nearest neighbor.
+\noindent where $d_i$ is the distance between $\text{stomate}_i$ and its nearest neighbor.
 We calculated $\mathrm{NNI}$ using the *R* package **spatialEco** version `r packageVersion("spatialEco")` [@evans_spatialeco_2023].
+The observed stomatal distribution is dispersed relative to a uniform random distribution if the observed $NNI$ is greater than 95% of the synthetic $\mathrm{NNI}$ values (one-tailed test).
 
@@
 
-\noindent $\mathrm{NNI}$ is calculated for each sample image as described above; $\text{median}(\mathrm{NNI_{random}})$ and $\text{median}(\mathrm{NNI_{uniform}})$ are calculated from the dry og synthetic data specific to each sample image as described above.
-We tested whether light treatment affects $\mathrm{DI}$ and $D_S$ using analysis of variance (ANOVA).
+\noindent $\mathrm{NNI}$ is calculated for each sample image as described above; $\text{median}(\mathrm{NNI_{random}})$ and $\text{median}(\mathrm{NNI_{uniform}})$ are calculated from the synthetic data specific to each sample image as described above.
+We tested whether light treatment affects $\mathrm{DI}$ and stomatla density ($D_S$) using analysis of variance (ANOVA).
 
@@
 
-Finally, we examined the relationship between stomatal zone area and stomatal length using a Bayesian generalized non-linear multilevel model with the R package **brms** version `r packageVersion("brms")` [@burkner_brms_2017; @burkner_advanced_2018].
+Finally, we examined the relationship between stomatal zone area and stomatal length using a Bayesian generalized non-linear multilevel model fit with the *R* package **brms** version `r fit$version$brms` [@burkner_brms_2017; @burkner_advanced_2018] and *Stan* version `r fit$version$cmdstan` [@stan_development_team_stan_2023].
 Stomatal zone area was calculated using Voronoi tessellation (e.g.
@@
 The stomatal zone area, $S_\text{area}$, is the region of the leaf surface whose distance to stomate, $S$, is less than the distance to any other stomate, $S$.
-Stomatal length was measured in `ImageJ` [@schneider_nih_2012].
+Stomatal length was measured in ImageJ [@schneider_nih_2012].
 
@@
 To test whether the position of ab- and adaxial stomata are coordinated we compared the observed distribution to a null distribution where the positions on each surface are random.
-For each pair of surfaces (observed or synthetic) we calculated the distance squared between each to the nearest stomatal centroid with the R package **raster** version `r packageVersion("raster")`.
+For each pair of surfaces (observed or synthetic) we calculated the distance squared between each to the nearest stomatal centroid with the *R* package **raster** version `r packageVersion("raster")`.
 Then we calculated the cell-wise Pearson correlation coefficient.
@@
 
-We modeled photosynthesis CO$_2$ assimilation rate using a spatially-explicit two-dimensional reaction diffusion model using a porous medium approximation (Parkhurst 1994).
+We modeled photosynthesis CO$_2$ assimilation rate using a spatially-explicit two-dimensional reaction diffusion model using a porous medium approximation [@parkhurst_intercellular_1994] using the finite element method (FEM) following @earles_excess_2017.
 Consider a two-dimensional leaf where stomata occur on each surface in a regular sequence with interstomatal distance $U$.
@@
 We arbitrarily set the adaxial stomate at $x = 0$ and toggled the abaxial stomata position between $x = U/2$ (offset) or $x = 0$ (below adaxial stomate).
-The advantage of offsetting stomatal position on each surface is the photosynthetic rate of the leaf with offset stomata compared to that with stomata aligned in the same $x$ position:
+The 'coordination advantage' of offsett stomatal position on each surface is the photosynthetic rate of the leaf with offset stomata compared to that with stomata aligned in the same $x$ position:
 
@@
 
-We modelled the coordination advantage over a range of leaf thicknesses, stomatal densities, photosynthetic capacities, and light environments to understand when offsetting stomatal position on each surface might deliver a significant photosynthetic advantage.
+We modeled the coordination advantage over a range of leaf thicknesses, stomatal densities, photosynthetic capacities, and light environments to understand when offsetting stomatal position on each surface might deliver a significant photosynthetic advantage (Table X).
+
+WORKING HERE - make table of parameter combinations!
 
 ### Light propogation model
-
-[NOTE: we should probably adjust light attenuation to be proportional to the chlorophyll concentration which is one of the parameters in the biochemical models.]
 
@@
 
-FYI - something is wrong with the model.
-I think the units are off somewhere because the result is very sensitive to grid size and the values often aren't reasonable.
-I'll need to troubleshoot more, but maybe something will be obvious to you.
-
-We used the `steady.2D()` function in the R package **rootSolve**.
-The function needs a function to calculate the time differential for node $ij$, $\frac{dC_{m,ij}}{dt}$.
-The R function below uses the light propagation, biochemical, and diffusional model to calculate a matrix of $\frac{dC_{m,ij}}{dt}$ values:
+We used the `steady.2D()` function in the *R* package **rootSolve** version `r packageVersion("rootSolve")` [@soetaert_practical_2009].
 

### 2023-10-16 r/12_table-model.R

@@ -1,0 +1,28 @@
+# testing code to make table of 2d fem model parameters
+source("r/header.R")
+
+model_output = read_rds("objects/model_output.rds") |>
+  map_dfr(\(.x) {
+    l = map(.x$parms, length)
+    .x$parms[l == 1] |>
+      as_tibble()
+  })
+
+# Select variables
+vars = model_output |>
+  as.list() |>
+  sapply(\(.x) length(unique(.x))) |>
+  sapply(\(.x) .x > 1)
+
+# Variable transformations
+var_transformation = list(
+  # convert from mol to umol
+  I_0 = function(.x) {.x * 1e6},
+
+)
+model_output[, vars] |>
+  as.list() |>
+  map(\(.x) {
+    glue::glue("${min}--{max}$", min = min(.x), max = max(.x))
+  })
+

### 2023-10-16 ms/ms.Rmd
added model_var_table

@@
 fit = readr::read_rds("objects/fit.rds")
+model_var_table = readr::read_rds("objects/model_var_table.rds")
 ```
@@
 
-We modeled the coordination advantage over a range of leaf thicknesses, stomatal densities, photosynthetic capacities, and light environments to understand when offsetting stomatal position on each surface might deliver a significant photosynthetic advantage (Table X).
+We modeled the coordination advantage over a range of leaf thicknesses, stomatal densities, photosynthetic capacities, and light environments to understand when offsetting stomatal position on each surface might deliver a significant photosynthetic advantage (Table \ref{tab:model_var}).
+
+```{r, model_var}
+knitr::kable(model_var_table)
+```
 

### 2023-10-16 ms/test.Rmd
added model_var_table

@@
 ```{r setup, include=FALSE}
-knitr::opts_chunk$set(echo = TRUE)
+knitr::opts_chunk$set(echo = FALSE, message = FALSE) 
+knitr::opts_chunk$set(fig.pos = 'p') # Places figures on pages separate from text
+knitr::opts_chunk$set(out.width = '100%', dpi = 300) # Figure resolution and size
+knitr::opts_chunk$set(fig.env = "figure") # Latex figure environment
+knitr::opts_knit$set(root.dir = "..")
 ```
 
-```{r dispersion, fig.cap = "Stomata are more dispersed than expected under the null model of uniform random position (dispersion index = 0) but far from a distribution that maximizes distance between stomata (dispersion index = 1)", echo = TRUE}
+```{r}
+source("r/header.R")
+fit = readr::read_rds("objects/fit.rds")
+model_var_table = readr::read_rds("objects/model_var_table.rds")
+```
 
-knitr::include_graphics("figures/dispersion.pdf")
-
+```{r, model_var}
+knitr::kable(model_var_table)
 ```

### 2023-10-16 r/12_table-model.R

@@ -8,21 +8,62 @@ model_output = read_rds("objects/model_output.rds") |>
       as_tibble()
   })
 
+pm_pars = readr::read_csv("raw-data/2d-pm-parameters.csv", show_col_types = FALSE)
+
 # Select variables
+## Find all elements that vary. It should include desired variables and
+## and additional response variables and redundant variables (e.g. T_leaf is
+## derived from n_z). The last step selects variables for table.
+focal_vars = c("I_0", "phi_pal", "T_leaf", "U")
+focal_vars_units = c(
+  I_0 = "$\\mu$mol m$^{-2}$ s$^{-1}$",
+  phi_pal = "m$^3$ airspace m$^{-3}$ leaf",
+  T_leaf = "$\\mu$m",
+  U = "$\\mu$m"
+)
+
+
 vars = model_output |>
   as.list() |>
   sapply(\(.x) length(unique(.x))) |>
-  sapply(\(.x) .x > 1)
+  sapply(\(.x) .x > 1) %>%
+  and(names(.) %in% focal_vars)
 
 # Variable transformations
 var_transformation = list(
   # convert from mol to umol
   I_0 = function(.x) {.x * 1e6},
-
+  # no conversion
+  phi_pal = function(.x) .x,
+  # convert from m to um
+  T_leaf = function(.x) {.x * 1e6},
+  # convert from m to um
+  U = function(.x) {.x * 1e6}
 )
+
+vars |>
+  which() |>
+  names() %>%
+  glue("model_output[, '{var}'] %<>% var_transformation${var}()", var = .) %>%
+  parse(text = .) |>
+  eval()
+
 model_output[, vars] |>
   as.list() |>
-  map(\(.x) {
-    glue::glue("${min}--{max}$", min = min(.x), max = max(.x))
-  })
+  map_dfr(\(.x) {
+    glue("${min}-{max}$", min = min(.x), max = max(.x))
+  }) |>
+  pivot_longer(
+    everything(),
+    names_to = "r",
+    values_to = "Parameter range"
+  ) |>
+  left_join(select(pm_pars, r, Variable = symbol), by = join_by(r)) |>
+  full_join(
+    tibble(r = focal_vars, Units = focal_vars_units[focal_vars]),
+    by = join_by(r)
+  ) |>
+  select(Variable, `Parameter range`, Units) |>
+  write_rds("objects/model_var_table.rds")
+
 

### 2023-10-16 r/header.R

@@ -25,6 +25,7 @@ library(ggplot2)
 library(ggpubr)
 library(ggsignif)
 library(ggvoronoi)
+library(glue)
 library(Hmisc)
 library(magrittr)
 library(maptools)

### 2023-10-16 ms/ms.Rmd
added autoref features to ms

--- 2023-10-16 ms/ms.Rmd
@@
  - \linenumbers # For line numbering
+ - \usepackage{hyperref} # For autoref
+ - \renewcommand{\figureautorefname}{Fig.}
 ---
@@
 fit = readr::read_rds("objects/fit.rds")
+pm_pars = readr::read_csv("raw-data/2d-pm-parameters.csv", show_col_types = FALSE)
+model_var = readr::read_rds("objects/model_var.rds")
 ```
@@
 We hypothesized that natural selection will favor stomatal patterning and distribution to minimize the diffusion path length.
-In amphistomatous leaves, this would be accomplished by 1) a uniform distribution of stomata on both abaxial and adaxial leaf surfaces and 2) coordinated stomatal spacing on each surface that offsets the position of stomata (Fig.
-\ref{fig:amphi_grid}).
+In amphistomatous leaves, this would be accomplished by 1) a uniform distribution of stomata on both abaxial and adaxial leaf surfaces and 2) coordinated stomatal spacing on each surface that offsets the position of stomata (\autoref{fig:ideal-amphi-grid}).
 Coordination between leaf surfaces is defined in this study as the occurrence of stomata in areas farthest from stomata on the opposite leaf surface.
@@
 We also used a 2-D porous medium approximation of CO$_2$ diffusion and photosynthesis to predict the photosynthetic advantage of optimal versus suboptimal coordination in stomatal coordination between surfaces.
-Specifically, we predicted that traits which affect diffusion path length (leaf thickness, stomatal density, leaf porosity, lateral-vertical diffusion rate ratio), diffusion rate (temperature, pressure), and CO$_2$ demand (Rubisco concentration, light) would modulate the advantage of optimal stomatal arrangement following the relationships outlined in Table \ref{tab:hypotheses}.
+Specifically, we predicted that traits which affect diffusion path length (leaf thickness, stomatal density, leaf porosity, lateral-vertical diffusion rate ratio), diffusion rate (temperature, pressure), and CO$_2$ demand (Rubisco concentration, light) would modulate the advantage of optimal stomatal arrangement following the relationships outlined in \autoref{tab:hypotheses}.
 Here, we integrate over reasonable parameter space to determine the ecophysiological context most likely to favor stomatal spatial coordination in amphistomatous leaves.
@@
 print(xtable::xtable(df, caption = "A summary of the hypothesized relationships between leaf traits and environmental conditions and photosynthetic advantage of stomatal spatial coordination in amphistomatous leaves.", label = "tab:hypotheses"),
-      comment = FALSE)
+      comment = FALSE, include.rownames = FALSE)
 ```
@@
 Stomatal zone area was calculated using Voronoi tessellation (e.g.
-Fig.
-\ref{fig:example}).
+\autoref{fig:tessellation-example}).
 The stomatal zone area, $S_\text{area}$, is the region of the leaf surface whose distance to stomate, $S$, is less than the distance to any other stomate, $S$.
@@
 If stomatal positions on each surface are coordinated to minimize the distance between mesophyll and the nearest stomate, then we expect a negative correlation.
-A cell that is far from a stomate on one surface should be near a stomate on the other surface (Fig.
-\ref{fig:amphi}).
+A cell that is far from a stomate on one surface should be near a stomate on the other surface (\autoref{fig:ideal-amphi-grid}).
 We generated a null distribution of the correlation coefficient by simulating $10^3$ synthetic data sets for each observed pair.
@@
 
-We modeled the coordination advantage over a range of leaf thicknesses, stomatal densities, photosynthetic capacities, and light environments to understand when offsetting stomatal position on each surface might deliver a significant photosynthetic advantage (Table X).
+We modeled the coordination advantage over a range of leaf thicknesses, stomatal densities, photosynthetic capacities, and light environments to understand when offsetting stomatal position on each surface might deliver a significant photosynthetic advantage (\autoref{tab:model_var}).
 
-WORKING HERE - make table of parameter combinations!
+```{r, model_var, results="asis"}
+#| label: tbl-model_var
+
+symbol_descriptions = left_join(rename(model_var, symbol = Variable),
+                                pm_pars,
+                                by = join_by(symbol)) |>
+  transmute(x = glue("{symbol}: {name}")) |>
+  pull(x) |>
+  str_c(collapse = "; ")
+
+# print(xtable::xtable(model_var, caption = paste0("The parameter range of model variables tested for their effect on coordination advantage (\\autoref{eq:coordination_advantage}) using a 2-D porous medium approximation.
+We used regularly spaced values within each range and simulated across all combinations.
+Here we converted model units to more conventional units (e.g.
+m to $\\mu$m).
+", symbol_descriptions), label = "tab:model_var"),
+#       comment = FALSE, include.rownames = FALSE, type = "latex")
+
+model_var_caption = paste0("The parameter range of model variables tested for their effect on coordination advantage (\\autoref{eq:coordination_advantage}) using a 2-D porous medium approximation.
+We used regularly spaced values within each range and simulated across all combinations.
+Here we converted model units to more conventional units (e.g.
+m to $\\mu$m).
+", symbol_descriptions)
+
+knitr::kable(
+  model_var, 
+  caption = model_var_caption,
+  label = "model_var",
+  booktabs = TRUE, format = "latex", escape = FALSE
+) #|>
+#   kableExtra::kable_styling()
+```
+
+WORKING HERE - ???
 
@@
 If our hypotheses that natural selection will act to reduce the diffusion path length of CO$_2$ while minimizing stomatal number are correct, then we would expect to find leaf surfaces with uniformly distributed stomata.
-However, to the contrary, we find that though 57 of the 132 (43.1%) leaf surfaces were significantly more uniformly dispersed than uniform random synthetic stomatal grids ($\alpha$ = 0.05), none of the leaf surfaces exhibited perfectly uniform stomatal patterning (dispersion index = 1) (Fig.
-\ref{fig:dispersion}).
+However, to the contrary, we find that though 57 of the 132 (43.1%) leaf surfaces were significantly more uniformly dispersed than uniform random synthetic stomatal grids ($\alpha$ = 0.05), none of the leaf surfaces exhibited perfectly uniform stomatal patterning (dispersion index = 1) (\autoref{fig:single-surface}).
 Additionally, we hypothesized that as CO$_2$ is more limiting to photosynthesis under high light, stomata would be more uniformly dispersed in plants grown in high light than plants grown in low and medium light.
@@
 Interestingly, adaxial leaf surfaces were more uniformly dispersed than associated abaxial leaf surfaces across all light treatments (F$_\text{1,126}$ = 28.8; p-value < 0.001).
-Rather than regulate stomatal patterning in response to light regimes, plants respond by increasing stomatal density (Fig.
-\ref{fig:density}).
+Rather than regulate stomatal patterning in response to light regimes, plants respond by increasing stomatal density (\autoref{fig:density}).
 Stomatal density drastically increased in plants grown under high light, validating the long held hypothesis that light strongly influences stomatal density (F$_\text{2,126}$ = 680.7; p-value < 0.001).
 
-Across all light treatments and leaf surfaces, stomatal length and stomatal area were weakly positively correlated, indicating some support for our hypothesis that stomata are selected for an optimal operational aperture (Fig.
-\ref{fig:length-area}).
+Across all light treatments and leaf surfaces, stomatal length and stomatal area were weakly positively correlated, indicating some support for our hypothesis that stomata are selected for an optimal operational aperture (\autoref{fig:length-area}).
 
@@
 
-As evidence against our hypothesis that natural selection should favor spatial coordination in the placement of stomata between abaxial and adaxial leaf surfaces, we found no correlation between paired abaxial and adaxial leaf surfaces (Fig.
-\ref{fig:dual-surface}).
+As evidence against our hypothesis that natural selection should favor spatial coordination in the placement of stomata between abaxial and adaxial leaf surfaces, we found no correlation between paired abaxial and adaxial leaf surfaces (\autoref{fig:dual-surface}).
 Light treatment had no effect on correlation between surfaces (F$_\text{2,63}$ = 2.28; p-value = 0.11).
@@
 We wrapped the above functions to adjust environmental and anatomical variables. We're going to simulate over a grid of parameters, but here is what the output looks like plotted:
-
-### DRAFT TEXT
-
-During a model run, CO$_2$ flux between neighboring nodes was determined according to the mesophyll CO$_2$ concentration ($C_m$) in each node, distance between cells, temperature, pressure, and leaf porosity.
-Assimilation rate ($A$) in each node was determined by $A_\text{max}$, $c_i$, irradiance, temperature, chlorophyll volume concentration, Rubisco octamer concentration per mol chlorophyll, and maximal oxygenation and carboxylation rates of Rubisco.
-Stomata were placed either 1) directly on top of one another to represent anticoordination or 2) offset from one another to represent coordination between leaf surfaces.
-$c_i$ and CO$_2$ assimilation rate (A) were set to 415ppm and 0 respectively and the model ran until $c_i$ and A reached equilibrium in each cell.
-Total A and average $c_i$ were then calculated for each parameter combination.
-
-A was calculated in terms of $A_\text{max}$ and $c_i$ following Eq. \ref{eq:eq4}
-
-\begin{equation}\label{eq:eq4}
-A = A_\text{max} * c_i / (c_i + K_c)
-\end{equation}
-
-where $K_c$ is the Michaelis constant for CO$_2$: 0.0184 mol $m^\text{-3}$.
-
-$A_\text{max}$ was calculated in terms of irradiance (i), light saturated maximum A ($A_\text{mm}$), and quantum yield ($\varphi$) following Eq. \ref{eq:eq5}
-
-\begin{equation}\label{eq:eq5}
-A_\text{max} = A_\text{mm} * \varphi * i / (A_\text{mm} + \varphi * i)
-\end{equation}
 
@@
 Significant differences between light treatments are indicated by asterisks according to analysis of variance followed by a post-hoc tukey honest significant difference test ($\alpha$ = 0.05).}
-\label{fig:dispersion}
+\label{fig:single-surface}
 \end{figure}
@@
 The zone defined by each stomate was calculated with voronoi tessellation and correlated with stomatal length in real leaves.}
-\label{fig:example}
+\label{fig:tessellation-example}
 \end{figure}
@@
 \caption{Idealized amphistomatous stomatal grid with uniform stomatal patterning and perfect abaxial-adaxial coordination.}
-\label{fig:amphi}
+\label{fig:ideal-amphi-grid}
 \end{figure}
@@
 Natural selection operates within developmental and physical constraints to drive each plant species toward its theoretical optimum.
-This study provides evidence that stomata in *Arbidopsis thaliana*, the model angiosperm, are non-randomly distributed, favoring dispersion over clustering (Fig.
-\ref{fig:dispersion}).
-However, stomata are not ideally dispersed in an equilateral triangular grid as would be optimal to minimize CO$_2$ diffusion path length and standardize the area supplied by each stomate (Fig.
-\ref{fig:example}).
+This study provides evidence that stomata in *Arbidopsis thaliana*, the model angiosperm, are non-randomly distributed, favoring dispersion over clustering (\autoref{fig:single-surface}).
+However, stomata are not ideally dispersed in an equilateral triangular grid as would be optimal to minimize CO$_2$ diffusion path length and standardize the area supplied by each stomate (\autoref{fig:tessellation-example}).
 Additionally, when grown in high light environments, *A.
-thaliana* exhibited increased stomatal density rather than increased stomatal dispersion (Fig.
-\ref{fig:density}), which suggests that natural selection has acted more strongly on developmental pathways that modulate stomatal density than those that control stomatal dispersion.
+thaliana* exhibited increased stomatal density rather than increased stomatal dispersion (\autoref{fig:density}), which suggests that natural selection has acted more strongly on developmental pathways that modulate stomatal density than those that control stomatal dispersion.
 In other words, plants optimize gas exchange by adding more stomata rather than dispersing them more evenly across the leaf surface.
-This study also demonstrates that stomata that supply larger leaf areas with CO$_2$ tend to be larger (Fig.
-\ref{fig:length-area}).
+This study also demonstrates that stomata that supply larger leaf areas with CO$_2$ tend to be larger (\autoref{fig:length-area}).
 These results could suggest that 1) the added energetic and hydraulic cost of non-ideally dispersed stomata is negligible and therefore not acted on by natural selection; 2) no developmental pathway exists to ensure the ideal placement of stomata on the leaf; or 3) the regulation of stomatal size limits the cost incurred by non-ideal stomatal dispersion.
@@
 In high light environments, amphistomy is favorable as high light photosynthesis is limited by CO$_2$ and amphistomy halves diffusion path length and boundary layer resistance, thereby reducing CO$_2$ limitation - increasing theoretical $A_\text{max}$.
-An optimal amphistomatous leaf has offset stomata such that stomata are more likely to appear on one leaf surface if there is not a stomata directly opposite it on the other surface as shown in figure \ref{fig:amphi}.
-However, our results show that leaf surfaces are not coordinated but are independent, regardless of light (Fig.
-\ref{fig:dual-surface}).
+An optimal amphistomatous leaf has offset stomata such that stomata are more likely to appear on one leaf surface if there is not a stomata directly opposite it on the other surface as shown in \autoref{fig:ideal-amphi-grid}.
+However, our results show that leaf surfaces are not coordinated but are independent, regardless of light (\autoref{fig:dual-surface}).
 Additionally, gas exchange models show little photosynthetic efficiency gain from abaxial-adaxial stomatal coordination compared to anticoordination (INSERT FIG FROM MODELING).
@@
 Our modelling results demonstrate that, even in ideal conditions, i.e.
-thick leaf, low stomatal densities, high light, low leaf porosity, high rubisco concentration, etc., the photosynthetic advantage of coordination is minimal.
+thick leaf, low stomatal densities, high light, low leaf porosity, high Rubisco concentration, etc., the photosynthetic advantage of coordination is minimal.
 We are not surprised by these results, but still highlight them here as we are the first to report this finding.

### 2023-10-16 r/12_table-model.R

@@ -64,6 +64,4 @@ model_output[, vars] |>
     by = join_by(r)
   ) |>
   select(Variable, `Parameter range`, Units) |>
-  write_rds("objects/model_var_table.rds")
-
-
+  write_rds("objects/model_var.rds")

### 2023-10-16 ms/ms.Rmd
updated caption in fig:density and removed extra model description from ms

@@
  - \renewcommand{\figureautorefname}{Fig.}
+ - \usepackage{booktabs} # for tables
 ---
@@
 We arbitrarily set the adaxial stomate at $x = 0$ and toggled the abaxial stomata position between $x = U/2$ (offset) or $x = 0$ (below adaxial stomate).
-The 'coordination advantage' of offsett stomatal position on each surface is the photosynthetic rate of the leaf with offset stomata compared to that with stomata aligned in the same $x$ position:
+The 'coordination advantage' of offset stomatal position on each surface is the photosynthetic rate of the leaf with offset stomata compared to that with stomata aligned in the same $x$ position:
 
@@
 We modeled the coordination advantage over a range of leaf thicknesses, stomatal densities, photosynthetic capacities, and light environments to understand when offsetting stomatal position on each surface might deliver a significant photosynthetic advantage (\autoref{tab:model_var}).
+The complete model description is available in the Supporting Information.
 
@@
   booktabs = TRUE, format = "latex", escape = FALSE
-) #|>
-#   kableExtra::kable_styling()
-```
-
-WORKING HERE - ???
-
-### Light propogation model
-
-Irradiance at depth $y$ in a leaf with thickness $L$ is modeled following Lloyd *et al.* (1992):
-
-\begin{equation}\label{eq:light_propogation}
-  I(y) = 1.1 I_0 e ^ {-2.4 y / L}
-\end{equation}
-
-where $I_0$ is photosynthetically active irradiance incident on the adaxial leaf surface.
-
-### Biochemical model
-
-All parameter symbols, units, descriptions, and values are described in Table X below.
-Following Gutschick (1984), we modeled photosynthetic rate per unit chlorophyll $A_\text{chl}$ then calculated the volumetric photosynthetic rate $A_\text{volume}$ by multiplying by the chlorophyll concentration $Q_\text{chl}$:
-
-$$A_\text{volume} = A_\text{chl} Q_\text{chl}$$
-A description of the model is given on page 553-556 of Gutschick (1984). The R code below is how we implemented the model.
-
-* `C_m` is a vector CO$_2$ concentrations in $\text{mmol m}^{-3}$ at different positions within the leaf
-* `I_m` is a vector of irradiances in $\mu \text{mol m}^{-2} \text{s}^{-1}$ at different positions within the leaf (same order as `C_m`).
-* `pars` is a list of parameters
-
-```{r, echo = TRUE, eval = FALSE}
-
-pars = list(
-
-    P = set_units(101.325, kPa), # air pressure at sea level
-    temp = set_units(298.15, K), # assume constant temperature
-    R_gas = set_units(8.314, J/K/mol), # ideal gas constant
-
-    # Calculations for C_a and O ued below
-    # 21% O2
-    # set_units(0.21 * P / (R_gas * temp), mol/m^3)
-    # 415 ppm
-    # set_units((415/1e6) * P / (R_gas * temp), mmol/m^3)
-
-    # Environmental
-    C_a = set_units(16.96367, mmol/m^3),
-    O = set_units(8.584027, mol/m^3),
-    PAR = set_units(1000, umol/m^2/s),
-
-    # Biochemical
-    E_t = set_units(0.01, mol/mol),
-    eta_t = set_units(0.59, 1),
-    k_c = set_units(20, mol/mol/s),
-    k_o = set_units(4.2, mol/mol/s),
-    K_c = set_units(0.0184, mol/m^3),
-    K_o = set_units(13.2, mol/m^3),
-    Q_chl = set_units(3.3, mol/m^3),
-    R_p = set_units(0.273, mol/mol),
-
-    # Diffusivity of CO2 in leaf airspace. From Gutschick 1984. Should be updated and include temperature dependence, effect of cell packing, tortuosity, etc.
-    D_mc = set_units(7e-7, m^2/s)
-
 )
-
-# strip units to speed up calculation
-upars = purrr::map(pars, ~ {if(inherits(.x, "units")) {drop_units(.x)}})
-
-# multiply/divide by 1e3 because C_m is mmol and parameters are in mol
-k_vc = upars[["k_c"]] / (1 + (1e3 * upars[["K_c"]] / C_m) *
-                           (1 + upars[["O"]] / upars[["K_o"]])) # [1/s]
-# Given light_propagation assumptions, k = 2.4/L assuming no scattering
-k_i = 2.4 / upars[["leaf_thickness"]] # [1/um] # I'm not sure sure this is right
-b_x = k_i * I_m # [mol/m^3/s]
-a_x = b_x / upars[["Q_chl"]] # [1/s]
-j = 0.5 * a_x * upars[["eta_t"]] # [1/s] # eqn 9
-phi = upars[["k_o"]] / upars[["k_c"]] * (upars[["O"]] / upars[["K_o"]]) /
-  (1e-3 * C_m / upars[["K_c"]]) # [1]
-v_cj = j / (4 + 4 * phi) # [1/s]
-v_cp = k_vc * upars[["R_p"]] # [1/s]
-v_cr = apply(cbind(v_cj, v_cp), 1, min) # [1/s]
-v_c = apply(cbind(k_vc * upars[["E_t"]], v_cr), 1, min) # [1/s]
-v_o = phi * v_c # [1/s]
-A_chl = v_c - 0.5 * v_o # [1/s] # i.e mol CO2 / mol Chl / s
-A_volume = A_chl * upars[["Q_chl"]] # [mol/m^3/s]
 
@@
 
-To model photosynthesis and CO$_2$ transport within a two-dimensional cross section of the leaf, we built a grid of nodes of dimensions leaf thickness by half the interstomatal distance.
-Nodes represent the leaf mesophyll where CO$_2$ diffusion and assimilation occur.
-
-**Table X** Glossary of mathematical symbols. The columns indicate the mathematical Symbol used in the paper, the associated symbol used in R scripts, scientific Units, and a verbal Description.
-
- Symbol              | Value(s) | Units            | Description
----------|--------|--------------------------|----------------------------------
-Biochemical Parameters |||
-$A_\text{volume}$    | NA    | $mol CO_2/m^3/s$    | volumetric assimilation rate
-$A_\text{chl}$       | NA    | $mol CO2/mol Chl/s$ | assimilation rate per mol chlorophyll
-$E_t$                | 0.01  | $mol / mol Chl$     | rubisco octamer concentration per chlorophyll  
-$eta_t$              | 0.59  | unitless            | quantum efficiency of photoexcitation transfer to reaction centers
-$k_c$                | 20    | $mol CO2/mol Rubisco/s$ at 25 C | maximal carboxylation velocity of Rubisco
-$k_o$                | 4.2   | $mol O2/mol Rubisco/s$ at 25 C | maximal oxygenation velocity of Rubisco
-$K_c$                | 0.0184 | $mol/m^3$ at 25 C  | Michaelis constant for CO2
-$K_o$                | 13.2  | $mol/m^3$ at 25 C   | Michaelis constant for O2
-$J_\text{max}$       | 0.253 | $mol e\text{-}/mol Chl/s$ | maximal electron transport rate per mol Chl
-$Q_\text{chl}$       | 3.3   | $mol/m^3$           | Chl volume concentration
-$v_c$                | NA    | $mol CO2/mol Chl/s$ | carboxylation rate
-$v_o$                | NA    | $mol O2/mol Chl/s$  | oxygenation rate
-$k_\text{vc}$        | NA    | $mol CO2/mol Rubisco/mol Chl$ | carboxylation velocity per Rubisco octamer at ambient CO2
-$R_p$                | 0.273 | $mol RuBP/mol Chl$  | RuBP pool size
-$k_i$                | 2.4   | $1/m^2$ or $1/L$    | light attenuation coefficient
-$D_\text{mc}$        | 7e-7  | $m^2/s$             | diffusivity of CO$_2$ in mesophyll airspace
-Environmental Parameters |||
-$P$                  | 101.325 | $kPa$             | air pressure at sea level
-$temp$               | 298.15 | $K$                | temperature
-$R_\text{gas}$       | 8.314 | $J/K/mol$           | ideal gas constant
-$C_a$                | 16.96367 | $mmol CO_2/m^3$  | atmospheric CO$_2$ concentration
-$O$                  | 8.584027 | $mol O_2/m^3$    | atmospheric $O_2$ concentration
-$PAR$                | 1000  | $umol photons/m^2/s$ | photosynthetically active radiation
-Variables |||
-$U_s$                | .05, .10, .15... | $mm$     | interstomatal distance
-$T_l$                | .1, .2, .3... | $mm$        | leaf thickness
-$g_\text{sc}$        | 0.2, 0.3, 0.4... | $mol/m^2/s$ | stomatal conductance
-
-## Two-dimensional model
-
-We used the `steady.2D()` function in the *R* package **rootSolve** version `r packageVersion("rootSolve")` [@soetaert_practical_2009].
-
-```{r, echo = TRUE, eval=FALSE}
-
-diffusion2D <- function (time, state, pars, stomata_offset) {
-
-  n_row = pars[["n_row"]]
-  n_col = pars[["n_col"]]
-
-  n_node = n_row * n_col
-
-  node_length_m = node_length * 1e-6 # node_length in [m]
-
-  # matrix of C_m values
-  C_m_mat = matrix(nrow = n_row, ncol = n_col, state)
-
-  # Photosynthetic demand
-  ## Light matrix
-  I_mat = seq(0, by = pars[["node_length"]], length.out = n_row) |>
-    light_propogation(pars[["leaf_thickness"]], pars[["PAR"]]) |>
-    matrix(n_row, ncol = n_col)
-
-  ## Biochemical model
-  # Units have to be correct for this to work, but unitless = TRUE should be faster
-  A = biochemical_model(C_m_mat, I_mat, pars = pars, unitless = TRUE)
-
-  ## Photosynthesis matrix
-  ## multiply by 1000 to convert from mol / m^3 / s to mmol / m^3 / s
-  A_mat = matrix(nrow = n_row, ncol = n_col, 1e3 * A)
-
-  # CO2 diffusion
-  C_a = drop_units(pars[["C_a"]])
-  flux_mat = matrix(nrow = n_row, ncol = n_col, 0)
-  flux_mat[1, 1] = pars[["g_sc"]] * (C_a - C_m_mat[1, 1])
-
-  ## 1. Flux through stomata
-  if (stomata_offset) {
-    flux_mat[n_row, n_col] = pars[["g_sc"]] * (C_a - C_m_mat[n_row, n_col])
-  } else {
-    flux_mat[n_row, 1] = pars[["g_sc"]] * (C_a - C_m_mat[n_row, 1])
-  }
-
-  zero_x <- rep(0, n_row)
-  zero_y <- rep(0, n_col)
-
-  ## 2. Mesophyll flux; zero fluxes near boundaries
-  flux_above = rbind(rep(0, n_col), C_m_mat[1:(n_row - 1), ] - C_m_mat[2:n_row, ])
-  flux_below = rbind(C_m_mat[2:n_row, ] - C_m_mat[1:(n_row - 1), ], rep(0, n_col))
-  flux_left  = cbind(rep(0, n_col), C_m_mat[, 1:(n_col - 1)] - C_m_mat[, 2:n_col])
-  flux_right = cbind(C_m_mat[, 2:n_col] - C_m_mat[, 1:(n_col - 1)], rep(0, n_col))
-
-  # g_mc [m/s] = D_mc [m^2/s] / node_length [m]
-  flux_mat = flux_mat + pars[["g_mc"]] * (flux_above + flux_below + flux_left + flux_right)
-
-    return(list(c(as.vector(flux_mat / node_length_m - A_mat))))
-
-}
-
-```
-
-## Global Results
+# Results
 
@@
 Leaves were amphistomatous with a mean stomatal ratio of 0.45.
+
+\begin{figure}[ht]
+\includegraphics[width = \textwidth]{figures/density.pdf}
+\caption{Stomatal density is higher in \textit{A.
+thaliana} plants grown under high light conditions.
+We determined statistical significance between light treatments using Tukey's post-hoc tests.
+$^*~0.05 > P \ge 0.01; ^{**}~0.01 > P \ge 0.001; ^{***}~0.0001 > P \ge 0.0001; ^{***}~ P <0.0001$.}
+\label{fig:density}
+\end{figure}
 
@@
 \label{fig:single-surface}
-\end{figure}
-
-\begin{figure}[ht]
-\includegraphics[width = \textwidth]{figures/density.pdf}
-\caption{Stomatal density is higher in plants grown under high light conditions.
-Significant differences between light treatments are indicated by asterisks according to analysis of variance followed by a post-hoc tukey honest significant difference test ($\alpha$ = 0.05).}
-\label{fig:density}
 \end{figure}

### 2023-10-17 ms/ms.Rmd
working on ms results

@@
 fit = readr::read_rds("objects/fit.rds")
+single_surface_results = readr::read_rds("objects/single_surface_results.rds")
 pm_pars = readr::read_csv("raw-data/2d-pm-parameters.csv", show_col_types = FALSE)
 model_var = readr::read_rds("objects/model_var.rds")
+
+# Summarize results on NNI
+nni = single_surface_results |>
+  dplyr::transmute(sig = p_value < 0.05) |>
+  dplyr::summarize(n1 = sum(sig), n2 = sum(!sig))
+
 ```
@@
 This way, each stomate can be optimally sized relative to the mesophyll volume it supplies.
+
+\begin{figure}[ht]
+\includegraphics[width = \textwidth]{figures/ideal-amphi-grid.pdf}
+\caption{Idealized amphistomatous stomatal grid with uniform stomatal patterning and perfect abaxial-adaxial coordination.}
+\label{fig:ideal-amphi-grid}
+\end{figure}
 
@@
 
-Stomatal density of *Arabidopsis thaliana* the 132 leaves measured ranged from 12 to 93 (units) with high light leaves ranging from 93 to 55 (units), medium light from 15 to 35 (units), and low light from 12 to 42 (units).
+Stomatal density of *Arabidopsis thaliana* the `r nrow(single_surface_results)` leaf surfaces measured ranged from 12 to 93 (units) with high light leaves ranging from 93 to 55 (units), medium light from 15 to 35 (units), and low light from 12 to 42 (units).
 Leaves were amphistomatous with a mean stomatal ratio of 0.45.
@@
 
-## Single Surface Results
+<!-- Copying Q's from Intro 
+1) a uniform distribution of stomata on both abaxial and adaxial leaf surfaces and 2) coordinated stomatal spacing on each surface that offsets the position of stomata (\autoref{fig:ideal-amphi-grid}).
+Coordination between leaf surfaces is defined in this study as the occurrence of stomata in areas farthest from stomata on the opposite leaf surface.
+Additionally, because CO$_2$ is more limiting for photosynthesis under high light, we hypothesize that in high light 3) there should be more stomata, and 4) stomata should be more uniformly distributed than in low light.
+Finally, as stomatal densities are selected for optimal operational aperture, we hypothesize that 5) stomatal length will be positively correlated with the area of the leaf surface to which it is closest.
+We refer to this as the 'stomatal zone', the leaf area surrounding a focal stomate closest to that stomate and therefore the zone it supplies with CO$_2$).
+This way, each stomate can be optimally sized relative to the mesophyll volume it supplies.
+-->
 
-If our hypotheses that natural selection will act to reduce the diffusion path length of CO$_2$ while minimizing stomatal number are correct, then we would expect to find leaf surfaces with uniformly distributed stomata.
-However, to the contrary, we find that though 57 of the 132 (43.1%) leaf surfaces were significantly more uniformly dispersed than uniform random synthetic stomatal grids ($\alpha$ = 0.05), none of the leaf surfaces exhibited perfectly uniform stomatal patterning (dispersion index = 1) (\autoref{fig:single-surface}).
-Additionally, we hypothesized that as CO$_2$ is more limiting to photosynthesis under high light, stomata would be more uniformly dispersed in plants grown in high light than plants grown in low and medium light.
+## Stomatal distribution is nonrandom, but far from ideal
+
+<!-- WORKING HERE -->
+Many leaf surface (`r nni[['n1']]` of `r nni[['n1']] + nni[['n2']]`, `r round(100 * nni[['n1']] / (nni[['n1']] + nni[['n2']]), 1)`%) are significantly overdispersed compared to a random uniform distribution, but none were close to an ideal hexagonal pattern (dispersion index = 1; \autoref{fig:single-surface}).
+Dispersion index differs significantly among light treatments (ANOVA, $F = `r$) is significantly greater Additionally, we hypothesized that as CO$_2$ is more limiting to photosynthesis under high light, the dispersion index would closer to 1 in high light...WORKING HERE would be more uniformly dispersed in plants grown in high light than plants grown in low and medium light.
 The data also fail to support this hypothesis as there is no strong, discernible trend between light and stomatal patterning.
@@
 All but one abaxial, adaxial surface pairs were independent ($\alpha$ = 0.05).
-
-### Example output
-
-We wrapped the above functions to adjust environmental and anatomical variables. We're going to simulate over a grid of parameters, but here is what the output looks like plotted:
 
@@
 ## Dual surface analysis
-
-\begin{figure}[ht]
-\includegraphics[width = \textwidth]{figures/ideal-amphi-grid.pdf}
-\caption{Idealized amphistomatous stomatal grid with uniform stomatal patterning and perfect abaxial-adaxial coordination.}
-\label{fig:ideal-amphi-grid}
-\end{figure}
 

### 2023-10-17 r/04_plot-single-surface.R

@@ -6,6 +6,7 @@ single_surface_results = readr::read_rds("objects/single_surface_results.rds") |
 # anova
 single_surface_anova = aov(dispersion ~ light * surface, data = single_surface_results)
 summary(single_surface_anova)
+write_rds(single_surface_anova, "objects/single_surface_anova.rds")
 
 # some preliminary ideas how to make a table for this
 tmp = single_surface_anova |>

### 2023-10-17 r/scratch-ANOVA.R

@@ -1,0 +1,27 @@
+# woorking on stuff to automate results for paper
+source("r/header.R")
+tmp = read_rds("objects/single_surface_results.rds")
+
+hist(tmp$p_value)
+plot(tmp$NNI_observed, log(tmp$p_value))
+
+nni = tmp |>
+  dplyr::transmute(sig = p_value < 0.05) |>
+  dplyr::summarize(n1 = sum(sig), n2 = sum(!sig))
+
+
+single_surface_anova = read_rds("objects/single_surface_anova.rds")
+summary(single_surface_anova)["light", ]
+
+str(single_surface_anova)
+
+single_surface_anova$df.residual
+
+m = single_surface_anova
+x = "light"
+
+glue(
+  "F_{{{df1},{df2}}} = {Fstat}, P = {pval}",
+  df1 = 1,
+  df2 = m$df.residual
+)

### 2023-10-17 ms/ms.Rmd
added multtest results

@@
 # Summarize results on NNI
-nni = single_surface_results |>
+nni1 = single_surface_results |>
   dplyr::transmute(sig = p_value < 0.05) |>
+  dplyr::summarize(n1 = sum(sig), n2 = sum(!sig))
+
+adjp = mt.rawp2adjp(single_surface_results$p_value, proc = "BH")
+
+nni2 = bind_cols(single_surface_results[adjp$index,], adjp$adjp) |>
+  dplyr::transmute(sig = BH < 0.05) |>
   dplyr::summarize(n1 = sum(sig), n2 = sum(!sig))
@@
 The observed stomatal distribution is dispersed relative to a uniform random distribution if the observed $NNI$ is greater than 95% of the synthetic $\mathrm{NNI}$ values (one-tailed test).
+We adjusted $P$-values to account for multiple comparisons using the Benjamini-Hochberg [@benjamini_controlling_1995] false discovery rate procedure implemented in the *R* package **multtest** version `r packageVersion("multtest")` [@pollard_multiple_2005].
 
@@
 <!-- WORKING HERE -->
-Many leaf surface (`r nni[['n1']]` of `r nni[['n1']] + nni[['n2']]`, `r round(100 * nni[['n1']] / (nni[['n1']] + nni[['n2']]), 1)`%) are significantly overdispersed compared to a random uniform distribution, but none were close to an ideal hexagonal pattern (dispersion index = 1; \autoref{fig:single-surface}).
-Dispersion index differs significantly among light treatments (ANOVA, $F = `r$) is significantly greater Additionally, we hypothesized that as CO$_2$ is more limiting to photosynthesis under high light, the dispersion index would closer to 1 in high light...WORKING HERE would be more uniformly dispersed in plants grown in high light than plants grown in low and medium light.
+Many leaf surfaces (`r nni2[['n1']]` of `r nni2[['n1']] + nni2[['n2']]`, `r round(100 * nni2[['n1']] / (nni2[['n1']] + nni2[['n2']]), 1)`%) are significantly overdispersed compared to a random uniform distribution, but none were close to an ideal hexagonal pattern (dispersion index = 1; \autoref{fig:single-surface}).
+Before controlling for multiple comparisons, `r round(100 * nni1[['n1']] / (nni1[['n1']] + nni1[['n2']]), 1)`% are significantly overdispersed.
+The dispersion index differs significantly among light treatments (ANOVA, $F = `r$) is significantly greater Additionally, we hypothesized that as CO$_2$ is more limiting to photosynthesis under high light, the dispersion index would closer to 1 in high light...WORKING HERE would be more uniformly dispersed in plants grown in high light than plants grown in low and medium light.
 The data also fail to support this hypothesis as there is no strong, discernible trend between light and stomatal patterning.

### 2023-10-17 r/header.R

@@ -31,6 +31,7 @@ library(magrittr)
 library(maptools)
 library(metR)
 library(mgcv)
+library(multtest)
 library(parallel)
 library(progress)
 library(progressr)

### 2023-10-17 r/scratch-ANOVA.R

@@ -1,6 +1,6 @@
 # woorking on stuff to automate results for paper
 source("r/header.R")
-tmp = read_rds("objects/single_surface_results.rds")
+single_surface_results = read_rds("objects/single_surface_results.rds")
 
 hist(tmp$p_value)
 plot(tmp$NNI_observed, log(tmp$p_value))
@@ -9,8 +9,15 @@ nni = tmp |>
   dplyr::transmute(sig = p_value < 0.05) |>
   dplyr::summarize(n1 = sum(sig), n2 = sum(!sig))
 
+adjp = mt.rawp2adjp(single_surface_results$p_value, proc = "BH")
+
+bind_cols(single_surface_results[adjp$index,], adjp$adjp)
 
 single_surface_anova = read_rds("objects/single_surface_anova.rds")
+
+
+single_surface_anova
+
 summary(single_surface_anova)["light", ]
 
 str(single_surface_anova)

### 2023-10-17 ms/ms.Rmd
added f statistics for single surface ANOVA results to ms

@@
 single_surface_results = readr::read_rds("objects/single_surface_results.rds")
+single_surface_anova = readr::read_rds("objects/single_surface_anova.rds")
 pm_pars = readr::read_csv("raw-data/2d-pm-parameters.csv", show_col_types = FALSE)
@@
 thaliana} plants grown under high light conditions.
-We determined statistical significance between light treatments using Tukey's post-hoc tests.
+We determined statistical significance between light treatments using Tukey post-hoc tests.
 $^*~0.05 > P \ge 0.01; ^{**}~0.01 > P \ge 0.001; ^{***}~0.0001 > P \ge 0.0001; ^{***}~ P <0.0001$.}
@@
 Before controlling for multiple comparisons, `r round(100 * nni1[['n1']] / (nni1[['n1']] + nni1[['n2']]), 1)`% are significantly overdispersed.
-The dispersion index differs significantly among light treatments (ANOVA, $F = `r$) is significantly greater Additionally, we hypothesized that as CO$_2$ is more limiting to photosynthesis under high light, the dispersion index would closer to 1 in high light...WORKING HERE would be more uniformly dispersed in plants grown in high light than plants grown in low and medium light.
-The data also fail to support this hypothesis as there is no strong, discernible trend between light and stomatal patterning.
-Interestingly, adaxial leaf surfaces were more uniformly dispersed than associated abaxial leaf surfaces across all light treatments (F$_\text{1,126}$ = 28.8; p-value < 0.001).
+The dispersion index differs significantly among light treatments (ANOVA, $`r report_fstat(single_surface_anova, 'light')`$) because the medium light treatment is significantly less than the low treatment (\autoref{fig:single-surface}).
+Dispersion index is consistently greater on adaxial leaf surfaces across all light treatments (ANOVA, $`r report_fstat(single_surface_anova, 'surface')`$; \autoref{fig:single-surface}).
+There is no evidence for an interaction between light treatment and surface (ANOVA, $`r report_fstat(single_surface_anova, 'light:surface')`$).
+
+\begin{figure}[ht]
+\includegraphics[width=\textwidth]{figures/single-surface.pdf}
+\caption{Stomata are more dispersed than expected under the null model of uniform random position (dispersion index = 0) but far from a distribution that maximizes distance between stomata (dispersion index = 1).
+We determined statistical significance between light treatments using Tukey post-hoc tests.
+$^*~0.05 > P \ge 0.01; ^{**}~0.01 > P \ge 0.001; ^{***}~0.0001 > P \ge 0.0001; ^{***}~ P <0.0001$.}
+\label{fig:single-surface}
+\end{figure}
+
 Rather than regulate stomatal patterning in response to light regimes, plants respond by increasing stomatal density (\autoref{fig:density}).
@@
 All but one abaxial, adaxial surface pairs were independent ($\alpha$ = 0.05).
-
-# Results
-
-## Single surface analysis
-
-\begin{figure}[ht]
-\includegraphics[width=\textwidth]{figures/single-surface.pdf}
-\caption{Stomata are more dispersed than expected under the null model of uniform random position (dispersion index = 0) but far from a distribution that maximizes distance between stomata (dispersion index = 1).
-Significant differences between light treatments are indicated by asterisks according to analysis of variance followed by a post-hoc tukey honest significant difference test ($\alpha$ = 0.05).}
-\label{fig:single-surface}
-\end{figure}
 

### 2023-10-17 r/functions.R

@@ -1501,3 +1501,21 @@ plot_photo_2d_pm = function(parms, soln = NULL, ...) {
     theme_void()
 
 }
+
+# Convenience function for reporting F-statistics
+report_fstat = function(m, x) {
+
+  assert_class(m, c("aov", "lm"))
+  s = summary(m)
+  r = str_detect(rownames(s[[1]]), glue("^{x}\\s*$"))
+  assert_true(sum(r) == 1)
+
+  glue(
+    "F_{{{df1},{df2}}} = {Fstat}, P = {pval}",
+    df1 = s[[1]][r, "Df"],
+    df2 = m$df.residual,
+    Fstat = scientize(s[[1]][r, "F value"]),
+    pval = scientize(s[[1]][r, "Pr(>F)"])
+  )
+
+}

### 2023-10-17 r/scratch-ANOVA.R

@@ -15,6 +15,7 @@ bind_cols(single_surface_results[adjp$index,], adjp$adjp)
 
 single_surface_anova = read_rds("objects/single_surface_anova.rds")
 
+TukeyHSD(single_surface_anova)
 
 single_surface_anova
 
@@ -27,8 +28,19 @@ single_surface_anova$df.residual
 m = single_surface_anova
 x = "light"
 
+report_fstat = function(m, x) {
+
+  assert_class(m, c("aov", "lm"))
+  s = summary(m)
+  r = str_detect(rownames(s[[1]]), glue("^{x}\\s*$"))
+  assert_true(sum(r) == 1)
+
 glue(
   "F_{{{df1},{df2}}} = {Fstat}, P = {pval}",
-  df1 = 1,
-  df2 = m$df.residual
+  df1 = s[[1]][r, "Df"],
+  df2 = m$df.residual,
+  Fstat = scientize(s[[1]][r, "F value"]),
+  pval = scientize(s[[1]][r, "Pr(>F)"])
 )
+
+}

### 2023-10-17 ms/ms.Rmd
added stomatal density results to ms

@@
 single_surface_anova = readr::read_rds("objects/single_surface_anova.rds")
+density_anova = readr::read_rds("objects/density_anova.rds")
 pm_pars = readr::read_csv("raw-data/2d-pm-parameters.csv", show_col_types = FALSE)
@@
 
-Stomatal density of *Arabidopsis thaliana* the `r nrow(single_surface_results)` leaf surfaces measured ranged from 12 to 93 (units) with high light leaves ranging from 93 to 55 (units), medium light from 15 to 35 (units), and low light from 12 to 42 (units).
-Leaves were amphistomatous with a mean stomatal ratio of 0.45.
+Stomatal density of *Arabidopsis thaliana* the `r nrow(single_surface_results)` leaf surfaces measured range from 12 to 93 (units) with high light leaves ranging from 93 to 55 (units), medium light from 15 to 35 (units), and low light from 12 to 42 (units).
+Stomatal density varies among light treatments (ANOVA, $`r report_fstat(density_anova, 'light')`$) because the density is much greater in the high light treatment (\autoref{fig:density}).
+Density is consistently greater on abaxial leaf surfaces across all light treatments (ANOVA, $`r report_fstat(density_anova, 'surface')`$; \autoref{fig:density}).
+There is no evidence for an interaction between light treatment and surface (ANOVA, $`r report_fstat(density_anova, 'light:surface')`$).
+Leaves are amphistomatous with a mean stomatal density ratio of 0.45.
 
@@
 
-Rather than regulate stomatal patterning in response to light regimes, plants respond by increasing stomatal density (\autoref{fig:density}).
-Stomatal density drastically increased in plants grown under high light, validating the long held hypothesis that light strongly influences stomatal density (F$_\text{2,126}$ = 680.7; p-value < 0.001).
-
+<!-- WORKING HERE -->
 Across all light treatments and leaf surfaces, stomatal length and stomatal area were weakly positively correlated, indicating some support for our hypothesis that stomata are selected for an optimal operational aperture (\autoref{fig:length-area}).

### 2023-10-17 r/04_plot-single-surface.R

@@ -5,20 +5,7 @@ single_surface_results = readr::read_rds("objects/single_surface_results.rds") |
 
 # anova
 single_surface_anova = aov(dispersion ~ light * surface, data = single_surface_results)
-summary(single_surface_anova)
 write_rds(single_surface_anova, "objects/single_surface_anova.rds")
-
-# some preliminary ideas how to make a table for this
-tmp = single_surface_anova |>
-  broom::tidy()
-  #mutate(p = formatC(tmp$p.value, format = "e", digits = 2))
-tmp
-tmp$p.value = format(tmp$p.value, format = "e", digits = 2)
-tmp
-  mutate(x = if.else(is.na(p.value), NA, scientize(p.value)))
-  rename(SS = sumsq, MS = meansq, `$F$` = statistic, `$P$` = p.value) |>
-
-  knitr::kable()
 
 stat.test <- aov(dispersion ~ light * surface, data = single_surface_results) |>
   tukey_hsd()
@@ -26,7 +13,7 @@ stat.test = stat.test[1:3,]
 
 # anova
 density_anova = aov(n_stomata ~ light * surface, data = single_surface_results)
-summary(density_anova)
+write_rds(density_anova, "objects/density_anova.rds")
 
 stat.test.density <- aov(n_stomata ~ light * surface, data = single_surface_results) |>
   tukey_hsd()

### 2023-10-17 ms/ms.Rmd
added dual surface results to ms

@@
 density_anova = readr::read_rds("objects/density_anova.rds")
+dual_surface_results = readr::read_rds("objects/dual_surface_results.rds")
+correlation_anova = readr::read_rds("objects/correlation_anova.rds")
 pm_pars = readr::read_csv("raw-data/2d-pm-parameters.csv", show_col_types = FALSE)
@@
 nni2 = bind_cols(single_surface_results[adjp$index,], adjp$adjp) |>
+  dplyr::transmute(sig = BH < 0.05) |>
+  dplyr::summarize(n1 = sum(sig), n2 = sum(!sig))
+
+# Summarize results on corr(NSD)
+nsd1 = dual_surface_results |>
+  ungroup() |>
+  dplyr::transmute(sig = p_value < 0.05) |>
+  dplyr::summarize(n1 = sum(sig), n2 = sum(!sig))
+
+adjp = mt.rawp2adjp(dual_surface_results$p_value, proc = "BH")
+
+nsd2 = bind_cols(dual_surface_results[adjp$index,], adjp$adjp) |>
+  ungroup() |>
   dplyr::transmute(sig = BH < 0.05) |>
@@
 
+\begin{figure}[ht]
+\includegraphics[width = \textwidth]{figures/tessellation-example.pdf}
+\caption{Examples of synthetic and real leaf surfaces.
+A) Uniform random synthetic leaf surface; B) Example of real leaf surface; C) Regularly distributed synthetic leaf surface.
+The zone defined by each stomate was calculated with voronoi tessellation and correlated with stomatal length in real leaves.}
+\label{fig:tessellation-example}
+\end{figure}
+
 ## Paired Abaxial and Adaxial Surface Analysis
@@
 To test whether the position of ab- and adaxial stomata are coordinated we compared the observed distribution to a null distribution where the positions on each surface are random.
-For each pair of surfaces (observed or synthetic) we calculated the distance squared between each to the nearest stomatal centroid with the *R* package **raster** version `r packageVersion("raster")`.
-Then we calculated the cell-wise Pearson correlation coefficient.
+For each pair of surfaces (observed or synthetic) we calculated the distance squared between each pixel to the nearest stomatal centroid with the *R* package **raster** version `r packageVersion("raster")`.
+We refer to this as the 'nearest stomatal distance' or NSD.
+Then we calculated the pixel-wise Pearson correlation coefficient.
 If stomatal positions on each surface are coordinated to minimize the distance between mesophyll and the nearest stomate, then we expect a negative correlation.
-A cell that is far from a stomate on one surface should be near a stomate on the other surface (\autoref{fig:ideal-amphi-grid}).
+A pixel that is far from a stomate on one surface should be near a stomate on the other surface (\autoref{fig:ideal-amphi-grid}).
 We generated a null distribution of the correlation coefficient by simulating $10^3$ synthetic data sets for each observed pair.
@@
 
-<!-- Copying Q's from Intro 
-1) a uniform distribution of stomata on both abaxial and adaxial leaf surfaces and 2) coordinated stomatal spacing on each surface that offsets the position of stomata (\autoref{fig:ideal-amphi-grid}).
-Coordination between leaf surfaces is defined in this study as the occurrence of stomata in areas farthest from stomata on the opposite leaf surface.
-Additionally, because CO$_2$ is more limiting for photosynthesis under high light, we hypothesize that in high light 3) there should be more stomata, and 4) stomata should be more uniformly distributed than in low light.
-Finally, as stomatal densities are selected for optimal operational aperture, we hypothesize that 5) stomatal length will be positively correlated with the area of the leaf surface to which it is closest.
-We refer to this as the 'stomatal zone', the leaf area surrounding a focal stomate closest to that stomate and therefore the zone it supplies with CO$_2$).
-This way, each stomate can be optimally sized relative to the mesophyll volume it supplies.
--->
-
 ## Stomatal distribution is nonrandom, but far from ideal
 
-<!-- WORKING HERE -->
 Many leaf surfaces (`r nni2[['n1']]` of `r nni2[['n1']] + nni2[['n2']]`, `r round(100 * nni2[['n1']] / (nni2[['n1']] + nni2[['n2']]), 1)`%) are significantly overdispersed compared to a random uniform distribution, but none were close to an ideal hexagonal pattern (dispersion index = 1; \autoref{fig:single-surface}).
@@
 
-<!-- WORKING HERE -->
-Across all light treatments and leaf surfaces, stomatal length and stomatal area were weakly positively correlated, indicating some support for our hypothesis that stomata are selected for an optimal operational aperture (\autoref{fig:length-area}).
+## No evidence for coordinated stomatal position between surfaces
 
-## Dual Surface Analysis
-
-As evidence against our hypothesis that natural selection should favor spatial coordination in the placement of stomata between abaxial and adaxial leaf surfaces, we found no correlation between paired abaxial and adaxial leaf surfaces (\autoref{fig:dual-surface}).
-Light treatment had no effect on correlation between surfaces (F$_\text{2,63}$ = 2.28; p-value = 0.11).
-All but one abaxial, adaxial surface pairs were independent ($\alpha$ = 0.05).
-
-\begin{figure}[ht]
-\includegraphics[width = \textwidth]{figures/tessellation-example.pdf}
-\caption{Examples of synthetic and real leaf surfaces.
-A) Uniform random synthetic leaf surface; B) Example of real leaf surface; C) Regularly distributed synthetic leaf surface.
-The zone defined by each stomate was calculated with voronoi tessellation and correlated with stomatal length in real leaves.}
-\label{fig:tessellation-example}
-\end{figure}
-
-\begin{figure}[ht]
-\includegraphics[width = \textwidth]{figures/length-area.pdf}
-\caption{Stomatal length and stomatal zone area.
-Lines of best fit computed using a bayesian generalized non-linear multilevel model.
-Each light level and leaf surface exhibits a unique, weakly positive relationship between stomatal zone area and length.}
-\label{fig:length-area}
-\end{figure}
-
-## Dual surface analysis
+There is no evidence of spatial coordination between abaxial and adaxial leaf surfaces.
+The pixel-wise correlation between nearest stomatal distance (NSD) squared on paired abaxial and adaxial leaf surfaces is not significantly less than zero in any of the `r nrow(dual_surface_results)` leaves (\autoref{fig:dual-surface}).
+Before controlling for multiple comparisons, `r round(100 * nsd1[['n1']] / (nsd1[['n1']] + nsd1[['n2']]), 1)`% are significantly  *positively* correlated.
+The NSD correlation is not different among light treatments (ANOVA, $`r report_fstat(correlation_anova, 'light')`$; \autoref{fig:dual-surface}).
 
@@
 \label{fig:dual-surface}
+\end{figure}
+
+<!-- WORKING HERE -->
+<!-- Copying Q's from Intro 
+1) a uniform distribution of stomata on both abaxial and adaxial leaf surfaces and 2) coordinated stomatal spacing on each surface that offsets the position of stomata (\autoref{fig:ideal-amphi-grid}).
+Coordination between leaf surfaces is defined in this study as the occurrence of stomata in areas farthest from stomata on the opposite leaf surface.
+Additionally, because CO$_2$ is more limiting for photosynthesis under high light, we hypothesize that in high light 3) there should be more stomata, and 4) stomata should be more uniformly distributed than in low light.
+Finally, as stomatal densities are selected for optimal operational aperture, we hypothesize that 5) stomatal length will be positively correlated with the area of the leaf surface to which it is closest.
+We refer to this as the 'stomatal zone', the leaf area surrounding a focal stomate closest to that stomate and therefore the zone it supplies with CO$_2$).
+This way, each stomate can be optimally sized relative to the mesophyll volume it supplies.
+-->
+Across all light treatments and leaf surfaces, stomatal length and stomatal area were weakly positively correlated, indicating some support for our hypothesis that stomata are selected for an optimal operational aperture (\autoref{fig:length-area}).
+
+\begin{figure}[ht]
+\includegraphics[width = \textwidth]{figures/length-area.pdf}
+\caption{Stomatal length and stomatal zone area.
+Lines of best fit computed using a bayesian generalized non-linear multilevel model.
+Each light level and leaf surface exhibits a unique, weakly positive relationship between stomatal zone area and length.}
+\label{fig:length-area}
 \end{figure}

### 2023-10-17 r/07_plot-dual-surface.R

@@ -15,11 +15,10 @@ dual_surface_results = readr::read_rds("objects/dual_surface_results.rds") |>
 
 # anova
 correlation_anova = aov(corr ~ light, data = dual_surface_results)
-summary(correlation_anova)
+write_rds(correlation_anova, "objects/correlation_anova.rds")
 
-stat.test.correlation <- aov(corr ~ light, data = dual_surface_results) |>
+stat.test.correlation = aov(corr ~ light, data = dual_surface_results) |>
   tukey_hsd()
-stat.test.correlation
 
 plot = ggplot(dual_surface_results, aes(light_reorder, corr)) +
   stat_boxplot() +

### 2023-10-17 r/scratch-ANOVA.R

@@ -1,46 +1,16 @@
 # woorking on stuff to automate results for paper
 source("r/header.R")
-single_surface_results = read_rds("objects/single_surface_results.rds")
 
-hist(tmp$p_value)
-plot(tmp$NNI_observed, log(tmp$p_value))
+dual_surface_results = readr::read_rds("objects/dual_surface_results.rds")
 
-nni = tmp |>
+nsd1 = dual_surface_results |>
+  ungroup() |>
   dplyr::transmute(sig = p_value < 0.05) |>
   dplyr::summarize(n1 = sum(sig), n2 = sum(!sig))
 
-adjp = mt.rawp2adjp(single_surface_results$p_value, proc = "BH")
+adjp = mt.rawp2adjp(dual_surface_results$p_value, proc = "BH")
 
-bind_cols(single_surface_results[adjp$index,], adjp$adjp)
-
-single_surface_anova = read_rds("objects/single_surface_anova.rds")
-
-TukeyHSD(single_surface_anova)
-
-single_surface_anova
-
-summary(single_surface_anova)["light", ]
-
-str(single_surface_anova)
-
-single_surface_anova$df.residual
-
-m = single_surface_anova
-x = "light"
-
-report_fstat = function(m, x) {
-
-  assert_class(m, c("aov", "lm"))
-  s = summary(m)
-  r = str_detect(rownames(s[[1]]), glue("^{x}\\s*$"))
-  assert_true(sum(r) == 1)
-
-glue(
-  "F_{{{df1},{df2}}} = {Fstat}, P = {pval}",
-  df1 = s[[1]][r, "Df"],
-  df2 = m$df.residual,
-  Fstat = scientize(s[[1]][r, "F value"]),
-  pval = scientize(s[[1]][r, "Pr(>F)"])
-)
-
-}
+nsd2 = bind_cols(dual_surface_results[adjp$index,], adjp$adjp) |>
+  ungroup() |>
+  dplyr::transmute(sig = BH < 0.05) |>
+  dplyr::summarize(n1 = sum(sig), n2 = sum(!sig))

### 2023-10-17 ms/ms.Rmd
added correlation_anova to objects

@@
 
-Finally, we examined the relationship between stomatal zone area and stomatal length using a Bayesian generalized non-linear multilevel model fit with the *R* package **brms** version `r fit$version$brms` [@burkner_brms_2017; @burkner_advanced_2018] and *Stan* version `r fit$version$cmdstan` [@stan_development_team_stan_2023].
+Finally, we examined the relationship between stomatal zone area and stomatal length using a Bayesian linear mixed-effects model fit with the *R* package **brms** version `r fit$version$brms` [@burkner_brms_2017; @burkner_advanced_2018] and *Stan* version `r fit$version$cmdstan` [@stan_development_team_stan_2023].
 Stomatal zone area was calculated using Voronoi tessellation (e.g.
@@
 \includegraphics[width = \textwidth]{figures/dual-surface.pdf}
-\caption{Correlation between paired abaxial and adaxial leaf surfaces.
-Dashed line indicates no correlation.
-Weak positive correlations are not significantly different from null simulations.
-No differences in abaxial-adaxial correlation were observed between light levels according to an analysis of variance ($\alpha$ = 0.05).}
+\caption{Pixel-wise correlation between near stomatal distance (NSD) squared on paired abaxial and adaxial leaf surfaces.
+Dashed line indicates zero correlation.
+Weak positive correlations are not significantly different from zero after correcting for multiple comparisons.
+The correlation does not differ among light treatments.}
 \label{fig:dual-surface}
@@
 
-<!-- WORKING HERE -->
-<!-- Copying Q's from Intro 
-1) a uniform distribution of stomata on both abaxial and adaxial leaf surfaces and 2) coordinated stomatal spacing on each surface that offsets the position of stomata (\autoref{fig:ideal-amphi-grid}).
-Coordination between leaf surfaces is defined in this study as the occurrence of stomata in areas farthest from stomata on the opposite leaf surface.
-Additionally, because CO$_2$ is more limiting for photosynthesis under high light, we hypothesize that in high light 3) there should be more stomata, and 4) stomata should be more uniformly distributed than in low light.
-Finally, as stomatal densities are selected for optimal operational aperture, we hypothesize that 5) stomatal length will be positively correlated with the area of the leaf surface to which it is closest.
-We refer to this as the 'stomatal zone', the leaf area surrounding a focal stomate closest to that stomate and therefore the zone it supplies with CO$_2$).
-This way, each stomate can be optimally sized relative to the mesophyll volume it supplies.
--->
-Across all light treatments and leaf surfaces, stomatal length and stomatal area were weakly positively correlated, indicating some support for our hypothesis that stomata are selected for an optimal operational aperture (\autoref{fig:length-area}).
+## Larger stomata supply larger mesophyll volumes
+
+Across all light treatments and leaf surfaces, stomatal length and stomatal area are weakly positively correlated, indicating some support for our hypothesis that stomata are selected for an optimal operational aperture (\autoref{fig:length-area}).
 
@@
 \includegraphics[width = \textwidth]{figures/length-area.pdf}
-\caption{Stomatal length and stomatal zone area.
-Lines of best fit computed using a bayesian generalized non-linear multilevel model.
-Each light level and leaf surface exhibits a unique, weakly positive relationship between stomatal zone area and length.}
+\caption{Stomatal length and stomatal zone area are positively correlated. Linear regression lines and 95\% confidence ribbons are from a Bayesian linear mixed-effects model.}
 \label{fig:length-area}

### 2023-10-17 r/08_fit-length-area.R

@@ -5,11 +5,12 @@ stomata = readr::read_rds("objects/stomata_position_length_area.rds") |>
   filter(length > 5, !reject, !is.na(area)) |>
   mutate(sqrt_area = sqrt(area))
 
-fit <- brm(bf(
+fit = brm(bf(
   sqrt_area ~ surface + length:light + (1 + surface:length|trt_leaf_number),
   sigma ~ light
 ),
-  data = stomata, backend = "cmdstanr", chains = 1L, seed = 226140189
+  data = stomata, backend = "cmdstanr", chains = 4L, cores = 4L,
+seed = 226140189, iter = 4e3, thin = 2, silent = 0
 )
 
 write_rds(fit, "objects/fit.rds")

### 2023-10-17 ms/ms.Rmd
revised length-area model and added to ms results

@@
  - \usepackage{booktabs} # for tables
+ - \usepackage[detect-none]{siunitx} # for \numrange{}{}
+ - \sisetup{range-phrase = \text{--}}
+
 ---
@@
 correlation_anova = readr::read_rds("objects/correlation_anova.rds")
+fit_emtrend = read_rds("objects/fit_emtrends.rds")
 pm_pars = readr::read_csv("raw-data/2d-pm-parameters.csv", show_col_types = FALSE)
@@
   dplyr::summarize(n1 = sum(sig), n2 = sum(!sig))
+
+# Summarize slopes of length against stomatal zone area
+fit_slopes = gsub(" ", "", print(fit_emtrend, export = TRUE)$summary) |>
+  as_tibble() %>%
+  set_colnames(gsub(" ", "", colnames(.))) |>
+  mutate(light = factor(light, levels = c("low", "medium", "high"))) |>
+  arrange(light, surface) |>
+  pmap_chr(~ {
+    glue("{light} light, {surface} surface: {s}",
+         light = ..1, surface = ..2, s = pretty_est(..3, ..4, ..5))
+  }) |> 
+  str_c(collapse = "; ")
 
@@
 The observed stomatal distribution is dispersed relative to a uniform random distribution if the observed $NNI$ is greater than 95% of the synthetic $\mathrm{NNI}$ values (one-tailed test).
-We adjusted $P$-values to account for multiple comparisons using the Benjamini-Hochberg [@benjamini_controlling_1995] false discovery rate procedure implemented in the *R* package **multtest** version `r packageVersion("multtest")` [@pollard_multiple_2005].
+We adjusted $P$-values to account for multiple comparisons using the Benjamini-Hochberg [@benjamini_controlling_1995] false discovery rate procedure implemented in the *R* package **multtest** version `r packageVersion("multtest")` [@wong_multiple_2005].
 
@@
 Stomatal length was measured in ImageJ [@schneider_nih_2012].
+We modeled fixed effects of surface, light treatment, stomatal length, and their 2- and 3-way interactions on $\sqrt{S_\text{area}}$.
+We included random intercepts, random effects of surface, random slopes, and random surface-by-slope interactions within both plant and individual to account for nonindependence of stomata within the same plant or individual.
+We also modeled residual variance as a function of light treatment.
+We sampled the posterior distribution from `r dim(fit$fit)[2]` chains with `r dim(fit$fit)[1]` iterations each after `r dim(fit$fit)[1]` warmup iterations.
+We calculated convergence diagnostics ($\hat{R}$) and effective sample sizes following @vehtari_rank-normalization_2021.
+We estimated the marginal slope and 95% highest posterior density (HPD) intervals between stomatal length and $\sqrt{S_\text{area}}$ using the *emtrends* function in the *R* package **emmeans** version `r packageVersion('emmeans')` [@lenth_emmeans_2023].
 
@@
 
-We modeled photosynthesis CO$_2$ assimilation rate using a spatially-explicit two-dimensional reaction diffusion model using a porous medium approximation [@parkhurst_intercellular_1994] using the finite element method (FEM) following @earles_excess_2017.
+We modeled photosynthesis CO$_2$ assimilation rate using a spatially-explicit two-dimensional reaction diffusion model using a porous medium approximation [@parkhurst_diffusion_1994] using the finite element method (FEM) following @earles_excess_2017.
 Consider a two-dimensional leaf where stomata occur on each surface in a regular sequence with interstomatal distance $U$.
@@
 
-Across all light treatments and leaf surfaces, stomatal length and stomatal area are weakly positively correlated, indicating some support for our hypothesis that stomata are selected for an optimal operational aperture (\autoref{fig:length-area}).
+All parameters converged ($\hat{R} < 1.01$) and effective sample sizes were exceeded $10^3$.
+Across all light treatments and leaf surfaces, stomatal length and stomatal area are weakly positively correlated (\autoref{fig:length-area}).
+The slope was significantly greater than zero for all abaxial surfaces, but not for the adaxial surface in low and medium light treatments.
+The estimated marginal slopes and 95% HPD intervals for each combination of light and surface is: `r fit_slopes`.
 

### 2023-10-17 r/08_fit-length-area.R

@@ -3,13 +3,14 @@ source("r/header.R")
 stomata = readr::read_rds("objects/stomata_position_length_area.rds") |>
   dplyr::ungroup() |>
   filter(length > 5, !reject, !is.na(area)) |>
-  mutate(sqrt_area = sqrt(area))
+  mutate(sqrt_area = sqrt(area), image = str_c(trt_leaf_number, image_number), sep = "-")
 
 fit = brm(bf(
-  sqrt_area ~ surface + length:light + (1 + surface:length|trt_leaf_number),
+  sqrt_area ~ surface * light * length + (surface * length | trt_leaf_number) +
+    (surface * length | image),
   sigma ~ light
 ),
-  data = stomata, backend = "cmdstanr", chains = 4L, cores = 4L,
+data = stomata, backend = "cmdstanr", chains = 4L, cores = 4L,
 seed = 226140189, iter = 4e3, thin = 2, silent = 0
 )
 

### 2023-10-17 r/09_plot-length-area.R

@@ -1,10 +1,11 @@
 source("r/header.R")
 
-fit <- read_rds("objects/fit.rds")
+fit = read_rds("objects/fit.rds")
 
-summary(fit)
+emtrends(fit, specs = c("light", "surface"), var = "length") |>
+  write_rds("objects/fit_emtrends.rds")
 
-df_new <- fit$data |>
+df_new = fit$data |>
   dplyr::group_by(light, surface) |>
   summarise(length_min = min(length), length_max = max(length)) |>
   rowwise() |>
@@ -12,20 +13,20 @@ df_new <- fit$data |>
     tibble(light = ..1, surface = ..2, length = seq(..3, ..4, length.out = 1e2))
   })
 
-df <- posterior_epred(fit, newdata = df_new, re_formula = NA)
+df = posterior_epred(fit, newdata = df_new, re_formula = NA)
 
-df_med <- df |>
+df_med = df |>
   apply(2, median) |>
   as_tibble(.name_repair = "minimal") |>
   rename(sqrt_area = value)
 
-df_qi <- df |>
+df_qi = df |>
   apply(2, ggdist::qi) |>
   t() |>
   as_tibble(.name_repair = "minimal") |>
   magrittr::set_colnames(c("lower", "upper"))
 
-df_pred <- df_new |>
+df_pred = df_new |>
   bind_cols(df_med, df_qi) |>
   arrange(light)
 

### 2023-10-17 r/functions.R

@@ -1519,3 +1519,12 @@ report_fstat = function(m, x) {
   )
 
 }
+
+# Function to print estimates and CIs
+pretty_est = function(estimate, lower, upper, ...) {
+  .e = prettyNum(estimate, ...)
+  .l = prettyNum(lower, ...)
+  .u = prettyNum(upper, ...)
+  # glue("{.e}~[{.l}-{.u}]")
+  glue("{.e} [\\numrange{{{.l}}}{{{.u}}}]")
+}

### 2023-10-17 r/header.R

@@ -16,6 +16,7 @@ library(checkmate)
 library(corrplot)
 library(cowplot)
 library(dplyr)
+library(emmeans)
 library(forcats)
 library(furrr)
 library(ggdist)

### 2023-10-17 r/scratch-ANOVA.R

@@ -1,16 +1,0 @@
-# woorking on stuff to automate results for paper
-source("r/header.R")
-
-dual_surface_results = readr::read_rds("objects/dual_surface_results.rds")
-
-nsd1 = dual_surface_results |>
-  ungroup() |>
-  dplyr::transmute(sig = p_value < 0.05) |>
-  dplyr::summarize(n1 = sum(sig), n2 = sum(!sig))
-
-adjp = mt.rawp2adjp(dual_surface_results$p_value, proc = "BH")
-
-nsd2 = bind_cols(dual_surface_results[adjp$index,], adjp$adjp) |>
-  ungroup() |>
-  dplyr::transmute(sig = BH < 0.05) |>
-  dplyr::summarize(n1 = sum(sig), n2 = sum(!sig))

### 2023-10-18 ms/figures/model_summary.tex
added results and figure for fem model

% Created by tikzDevice version 0.12.5 on 2023-10-17 20:58:55
% !TEX encoding = UTF-8 Unicode
\documentclass[10pt]{article}
\usepackage{tikz}

\usepackage[active,tightpage,psfixbb]{preview}

\PreviewEnvironment{pgfpicture}

\setlength\PreviewBorder{0pt}
\begin{document}

\begin{tikzpicture}[x=1pt,y=1pt]
\definecolor{fillColor}{RGB}{255,255,255}
\path[use as bounding box,fill=fillColor,fill opacity=0.00] (0,0) rectangle (361.35,361.35);
\begin{scope}
\path[clip] ( 50.69,222.89) rectangle (141.38,307.35);
\definecolor{drawColor}{RGB}{248,118,109}
\definecolor{fillColor}{RGB}{248,118,109}

\path[draw=drawColor,line width= 0.4pt,line join=round,line cap=round,fill=fillColor] ( 58.94,226.84) circle (  1.96);
\definecolor{drawColor}{RGB}{0,186,56}
\definecolor{fillColor}{RGB}{0,186,56}

\path[draw=drawColor,line width= 0.4pt,line join=round,line cap=round,fill=fillColor] ( 58.94,226.84) circle (  1.96);
\definecolor{drawColor}{RGB}{97,156,255}
\definecolor{fillColor}{RGB}{97,156,255}

\path[draw=drawColor,line width= 0.4pt,line join=round,line cap=round,fill=fillColor] ( 58.94,226.84) circle (  1.96);
\definecolor{drawColor}{RGB}{248,118,109}
\definecolor{fillColor}{RGB}{248,118,109}

\path[draw=drawColor,line width= 0.4pt,line join=round,line cap=round,fill=fillColor] ( 75.42,226.84) circle (  1.96);
\definecolor{drawColor}{RGB}{0,186,56}
\definecolor{fillColor}{RGB}{0,186,56}

\path[draw=drawColor,line width= 0.4pt,line join=round,line cap=round,fill=fillColor] ( 75.42,226.84) circle (  1.96);
\definecolor{drawColor}{RGB}{97,156,255}
\definecolor{fillColor}{RGB}{97,156,255}

\path[draw=drawColor,line width= 0.4pt,line join=round,line cap=round,fill=fillColor] ( 75.42,226.84) circle (  1.96);
\definecolor{drawColor}{RGB}{248,118,109}
\definecolor{fillColor}{RGB}{248,118,109}

\path[draw=drawColor,line width= 0.4pt,line join=round,line cap=round,fill=fillColor] (137.26,226.84) circle (  1.96);
\definecolor{drawColor}{RGB}{0,186,56}
\definecolor{fillColor}{RGB}{0,186,56}

\path[draw=drawColor,line width= 0.4pt,line join=round,line cap=round,fill=fillColor] (137.26,226.84) circle (  1.96);
\definecolor{drawColor}{RGB}{97,156,255}
\definecolor{fillColor}{RGB}{97,156,255}

\path[draw=drawColor,line width= 0.4pt,line join=round,line cap=round,fill=fillColor] (137.26,226.84) circle (  1.96);
\end{scope}
\begin{scope}
\path[clip] ( 50.69,131.42) rectangle (141.38,215.89);
\definecolor{drawColor}{RGB}{248,118,109}
\definecolor{fillColor}{RGB}{248,118,109}

\path[draw=drawColor,line width= 0.4pt,line join=round,line cap=round,fill=fillColor] ( 58.94,135.37) circle (  1.96);
\definecolor{drawColor}{RGB}{0,186,56}
\definecolor{fillColor}{RGB}{0,186,56}

\path[draw=drawColor,line width= 0.4pt,line join=round,line cap=round,fill=fillColor] ( 58.94,135.37) circle (  1.96);
\definecolor{drawColor}{RGB}{97,156,255}
\definecolor{fillColor}{RGB}{97,156,255}

\path[draw=drawColor,line width= 0.4pt,line join=round,line cap=round,fill=fillColor] ( 58.94,135.37) circle (  1.96);
\definecolor{drawColor}{RGB}{248,118,109}
\definecolor{fillColor}{RGB}{248,118,109}

\path[draw=drawColor,line width= 0.4pt,line join=round,line cap=round,fill=fillColor] ( 75.42,135.37) circle (  1.96);
\definecolor{drawColor}{RGB}{0,186,56}
\definecolor{fillColor}{RGB}{0,186,56}

\path[draw=drawColor,line width= 0.4pt,line join=round,line cap=round,fill=fillColor] ( 75.42,135.37) circle (  1.96);
\definecolor{drawColor}{RGB}{97,156,255}
\definecolor{fillColor}{RGB}{97,156,255}

\path[draw=drawColor,line width= 0.4pt,line join=round,line cap=round,fill=fillColor] ( 75.42,135.37) circle (  1.96);
\definecolor{drawColor}{RGB}{248,118,109}
\definecolor{fillColor}{RGB}{248,118,109}

\path[draw=drawColor,line width= 0.4pt,line join=round,line cap=round,fill=fillColor] (137.26,135.37) circle (  1.96);
\definecolor{drawColor}{RGB}{0,186,56}
\definecolor{fillColor}{RGB}{0,186,56}

\path[draw=drawColor,line width= 0.4pt,line join=round,line cap=round,fill=fillColor] (137.26,135.37) circle (  1.96);
\definecolor{drawColor}{RGB}{97,156,255}
\definecolor{fillColor}{RGB}{97,156,255}

\path[draw=drawColor,line width= 0.4pt,line join=round,line cap=round,fill=fillColor] (137.26,135.37) circle (  1.96);
\end{scope}
\begin{scope}
\path[clip] ( 50.69, 39.96) rectangle (141.38,124.42);
\definecolor{drawColor}{RGB}{248,118,109}
\definecolor{fillColor}{RGB}{248,118,109}

\path[draw=drawColor,line width= 0.4pt,line join=round,line cap=round,fill=fillColor] ( 58.94, 43.91) circle (  1.96);
\definecolor{drawColor}{RGB}{0,186,56}
\definecolor{fillColor}{RGB}{0,186,56}

\path[draw=drawColor,line width= 0.4pt,line join=round,line cap=round,fill=fillColor] ( 58.94, 43.91) circle (  1.96);
\definecolor{drawColor}{RGB}{97,156,255}
\definecolor{fillColor}{RGB}{97,156,255}

\path[draw=drawColor,line width= 0.4pt,line join=round,line cap=round,fill=fillColor] ( 58.94, 43.91) circle (  1.96);
\definecolor{drawColor}{RGB}{248,118,109}
\definecolor{fillColor}{RGB}{248,118,109}

\path[draw=drawColor,line width= 0.4pt,line join=round,line cap=round,fill=fillColor] ( 75.42, 43.91) circle (  1.96);
\definecolor{drawColor}{RGB}{0,186,56}
\definecolor{fillColor}{RGB}{0,186,56}

\path[draw=drawColor,line width= 0.4pt,line join=round,line cap=round,fill=fillColor] ( 75.42, 43.91) circle (  1.96);
\definecolor{drawColor}{RGB}{97,156,255}
\definecolor{fillColor}{RGB}{97,156,255}

\path[draw=drawColor,line width= 0.4pt,line join=round,line cap=round,fill=fillColor] ( 75.42, 43.91) circle (  1.96);
\definecolor{drawColor}{RGB}{248,118,109}
\definecolor{fillColor}{RGB}{248,118,109}

\path[draw=drawColor,line width= 0.4pt,line join=round,line cap=round,fill=fillColor] (137.26, 43.91) circle (  1.96);
\definecolor{drawColor}{RGB}{0,186,56}
\definecolor{fillColor}{RGB}{0,186,56}

\path[draw=drawColor,line width= 0.4pt,line join=round,line cap=round,fill=fillColor] (137.26, 43.91) circle (  1.96);
\definecolor{drawColor}{RGB}{97,156,255}
\definecolor{fillColor}{RGB}{97,156,255}

\path[draw=drawColor,line width= 0.4pt,line join=round,line cap=round,fill=fillColor] (137.26, 43.91) circle (  1.96);
\end{scope}
\begin{scope}
\path[clip] (148.38,222.89) rectangle (239.07,307.35);
\definecolor{drawColor}{RGB}{248,118,109}
\definecolor{fillColor}{RGB}{248,118,109}

\path[draw=drawColor,line width= 0.4pt,line join=round,line cap=round,fill=fillColor] (156.62,226.86) circle (  1.96);
\definecolor{drawColor}{RGB}{0,186,56}
\definecolor{fillColor}{RGB}{0,186,56}

\path[draw=drawColor,line width= 0.4pt,line join=round,line cap=round,fill=fillColor] (156.62,226.85) circle (  1.96);
\definecolor{drawColor}{RGB}{97,156,255}
\definecolor{fillColor}{RGB}{97,156,255}

\path[draw=drawColor,line width= 0.4pt,line join=round,line cap=round,fill=fillColor] (156.62,226.84) circle (  1.96);
\definecolor{drawColor}{RGB}{248,118,109}
\definecolor{fillColor}{RGB}{248,118,109}

\path[draw=drawColor,line width= 0.4pt,line join=round,line cap=round,fill=fillColor] (173.11,226.98) circle (  1.96);
\definecolor{drawColor}{RGB}{0,186,56}
\definecolor{fillColor}{RGB}{0,186,56}

\path[draw=drawColor,line width= 0.4pt,line join=round,line cap=round,fill=fillColor] (173.11,226.91) circle (  1.96);
\definecolor{drawColor}{RGB}{97,156,255}
\definecolor{fillColor}{RGB}{97,156,255}

\path[draw=drawColor,line width= 0.4pt,line join=round,line cap=round,fill=fillColor] (173.11,226.88) circle (  1.96);
\definecolor{drawColor}{RGB}{248,118,109}
\definecolor{fillColor}{RGB}{248,118,109}

\path[draw=drawColor,line width= 0.4pt,line join=round,line cap=round,fill=fillColor] (234.94,227.39) circle (  1.96);
\definecolor{drawColor}{RGB}{0,186,56}
\definecolor{fillColor}{RGB}{0,186,56}

\path[draw=drawColor,line width= 0.4pt,line join=round,line cap=round,fill=fillColor] (234.94,227.14) circle (  1.96);
\definecolor{drawColor}{RGB}{97,156,255}
\definecolor{fillColor}{RGB}{97,156,255}

\path[draw=drawColor,line width= 0.4pt,line join=round,line cap=round,fill=fillColor] (234.94,227.05) circle (  1.96);
\end{scope}
\begin{scope}
\path[clip] (148.38,131.42) rectangle (239.07,215.89);
\definecolor{drawColor}{RGB}{248,118,109}
\definecolor{fillColor}{RGB}{248,118,109}

\path[draw=drawColor,line width= 0.4pt,line join=round,line cap=round,fill=fillColor] (156.62,135.37) circle (  1.96);
\definecolor{drawColor}{RGB}{0,186,56}
\definecolor{fillColor}{RGB}{0,186,56}

\path[draw=drawColor,line width= 0.4pt,line join=round,line cap=round,fill=fillColor] (156.62,135.37) circle (  1.96);
\definecolor{drawColor}{RGB}{97,156,255}
\definecolor{fillColor}{RGB}{97,156,255}

\path[draw=drawColor,line width= 0.4pt,line join=round,line cap=round,fill=fillColor] (156.62,135.37) circle (  1.96);
\definecolor{drawColor}{RGB}{248,118,109}
\definecolor{fillColor}{RGB}{248,118,109}

\path[draw=drawColor,line width= 0.4pt,line join=round,line cap=round,fill=fillColor] (173.11,135.37) circle (  1.96);
\definecolor{drawColor}{RGB}{0,186,56}
\definecolor{fillColor}{RGB}{0,186,56}

\path[draw=drawColor,line width= 0.4pt,line join=round,line cap=round,fill=fillColor] (173.11,135.37) circle (  1.96);
\definecolor{drawColor}{RGB}{97,156,255}
\definecolor{fillColor}{RGB}{97,156,255}

\path[draw=drawColor,line width= 0.4pt,line join=round,line cap=round,fill=fillColor] (173.11,135.37) circle (  1.96);
\definecolor{drawColor}{RGB}{248,118,109}
\definecolor{fillColor}{RGB}{248,118,109}

\path[draw=drawColor,line width= 0.4pt,line join=round,line cap=round,fill=fillColor] (234.94,135.37) circle (  1.96);
\definecolor{drawColor}{RGB}{0,186,56}
\definecolor{fillColor}{RGB}{0,186,56}

\path[draw=drawColor,line width= 0.4pt,line join=round,line cap=round,fill=fillColor] (234.94,135.37) circle (  1.96);
\definecolor{drawColor}{RGB}{97,156,255}
\definecolor{fillColor}{RGB}{97,156,255}

\path[draw=drawColor,line width= 0.4pt,line join=round,line cap=round,fill=fillColor] (234.94,135.37) circle (  1.96);
\end{scope}
\begin{scope}
\path[clip] (148.38, 39.96) rectangle (239.07,124.42);
\definecolor{drawColor}{RGB}{248,118,109}
\definecolor{fillColor}{RGB}{248,118,109}

\path[draw=drawColor,line width= 0.4pt,line join=round,line cap=round,fill=fillColor] (156.62, 43.91) circle (  1.96);
\definecolor{drawColor}{RGB}{0,186,56}
\definecolor{fillColor}{RGB}{0,186,56}

\path[draw=drawColor,line width= 0.4pt,line join=round,line cap=round,fill=fillColor] (156.62, 43.91) circle (  1.96);
\definecolor{drawColor}{RGB}{97,156,255}
\definecolor{fillColor}{RGB}{97,156,255}

\path[draw=drawColor,line width= 0.4pt,line join=round,line cap=round,fill=fillColor] (156.62, 43.91) circle (  1.96);
\definecolor{drawColor}{RGB}{248,118,109}
\definecolor{fillColor}{RGB}{248,118,109}

\path[draw=drawColor,line width= 0.4pt,line join=round,line cap=round,fill=fillColor] (173.11, 43.91) circle (  1.96);
\definecolor{drawColor}{RGB}{0,186,56}
\definecolor{fillColor}{RGB}{0,186,56}

\path[draw=drawColor,line width= 0.4pt,line join=round,line cap=round,fill=fillColor] (173.11, 43.91) circle (  1.96);
\definecolor{drawColor}{RGB}{97,156,255}
\definecolor{fillColor}{RGB}{97,156,255}

\path[draw=drawColor,line width= 0.4pt,line join=round,line cap=round,fill=fillColor] (173.11, 43.91) circle (  1.96);
\definecolor{drawColor}{RGB}{248,118,109}
\definecolor{fillColor}{RGB}{248,118,109}

\path[draw=drawColor,line width= 0.4pt,line join=round,line cap=round,fill=fillColor] (234.94, 43.91) circle (  1.96);
\definecolor{drawColor}{RGB}{0,186,56}
\definecolor{fillColor}{RGB}{0,186,56}

\path[draw=drawColor,line width= 0.4pt,line join=round,line cap=round,fill=fillColor] (234.94, 43.91) circle (  1.96);
\definecolor{drawColor}{RGB}{97,156,255}
\definecolor{fillColor}{RGB}{97,156,255}

\path[draw=drawColor,line width= 0.4pt,line join=round,line cap=round,fill=fillColor] (234.94, 43.91) circle (  1.96);
\end{scope}
\begin{scope}
\path[clip] (246.07,222.89) rectangle (336.75,307.35);
\definecolor{drawColor}{RGB}{248,118,109}
\definecolor{fillColor}{RGB}{248,118,109}

\path[draw=drawColor,line width= 0.4pt,line join=round,line cap=round,fill=fillColor] (254.31,230.68) circle (  1.96);
\definecolor{drawColor}{RGB}{0,186,56}
\definecolor{fillColor}{RGB}{0,186,56}

\path[draw=drawColor,line width= 0.4pt,line join=round,line cap=round,fill=fillColor] (254.31,228.76) circle (  1.96);
\definecolor{drawColor}{RGB}{97,156,255}
\definecolor{fillColor}{RGB}{97,156,255}

\path[draw=drawColor,line width= 0.4pt,line join=round,line cap=round,fill=fillColor] (254.31,228.14) circle (  1.96);
\definecolor{drawColor}{RGB}{248,118,109}
\definecolor{fillColor}{RGB}{248,118,109}

\path[draw=drawColor,line width= 0.4pt,line join=round,line cap=round,fill=fillColor] (270.80,260.66) circle (  1.96);
\definecolor{drawColor}{RGB}{0,186,56}
\definecolor{fillColor}{RGB}{0,186,56}

\path[draw=drawColor,line width= 0.4pt,line join=round,line cap=round,fill=fillColor] (270.80,242.59) circle (  1.96);
\definecolor{drawColor}{RGB}{97,156,255}
\definecolor{fillColor}{RGB}{97,156,255}

\path[draw=drawColor,line width= 0.4pt,line join=round,line cap=round,fill=fillColor] (270.80,237.10) circle (  1.96);
\definecolor{drawColor}{RGB}{248,118,109}
\definecolor{fillColor}{RGB}{248,118,109}

\path[draw=drawColor,line width= 0.4pt,line join=round,line cap=round,fill=fillColor] (332.63,303.51) circle (  1.96);
\definecolor{drawColor}{RGB}{0,186,56}
\definecolor{fillColor}{RGB}{0,186,56}

\path[draw=drawColor,line width= 0.4pt,line join=round,line cap=round,fill=fillColor] (332.63,277.32) circle (  1.96);
\definecolor{drawColor}{RGB}{97,156,255}
\definecolor{fillColor}{RGB}{97,156,255}

\path[draw=drawColor,line width= 0.4pt,line join=round,line cap=round,fill=fillColor] (332.63,264.42) circle (  1.96);
\end{scope}
\begin{scope}
\path[clip] (246.07,131.42) rectangle (336.75,215.89);
\definecolor{drawColor}{RGB}{248,118,109}
\definecolor{fillColor}{RGB}{248,118,109}

\path[draw=drawColor,line width= 0.4pt,line join=round,line cap=round,fill=fillColor] (254.31,135.46) circle (  1.96);
\definecolor{drawColor}{RGB}{0,186,56}
\definecolor{fillColor}{RGB}{0,186,56}

\path[draw=drawColor,line width= 0.4pt,line join=round,line cap=round,fill=fillColor] (254.31,135.42) circle (  1.96);
\definecolor{drawColor}{RGB}{97,156,255}
\definecolor{fillColor}{RGB}{97,156,255}

\path[draw=drawColor,line width= 0.4pt,line join=round,line cap=round,fill=fillColor] (254.31,135.40) circle (  1.96);
\definecolor{drawColor}{RGB}{248,118,109}
\definecolor{fillColor}{RGB}{248,118,109}

\path[draw=drawColor,line width= 0.4pt,line join=round,line cap=round,fill=fillColor] (270.80,136.17) circle (  1.96);
\definecolor{drawColor}{RGB}{0,186,56}
\definecolor{fillColor}{RGB}{0,186,56}

\path[draw=drawColor,line width= 0.4pt,line join=round,line cap=round,fill=fillColor] (270.80,135.75) circle (  1.96);
\definecolor{drawColor}{RGB}{97,156,255}
\definecolor{fillColor}{RGB}{97,156,255}

\path[draw=drawColor,line width= 0.4pt,line join=round,line cap=round,fill=fillColor] (270.80,135.62) circle (  1.96);
\definecolor{drawColor}{RGB}{248,118,109}
\definecolor{fillColor}{RGB}{248,118,109}

\path[draw=drawColor,line width= 0.4pt,line join=round,line cap=round,fill=fillColor] (332.63,137.15) circle (  1.96);
\definecolor{drawColor}{RGB}{0,186,56}
\definecolor{fillColor}{RGB}{0,186,56}

\path[draw=drawColor,line width= 0.4pt,line join=round,line cap=round,fill=fillColor] (332.63,136.54) circle (  1.96);
\definecolor{drawColor}{RGB}{97,156,255}
\definecolor{fillColor}{RGB}{97,156,255}

\path[draw=drawColor,line width= 0.4pt,line join=round,line cap=round,fill=fillColor] (332.63,136.25) circle (  1.96);
\end{scope}
\begin{scope}
\path[clip] (246.07, 39.96) rectangle (336.75,124.42);
\definecolor{drawColor}{RGB}{248,118,109}
\definecolor{fillColor}{RGB}{248,118,109}

\path[draw=drawColor,line width= 0.4pt,line join=round,line cap=round,fill=fillColor] (254.31, 43.91) circle (  1.96);
\definecolor{drawColor}{RGB}{0,186,56}
\definecolor{fillColor}{RGB}{0,186,56}

\path[draw=drawColor,line width= 0.4pt,line join=round,line cap=round,fill=fillColor] (254.31, 43.91) circle (  1.96);
\definecolor{drawColor}{RGB}{97,156,255}
\definecolor{fillColor}{RGB}{97,156,255}

\path[draw=drawColor,line width= 0.4pt,line join=round,line cap=round,fill=fillColor] (254.31, 43.91) circle (  1.96);
\definecolor{drawColor}{RGB}{248,118,109}
\definecolor{fillColor}{RGB}{248,118,109}

\path[draw=drawColor,line width= 0.4pt,line join=round,line cap=round,fill=fillColor] (270.80, 43.93) circle (  1.96);
\definecolor{drawColor}{RGB}{0,186,56}
\definecolor{fillColor}{RGB}{0,186,56}

\path[draw=drawColor,line width= 0.4pt,line join=round,line cap=round,fill=fillColor] (270.80, 43.92) circle (  1.96);
\definecolor{drawColor}{RGB}{97,156,255}
\definecolor{fillColor}{RGB}{97,156,255}

\path[draw=drawColor,line width= 0.4pt,line join=round,line cap=round,fill=fillColor] (270.80, 43.91) circle (  1.96);
\definecolor{drawColor}{RGB}{248,118,109}
\definecolor{fillColor}{RGB}{248,118,109}

\path[draw=drawColor,line width= 0.4pt,line join=round,line cap=round,fill=fillColor] (332.63, 43.95) circle (  1.96);
\definecolor{drawColor}{RGB}{0,186,56}
\definecolor{fillColor}{RGB}{0,186,56}

\path[draw=drawColor,line width= 0.4pt,line join=round,line cap=round,fill=fillColor] (332.63, 43.94) circle (  1.96);
\definecolor{drawColor}{RGB}{97,156,255}
\definecolor{fillColor}{RGB}{97,156,255}

\path[draw=drawColor,line width= 0.4pt,line join=round,line cap=round,fill=fillColor] (332.63, 43.80) circle (  1.96);
\end{scope}
\begin{scope}
\path[clip] ( 50.69,307.35) rectangle (141.38,324.95);
\definecolor{fillColor}{gray}{0.80}

\path[fill=fillColor] ( 50.69,307.35) rectangle (141.38,324.95);
\definecolor{drawColor}{RGB}{0,0,0}

\node[text=drawColor,anchor=base,inner sep=0pt, outer sep=0pt, scale=  1.20] at ( 96.04,312.02) {$U = 34~\mu \mathrm{m}$};
\end{scope}
\begin{scope}
\path[clip] (148.38,307.35) rectangle (239.07,324.95);
\definecolor{fillColor}{gray}{0.80}

\path[fill=fillColor] (148.38,307.35) rectangle (239.07,324.95);
\definecolor{drawColor}{RGB}{0,0,0}

\node[text=drawColor,anchor=base,inner sep=0pt, outer sep=0pt, scale=  1.20] at (193.72,312.02) {$U = 106~\mu \mathrm{m}$};
\end{scope}
\begin{scope}
\path[clip] (246.07,307.35) rectangle (336.75,324.95);
\definecolor{fillColor}{gray}{0.80}

\path[fill=fillColor] (246.07,307.35) rectangle (336.75,324.95);
\definecolor{drawColor}{RGB}{0,0,0}

\node[text=drawColor,anchor=base,inner sep=0pt, outer sep=0pt, scale=  1.20] at (291.41,312.02) {$U = 338~\mu \mathrm{m}$};
\end{scope}
\begin{scope}
\path[clip] (336.75,222.89) rectangle (354.35,307.35);
\definecolor{fillColor}{gray}{0.80}

\path[fill=fillColor] (336.75,222.89) rectangle (354.35,307.35);
\definecolor{drawColor}{RGB}{0,0,0}

\node[text=drawColor,rotate=-90.00,anchor=base,inner sep=0pt, outer sep=0pt, scale=  1.20] at (341.42,265.12) {$T_\mathrm{leaf} = 100~\mu \mathrm{m}$};
\end{scope}
\begin{scope}
\path[clip] (336.75,131.42) rectangle (354.35,215.89);
\definecolor{fillColor}{gray}{0.80}

\path[fill=fillColor] (336.75,131.42) rectangle (354.35,215.89);
\definecolor{drawColor}{RGB}{0,0,0}

\node[text=drawColor,rotate=-90.00,anchor=base,inner sep=0pt, outer sep=0pt, scale=  1.20] at (341.42,173.66) {$T_\mathrm{leaf} = 300~\mu \mathrm{m}$};
\end{scope}
\begin{scope}
\path[clip] (336.75, 39.96) rectangle (354.35,124.42);
\definecolor{fillColor}{gray}{0.80}

\path[fill=fillColor] (336.75, 39.96) rectangle (354.35,124.42);
\definecolor{drawColor}{RGB}{0,0,0}

\node[text=drawColor,rotate=-90.00,anchor=base,inner sep=0pt, outer sep=0pt, scale=  1.20] at (341.42, 82.19) {$T_\mathrm{leaf} = 500~\mu \mathrm{m}$};
\end{scope}
\begin{scope}
\path[clip] (  0.00,  0.00) rectangle (361.35,361.35);
\definecolor{drawColor}{RGB}{0,0,0}

\path[draw=drawColor,line width= 0.6pt,line join=round,line cap=rect] ( 50.69, 39.96) --
	(141.38, 39.96);
\end{scope}
\begin{scope}
\path[clip] (  0.00,  0.00) rectangle (361.35,361.35);
\definecolor{drawColor}{RGB}{0,0,0}

\path[draw=drawColor,line width= 0.6pt,line join=round] ( 54.81, 36.46) --
	( 54.81, 39.96);

\path[draw=drawColor,line width= 0.6pt,line join=round] ( 96.04, 36.46) --
	( 96.04, 39.96);

\path[draw=drawColor,line width= 0.6pt,line join=round] (137.26, 36.46) --
	(137.26, 39.96);
\end{scope}
\begin{scope}
\path[clip] (  0.00,  0.00) rectangle (361.35,361.35);
\definecolor{drawColor}{RGB}{0,0,0}

\node[text=drawColor,anchor=base,inner sep=0pt, outer sep=0pt, scale=  1.20] at ( 54.81, 25.20) {0};

\node[text=drawColor,anchor=base,inner sep=0pt, outer sep=0pt, scale=  1.20] at ( 96.04, 25.20) {500};

\node[text=drawColor,anchor=base,inner sep=0pt, outer sep=0pt, scale=  1.20] at (137.26, 25.20) {1000};
\end{scope}
\begin{scope}
\path[clip] (  0.00,  0.00) rectangle (361.35,361.35);
\definecolor{drawColor}{RGB}{0,0,0}

\path[draw=drawColor,line width= 0.6pt,line join=round,line cap=rect] (148.38, 39.96) --
	(239.07, 39.96);
\end{scope}
\begin{scope}
\path[clip] (  0.00,  0.00) rectangle (361.35,361.35);
\definecolor{drawColor}{RGB}{0,0,0}

\path[draw=drawColor,line width= 0.6pt,line join=round] (152.50, 36.46) --
	(152.50, 39.96);

\path[draw=drawColor,line width= 0.6pt,line join=round] (193.72, 36.46) --
	(193.72, 39.96);

\path[draw=drawColor,line width= 0.6pt,line join=round] (234.94, 36.46) --
	(234.94, 39.96);
\end{scope}
\begin{scope}
\path[clip] (  0.00,  0.00) rectangle (361.35,361.35);
\definecolor{drawColor}{RGB}{0,0,0}

\node[text=drawColor,anchor=base,inner sep=0pt, outer sep=0pt, scale=  1.20] at (152.50, 25.20) {0};

\node[text=drawColor,anchor=base,inner sep=0pt, outer sep=0pt, scale=  1.20] at (193.72, 25.20) {500};

\node[text=drawColor,anchor=base,inner sep=0pt, outer sep=0pt, scale=  1.20] at (234.94, 25.20) {1000};
\end{scope}
\begin{scope}
\path[clip] (  0.00,  0.00) rectangle (361.35,361.35);
\definecolor{drawColor}{RGB}{0,0,0}

\path[draw=drawColor,line width= 0.6pt,line join=round,line cap=rect] (246.07, 39.96) --
	(336.75, 39.96);
\end{scope}
\begin{scope}
\path[clip] (  0.00,  0.00) rectangle (361.35,361.35);
\definecolor{drawColor}{RGB}{0,0,0}

\path[draw=drawColor,line width= 0.6pt,line join=round] (250.19, 36.46) --
	(250.19, 39.96);

\path[draw=drawColor,line width= 0.6pt,line join=round] (291.41, 36.46) --
	(291.41, 39.96);

\path[draw=drawColor,line width= 0.6pt,line join=round] (332.63, 36.46) --
	(332.63, 39.96);
\end{scope}
\begin{scope}
\path[clip] (  0.00,  0.00) rectangle (361.35,361.35);
\definecolor{drawColor}{RGB}{0,0,0}

\node[text=drawColor,anchor=base,inner sep=0pt, outer sep=0pt, scale=  1.20] at (250.19, 25.20) {0};

\node[text=drawColor,anchor=base,inner sep=0pt, outer sep=0pt, scale=  1.20] at (291.41, 25.20) {500};

\node[text=drawColor,anchor=base,inner sep=0pt, outer sep=0pt, scale=  1.20] at (332.63, 25.20) {1000};
\end{scope}
\begin{scope}
\path[clip] (  0.00,  0.00) rectangle (361.35,361.35);
\definecolor{drawColor}{RGB}{0,0,0}

\path[draw=drawColor,line width= 0.6pt,line join=round,line cap=rect] ( 50.69,222.89) --
	( 50.69,307.35);
\end{scope}
\begin{scope}
\path[clip] (  0.00,  0.00) rectangle (361.35,361.35);
\definecolor{drawColor}{RGB}{0,0,0}

\node[text=drawColor,anchor=base east,inner sep=0pt, outer sep=0pt, scale=  1.20] at ( 44.19,222.70) {0.00};

\node[text=drawColor,anchor=base east,inner sep=0pt, outer sep=0pt, scale=  1.20] at ( 44.19,242.28) {0.01};

\node[text=drawColor,anchor=base east,inner sep=0pt, outer sep=0pt, scale=  1.20] at ( 44.19,261.85) {0.02};

\node[text=drawColor,anchor=base east,inner sep=0pt, outer sep=0pt, scale=  1.20] at ( 44.19,281.42) {0.03};

\node[text=drawColor,anchor=base east,inner sep=0pt, outer sep=0pt, scale=  1.20] at ( 44.19,300.99) {0.04};
\end{scope}
\begin{scope}
\path[clip] (  0.00,  0.00) rectangle (361.35,361.35);
\definecolor{drawColor}{RGB}{0,0,0}

\path[draw=drawColor,line width= 0.6pt,line join=round] ( 47.19,226.84) --
	( 50.69,226.84);

\path[draw=drawColor,line width= 0.6pt,line join=round] ( 47.19,246.41) --
	( 50.69,246.41);

\path[draw=drawColor,line width= 0.6pt,line join=round] ( 47.19,265.98) --
	( 50.69,265.98);

\path[draw=drawColor,line width= 0.6pt,line join=round] ( 47.19,285.55) --
	( 50.69,285.55);

\path[draw=drawColor,line width= 0.6pt,line join=round] ( 47.19,305.12) --
	( 50.69,305.12);
\end{scope}
\begin{scope}
\path[clip] (  0.00,  0.00) rectangle (361.35,361.35);
\definecolor{drawColor}{RGB}{0,0,0}

\path[draw=drawColor,line width= 0.6pt,line join=round,line cap=rect] ( 50.69,131.42) --
	( 50.69,215.89);
\end{scope}
\begin{scope}
\path[clip] (  0.00,  0.00) rectangle (361.35,361.35);
\definecolor{drawColor}{RGB}{0,0,0}

\node[text=drawColor,anchor=base east,inner sep=0pt, outer sep=0pt, scale=  1.20] at ( 44.19,131.24) {0.00};

\node[text=drawColor,anchor=base east,inner sep=0pt, outer sep=0pt, scale=  1.20] at ( 44.19,150.81) {0.01};

\node[text=drawColor,anchor=base east,inner sep=0pt, outer sep=0pt, scale=  1.20] at ( 44.19,170.38) {0.02};

\node[text=drawColor,anchor=base east,inner sep=0pt, outer sep=0pt, scale=  1.20] at ( 44.19,189.96) {0.03};

\node[text=drawColor,anchor=base east,inner sep=0pt, outer sep=0pt, scale=  1.20] at ( 44.19,209.53) {0.04};
\end{scope}
\begin{scope}
\path[clip] (  0.00,  0.00) rectangle (361.35,361.35);
\definecolor{drawColor}{RGB}{0,0,0}

\path[draw=drawColor,line width= 0.6pt,line join=round] ( 47.19,135.37) --
	( 50.69,135.37);

\path[draw=drawColor,line width= 0.6pt,line join=round] ( 47.19,154.94) --
	( 50.69,154.94);

\path[draw=drawColor,line width= 0.6pt,line join=round] ( 47.19,174.52) --
	( 50.69,174.52);

\path[draw=drawColor,line width= 0.6pt,line join=round] ( 47.19,194.09) --
	( 50.69,194.09);

\path[draw=drawColor,line width= 0.6pt,line join=round] ( 47.19,213.66) --
	( 50.69,213.66);
\end{scope}
\begin{scope}
\path[clip] (  0.00,  0.00) rectangle (361.35,361.35);
\definecolor{drawColor}{RGB}{0,0,0}

\path[draw=drawColor,line width= 0.6pt,line join=round,line cap=rect] ( 50.69, 39.96) --
	( 50.69,124.42);
\end{scope}
\begin{scope}
\path[clip] (  0.00,  0.00) rectangle (361.35,361.35);
\definecolor{drawColor}{RGB}{0,0,0}

\node[text=drawColor,anchor=base east,inner sep=0pt, outer sep=0pt, scale=  1.20] at ( 44.19, 39.78) {0.00};

\node[text=drawColor,anchor=base east,inner sep=0pt, outer sep=0pt, scale=  1.20] at ( 44.19, 59.35) {0.01};

\node[text=drawColor,anchor=base east,inner sep=0pt, outer sep=0pt, scale=  1.20] at ( 44.19, 78.92) {0.02};

\node[text=drawColor,anchor=base east,inner sep=0pt, outer sep=0pt, scale=  1.20] at ( 44.19, 98.49) {0.03};

\node[text=drawColor,anchor=base east,inner sep=0pt, outer sep=0pt, scale=  1.20] at ( 44.19,118.06) {0.04};
\end{scope}
\begin{scope}
\path[clip] (  0.00,  0.00) rectangle (361.35,361.35);
\definecolor{drawColor}{RGB}{0,0,0}

\path[draw=drawColor,line width= 0.6pt,line join=round] ( 47.19, 43.91) --
	( 50.69, 43.91);

\path[draw=drawColor,line width= 0.6pt,line join=round] ( 47.19, 63.48) --
	( 50.69, 63.48);

\path[draw=drawColor,line width= 0.6pt,line join=round] ( 47.19, 83.05) --
	( 50.69, 83.05);

\path[draw=drawColor,line width= 0.6pt,line join=round] ( 47.19,102.62) --
	( 50.69,102.62);

\path[draw=drawColor,line width= 0.6pt,line join=round] ( 47.19,122.20) --
	( 50.69,122.20);
\end{scope}
\begin{scope}
\path[clip] (  0.00,  0.00) rectangle (361.35,361.35);
\definecolor{drawColor}{RGB}{0,0,0}

\node[text=drawColor,anchor=base,inner sep=0pt, outer sep=0pt, scale=  1.40] at (193.72,  9.72) {$I_0~[\mu \mathrm{mol}~\mathrm{m}^{-2}~\mathrm{s}^{-1}]$};
\end{scope}
\begin{scope}
\path[clip] (  0.00,  0.00) rectangle (361.35,361.35);
\definecolor{drawColor}{RGB}{0,0,0}

\node[text=drawColor,rotate= 90.00,anchor=base,inner sep=0pt, outer sep=0pt, scale=  1.40] at ( 16.64,173.66) {coordination advantage};
\end{scope}
\begin{scope}
\path[clip] (  0.00,  0.00) rectangle (361.35,361.35);
\definecolor{drawColor}{RGB}{0,0,0}

\node[text=drawColor,anchor=base west,inner sep=0pt, outer sep=0pt, scale=  1.40] at ( 50.69,341.83) {$\varphi_\mathrm{pal}$};
\end{scope}
\begin{scope}
\path[clip] (  0.00,  0.00) rectangle (361.35,361.35);
\definecolor{drawColor}{RGB}{248,118,109}
\definecolor{fillColor}{RGB}{248,118,109}

\path[draw=drawColor,line width= 0.4pt,line join=round,line cap=round,fill=fillColor] ( 90.18,346.65) circle (  1.96);
\end{scope}
\begin{scope}
\path[clip] (  0.00,  0.00) rectangle (361.35,361.35);
\definecolor{drawColor}{RGB}{0,186,56}
\definecolor{fillColor}{RGB}{0,186,56}

\path[draw=drawColor,line width= 0.4pt,line join=round,line cap=round,fill=fillColor] (134.91,346.65) circle (  1.96);
\end{scope}
\begin{scope}
\path[clip] (  0.00,  0.00) rectangle (361.35,361.35);
\definecolor{drawColor}{RGB}{97,156,255}
\definecolor{fillColor}{RGB}{97,156,255}

\path[draw=drawColor,line width= 0.4pt,line join=round,line cap=round,fill=fillColor] (179.64,346.65) circle (  1.96);
\end{scope}
\begin{scope}
\path[clip] (  0.00,  0.00) rectangle (361.35,361.35);
\definecolor{drawColor}{RGB}{0,0,0}

\node[text=drawColor,anchor=base west,inner sep=0pt, outer sep=0pt, scale=  1.20] at (104.88,342.52) {0.1};
\end{scope}
\begin{scope}
\path[clip] (  0.00,  0.00) rectangle (361.35,361.35);
\definecolor{drawColor}{RGB}{0,0,0}

\node[text=drawColor,anchor=base west,inner sep=0pt, outer sep=0pt, scale=  1.20] at (149.61,342.52) {0.2};
\end{scope}
\begin{scope}
\path[clip] (  0.00,  0.00) rectangle (361.35,361.35);
\definecolor{drawColor}{RGB}{0,0,0}

\node[text=drawColor,anchor=base west,inner sep=0pt, outer sep=0pt, scale=  1.20] at (194.34,342.52) {0.3};
\end{scope}
\end{tikzpicture}

\end{document}

### 2023-10-18 ms/ms.Rmd
added results and figure for fem model

@@
   This is the abstract.
-
   It consists of two paragraphs.
@@
   This is an acknowledgement.
-
   It consists of two paragraphs.
@@
     latex_engine: xelatex
+    oup_version: 1 # 1 = 2020 CTAN OUP CLS package; 0 = 2009 OUP CLS package
+    # journal: "AoB PLANTS"
+    document_style: "modern" # Can be contemporary, modern, traditional
+    papersize: "large" # Can be large, medium, small
+    #citation_package: "default" # Uncomment when using a CSL; default "natbib"
+    namedate: TRUE # Set FALSE to use numeric refs; Default FALSE
+    number_sections: FALSE # Uncomment to not number sections; default TRUE
+    #number_lines: TRUE # Use `lineno` package - Default FALSE
+    #number_lines_options: ["mathlines","switch"]  # Options for latex lineno package.
+    onecolumn: TRUE # Uncomment for one column format; default FALSE
+    extra_dependencies:
+      - booktabs # to use with knitr::kable() example below
 #urlcolor: orange
@@
  - \renewcommand{\figureautorefname}{Fig.}
- - \usepackage{booktabs} # for tables
  - \usepackage[detect-none]{siunitx} # for \numrange{}{}
  - \sisetup{range-phrase = \text{--}}
+ - \usepackage{caption} # for \captionsetup{}
 
@@
 
-# print(xtable::xtable(model_var, caption = paste0("The parameter range of model variables tested for their effect on coordination advantage (\\autoref{eq:coordination_advantage}) using a 2-D porous medium approximation.
-We used regularly spaced values within each range and simulated across all combinations.
-Here we converted model units to more conventional units (e.g.
-m to $\\mu$m).
-", symbol_descriptions), label = "tab:model_var"),
-#       comment = FALSE, include.rownames = FALSE, type = "latex")
-
 model_var_caption = paste0("The parameter range of model variables tested for their effect on coordination advantage (\\autoref{eq:coordination_advantage}) using a 2-D porous medium approximation.
@@
 
+## Little benefit of coordinated stomatal arrangement
+
+We used the finite element method (FEM) to model CO$_2$ diffusion within the leaf and photosynthesis as a 2-D porous medium.
+Across all realistic parts of parameter space, the coordination advantage is much less than 0.01 (\autoref{fig:model_summary}).
+For reference, a log-response of ratio is 0.01 is approximately 1%.
+The only exception was for thin leaves ($T_\text{leaf} = 100~\mu \text{m}$) with few stomata ($U = 338~\mu \text{m}$, which corresponds to a stomatal density of $\approx 10~\text{mm}^{-2}$), where lateral diffusion is major constraint on CO$_2$ supply.
+However, such thin leaves with so few stomata are uncommon among C$_3$ plants (some CAM plants have low stomatal density [@males_stomatal_2017]).
+In other areas of parameter space, lateral diffusion limitations were small relative to those along the ab-adaxial axis (see \autoref{fig:model_example} for a representative model solution).
+
+\begin{figure}[ht]
+\includegraphics[width = \textwidth]{figures/model_summary.pdf}
+\caption{There is little photosynthetic benefit of offsetting stomatal position each surface based on a 2-D model of photosynthesis.
+The coordination advantage (\autoref{eq:coordination_advantage}) is close to zero under nearly all of the parameter space \autoref{tab:model_var}, meaning that the photosynthetic rate of amphistomatous leaves with stomata optimally offset is nearly equal to leaves with stomata on each surface in the same position along the leaf plane.
+`r symbol_descriptions`.}
+\label{fig:model_summary}
+\end{figure}
+
 # Discussion
 
-Stomata are expensive.
+Stomata cost resources to maintain [@deans_optimization_2020] and expose leaves to risks such as hydraulic failure [@wang_theoretical_2020] or infection by plant pathogens [@melotto_stomatal_2017].
 A theoretical, optimized plant would minimize stomatal density while also allowing competitive gas exchange rates for its environment so as to maximize C assimilation per unit investment in stomata.
 Natural selection operates within developmental and physical constraints to drive each plant species toward its theoretical optimum.
-This study provides evidence that stomata in *Arbidopsis thaliana*, the model angiosperm, are non-randomly distributed, favoring dispersion over clustering (\autoref{fig:single-surface}).
+This study provides evidence that stomata in *Arabidopsis thaliana*, the model angiosperm, are non-randomly distributed, favoring dispersion over clustering (\autoref{fig:single-surface}).
 However, stomata are not ideally dispersed in an equilateral triangular grid as would be optimal to minimize CO$_2$ diffusion path length and standardize the area supplied by each stomate (\autoref{fig:tessellation-example}).
@@
 # References
+
+<!-- Placeholders for supplemental figures
+
+\renewcommand\thefigure{S\arabic{figure}}    
+\renewcommand\thetable{S\arabic{table}}    
+\renewcommand\theequation{S\arabic{equation}}    
+\setcounter{figure}{0}    
+\setcounter{table}{0}    
+\setcounter{equation}{0}
+
+\begin{figure}
+  \captionsetup{labelformat=empty}
+  \caption{}
+  \label{fig:model_example}
+\end{figure}
+-->

### 2023-10-18 ms/ms.tex
added results and figure for fem model

% Options for packages loaded elsewhere
\PassOptionsToPackage{unicode}{hyperref}

% Set document class options
\documentclass[webpdf,large,modern,unnumsec,namedate]{oup-authoring-template}

% one column
\onecolumn

%\usepackage{showframe}

% line numbers

% use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}

% From Pandoc template for its feature
\usepackage{xcolor}
\usepackage{hyperref}

\hypersetup{
  pdftitle={Spatial Coordination Of Stomatal Patterning Between Leaf Surfaces In Amphistomatous Arabidopsis thaliana Incurs No Photosynthetic Advantage},
  pdfkeywords={amphistomy, Arabidopsis thaliana, CO\(_2\)
diffusion, finite element method, optimality, photosynthesis, stomata},
  breaklinks=true,
  bookmarks=true,
  hidelinks,
  pdfcreator={LaTeX via pandoc}}

% tightlist command for lists without linebreak
\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}

% Counters for addresses and footnotes
\newcounter{correspcnt} % For author footnotes
\renewcommand*{\thecorrespcnt}{\fnsymbol{correspcnt}}
\newcounter{addrcnt} % For author addresses

% Macros for dealing with affiliations, footnotes, etc.
\makeatletter

\def\MyNewLabel#1#2#3{\expandafter\gdef\csname #1@#2\endcsname{#3}}

\def\MyRef#1#2{\@ifundefined{#1@#2}{???}{\csname #1@#2\endcsname}}

\newcommand*\ifcounter[1]{%
  \ifcsname c@#1\endcsname
    \expandafter\@firstoftwo
  \else
    \expandafter\@secondoftwo
  \fi
}

\newcommand*\addrlblbycode[1]{%
  \ifcounter{ADDRLBL@#1}
    {}
    {\refstepcounter{addrcnt}\newcounter{ADDRLBL@#1}\setcounter{ADDRLBL@#1}{\value{addrcnt}}}%
    \arabic{ADDRLBL@#1}%
}

\newcommand*\addrbycode[1]{%
  \ifcounter{ADDR@#1}
    {}
    {\newcounter{ADDR@#1}%
     \address[\addrlblbycode{#1}]{\MyRef{ADDRTXT}{#1}}}%
}

\newcommand*\corresplblbycode[1]{%
  \ifcounter{CORRESPLBL@#1}
    {}
    {\refstepcounter{correspcnt}\newcounter{CORRESPLBL@#1}\setcounter{CORRESPLBL@#1}{\value{correspcnt}}}%
    \fnsymbol{CORRESPLBL@#1}%
}

\newcommand*\correspbycode[1]{%
  \ifcounter{CORRESP@#1}
    {}
    {\newcounter{CORRESP@#1}%
     \corresp[\corresplblbycode{#1}]{\MyRef{CORRESPTXT}{#1}}}%
}

\makeatother

% Add missing \city command mentioned in documentation but absent from cls
\providecommand\city[1]{#1}

% Create labels for Addresses if the are given in Elsevier format
   \MyNewLabel{ADDRTXT}{Colgate University}{%
  13 Oak Drive, Hamilton, NY 13346%
 }
   \MyNewLabel{ADDRTXT}{University of Hawaii at Manoa}{%
  2500 Campus Rd, Honolulu, HI 96822%
 }

% Create labels for Footnotes if they are given in Elsevier format
\MyNewLabel{CORRESPTXT}{1}{Equal contribution}
\MyNewLabel{CORRESPTXT}{2}{Current email address:
\href{mailto:cat@example.com}{cat@example.com}}

% Pandoc header-include feature
\usepackage[nomarkers,tablesfirst]{endfloat}
\usepackage{lineno}
\linenumbers
\usepackage{hyperref}
\renewcommand{\figureautorefname}{Fig.}
\usepackage[detect-none]{siunitx}
\sisetup{range-phrase = \text{--}}
\usepackage{caption}
% Pandoc header-include feature
\usepackage{booktabs}

\begin{document}

\journaltitle{Journal Title Here}
\DOI{DOI HERE}
\copyrightyear{YYYY}
\pubyear{YYYY}
\access{Advance Access Publication Date: Day Month Year}
\appnotes{Paper}

\firstpage{1}

\title[]{Spatial Coordination Of Stomatal Patterning Between Leaf
Surfaces In Amphistomatous \emph{Arabidopsis thaliana} Incurs No
Photosynthetic Advantage}

\newcounter{thisauthcorresp} % For storage if author is corresponding author
\newcounter{thisauththanks} % For storage if author has thanks

\author[%
\addrlblbycode{Colgate University}%
,\refstepcounter{correspcnt}\setcounter{thisauthcorresp}{\value{correspcnt}}\fnsymbol{thisauthcorresp}%
%
%
]{Jacob L. Watts}

\addrbycode{Colgate University}

\corresp[\fnsymbol{thisauthcorresp}]{Corresponding author. \href{mailto:jwatts@colgate.edu}{\nolinkurl{jwatts@colgate.edu}}}

\author[%
\addrlblbycode{ETH}%
%
%
%
]{Graham Dow}

\addrbycode{ETH}

\author[%
\addrlblbycode{University of California Davis}%
%
%
,\corresplblbycode{1}%
]{Thomas N. Buckley}

\addrbycode{University of California Davis}

\correspbycode{1}

\author[%
\addrlblbycode{University of Hawaii at Manoa}%
%
%
,\corresplblbycode{1}%
]{Christopher D. Muir}

\addrbycode{University of Hawaii at Manoa}

\correspbycode{1}

% Add author mark
\authormark{Jacob L. Watts et al.}

\received{Date}{0}{Year}
\revised{Date}{0}{Year}
\accepted{Date}{0}{Year}

%\editor{Associate Editor: Name}

\abstract{
This is the abstract. It consists of two paragraphs.}

\keywords{amphistomy; Arabidopsis thaliana; CO\(_2\) diffusion; finite
element method; optimality; photosynthesis; stomata}

\maketitle

\hypertarget{introduction}{%
\section{Introduction}\label{introduction}}

Stomatal anatomy (e.g.~size, density, distribution, and patterning) and
movement regulate gas exchange during photosynthesis, namely CO\(_2\)
assimilation and water loss through transpiration. Since waxy cuticles
are mostly impermeable to CO\(_2\) and H\(_2\)O, stomata are the primary
entry points through which gas exchange occurs despite making up a small
percentage of the leaf area \citep{lange_responses_1971}. Stomata
consist of two guard cells which open and close upon changes in turgor
pressure or hormonal cues \citep{mcadam_linking_2016}. The stomatal pore
leads to an internal space known as the substomatal cavity where gases
contact the mesophyll. Once in the mesophyll, CO\(_2\) diffuses
throughout a network of intercellular air space (IAS) and into mesophyll
cells where CO\(_2\) assimilation (\(A\)) occurs within the chloroplasts
\citep{lee_diffusion_1964}. Stomatal conductance and transpiration are
determined by numerous environmental and anatomical parameters such as
vapor pressure deficit (VPD), irradiance, temperature, wind speed, leaf
water potential, IAS geometry, mesophyll cell anatomy, and stomatal
anatomy.

Many successful predictions about stomata and other leaf traits can be
made by hypothesizing that natural selection should optimizes CO\(_2\)
gain per unit of water loss
\citep{cowan_stomatal_1977, buckley_optimal_2017, sperry_predicting_2017}.
However, stomatal anatomy may be partially constrained by physical and
developments limits on phenotypic expression
\citep{croxdale_stomatal_2000, harrison_influence_2020, muir_how_2023}.
Sometimes optimization leads to similar phenotypes across many disparate
species. For example, almost all stomata follow the one cell spacing
rule to maintain proper stomatal functioning
\citep{geisler_oriented_2000, dow_physiological_2014}; however some
species (notably in \emph{Begonia}) appear to benefit from overlapping
vapor shells caused by stomatal clustering
\citep{yi_gan_stomatal_2010, lehmann_effects_2015, papanatsiou_stomatal_2017}.
Stomatal traits also vary adaptively in different environments. Stomatal
density positively co-varies with irradiance during leaf development and
negatively co-varies with CO\(_2\) concentration
\citep{gay_influence_1975, schoch_dependence_1980, woodward_stomatal_1987, royer_stomatal_2001},
consistent with optimality predictions. Stomatal size is jointly
controlled by genome size, light, and stomatal density
\citep{jordan_environmental_2015}. Size positively co-varies with genome
size \citep{roddy_scaling_2020} and negatively co-varies with stomatal
density \citep{camargo_density_2011}. Total stomatal area (size
\(\times\) density) is optimized for operational conductance
(\(gs_\text{s,op}\)) rather than maximum conductance
(\(gs_\text{s,max}\)) such that stomatal apertures are most responsive
to changes in the environment at their operational aperture
\citep{franks_physiological_2012, liu_scaling_2021}. Stomatal aperture
can compensate for maladaptive stomatal densities to an extent
\citep{bussis_stomatal_2006}, but stomatal density and size ultimately
determine a leaf's theoretical \(gs_\text{s,max}\)
\citep{sack_developmental_2016}, which is proportional to
\(gs_\text{s,op}\) \citep{murray_consistent_2020}. Additionally, low
stomatal densities lead to irregular and insufficient CO\(_2\) supply
and reduced photosynthetic efficiency in areas far from stomata
\citep{pieruschka_lateral_2006, morison_lateral_2005}, while high
stomatal densities can reduce water use efficiency (WUE)
\citep{bussis_stomatal_2006} and incur excessive metabolic costs
\citep{deans_optimization_2020}. In most species, stomata occur on the
abaxial (usually lower) leaf surface; but amphistomy, the occurrence of
stomata on both abaxial and adaxial leaf surfaces, is also prevalent in
high light environments with constant or intermittent access to
sufficient water
\citep{mott_adaptive_1982, jordan_using_2014, muir_light_2018, drake_two_2019, muir_is_2019}.
Amphistomy effectively halves the CO\(_2\) diffusion path length and
boundary layer resistance by doubling boundary layer conductance
\citep{parkhurst_adaptive_1978, harrison_influence_2020, mott_amphistomy_1991}.
Historically, stomatal patterning in dicot angiosperms was thought to be
random with an exclusionary distance surrounding each stomate
\citep{sachs_developmental_1974}; however, the developmental controls of
stomatal patterning are poorly understood and likely more complex than
random development along the leaf surface.
\citet{croxdale_stomatal_2000}{]} reviews three developmental theories
which attempt to explain stomatal patterning in angiosperms: inhibition,
cell lineage, and cell cycle, ultimately arguing for a cell cycle based
control of stomatal patterning.

The patterning and spacing of stomata on the leaf affects photosynthesis
in \(C_3\) leaves by altering the CO\(_2\) diffusion path length from
stomata to sites of carboxylation in the mesophyll. Maximum
photosynthetic rate (\(A_\text{max}\)) in \(C_3\) plants is generally
co-limited by biochemistry and diffusion, but modulated by light
availability
\citep{parkhurst_intercellular_1990, manter_ci_2004, carriqui_diffusional_2015}.
Low light decreasing CO\(_2\) demand by limiting electron transport
rate, leading to relatively high internal CO\(_2\) concentration
(\(C_\text{i}\)) and low \(A_\text{max}\) \citep{kaiser_metabolic_2016}.
In contrast, well hydrated leaves with open stomata in high light,
photosynthesis is often limited by CO\(_2\) supply as resistances from
the boundary layer, stomatal pore, and mesophyll can result in
insufficient CO\(C_2\) supply at the chloroplast to maxmimize
photosynthesis \citep{farquhar_biochemical_1980, lehmeier_cell_2017}. In
this study, we focus primarily on how stomatal patterning affects
diffusion, ignoring boundary layer and mesophyll resistances.

To maximize CO\(_2\) supply from the stomatal pore to chloroplasts,
stomata should be uniformly distributed in an equilateral triangular
grid on the leaf surface so as to minimize stomatal number and CO\(_2\)
diffusion path length \citep{parkhurst_diffusion_1994}. As the diffusion
rate of CO\(_2\) though liquid is approximately \(10^4\times\) slower
than CO\(_2\) diffusion through air, mesophyll resistance is generally
thought to be primarily limited by liquid diffusion
\citep{aalto_three-dimensional_2002, evans_resistances_2009}, but
diffusion through the IAS has also been shown to be a rate limiting
process because the tortuous, disjunct nature of the IAS can greatly
increase diffusion path lengths \citep{harwood_understanding_2021}.
Additionally, tortuosity is higher in horizontal directions (parallel to
leaf surface) than vertical directions (perpendicular to leaf surface)
because of the cylindrical shape and vertical arrangement of pallisade
mesophyll cells \citep{earles_beyond_2018, harwood_understanding_2021}.
However, the ratio of lateral to vertical diffusion rate is still
largely unknown
\citep{morison_lateral_2005, pieruschka_lateral_2005, pieruschka_lateral_2006}.
Depending on the thickness of the leaf, porosity of the leaf mesophyll,
tortuosity of the IAS, and lateral to vertical diffusion rate ratio,
minimizing diffusion path length for CO\(_2\) via optimally distributed
stomata may yield significant increases in CO\(_2\) supply for
photosynthesis and higher \(A_\text{max}\).

We hypothesized that natural selection will favor stomatal patterning
and distribution to minimize the diffusion path length. In
amphistomatous leaves, this would be accomplished by 1) a uniform
distribution of stomata on both abaxial and adaxial leaf surfaces and 2)
coordinated stomatal spacing on each surface that offsets the position
of stomata (\autoref{fig:ideal-amphi-grid}). Coordination between leaf
surfaces is defined in this study as the occurrence of stomata in areas
farthest from stomata on the opposite leaf surface. Additionally,
because CO\(_2\) is more limiting for photosynthesis under high light,
we hypothesize that in high light 3) there should be more stomata, and
4) stomata should be more uniformly distributed than in low light.
Finally, as stomatal densities are selected for optimal operational
aperture, we hypothesize that 5) stomatal length will be positively
correlated with the area of the leaf surface to which it is closest. We
refer to this as the `stomatal zone', the leaf area surrounding a focal
stomate closest to that stomate and therefore the zone it supplies with
CO\(_2\)). This way, each stomate can be optimally sized relative to the
mesophyll volume it supplies.

\begin{figure}[ht]
\includegraphics[width = \textwidth]{figures/ideal-amphi-grid.pdf}
\caption{Idealized amphistomatous stomatal grid with uniform stomatal patterning and perfect abaxial-adaxial coordination.}
\label{fig:ideal-amphi-grid}
\end{figure}

To test these hypotheses, we grew the model plant \emph{Arabidopsis
thaliana} in high, medium, and low light and measured stomatal density,
size, and patterning on both leaf surfaces, and spatial coordination
between them. We use Voronoi tessellation techniques to calculate
stomatal zones. We also used a 2-D porous medium approximation of
CO\(_2\) diffusion and photosynthesis to predict the photosynthetic
advantage of optimal versus suboptimal coordination in stomatal
coordination between surfaces. Specifically, we predicted that traits
which affect diffusion path length (leaf thickness, stomatal density,
leaf porosity, lateral-vertical diffusion rate ratio), diffusion rate
(temperature, pressure), and CO\(_2\) demand (Rubisco concentration,
light) would modulate the advantage of optimal stomatal arrangement
following the relationships outlined in \autoref{tab:hypotheses}. Here,
we integrate over reasonable parameter space to determine the
ecophysiological context most likely to favor stomatal spatial
coordination in amphistomatous leaves.

\begin{table}[ht]
\centering
\begin{tabular}{ll}
  \hline
trait & relationship \\ 
  \hline
leaf thickness & + \\ 
  stomatal density & - \\ 
  leaf porosity & - \\ 
  lat.-vert. diffusion ratio & - \\ 
  temperature & - \\ 
  pressure & - \\ 
  Rubisco concentration & + \\ 
  light & + \\ 
   \hline
\end{tabular}
\caption{A summary of the hypothesized relationships between leaf traits and environmental conditions and photosynthetic advantage of stomatal spatial coordination in amphistomatous leaves.} 
\label{tab:hypotheses}
\end{table}

\hypertarget{materials-and-methods}{%
\section{Materials and methods}\label{materials-and-methods}}

\hypertarget{data-preparation}{%
\subsection{Data Preparation}\label{data-preparation}}

{[}CDM: Graham Dow provided these images. We'll need to add him as a
co-author and ask him to write methods on image acquisition.{]}

\emph{Arabidopsis thaliana} plants were grown in three different light
environments: low light (PAR = 50
\(\mu \text{mol}~\text{m}^{-2}~\text{s}^{-1}\)), medium light (100
\(\mu \text{mol}~\text{m}^{-2}~\text{s}^{-1}\)), and high light (200
\(\mu \text{mol}~\text{m}^{-2}~\text{s}^{-1}\)). PAR stands for
photosynthetically active radiation. Once leaves were mature, we
captured images of the abaxial and adaxial leaf surfaces using XXX. We
captured 132 images in total, making 66 abaxial-adaxial image pairs. We
measured stomatal position and size using ImageJ
\citep{schneider_nih_2012}.

\hypertarget{single-surface-analyses}{%
\subsection{Single surface analyses}\label{single-surface-analyses}}

We tested whether stomata are non-randomly distributed by comparing the
observed stomatal patterning to a random uniform pattern. For each leaf
surface image with \(n\) stomata we generated \(10^3\) synthetic
surfaces with \(n\) stomata uniformly randomly distributed on the
surface. For each sample image, we compared the observed Nearest
Neighbor Index (\(\mathrm{NNI}\)) to the null distribution of
\(\mathrm{NNI}\) values calculated from the synthetic data set.
\(\mathrm{NNI}\) is the ratio of observed mean distance
(\(\overline{D}_O\)) to the expected mean distance (\(\overline{D}_E\))
where \(\overline{D}_E\) is:

\begin{equation}\label{eq:emd}
  \overline{D}_E = \frac{0.5}{\sqrt{A_\text{leaf} / n_\text{stomata}}}.
\end{equation}

\noindent \(A_\text{leaf}\) is leaf area visible in the sampled field
and \(n_\text{stomata}\) the number of stomata. \(\overline{D}_E\) is
the theoretical average distance to the nearest neighbor of each stomate
if stomata were uniformly randomly distributed
\citep{clark_distance_1954}. \(\overline{D}_O\) calculated for each
synthetic data set is:

\begin{equation}\label{eq:omd}
  \overline{D}_O = \frac{\sum_{i=1}^{n_\text{stomata}}d_i}{n_\text{stomata}},
\end{equation}

\noindent where \(d_i\) is the distance between \(\text{stomate}_i\) and
its nearest neighbor. We calculated \(\mathrm{NNI}\) using the \emph{R}
package \textbf{spatialEco} version 2.0.1 \citep{evans_spatialeco_2023}.
The observed stomatal distribution is dispersed relative to a uniform
random distribution if the observed \(NNI\) is greater than 95\% of the
synthetic \(\mathrm{NNI}\) values (one-tailed test). We adjusted
\(P\)-values to account for multiple comparisons using the
Benjamini-Hochberg \citep{benjamini_controlling_1995} false discovery
rate procedure implemented in the \emph{R} package \textbf{multtest}
version 2.56.0 \citep{wong_multiple_2005}.

For each sample image, we also simulated \(10^3\) synthetic data with
\(n\) stomata ideally dispersed in an equilateral triangular grid. For
these grids, we integrated over plausible stomatal densities and then
conditioned on stomatal grids with exactly \(n\) stomata. The simulated
stomatal count was drawn from a Poisson distribution with the mean
parameter \(\lambda\) drawn from a Gamma distribution with shape \(n\)
and scale 1 \(\lambda \sim \Gamma(n, 1)\). \(\Gamma(n, 1)\) is the
posterior distribution of \(\lambda\) with a flat prior distribution.
This allows us to integrate over uncertainty in the stomatal density
from the sample image.

We developed a dispersion index \(\mathrm{DI}\) to quantify how close
observed stomatal distributions are to random uniform versus maximally
dispered in an equilateral triangular grid. \(\mathrm{DI}\) varies from
zero to one, where zero is uniformly random and one is ideally
dispersed:

\begin{equation}\label{eq:disp}
  \mathrm{DI} = \frac{\mathrm{NNI} - \text{median}(\mathrm{NNI_{random}})}{\text{median}(\mathrm{NNI_{uniform}}) - \text{median}(\mathrm{NNI_{random}})}
\end{equation}

\noindent \(\mathrm{NNI}\) is calculated for each sample image as
described above; \(\text{median}(\mathrm{NNI_{random}})\) and
\(\text{median}(\mathrm{NNI_{uniform}})\) are calculated from the
synthetic data specific to each sample image as described above. We
tested whether light treatment affects \(\mathrm{DI}\) and stomatla
density (\(D_S\)) using analysis of variance (ANOVA).

Finally, we examined the relationship between stomatal zone area and
stomatal length using a Bayesian linear mixed-effects model fit with the
\emph{R} package \textbf{brms} version 2.20.4
\citep{burkner_brms_2017, burkner_advanced_2018} and \emph{Stan} version
2.33.1 \citep{stan_development_team_stan_2023}. Stomatal zone area was
calculated using Voronoi tessellation
(e.g.~\autoref{fig:tessellation-example}). The stomatal zone area,
\(S_\text{area}\), is the region of the leaf surface whose distance to
stomate, \(S\), is less than the distance to any other stomate, \(S\).
Stomatal length was measured in ImageJ \citep{schneider_nih_2012}. We
modeled fixed effects of surface, light treatment, stomatal length, and
their 2- and 3-way interactions on \(\sqrt{S_\text{area}}\). We included
random intercepts, random effects of surface, random slopes, and random
surface-by-slope interactions within both plant and individual to
account for nonindependence of stomata within the same plant or
individual. We also modeled residual variance as a function of light
treatment. We sampled the posterior distribution from 4 chains with 1000
iterations each after 1000 warmup iterations. We calculated convergence
diagnostics (\(\hat{R}\)) and effective sample sizes following
\citet{vehtari_rank-normalization_2021}. We estimated the marginal slope
and 95\% highest posterior density (HPD) intervals between stomatal
length and \(\sqrt{S_\text{area}}\) using the \emph{emtrends} function
in the \emph{R} package \textbf{emmeans} version 1.8.8
\citep{lenth_emmeans_2023}.

\begin{figure}[ht]
\includegraphics[width = \textwidth]{figures/tessellation-example.pdf}
\caption{Examples of synthetic and real leaf surfaces.  A) Uniform random synthetic leaf surface; B) Example of real leaf surface; C) Regularly distributed synthetic leaf surface. The zone defined by each stomate was calculated with voronoi tessellation and correlated with stomatal length in real leaves.}
\label{fig:tessellation-example}
\end{figure}

\hypertarget{paired-abaxial-and-adaxial-surface-analysis}{%
\subsection{Paired Abaxial and Adaxial Surface
Analysis}\label{paired-abaxial-and-adaxial-surface-analysis}}

To test whether the position of ab- and adaxial stomata are coordinated
we compared the observed distribution to a null distribution where the
positions on each surface are random. For each pair of surfaces
(observed or synthetic) we calculated the distance squared between each
pixel to the nearest stomatal centroid with the \emph{R} package
\textbf{raster} version 3.6.26. We refer to this as the `nearest
stomatal distance' or NSD. Then we calculated the pixel-wise Pearson
correlation coefficient. If stomatal positions on each surface are
coordinated to minimize the distance between mesophyll and the nearest
stomate, then we expect a negative correlation. A pixel that is far from
a stomate on one surface should be near a stomate on the other surface
(\autoref{fig:ideal-amphi-grid}). We generated a null distribution of
the correlation coefficient by simulating \(10^3\) synthetic data sets
for each observed pair. For each synthetic data set, we simulated
stomatal position using a random uniform distribution, as described
above, matching the number of stomata on abaxial and adaxial leaf
surfaces. Stomatal positions on each surface are coordinated if the
correlation coefficient is greater than 95\% of the synthetic
correlation values (one-tailed test).

\hypertarget{modeling-photosynthesis}{%
\subsection{Modeling Photosynthesis}\label{modeling-photosynthesis}}

We modeled photosynthesis CO\(_2\) assimilation rate using a
spatially-explicit two-dimensional reaction diffusion model using a
porous medium approximation \citep{parkhurst_diffusion_1994} using the
finite element method (FEM) following \citet{earles_excess_2017}.
Consider a two-dimensional leaf where stomata occur on each surface in a
regular sequence with interstomatal distance \(U\). The main outcome we
assessed is the advantage of offsetting the position of stomata on each
surface compared to have stomata on the same \(x\) position on each
surface. With these assumptions, by symmetry, we only need to model two
stomata, one abaxial and one adaxial, from \(x = 0\) to \(x = U/2\) and
from the adaxial surface at \(y = 0\) to the abaxial surface at
\(y = L\), the leaf thickness. We arbitrarily set the adaxial stomate at
\(x = 0\) and toggled the abaxial stomata position between \(x = U/2\)
(offset) or \(x = 0\) (below adaxial stomate). The `coordination
advantage' of offset stomatal position on each surface is the
photosynthetic rate of the leaf with offset stomata compared to that
with stomata aligned in the same \(x\) position:

\begin{equation} \label{eq:coordination_advantage}
  \text{coordination advantage} = \frac{A_\text{offset}}{A_\text{aligned}}
\end{equation}

We modeled the coordination advantage over a range of leaf thicknesses,
stomatal densities, photosynthetic capacities, and light environments to
understand when offsetting stomatal position on each surface might
deliver a significant photosynthetic advantage
(\autoref{tab:model_var}). The complete model description is available
in the Supporting Information.

\begin{table}

\caption{\label{tab:model_var}The parameter range of model variables tested for their effect on coordination advantage (\autoref{eq:coordination_advantage}) using a 2-D porous medium approximation. We used regularly spaced values within each range and simulated across all combinations. Here we converted model units to more conventional units (e.g. m to $\mu$m). $I_0$: PPFD incident on the leaf surface; $\varphi_\text{pal}$: Fraction of intercellular airspace (aka porosity), palisade; $T_\text{leaf}$: Leaf thickness; $U$: Interstomatal distance}
\centering
\begin{tabular}[t]{lll}
\toprule
Variable & Parameter range & Units\\
\midrule
$I_0$ & $50-1000$ & $\mu$mol m$^{-2}$ s$^{-1}$\\
$\varphi_\text{pal}$ & $0.1-0.3$ & m$^3$ airspace m$^{-3}$ leaf\\
$T_\text{leaf}$ & $101-501$ & $\mu$m\\
$U$ & $17-169$ & $\mu$m\\
\bottomrule
\end{tabular}
\end{table}

\hypertarget{results}{%
\section{Results}\label{results}}

Stomatal density of \emph{Arabidopsis thaliana} the 132 leaf surfaces
measured range from 12 to 93 (units) with high light leaves ranging from
93 to 55 (units), medium light from 15 to 35 (units), and low light from
12 to 42 (units). Stomatal density varies among light treatments (ANOVA,
\(F_{2,126} = 681, P = 2.88 \times 10^{-68}\)) because the density is
much greater in the high light treatment (\autoref{fig:density}).
Density is consistently greater on abaxial leaf surfaces across all
light treatments (ANOVA, \(F_{1,126} = 44.2, P = 8.21 \times 10^{-10}\);
\autoref{fig:density}). There is no evidence for an interaction between
light treatment and surface (ANOVA,
\(F_{2,126} = 2.75 \times 10^{-2}, P = 0.973\)). Leaves are
amphistomatous with a mean stomatal density ratio of 0.45.

\begin{figure}[ht]
\includegraphics[width = \textwidth]{figures/density.pdf}
\caption{Stomatal density is higher in \textit{A. thaliana} plants grown under high light conditions. We determined statistical significance between light treatments using Tukey post-hoc tests. $^*~0.05 > P \ge 0.01; ^{**}~0.01 > P \ge 0.001; ^{***}~0.0001 > P \ge 0.0001; ^{***}~ P <0.0001$.}
\label{fig:density}
\end{figure}

\hypertarget{stomatal-distribution-is-nonrandom-but-far-from-ideal}{%
\subsection{Stomatal distribution is nonrandom, but far from
ideal}\label{stomatal-distribution-is-nonrandom-but-far-from-ideal}}

Many leaf surfaces (37 of 132, 28\%) are significantly overdispersed
compared to a random uniform distribution, but none were close to an
ideal hexagonal pattern (dispersion index = 1;
\autoref{fig:single-surface}). Before controlling for multiple
comparisons, 43.2\% are significantly overdispersed. The dispersion
index differs significantly among light treatments (ANOVA,
\(F_{2,126} = 8.55, P = 3.30 \times 10^{-4}\)) because the medium light
treatment is significantly less than the low treatment
(\autoref{fig:single-surface}). Dispersion index is consistently greater
on adaxial leaf surfaces across all light treatments (ANOVA,
\(F_{1,126} = 28.8, P = 3.67 \times 10^{-7}\);
\autoref{fig:single-surface}). There is no evidence for an interaction
between light treatment and surface (ANOVA,
\(F_{2,126} = 0.577, P = 0.563\)).

\begin{figure}[ht]
\includegraphics[width=\textwidth]{figures/single-surface.pdf}
\caption{Stomata are more dispersed than expected under the null model of uniform random position (dispersion index = 0) but far from a distribution that maximizes distance between stomata (dispersion index = 1). We determined statistical significance between light treatments using Tukey post-hoc tests. $^*~0.05 > P \ge 0.01; ^{**}~0.01 > P \ge 0.001; ^{***}~0.0001 > P \ge 0.0001; ^{***}~ P <0.0001$.}
\label{fig:single-surface}
\end{figure}

\hypertarget{no-evidence-for-coordinated-stomatal-position-between-surfaces}{%
\subsection{No evidence for coordinated stomatal position between
surfaces}\label{no-evidence-for-coordinated-stomatal-position-between-surfaces}}

There is no evidence of spatial coordination between abaxial and adaxial
leaf surfaces. The pixel-wise correlation between nearest stomatal
distance (NSD) squared on paired abaxial and adaxial leaf surfaces is
not significantly less than zero in any of the 66 leaves
(\autoref{fig:dual-surface}). Before controlling for multiple
comparisons, 3\% are significantly \emph{positively} correlated. The NSD
correlation is not different among light treatments (ANOVA,
\(F_{2,63} = 2.28, P = 0.111\); \autoref{fig:dual-surface}).

\begin{figure}[ht]
\includegraphics[width = \textwidth]{figures/dual-surface.pdf}
\caption{Pixel-wise correlation between near stomatal distance (NSD) squared on paired abaxial and adaxial leaf surfaces. Dashed line indicates zero correlation. Weak positive correlations are not significantly different from zero after correcting for multiple comparisons. The correlation does not differ among light treatments.}
\label{fig:dual-surface}
\end{figure}

\hypertarget{larger-stomata-supply-larger-mesophyll-volumes}{%
\subsection{Larger stomata supply larger mesophyll
volumes}\label{larger-stomata-supply-larger-mesophyll-volumes}}

All parameters converged (\(\hat{R} < 1.01\)) and effective sample sizes
were exceeded \(10^3\). Across all light treatments and leaf surfaces,
stomatal length and stomatal area are weakly positively correlated
(\autoref{fig:length-area}). The slope was significantly greater than
zero for all abaxial surfaces, but not for the adaxial surface in low
and medium light treatments. The estimated marginal slopes and 95\% HPD
intervals for each combination of light and surface is: low light,
abaxial surface: 1.928 {[}\numrange{0.779}{3.133}{]}; low light, adaxial
surface: 1.745 {[}\numrange{-0.041}{3.373}{]}; medium light, abaxial
surface: 1.085 {[}\numrange{0.328}{1.957}{]}; medium light, adaxial
surface: 0.656 {[}\numrange{-0.399}{1.691}{]}; high light, abaxial
surface: 0.597 {[}\numrange{0.316}{0.911}{]}; high light, adaxial
surface: 1.269 {[}\numrange{0.831}{1.721}{]}.

\begin{figure}[ht]
\includegraphics[width = \textwidth]{figures/length-area.pdf}
\caption{Stomatal length and stomatal zone area are positively correlated. Linear regression lines and 95\% confidence ribbons are from a Bayesian linear mixed-effects model.}
\label{fig:length-area}
\end{figure}

\hypertarget{little-benefit-of-coordinated-stomatal-arrangement}{%
\subsection{Little benefit of coordinated stomatal
arrangement}\label{little-benefit-of-coordinated-stomatal-arrangement}}

We used the finite element method (FEM) to model CO\(_2\) diffusion
within the leaf and photosynthesis as a 2-D porous medium. Across all
realistic parts of parameter space, the coordination advantage is much
less than 0.01 (\autoref{fig:model_summary}). For reference, a
log-response of ratio is 0.01 is approximately 1\%. The only exception
was for thin leaves (\(T_\text{leaf} = 100~\mu \text{m}\)) with few
stomata (\(U = 338~\mu \text{m}\), which corresponds to a stomatal
density of \(\approx 10~\text{mm}^{-2}\)), where lateral diffusion is
major constraint on CO\(_2\) supply. However, such thin leaves with so
few stomata are uncommon among C\(_3\) plants (some CAM plants have low
stomatal density \citep{males_stomatal_2017}). In other areas of
parameter space, lateral diffusion limitations were small relative to
those along the ab-adaxial axis (see \autoref{fig:model_example} for a
representative model solution).

\begin{figure}[ht]
\includegraphics[width = \textwidth]{figures/model_summary.pdf}
\caption{There is little photosynthetic benefit of offsetting stomatal position each surface based on a 2-D model of photosynthesis. The coordination advantage (\autoref{eq:coordination_advantage}) is close to zero under nearly all of the parameter space \autoref{tab:model_var}, meaning that the photosynthetic rate of amphistomatous leaves with stomata optimally offset is nearly equal to leaves with stomata on each surface in the same position along the leaf plane. $I_0$: PPFD incident on the leaf surface; $\varphi_\text{pal}$: Fraction of intercellular airspace (aka porosity), palisade; $T_\text{leaf}$: Leaf thickness; $U$: Interstomatal distance.}
\label{fig:model_summary}
\end{figure}

\hypertarget{discussion}{%
\section{Discussion}\label{discussion}}

Stomata cost resources to maintain \citep{deans_optimization_2020} and
expose leaves to risks such as hydraulic failure
\citep{wang_theoretical_2020} or infection by plant pathogens
\citep{melotto_stomatal_2017}. A theoretical, optimized plant would
minimize stomatal density while also allowing competitive gas exchange
rates for its environment so as to maximize C assimilation per unit
investment in stomata. Natural selection operates within developmental
and physical constraints to drive each plant species toward its
theoretical optimum. This study provides evidence that stomata in
\emph{Arabidopsis thaliana}, the model angiosperm, are non-randomly
distributed, favoring dispersion over clustering
(\autoref{fig:single-surface}). However, stomata are not ideally
dispersed in an equilateral triangular grid as would be optimal to
minimize CO\(_2\) diffusion path length and standardize the area
supplied by each stomate (\autoref{fig:tessellation-example}).
Additionally, when grown in high light environments, \emph{A. thaliana}
exhibited increased stomatal density rather than increased stomatal
dispersion (\autoref{fig:density}), which suggests that natural
selection has acted more strongly on developmental pathways that
modulate stomatal density than those that control stomatal dispersion.
In other words, plants optimize gas exchange by adding more stomata
rather than dispersing them more evenly across the leaf surface. This
study also demonstrates that stomata that supply larger leaf areas with
CO\(_2\) tend to be larger (\autoref{fig:length-area}). These results
could suggest that 1) the added energetic and hydraulic cost of
non-ideally dispersed stomata is negligible and therefore not acted on
by natural selection; 2) no developmental pathway exists to ensure the
ideal placement of stomata on the leaf; or 3) the regulation of stomatal
size limits the cost incurred by non-ideal stomatal dispersion.

In high light environments, amphistomy is favorable as high light
photosynthesis is limited by CO\(_2\) and amphistomy halves diffusion
path length and boundary layer resistance, thereby reducing CO\(_2\)
limitation - increasing theoretical \(A_\text{max}\). An optimal
amphistomatous leaf has offset stomata such that stomata are more likely
to appear on one leaf surface if there is not a stomata directly
opposite it on the other surface as shown in
\autoref{fig:ideal-amphi-grid}. However, our results show that leaf
surfaces are not coordinated but are independent, regardless of light
(\autoref{fig:dual-surface}). Additionally, gas exchange models show
little photosynthetic efficiency gain from abaxial-adaxial stomatal
coordination compared to anticoordination (INSERT FIG FROM MODELING). We
posit that this marginal gain is not sufficient to be acted upon
strongly by natural selection. Thus, amphistomatous plants do not
exhibit abaxial-adaxial stomatal coordination for there is little
selective advantage of it.

Our study corroborates previous studies which demonstrate that stomata
are non-randomly distributed along the leaf surface as a result of
developmental mechanisms such as spatially biased arrest of stomatal
initials \citep{boetsch_arrest_1995}, oriented asymmetric cell division
\citep{geisler_oriented_2000}, and cell cycle controls
\citep{croxdale_stomatal_2000}. We do not investigate the potential
developmental pathways that influence stomatal dispersion in this study;
however, they are important to consider as these pathways could limit
plants from reaching the theoretical peak in the adaptive landscape:
uniform stomatal dispersion. Instead, as this study suggests, plants may
simply compensate with higher stomatal density and by fitting stomatal
size to the area that they supply with CO\(_2\). To understand why
stomata are not ideally dispersed, more modelling should be done to
estimate the fitness gain of stomatal dispersion. Additionally, genetic
manipulation studies should attempt to create mutants with clustered and
ideally dispersed stomata for a comparison of their photosynthetic
traits. This could have extremely important implications for maximum
assimilation rates in crops as most crop species are grown in high light
where CO\(_2\) is often limiting. In drought-prone environments,
increased stomatal dispersion may increase water use efficiency by
reducing the number of stomata needed to achieve the same internal
CO\(_2\) concentration, \(C_\text{i}\).

Beyond disperson on a single surface, gas exchange can be optimized via
stomatal coordination of abaxial and adaxial surfaces in amphistomatous
leaves. Given that leaf thicknesses are generally multiple times greater
than interstomatal distance (GIVE DISTANCES HERE). As a result,
abaxial-adaxial stomatal coordination reduces CO\(_2\) diffusion path
length far less than single surface dispersion, so we hypothesize this
strategy to afford less photosynthetic advantage to the leaf. Our
modelling results demonstrate that, even in ideal conditions, i.e.~thick
leaf, low stomatal densities, high light, low leaf porosity, high
Rubisco concentration, etc., the photosynthetic advantage of
coordination is minimal. We are not surprised by these results, but
still highlight them here as we are the first to report this finding.

Amphistomy is a unique and important adaptation found around the world
across many plant lineages \citep{muir_light_2018}, yet much of the
dynamics of amphistomy remain poorly understood. Here, we show that in
\emph{Arabidopsis thaliana} 1) stomata in are non-randomly dispersed,
but not ideally dispersed; 2) stomatal size and density are modulated by
light; 3) stomatal size is positively correlated with the area to which
it supplies CO\(_2\); and 4) abaxial-adaxial stomatal coordination is
not exhibited and is not shown to provide a strong photosynthetic
advantage using CO\(_2\) diffusion models. Interestingly, these findings
did not validate many of our hypotheses which were based on first
principles, suggesting that there may be limits on plants' ability to
control stomatal placement. Future studies which elucidate these
limitations may have important implications for agriculturaly
productivity in a rapidly changing world.

\hypertarget{references}{%
\section{References}\label{references}}

\bibliographystyle{abbrvnat}
\bibliography{stomata-spacing.bib}

%% Author bio-pics with images

\end{document}

### 2023-10-18 r/11_plot-model.R

@@ -3,20 +3,35 @@ source("r/photo_2d_pm.R")
 
 ph2d = read_rds("objects/model_output.rds")
 
-# STUFF for processing
 vars = c("stomatal_arrangement", "n_z", "n_x", "I_0", "A_n", "phi_pal")
 
 ph2d_vars = ph2d |>
   map("parms") |>
   map_dfr(\(.x) as_tibble(.x[vars]))
 
+tikz("ms/figures/model_summary.tex", standAlone = TRUE, width = 5, height = 5)
+options(tikzLatexPackages = c(getOption("tikzLatexPackages")))
+
 ph2d_vars |>
   pivot_wider(names_from = "stomatal_arrangement", values_from = "A_n") |>
   # coordination advantage
-  mutate(ca = log(offset / aligned)) |>
-  ggplot(aes(I_0, ca, color = phi_pal)) +
-  facet_grid(n_z ~ n_x) +
-  geom_point()
+  mutate(
+    ca = log(offset / aligned),
+    U = glue("$U = {U}~\\mu \\mathrm{{m}}$", U = n_x * 2),
+    T_leaf = glue("$T_\\mathrm{{leaf}} = {T_leaf}~\\mu \\mathrm{{m}}$", T_leaf = n_z - 1),
+    U = fct_reorder(U, n_x),
+    T_leaf = fct_reorder(T_leaf, n_z),
+    `$\\varphi_\\mathrm{pal}$` = as.factor(phi_pal)
+  ) |>
+  ggplot(aes(1e6 * I_0, ca, color = `$\\varphi_\\mathrm{pal}$`)) +
+  facet_grid(T_leaf ~ U) +
+  geom_point() +
+  scale_x_continuous(limits = c(0, 1000), breaks = c(0, 500, 1000)) +
+  xlab("$I_0~[\\mu \\mathrm{mol}~\\mathrm{m}^{-2}~\\mathrm{s}^{-1}]$") +
+  ylab("coordination advantage") +
+  theme(legend.position = "top")
 
-ph2d_vars |>
-  filter(n_z == 101, n_x == 169)
+dev.off()
+system("pdflatex -output-directory=ms/figures/ ms/figures/model_summary.tex")
+
+file.remove(paste0("ms/figures/model_summary.", c("aux", "log")))

### 2023-10-18 r/header.R

@@ -52,6 +52,7 @@ library(stringr)
 library(stats)
 library(terra)
 library(tibble)
+library(tikzDevice)
 library(tidyr)
 library(tripack)
 library(units)

### 2023-10-18 ms/ms.Rmd
revised discussion

@@
 ---
-# This template is largely functionally equivalent to that for elsevier
-# The two fields that are not in the elsevier format are:
-#      `corresponding_author` and `acknowledgements`
-title: Spatial Coordination Of Stomatal Patterning Between Leaf Surfaces In Amphistomatous *Arabidopsis thaliana* Incurs No Photosynthetic Advantage
-#date: "`r Sys.Date()`"
+title: Does stomatal patterning in amphistomatous leaves minimize the CO$_2$ diffusion path length with leaves of *Arabidopsis thaliana*?
+date: "`r Sys.Date()`"
 author:
   - name: Jacob L. Watts
-    email: jwatts@colgate.edu
+    email: Jacob.Watts-1@colorado.edu
     corresponding_author: yes # Footnote using corresponding_author
-    affiliation: Colgate University
+    affiliation: 
+      - UHM
+      - CU
   - name: Graham Dow
-    email: graham.dow@usys.ethz.ch
-    affiliation: ETH
+    email: graham.dow@niab.com
+    affiliation: NIAB
   - name: Thomas N. Buckley
     email: tnbuckley@ucdavis.edu
-    affiliation: University of California Davis
-    footnote: 1 # Footnote using reference
+    affiliation: UCD
   - name: Christopher D. Muir
-    email: cdmuir@hawaii.edu
-    affiliation: University of Hawaii at Manoa
-    footnote: 1 # Footnote using reference
+    email: cdmuir@wisc.edu
+    affiliation: 
+      - UHM
+      - UWM
+    # footnote: 1 # Footnote using reference
 address:
-  - code: Colgate University
-    address: 13 Oak Drive, Hamilton, NY 13346
-  - code: University of Hawaii at Manoa
-    address: 2500 Campus Rd, Honolulu, HI 96822
-footnote:
-  - code: 1
-    text: "Equal contribution"
-  - code: 2
-    text: "Current email address: \\href{mailto:cat@example.com}{cat@example.com}"
-abstract: |
-  This is the abstract.
-  It consists of two paragraphs.
-acknowledgements: |
-  This is an acknowledgement.
-  It consists of two paragraphs.
+  - code: UHM
+    address: School of Life Sciences, University of Hawaiʻi at Mānoa, Honolulu, HI 96822
+  - code: CU
+    address: Ecology and Evolutionary Biology, Universtity of Colorado, Boulder, CO 80309
+  - code: NIAB
+    address: National Institute of Agricultural Botany, Cambridge, United Kingdom CB3 0LE
+  - code: UCD
+    address: Department of Plant Sciences, University of California, Davis, CA 95616
+  - code: UWM
+    address: Department of Botany, University of Wisconsin, Madison, WI 53706
+# footnote:
+#   - code: 1
+#     text: "Current email address: \\href{mailto:cdmuir@wisc.edu}{cdmuir@wisc.edu}"
+abstract:
+  - "We will write abstract after Discussion and Conclusions are complete"
 keywords:
@@
 fontsize: 12pt
-#spacing: halfline # could also be oneline
-#classoptions:
-#  - endnotes
+competing_interests: |
+  The authors declare no competing interests.
+author_contributions: |
+  JLW and CDM conceived of the project, analyzed data, and wrote the manuscript. GW provided data. TNB contributed to model development and helped edit the manuscript.
+acknowledgements: |
+  This work was supported by NSF Award 1929167 to CDM and NSF Award 2307341 to TNB.
 bibliography: stomata-spacing.bib
-#csl: mycitationstyle.csl
 link-citations: yes
@@
       - booktabs # to use with knitr::kable() example below
-#urlcolor: orange
-#linkcolor: green
-#citecolor: red
 header-includes:
@@
  - \usepackage{caption} # for \captionsetup{}
+ - \usepackage{newunicodechar,graphicx}
+ - \DeclareRobustCommand{\okina}{\raisebox{\dimexpr\fontcharht\font`A-\height}{\scalebox{0.8}{`}}}
+ - \newunicodechar{ʻ}{\okina}
 
@@
 ```{r setup, include=FALSE}
-knitr::opts_chunk$set(echo = FALSE, message = FALSE) 
-knitr::opts_chunk$set(fig.pos = 'p') # Places figures on pages separate from text
-knitr::opts_chunk$set(out.width = '100%', dpi = 300) # Figure resolution and size
-knitr::opts_chunk$set(fig.env = "figure") # Latex figure environment
+knitr::opts_chunk$set(
+  echo = FALSE, 
+  message = FALSE,
+  fig.pos = 'th',
+  out.width = '100%', dpi = 300,
+  fig.env = "figure"
+)
 knitr::opts_knit$set(root.dir = "..")
@@
 Size positively co-varies with genome size [@roddy_scaling_2020] and negatively co-varies with stomatal density [@camargo_density_2011].
-Total stomatal area (size $\times$ density) is optimized for operational conductance ($gs_\text{s,op}$) rather than maximum conductance ($gs_\text{s,max}$) such that stomatal apertures are most responsive to changes in the environment at their operational aperture [@franks_physiological_2012; @liu_scaling_2021].
-Stomatal aperture can compensate for maladaptive stomatal densities to an extent [@bussis_stomatal_2006], but stomatal density and size ultimately determine a leaf's theoretical $gs_\text{s,max}$ [@sack_developmental_2016], which is proportional to $gs_\text{s,op}$ [@murray_consistent_2020].
+Total stomatal area (size $\times$ density) is optimized for operational conductance ($g_\text{s,op}$) rather than maximum conductance ($g_\text{s,max}$) such that stomatal apertures are most responsive to changes in the environment at their operational aperture [@franks_physiological_2012; @liu_scaling_2021].
+Stomatal aperture can compensate for maladaptive stomatal densities to an extent [@bussis_stomatal_2006], but stomatal density and size ultimately determine a leaf's theoretical $g_\text{s,max}$ [@sack_developmental_2016], which is proportional to $g_\text{s,op}$ [@murray_consistent_2020].
 Additionally, low stomatal densities lead to irregular and insufficient CO$_2$ supply and reduced photosynthetic efficiency in areas far from stomata [@pieruschka_lateral_2006; @morison_lateral_2005], while high stomatal densities can reduce water use efficiency (WUE) [@bussis_stomatal_2006] and incur excessive metabolic costs [@deans_optimization_2020].
@@
 We hypothesized that natural selection will favor stomatal patterning and distribution to minimize the diffusion path length.
-In amphistomatous leaves, this would be accomplished by 1) a uniform distribution of stomata on both abaxial and adaxial leaf surfaces and 2) coordinated stomatal spacing on each surface that offsets the position of stomata (\autoref{fig:ideal-amphi-grid}).
+In amphistomatous leaves, this would be accomplished by 1) a dispersed, equilateral triangular distribution of stomata on both abaxial and adaxial leaf surfaces and 2) coordinated stomatal spacing on each surface that offsets the position of stomata (\autoref{fig:ideal-amphi-grid}).
 Coordination between leaf surfaces is defined in this study as the occurrence of stomata in areas farthest from stomata on the opposite leaf surface.
@@
 \begin{figure}[ht]
-\includegraphics[width = \textwidth]{figures/ideal-amphi-grid.pdf}
+\includegraphics[width = 4in]{figures/ideal-amphi-grid.pdf}
 \caption{Idealized amphistomatous stomatal grid with uniform stomatal patterning and perfect abaxial-adaxial coordination.}
@@
   "leaf thickness", "+",
-  "stomatal density", "-",
+  "interstomatal distance", "+",
   "leaf porosity", "-", 
-  "lat.-vert. diffusion ratio", "-", 
-  "temperature", "-",
-  "pressure", "-",
-  "Rubisco concentration", "+",
+  # "lat.-vert. diffusion ratio", "-", 
+  # "temperature", "-",
+  # "pressure", "-",
+  # "Rubisco concentration", "+",
   "light", "+"
@@
 
-[CDM: Graham Dow provided these images. We'll need to add him as a co-author and ask him to write methods on image acquisition.]
+[Graham - can you fill in details on growth conditions, image aquisitiion, and image dimensions?]
 
@@
 For these grids, we integrated over plausible stomatal densities and then conditioned on stomatal grids with exactly $n$ stomata.
-The simulated stomatal count was drawn from a Poisson distribution with the mean parameter $\lambda$ drawn from a Gamma distribution with shape $n$ and scale 1 $\lambda \sim \Gamma(n, 1)$.
+The simulated stomatal count was drawn from a Poisson distribution with the mean parameter $\lambda$ drawn from a Gamma distribution with shape $n$ and scale 1 ($\lambda \sim \Gamma(n, 1)$).
 $\Gamma(n, 1)$ is the posterior distribution of $\lambda$ with a flat prior distribution.
@@
 \begin{figure}[ht]
-\includegraphics[width = \textwidth]{figures/tessellation-example.pdf}
+\includegraphics[height = \textheight]{figures/tessellation-example.pdf}
 \caption{Examples of synthetic and real leaf surfaces.
@@
 \begin{figure}[ht]
-\includegraphics[width = \textwidth]{figures/density.pdf}
+\includegraphics[width = 4in]{figures/density.pdf}
 \caption{Stomatal density is higher in \textit{A.
@@
 \begin{figure}[ht]
-\includegraphics[width=\textwidth]{figures/single-surface.pdf}
+\includegraphics[width=4in]{figures/single-surface.pdf}
 \caption{Stomata are more dispersed than expected under the null model of uniform random position (dispersion index = 0) but far from a distribution that maximizes distance between stomata (dispersion index = 1).
@@
 \begin{figure}[ht]
-\includegraphics[width = \textwidth]{figures/dual-surface.pdf}
+\includegraphics[width = 4in]{figures/dual-surface.pdf}
 \caption{Pixel-wise correlation between near stomatal distance (NSD) squared on paired abaxial and adaxial leaf surfaces.
@@
 \begin{figure}[ht]
-\includegraphics[width = \textwidth]{figures/length-area.pdf}
+\includegraphics[width = 4in]{figures/length-area.pdf}
 \caption{Stomatal length and stomatal zone area are positively correlated. Linear regression lines and 95\% confidence ribbons are from a Bayesian linear mixed-effects model.}
@@
 \begin{figure}[ht]
-\includegraphics[width = \textwidth]{figures/model_summary.pdf}
+\includegraphics[width = 5in]{figures/model_summary.pdf}
 \caption{There is little photosynthetic benefit of offsetting stomatal position each surface based on a 2-D model of photosynthesis.
@@
 Stomata cost resources to maintain [@deans_optimization_2020] and expose leaves to risks such as hydraulic failure [@wang_theoretical_2020] or infection by plant pathogens [@melotto_stomatal_2017].
-A theoretical, optimized plant would minimize stomatal density while also allowing competitive gas exchange rates for its environment so as to maximize C assimilation per unit investment in stomata.
-Natural selection operates within developmental and physical constraints to drive each plant species toward its theoretical optimum.
-This study provides evidence that stomata in *Arabidopsis thaliana*, the model angiosperm, are non-randomly distributed, favoring dispersion over clustering (\autoref{fig:single-surface}).
-However, stomata are not ideally dispersed in an equilateral triangular grid as would be optimal to minimize CO$_2$ diffusion path length and standardize the area supplied by each stomate (\autoref{fig:tessellation-example}).
-Additionally, when grown in high light environments, *A.
-thaliana* exhibited increased stomatal density rather than increased stomatal dispersion (\autoref{fig:density}), which suggests that natural selection has acted more strongly on developmental pathways that modulate stomatal density than those that control stomatal dispersion.
-In other words, plants optimize gas exchange by adding more stomata rather than dispersing them more evenly across the leaf surface.
-This study also demonstrates that stomata that supply larger leaf areas with CO$_2$ tend to be larger (\autoref{fig:length-area}).
-These results could suggest that 1) the added energetic and hydraulic cost of non-ideally dispersed stomata is negligible and therefore not acted on by natural selection; 2) no developmental pathway exists to ensure the ideal placement of stomata on the leaf; or 3) the regulation of stomatal size limits the cost incurred by non-ideal stomatal dispersion.
+Therefore leaves should develop enough stomata to adequately supply CO$_2$ to chloroplasts, but not overinvest.
+A widespread hypothesis in plant ecophysiology is that natural selection optimizes traits like stomatal size, density, and distribution to maximize carbon gain relative to any costs in a given environmental context.
+In principle, spacing stomata to minimize the average distance between stomatal pores and chloroplasts within the mesophyll should increase carbon gain, all else being equal.
+However, reducing this distance to its absolute minimum may be constrained by developmental processes, or the photosynthetic benefit may be too small to be 'seen' by natural selection (i.e.
+the selection coefficient is less than drift barrier [@sung_drift-barrier_2012]).
 
-In high light environments, amphistomy is favorable as high light photosynthesis is limited by CO$_2$ and amphistomy halves diffusion path length and boundary layer resistance, thereby reducing CO$_2$ limitation - increasing theoretical $A_\text{max}$.
-An optimal amphistomatous leaf has offset stomata such that stomata are more likely to appear on one leaf surface if there is not a stomata directly opposite it on the other surface as shown in \autoref{fig:ideal-amphi-grid}.
-However, our results show that leaf surfaces are not coordinated but are independent, regardless of light (\autoref{fig:dual-surface}).
-Additionally, gas exchange models show little photosynthetic efficiency gain from abaxial-adaxial stomatal coordination compared to anticoordination (INSERT FIG FROM MODELING).
-We posit that this marginal gain is not sufficient to be acted upon strongly by natural selection.
-Thus, amphistomatous plants do not exhibit abaxial-adaxial stomatal coordination for there is little selective advantage of it.
+We tested five related hypotheses about stomatal spacing in amphistomatous leaves using the model angiosperm *Arabidopsis thaliana* grown under different light intensities.
+First, we predicted that stomata on each surface are overdispersed relative to a random uniform distribution, which should increase CO$_2$ supply.
+Stomata on each surface are overdispersed (\autoref{fig:single-surface}), but are not ideally dispersed in an equilateral triangular grid as would be optimal to minimize CO$_2$ diffusion path length and equalize the area supplied by each stomate (\autoref{fig:tessellation-example}).
+Second, we predicted that an optimal amphistomatous leaf has offset stomata such that stomata are more likely to appear on one leaf surface if there is not a stomata directly opposite it on the other surface as shown in \autoref{fig:ideal-amphi-grid}.
+However, there is no evidence for coordination and the positions on each surface appear independent, regardless of light treatment (\autoref{fig:dual-surface}).
+Third, we predicted that plants respond plastically to higher light intensity by increasing stomatal density.
+*Arabidopsis* plants grown under high light had higher stomatal density than the same genotype growns under low and medium light intensity (\autoref{fig:density}).
+However, we found no support for our fourth prediction that stomata would be more evenly dispersed at high light intensity (\autoref{fig:single-surface}).
+Finally, we predicted that within leaf variation in stomatal size would correlate with stomatal spacing, as larger stomata can supply larger volumes of adjacent mesophyll.
+In all three light treatments, stomatal size positively covaried with the stomatal zone, i.e.
+adjacent region of mesophyll that would be supplied by that stomate (\autoref{fig:length-area}).
 
-Our study corroborates previous studies which demonstrate that stomata are non-randomly distributed along the leaf surface as a result of developmental mechanisms such as spatially biased arrest of stomatal initials [@boetsch_arrest_1995], oriented asymmetric cell division [@geisler_oriented_2000], and cell cycle controls [@croxdale_stomatal_2000].
+Stomatal spacing on *A.
+thaliana* leaves partially supports our overall hypothesis that natural selection minimizes the average distance between stomata and chloroplasts, for a given overall stomatal density.
+There are three nonmutually exclusive hypotheses for why several of our predictions were wrong.
+First, our predictions are wrong becuase they are based on overly simplistic assumptions about epidermal and mesophyll anatomy.
+Second, natural selection may be constrained by developmental processes that prevent phenotypes from reaching their adaptive optima.
+Third, the benefit of some traits may be of too little consequence to result in fitness differences large enough to be respond to selection.
+We consider the plausibility of these alternative hypotheses below and present ideas for future work to test them.
+
+[this paragraph should discuss why our assumptions may be too simplistic]
+We assume an idealized leaf epidermal and mesophyll structure that is homogenous and unconstrained by other tradeoffs.
+Real leaves not only provide pathways for CO$_2$ diffusion, but must supply water, intercept light, and deter herbivores and pathogens.
+These competing interests results in nonuniform epidermal and mesophyll structure that could alter our predictions about optimal stomatal spacing.
+[explain some examples]
+
+[this paragraph should discuss developmental constraint] No developmental pathway exists to ensure the ideal placement of stomata on the leaf. [add more explanation]
+
+[this paragraph should discuss limits on natural selection] The gas exchange model demonstrate little photosynthetic gain from abaxial-adaxial stomatal coordination (\autoref{fig:model_summary}).
+Even though lateral diffusion may limit photosynthesis [@morison_lateral_2005], the marginal gain from optimally offsetting stomata is not sufficient to generate fitness differences relative to the strength of genetic drift (i.e.
+the drift-barrier).
+We can similarly extrapolate that an ideal, equilateral triangular stomatal spacing is only slightly better than a suboptimal overdispered pattern.
+Explaining these observations as the result of weak selection is in tension with the finding that stomatal size and zone positively covary, which would suggest that small changes in lateral diffusion distance are significant.
+As described above, the positive correlation between stomatal size and zone may be explained by common developmental processes rather than as an adaptation to maximize CO$_2$ diffusion.
+
+[discuss significance of our study and future directions] Our study corroborates previous studies which demonstrate that stomata are non-randomly distributed along the leaf surface as a result of developmental mechanisms such as spatially biased arrest of stomatal initials [@boetsch_arrest_1995], oriented asymmetric cell division [@geisler_oriented_2000], and cell cycle controls [@croxdale_stomatal_2000].
 We do not investigate the potential developmental pathways that influence stomatal dispersion in this study; however, they are important to consider as these pathways could limit plants from reaching the theoretical peak in the adaptive landscape: uniform stomatal dispersion.
@@
 
-Beyond disperson on a single surface, gas exchange can be optimized via stomatal coordination of abaxial and adaxial surfaces in amphistomatous leaves.
-Given that leaf thicknesses are generally multiple times greater than interstomatal distance (GIVE DISTANCES HERE).
-As a result, abaxial-adaxial stomatal coordination reduces CO$_2$ diffusion path length far less than single surface dispersion, so we hypothesize this strategy to afford less photosynthetic advantage to the leaf.
-Our modelling results demonstrate that, even in ideal conditions, i.e.
-thick leaf, low stomatal densities, high light, low leaf porosity, high Rubisco concentration, etc., the photosynthetic advantage of coordination is minimal.
-We are not surprised by these results, but still highlight them here as we are the first to report this finding.
-
-Amphistomy is a unique and important adaptation found around the world across many plant lineages [@muir_light_2018], yet much of the dynamics of amphistomy remain poorly understood.
-Here, we show that in *Arabidopsis thaliana* 1) stomata in are non-randomly dispersed, but not ideally dispersed; 2) stomatal size and density are modulated by light; 3) stomatal size is positively correlated with the area to which it supplies CO$_2$; and 4) abaxial-adaxial stomatal coordination is not exhibited and is not shown to provide a strong photosynthetic advantage using CO$_2$ diffusion models.
-Interestingly, these findings did not validate many of our hypotheses which were based on first principles, suggesting that there may be limits on plants' ability to control stomatal placement.
-Future studies which elucidate these limitations may have important implications for agriculturaly productivity in a rapidly changing world.
+[Conclusion] Our results suggest that after optimiziing stomatal density and having developmental rules for spacing stomata relatively evenly, there may be limited gains to further optimization.
+Therefore, developmental constraints may be necessary to make sense of some features of stomatal spacing and distribution.
 

### 2023-10-18 ms/ms.tex
revised discussion

@@
 \hypersetup{
-  pdftitle={Spatial Coordination Of Stomatal Patterning Between Leaf Surfaces In Amphistomatous Arabidopsis thaliana Incurs No Photosynthetic Advantage},
+  pdftitle={Does stomatal patterning in amphistomatous leaves minimize the CO\_2 diffusion path length with leaves of Arabidopsis thaliana?},
   pdfkeywords={amphistomy, Arabidopsis thaliana, CO\(_2\)
@@
 % Create labels for Addresses if the are given in Elsevier format
-   \MyNewLabel{ADDRTXT}{Colgate University}{%
-  13 Oak Drive, Hamilton, NY 13346%
+   \MyNewLabel{ADDRTXT}{UHM}{%
+  School of Life Sciences, University of Hawaiʻi at Mānoa, Honolulu, HI
+96822%
  }
-   \MyNewLabel{ADDRTXT}{University of Hawaii at Manoa}{%
-  2500 Campus Rd, Honolulu, HI 96822%
+   \MyNewLabel{ADDRTXT}{CU}{%
+  Ecology and Evolutionary Biology, Universtity of Colorado, Boulder, CO
+80309%
+ }
+   \MyNewLabel{ADDRTXT}{NIAB}{%
+  National Institute of Agricultural Botany, Cambridge, United Kingdom
+CB3 0LE%
+ }
+   \MyNewLabel{ADDRTXT}{UCD}{%
+  Department of Plant Sciences, University of California, Davis, CA
+95616%
+ }
+   \MyNewLabel{ADDRTXT}{UWM}{%
+  Department of Botany, University of Wisconsin, Madison, WI 53706%
  }
@@
 % Create labels for Footnotes if they are given in Elsevier format
-\MyNewLabel{CORRESPTXT}{1}{Equal contribution}
-\MyNewLabel{CORRESPTXT}{2}{Current email address:
-\href{mailto:cat@example.com}{cat@example.com}}
 
@@
 \usepackage{caption}
+\usepackage{newunicodechar,graphicx}
+\DeclareRobustCommand{\okina}{\raisebox{\dimexpr\fontcharht\font`A-\height}{\scalebox{0.8}{`}}}
+\newunicodechar{ʻ}{\okina}
 % Pandoc header-include feature
@@
 
-\title[]{Spatial Coordination Of Stomatal Patterning Between Leaf
-Surfaces In Amphistomatous \emph{Arabidopsis thaliana} Incurs No
-Photosynthetic Advantage}
+\title[]{Does stomatal patterning in amphistomatous leaves minimize the
+CO\(_2\) diffusion path length with leaves of \emph{Arabidopsis
+thaliana}?}
 
@@
 \author[%
-\addrlblbycode{Colgate University}%
+\addrlblbycode{UHM},\addrlblbycode{CU}%
 ,\refstepcounter{correspcnt}\setcounter{thisauthcorresp}{\value{correspcnt}}\fnsymbol{thisauthcorresp}%
@@
 
-\addrbycode{Colgate University}
+\addrbycode{UHM}
+\addrbycode{CU}
 
-\corresp[\fnsymbol{thisauthcorresp}]{Corresponding author. \href{mailto:jwatts@colgate.edu}{\nolinkurl{jwatts@colgate.edu}}}
+\corresp[\fnsymbol{thisauthcorresp}]{Corresponding author. \href{mailto:Jacob.Watts-1@colorado.edu}{\nolinkurl{Jacob.Watts-1@colorado.edu}}}
 
 \author[%
-\addrlblbycode{ETH}%
+\addrlblbycode{NIAB}%
 %
@@
 
-\addrbycode{ETH}
+\addrbycode{NIAB}
 
 \author[%
-\addrlblbycode{University of California Davis}%
+\addrlblbycode{UCD}%
 %
 %
-,\corresplblbycode{1}%
+%
 ]{Thomas N. Buckley}
 
-\addrbycode{University of California Davis}
-
-\correspbycode{1}
+\addrbycode{UCD}
 
 \author[%
-\addrlblbycode{University of Hawaii at Manoa}%
+\addrlblbycode{UHM},\addrlblbycode{UWM}%
 %
 %
-,\corresplblbycode{1}%
+%
 ]{Christopher D. Muir}
 
-\addrbycode{University of Hawaii at Manoa}
-
-\correspbycode{1}
+\addrbycode{UHM}
+\addrbycode{UWM}
 
@@
 \abstract{
-This is the abstract. It consists of two paragraphs.}
+We will write abstract after Discussion and Conclusions are complete}
 
@@
 \(\times\) density) is optimized for operational conductance
-(\(gs_\text{s,op}\)) rather than maximum conductance
-(\(gs_\text{s,max}\)) such that stomatal apertures are most responsive
-to changes in the environment at their operational aperture
+(\(g_\text{s,op}\)) rather than maximum conductance (\(g_\text{s,max}\))
+such that stomatal apertures are most responsive to changes in the
+environment at their operational aperture
 \citep{franks_physiological_2012, liu_scaling_2021}. Stomatal aperture
@@
 \citep{bussis_stomatal_2006}, but stomatal density and size ultimately
-determine a leaf's theoretical \(gs_\text{s,max}\)
+determine a leaf's theoretical \(g_\text{s,max}\)
 \citep{sack_developmental_2016}, which is proportional to
-\(gs_\text{s,op}\) \citep{murray_consistent_2020}. Additionally, low
+\(g_\text{s,op}\) \citep{murray_consistent_2020}. Additionally, low
 stomatal densities lead to irregular and insufficient CO\(_2\) supply
@@
 and distribution to minimize the diffusion path length. In
-amphistomatous leaves, this would be accomplished by 1) a uniform
-distribution of stomata on both abaxial and adaxial leaf surfaces and 2)
-coordinated stomatal spacing on each surface that offsets the position
-of stomata (\autoref{fig:ideal-amphi-grid}). Coordination between leaf
-surfaces is defined in this study as the occurrence of stomata in areas
-farthest from stomata on the opposite leaf surface. Additionally,
-because CO\(_2\) is more limiting for photosynthesis under high light,
-we hypothesize that in high light 3) there should be more stomata, and
-4) stomata should be more uniformly distributed than in low light.
-Finally, as stomatal densities are selected for optimal operational
-aperture, we hypothesize that 5) stomatal length will be positively
-correlated with the area of the leaf surface to which it is closest. We
-refer to this as the `stomatal zone', the leaf area surrounding a focal
-stomate closest to that stomate and therefore the zone it supplies with
-CO\(_2\)). This way, each stomate can be optimally sized relative to the
-mesophyll volume it supplies.
+amphistomatous leaves, this would be accomplished by 1) a dispersed,
+equilateral triangular distribution of stomata on both abaxial and
+adaxial leaf surfaces and 2) coordinated stomatal spacing on each
+surface that offsets the position of stomata
+(\autoref{fig:ideal-amphi-grid}). Coordination between leaf surfaces is
+defined in this study as the occurrence of stomata in areas farthest
+from stomata on the opposite leaf surface. Additionally, because
+CO\(_2\) is more limiting for photosynthesis under high light, we
+hypothesize that in high light 3) there should be more stomata, and 4)
+stomata should be more uniformly distributed than in low light. Finally,
+as stomatal densities are selected for optimal operational aperture, we
+hypothesize that 5) stomatal length will be positively correlated with
+the area of the leaf surface to which it is closest. We refer to this as
+the `stomatal zone', the leaf area surrounding a focal stomate closest
+to that stomate and therefore the zone it supplies with CO\(_2\)). This
+way, each stomate can be optimally sized relative to the mesophyll
+volume it supplies.
 
 \begin{figure}[ht]
-\includegraphics[width = \textwidth]{figures/ideal-amphi-grid.pdf}
+\includegraphics[width = 4in]{figures/ideal-amphi-grid.pdf}
 \caption{Idealized amphistomatous stomatal grid with uniform stomatal patterning and perfect abaxial-adaxial coordination.}
@@
 leaf thickness & + \\ 
-  stomatal density & - \\ 
+  interstomatal distance & + \\ 
   leaf porosity & - \\ 
-  lat.-vert. diffusion ratio & - \\ 
-  temperature & - \\ 
-  pressure & - \\ 
-  Rubisco concentration & + \\ 
   light & + \\ 
@@
 
-{[}CDM: Graham Dow provided these images. We'll need to add him as a
-co-author and ask him to write methods on image acquisition.{]}
+{[}Graham - can you fill in details on growth conditions, image
+aquisitiion, and image dimensions?{]}
 
@@
 parameter \(\lambda\) drawn from a Gamma distribution with shape \(n\)
-and scale 1 \(\lambda \sim \Gamma(n, 1)\). \(\Gamma(n, 1)\) is the
+and scale 1 (\(\lambda \sim \Gamma(n, 1)\)). \(\Gamma(n, 1)\) is the
 posterior distribution of \(\lambda\) with a flat prior distribution.
@@
 \citep{wang_theoretical_2020} or infection by plant pathogens
-\citep{melotto_stomatal_2017}. A theoretical, optimized plant would
-minimize stomatal density while also allowing competitive gas exchange
-rates for its environment so as to maximize C assimilation per unit
-investment in stomata. Natural selection operates within developmental
-and physical constraints to drive each plant species toward its
-theoretical optimum. This study provides evidence that stomata in
-\emph{Arabidopsis thaliana}, the model angiosperm, are non-randomly
-distributed, favoring dispersion over clustering
-(\autoref{fig:single-surface}). However, stomata are not ideally
-dispersed in an equilateral triangular grid as would be optimal to
-minimize CO\(_2\) diffusion path length and standardize the area
-supplied by each stomate (\autoref{fig:tessellation-example}).
-Additionally, when grown in high light environments, \emph{A. thaliana}
-exhibited increased stomatal density rather than increased stomatal
-dispersion (\autoref{fig:density}), which suggests that natural
-selection has acted more strongly on developmental pathways that
-modulate stomatal density than those that control stomatal dispersion.
-In other words, plants optimize gas exchange by adding more stomata
-rather than dispersing them more evenly across the leaf surface. This
-study also demonstrates that stomata that supply larger leaf areas with
-CO\(_2\) tend to be larger (\autoref{fig:length-area}). These results
-could suggest that 1) the added energetic and hydraulic cost of
-non-ideally dispersed stomata is negligible and therefore not acted on
-by natural selection; 2) no developmental pathway exists to ensure the
-ideal placement of stomata on the leaf; or 3) the regulation of stomatal
-size limits the cost incurred by non-ideal stomatal dispersion.
+\citep{melotto_stomatal_2017}. Therefore leaves should develop enough
+stomata to adequately supply CO\(_2\) to chloroplasts, but not
+overinvest. A widespread hypothesis in plant ecophysiology is that
+natural selection optimizes traits like stomatal size, density, and
+distribution to maximize carbon gain relative to any costs in a given
+environmental context. In principle, spacing stomata to minimize the
+average distance between stomatal pores and chloroplasts within the
+mesophyll should increase carbon gain, all else being equal. However,
+reducing this distance to its absolute minimum may be constrained by
+developmental processes, or the photosynthetic benefit may be too small
+to be `seen' by natural selection (i.e.~the selection coefficient is
+less than drift barrier \citep{sung_drift-barrier_2012}).
 
-In high light environments, amphistomy is favorable as high light
-photosynthesis is limited by CO\(_2\) and amphistomy halves diffusion
-path length and boundary layer resistance, thereby reducing CO\(_2\)
-limitation - increasing theoretical \(A_\text{max}\). An optimal
-amphistomatous leaf has offset stomata such that stomata are more likely
-to appear on one leaf surface if there is not a stomata directly
-opposite it on the other surface as shown in
-\autoref{fig:ideal-amphi-grid}. However, our results show that leaf
-surfaces are not coordinated but are independent, regardless of light
-(\autoref{fig:dual-surface}). Additionally, gas exchange models show
-little photosynthetic efficiency gain from abaxial-adaxial stomatal
-coordination compared to anticoordination (INSERT FIG FROM MODELING). We
-posit that this marginal gain is not sufficient to be acted upon
-strongly by natural selection. Thus, amphistomatous plants do not
-exhibit abaxial-adaxial stomatal coordination for there is little
-selective advantage of it.
+We tested five related hypotheses about stomatal spacing in
+amphistomatous leaves using the model angiosperm \emph{Arabidopsis
+thaliana} grown under different light intensities. First, we predicted
+that stomata on each surface are overdispersed relative to a random
+uniform distribution, which should increase CO\(_2\) supply. Stomata on
+each surface are overdispersed (\autoref{fig:single-surface}), but are
+not ideally dispersed in an equilateral triangular grid as would be
+optimal to minimize CO\(_2\) diffusion path length and equalize the area
+supplied by each stomate (\autoref{fig:tessellation-example}). Second,
+we predicted that an optimal amphistomatous leaf has offset stomata such
+that stomata are more likely to appear on one leaf surface if there is
+not a stomata directly opposite it on the other surface as shown in
+\autoref{fig:ideal-amphi-grid}. However, there is no evidence for
+coordination and the positions on each surface appear independent,
+regardless of light treatment (\autoref{fig:dual-surface}). Third, we
+predicted that plants respond plastically to higher light intensity by
+increasing stomatal density. \emph{Arabidopsis} plants grown under high
+light had higher stomatal density than the same genotype growns under
+low and medium light intensity (\autoref{fig:density}). However, we
+found no support for our fourth prediction that stomata would be more
+evenly dispersed at high light intensity (\autoref{fig:single-surface}).
+Finally, we predicted that within leaf variation in stomatal size would
+correlate with stomatal spacing, as larger stomata can supply larger
+volumes of adjacent mesophyll. In all three light treatments, stomatal
+size positively covaried with the stomatal zone, i.e.~adjacent region of
+mesophyll that would be supplied by that stomate
+(\autoref{fig:length-area}).
 
-Our study corroborates previous studies which demonstrate that stomata
-are non-randomly distributed along the leaf surface as a result of
+Stomatal spacing on \emph{A. thaliana} leaves partially supports our
+overall hypothesis that natural selection minimizes the average distance
+between stomata and chloroplasts, for a given overall stomatal density.
+There are three nonmutually exclusive hypotheses for why several of our
+predictions were wrong. First, our predictions are wrong becuase they
+are based on overly simplistic assumptions about epidermal and mesophyll
+anatomy. Second, natural selection may be constrained by developmental
+processes that prevent phenotypes from reaching their adaptive optima.
+Third, the benefit of some traits may be of too little consequence to
+result in fitness differences large enough to be respond to selection.
+We consider the plausibility of these alternative hypotheses below and
+present ideas for future work to test them.
+
+{[}this paragraph should discuss why our assumptions may be too
+simplistic{]} We assume an idealized leaf epidermal and mesophyll
+structure that is homogenous and unconstrained by other tradeoffs. Real
+leaves not only provide pathways for CO\(_2\) diffusion, but must supply
+water, intercept light, and deter herbivores and pathogens. These
+competing interests results in nonuniform epidermal and mesophyll
+structure that could alter our predictions about optimal stomatal
+spacing. {[}explain some examples{]}
+
+{[}this paragraph should discuss developmental constraint{]} No
+developmental pathway exists to ensure the ideal placement of stomata on
+the leaf. {[}add more explanation{]}
+
+{[}this paragraph should discuss limits on natural selection{]} The gas
+exchange model demonstrate little photosynthetic gain from
+abaxial-adaxial stomatal coordination (\autoref{fig:model_summary}).
+Even though lateral diffusion may limit photosynthesis
+\citep{morison_lateral_2005}, the marginal gain from optimally
+offsetting stomata is not sufficient to generate fitness differences
+relative to the strength of genetic drift (i.e.~the drift-barrier). We
+can similarly extrapolate that an ideal, equilateral triangular stomatal
+spacing is only slightly better than a suboptimal overdispered pattern.
+Explaining these observations as the result of weak selection is in
+tension with the finding that stomatal size and zone positively covary,
+which would suggest that small changes in lateral diffusion distance are
+significant. As described above, the positive correlation between
+stomatal size and zone may be explained by common developmental
+processes rather than as an adaptation to maximize CO\(_2\) diffusion.
+
+{[}discuss significance of our study and future directions{]} Our study
+corroborates previous studies which demonstrate that stomata are
+non-randomly distributed along the leaf surface as a result of
 developmental mechanisms such as spatially biased arrest of stomatal
@@
 
-Beyond disperson on a single surface, gas exchange can be optimized via
-stomatal coordination of abaxial and adaxial surfaces in amphistomatous
-leaves. Given that leaf thicknesses are generally multiple times greater
-than interstomatal distance (GIVE DISTANCES HERE). As a result,
-abaxial-adaxial stomatal coordination reduces CO\(_2\) diffusion path
-length far less than single surface dispersion, so we hypothesize this
-strategy to afford less photosynthetic advantage to the leaf. Our
-modelling results demonstrate that, even in ideal conditions, i.e.~thick
-leaf, low stomatal densities, high light, low leaf porosity, high
-Rubisco concentration, etc., the photosynthetic advantage of
-coordination is minimal. We are not surprised by these results, but
-still highlight them here as we are the first to report this finding.
-
-Amphistomy is a unique and important adaptation found around the world
-across many plant lineages \citep{muir_light_2018}, yet much of the
-dynamics of amphistomy remain poorly understood. Here, we show that in
-\emph{Arabidopsis thaliana} 1) stomata in are non-randomly dispersed,
-but not ideally dispersed; 2) stomatal size and density are modulated by
-light; 3) stomatal size is positively correlated with the area to which
-it supplies CO\(_2\); and 4) abaxial-adaxial stomatal coordination is
-not exhibited and is not shown to provide a strong photosynthetic
-advantage using CO\(_2\) diffusion models. Interestingly, these findings
-did not validate many of our hypotheses which were based on first
-principles, suggesting that there may be limits on plants' ability to
-control stomatal placement. Future studies which elucidate these
-limitations may have important implications for agriculturaly
-productivity in a rapidly changing world.
+{[}Conclusion{]} Our results suggest that after optimiziing stomatal
+density and having developmental rules for spacing stomata relatively
+evenly, there may be limited gains to further optimization. Therefore,
+developmental constraints may be necessary to make sense of some
+features of stomatal spacing and distribution.
 
@@
 \section{References}\label{references}}
+
+\section{Competing interests}
+
+The authors declare no competing interests.
+
+\section{Author contributions statement}
+
+JLW and CDM conceived of the project, analyzed data, and wrote the
+manuscript. GW provided data. TNB contributed to model development and
+helped edit the manuscript.
 

### 2023-10-18 r/13_prepare-submission.R

@@ -1,0 +1,30 @@
+# Move lines around so okina formats properly and then recompile
+ms = readr::read_lines("ms/ms.tex")
+x = which(ms %in% c(
+  "\\usepackage{newunicodechar,graphicx}",
+  "\\DeclareRobustCommand{\\okina}{\\raisebox{\\dimexpr\\fontcharht\\font`A-\\height}{\\scalebox{0.8}{`}}}",
+  "\\newunicodechar{ʻ}{\\okina}"
+))
+
+readr::write_lines(c(ms[1:6], ms[x], ms[7:(min(x) - 1)], ms[(max(x) + 1): length(ms)]), "ms/ms.tex")
+system("cd ms; xelatex ms.tex")
+
+# Copy and rename files for submission
+file.copy("ms/ms.pdf", "aobp1/ms.pdf", overwrite = TRUE)
+file.copy("ms/si.pdf", "aobp1/si.pdf", overwrite = TRUE)
+# file.copy("figures/study-system.pdf", "aobp1/fig1.pdf", overwrite = TRUE)
+# file.copy("figures/habitat-aa.pdf", "aobp1/fig2.pdf", overwrite = TRUE)
+# file.copy("figures/traits-aa.pdf", "aobp1/fig3.pdf", overwrite = TRUE)
+# file.copy("figures/licor.pdf", "aobp1/figS1.pdf", overwrite = TRUE)
+# file.copy("figures/pp-licor.pdf", "aobp1/figS2.pdf", overwrite = TRUE)
+# file.copy("figures/habitat-Ags.pdf", "aobp1/figS3.pdf", overwrite = TRUE)
+# file.copy("figures/habitat-gmaxratio.pdf", "aobp1/figS4.pdf", overwrite = TRUE)
+
+# zip("aobp1/dryad.zip", list.files("dryad", full.names = TRUE))
+
+system("cd aobp1; git clone git@github.com:cdmuir/stomata-spacing.git")
+
+zip("aobp1/github.zip", list.files("aobp1/stomata-spacing", full.names = TRUE,
+                                  recursive = TRUE))
+
+system("rm -r aobp1/stomata-spacing")

### 2023-10-19 ms/ms.Rmd
fixed scale following Graham's email on 2023-10-19

@@
       - CU
-  - name: Graham Dow
+  - name: Graham J. Dow
     email: graham.dow@niab.com
@@
   - code: NIAB
-    address: National Institute of Agricultural Botany, Cambridge, United Kingdom CB3 0LE
+    address: Department of Crop Science and Production Systems, NIAB, Cambridge, CB3 0LE, UK
   - code: UCD
@@
 
+# Mean stomatal density ratio
+mean_sr = single_surface_results |>
+  mutate(rep = str_extract(image_number, "[0-9]+")) |>
+  select(treatment, light, leaf_number, rep, surface, n_stomata) |>
+  pivot_wider(values_from = "n_stomata", names_from = surface) |>
+  mutate(sr = adaxial / (abaxial + adaxial)) |>
+  pull(sr) |>
+  mean()
+
 ```
@@
 
-[Graham - can you fill in details on growth conditions, image aquisitiion, and image dimensions?]
-
-*Arabidopsis thaliana* plants were grown in three different light environments: low light (PAR = 50 $\mu \text{mol}~\text{m}^{-2}~\text{s}^{-1}$), medium light (100 $\mu \text{mol}~\text{m}^{-2}~\text{s}^{-1}$), and high light (200 $\mu \text{mol}~\text{m}^{-2}~\text{s}^{-1}$).
+Plant material, growth conditions, and three-dimensional confocal imaging are described in @dow_disruption_2017.
+Briefly, Columbia (Col-0) ecotype of *Arabidopsis thaliana* (L.) Heynh.
+plants were grown in three different light environments: low light (PAR = 50 $\mu \text{mol}~\text{m}^{-2}~\text{s}^{-1}$), medium light (100 $\mu \text{mol}~\text{m}^{-2}~\text{s}^{-1}$), and high light (200 $\mu \text{mol}~\text{m}^{-2}~\text{s}^{-1}$).
 PAR stands for photosynthetically active radiation.
-Once leaves were mature, we captured images of the abaxial and adaxial leaf surfaces using XXX.
+Seeds were surface-sterilized and stratified at 4°C for 3–5 d in 0.15% agarose solution and then sown directly into Pro-Mix HP soil (Premier Horticulture; Quakerstown, PA, USA) and supplemented with Scott's Osmocote Classic 14-14-14 fertilizer (Scotts-Sierra, Marysville, OH, USA).
+At 10–14 d, seedlings were thinned so only one seedling per container remained.
+Plants were grown to maturity in growth chambers where the conditions were as follows: 16 : 8 h, 22 : 20°C, day : night cycle.
+Imaging of the epidermis and internal leaf structures was performed using a Leica SP5 confocal microscope with the protocol developed by @wuyts_high-contrast_2010 with additional modification described in @dow_disruption_2017.
 We captured 132 images in total, making 66 abaxial-adaxial image pairs.
@@
 
-Stomatal density of *Arabidopsis thaliana* the `r nrow(single_surface_results)` leaf surfaces measured range from 12 to 93 (units) with high light leaves ranging from 93 to 55 (units), medium light from 15 to 35 (units), and low light from 12 to 42 (units).
-Stomatal density varies among light treatments (ANOVA, $`r report_fstat(density_anova, 'light')`$) because the density is much greater in the high light treatment (\autoref{fig:density}).
+Stomatal density of *Arabidopsis thaliana* varies among light treatments (ANOVA, $`r report_fstat(density_anova, 'light')`$) because the density is much greater in the high light treatment (\autoref{fig:density}).
 Density is consistently greater on abaxial leaf surfaces across all light treatments (ANOVA, $`r report_fstat(density_anova, 'surface')`$; \autoref{fig:density}).
 There is no evidence for an interaction between light treatment and surface (ANOVA, $`r report_fstat(density_anova, 'light:surface')`$).
-Leaves are amphistomatous with a mean stomatal density ratio of 0.45.
+Leaves are amphistomatous with a mean stomatal density ratio of `r prettyNum(mean_sr, digits = 2)`.
 

### 2023-10-19 ms/ms.tex
fixed scale following Graham's email on 2023-10-19

@@
    \MyNewLabel{ADDRTXT}{NIAB}{%
-  National Institute of Agricultural Botany, Cambridge, United Kingdom
-CB3 0LE%
+  Department of Crop Science and Production Systems, NIAB, Cambridge,
+CB3 0LE, UK%
  }
@@
 %
-]{Graham Dow}
+]{Graham J. Dow}
 
@@
 
-{[}Graham - can you fill in details on growth conditions, image
-aquisitiion, and image dimensions?{]}
-
-\emph{Arabidopsis thaliana} plants were grown in three different light
-environments: low light (PAR = 50
+Plant material, growth conditions, and three-dimensional confocal
+imaging are described in \citet{dow_disruption_2017}. Briefly, Columbia
+(Col-0) ecotype of \emph{Arabidopsis thaliana} (L.) Heynh. plants were
+grown in three different light environments: low light (PAR = 50
 \(\mu \text{mol}~\text{m}^{-2}~\text{s}^{-1}\)), medium light (100
@@
 \(\mu \text{mol}~\text{m}^{-2}~\text{s}^{-1}\)). PAR stands for
-photosynthetically active radiation. Once leaves were mature, we
-captured images of the abaxial and adaxial leaf surfaces using XXX. We
-captured 132 images in total, making 66 abaxial-adaxial image pairs. We
-measured stomatal position and size using ImageJ
-\citep{schneider_nih_2012}.
+photosynthetically active radiation. Seeds were surface-sterilized and
+stratified at 4°C for 3--5 d in 0.15\% agarose solution and then sown
+directly into Pro-Mix HP soil (Premier Horticulture; Quakerstown, PA,
+USA) and supplemented with Scott's Osmocote Classic 14-14-14 fertilizer
+(Scotts-Sierra, Marysville, OH, USA). At 10--14 d, seedlings were
+thinned so only one seedling per container remained. Plants were grown
+to maturity in growth chambers where the conditions were as follows: 16
+: 8 h, 22 : 20°C, day : night cycle. Imaging of the epidermis and
+internal leaf structures was performed using a Leica SP5 confocal
+microscope with the protocol developed by
+\citet{wuyts_high-contrast_2010} with additional modification described
+in \citet{dow_disruption_2017}. We captured 132 images in total, making
+66 abaxial-adaxial image pairs. We measured stomatal position and size
+using ImageJ \citep{schneider_nih_2012}.
 
@@
 
-Stomatal density of \emph{Arabidopsis thaliana} the 132 leaf surfaces
-measured range from 12 to 93 (units) with high light leaves ranging from
-93 to 55 (units), medium light from 15 to 35 (units), and low light from
-12 to 42 (units). Stomatal density varies among light treatments (ANOVA,
-\(F_{2,126} = 681, P = 2.88 \times 10^{-68}\)) because the density is
-much greater in the high light treatment (\autoref{fig:density}).
-Density is consistently greater on abaxial leaf surfaces across all
-light treatments (ANOVA, \(F_{1,126} = 44.2, P = 8.21 \times 10^{-10}\);
-\autoref{fig:density}). There is no evidence for an interaction between
-light treatment and surface (ANOVA,
-\(F_{2,126} = 2.75 \times 10^{-2}, P = 0.973\)). Leaves are
-amphistomatous with a mean stomatal density ratio of 0.45.
+Stomatal density of \emph{Arabidopsis thaliana} varies among light
+treatments (ANOVA, \(F_{2,126} = 681, P = 2.88 \times 10^{-68}\))
+because the density is much greater in the high light treatment
+(\autoref{fig:density}). Density is consistently greater on abaxial leaf
+surfaces across all light treatments (ANOVA,
+\(F_{1,126} = 44.2, P = 8.21 \times 10^{-10}\); \autoref{fig:density}).
+There is no evidence for an interaction between light treatment and
+surface (ANOVA, \(F_{2,126} = 2.75 \times 10^{-2}, P = 0.973\)). Leaves
+are amphistomatous with a mean stomatal density ratio of 0.44.
 

### 2023-10-19 r/04_plot-single-surface.R

@@ -41,19 +41,24 @@ ggsave("single-surface.pdf", plot = plot, path = "ms/figures",
        width = 4, height = 4)
 
 # 2. plot stomatal density ----
+# Each field is 621.21 um^2 = 0.62121^2 mm^2 = 0.3859019 mm ^ 2
+# so multiply by (1 / 0.62121^2) to get stomata per mm^2
+
 density_plot = ggplot(
     single_surface_results,
     aes(light_reorder, n_stomata, fill = surface)
   ) +
   stat_boxplot() +
   xlab("Light treatment") +
-  ylab(expression(paste("Stomatal density [pores are", a^-2, "]"))) +
+  ylab(expression(paste("Stomatal density [pores m", m^-2, "]"))) +
   scale_fill_grey(start = 0.3) +
   stat_pvalue_manual(
     stat.test.density, label = "p.adj.signif",
     y.position = c(100, 110, 50),
     inherit.aes = FALSE
   ) +
+  scale_y_continuous(labels = scales::number_format(scale = 1 / 0.62121 ^ 2),
+                     breaks = seq(50, 250, 50) * (0.62121 ^ 2)) +
   theme_pubr()
 
 ggsave("density.pdf", plot = density_plot, path = "ms/figures",

### 2023-10-19 r/09_plot-length-area.R

@@ -40,8 +40,12 @@ plot = ggplot(
   geom_point(alpha = 0.2) +
   scale_color_viridis_d() +
   scale_fill_viridis_d() +
-  xlab("Stomata length [UNITS]") +
-  ylab(expression(paste(sqrt(Tessellation~zone), " [UNITS]"))) +
+  xlab(expression(paste("Stomata length [", mu, "m]"))) +
+  ylab(expression(paste(sqrt(Tessellation~zone~area)~bgroup('[', paste(mu, "m"), ']')))) +
+  scale_x_continuous(labels = scales::number_format(scale = pixels_per_um),
+                     breaks = seq(15, 35, 5) / pixels_per_um) +
+  scale_y_continuous(labels = scales::number_format(scale = pixels_per_um),
+                     breaks = seq(50, 200, 50) / pixels_per_um) +
   theme_pubr() +
   theme(legend.position = "right", panel.background = element_rect(fill = "grey"))
 

### 2023-10-19 r/header.R

@@ -59,6 +59,9 @@ library(units)
 
 theme_set(theme_cowplot())
 
+# Scale for conversion from pixel to um according to image metadata
+pixels_per_um = 621.21 / 512
+
 # Fit Rubisco model
 x_depth = c(145, 830.5)
 y_depth = c(0, 700)
@@ -75,38 +78,3 @@ nishio_carbon_1993_fig5 = read_csv("raw-data/nishio_carbon_1993_fig5.csv",
   )
 
 fit_rubisco = gam(rubisco ~ s(rel_depth), data = nishio_carbon_1993_fig5)
-
-# Resolve conflicts
-# library(conflicted)
-# conflict_prefer("abbreviate", "arules")
-# conflict_prefer("write", "arules")
-# conflict_prefer("intersect", "dplyr")
-# conflict_prefer("recode", "dplyr")
-# conflict_prefer("setdiff", "dplyr")
-# conflict_prefer("setequal", "dplyr")
-# conflict_prefer("union", "dplyr")
-# conflict_prefer("filter", "dplyr")
-# conflict_prefer("lag", "dplyr")
-# conflict_prefer("src", "dplyr")
-# conflict_prefer("summarize", "dplyr")
-# conflict_prefer("select", "dplyr")
-# conflict_prefer("combine", "dplyr")
-# conflict_prefer("format.pval", "base")
-# conflict_prefer("units", "base")
-# conflict_prefer("label", "Hmisc")
-# conflict_prefer("lift", "purrr")
-# conflict_prefer("mask", "raster")
-# conflict_prefer("zoom", "raster")
-# conflict_prefer("shift", "raster")
-# conflict_prefer("rotate", "raster")
-# conflict_prefer("cluster", "survival")
-# conflict_prefer("concordance", "survival")
-# conflict_prefer("getData", "raster")
-# conflict_prefer("collapse", "dplyr")
-# conflict_prefer("panel.histogram", "lattice")
-# conflict_prefer("cells", "terra")
-# conflict_prefer("rescale", "terra")
-# conflict_prefer("project", "rgdal")
-# conflict_prefer("describe", "Hmisc")
-# conflict_prefer("size", "arules")
-

### 2023-11-06 ms/ms.Rmd
Finished paper draft. Added new .bib file to reflect new citations

@@
   - code: CU
-    address: Ecology and Evolutionary Biology, Universtity of Colorado, Boulder, CO 80309
+    address: Ecology and Evolutionary Biology, University of Colorado, Boulder, CO 80309
   - code: NIAB
@@
 abstract:
-  - "We will write abstract after Discussion and Conclusions are complete"
+  - "Photosynthesis is co-limited by multiple factors depending on the plant and its environment.
+These include biochemical rate limitations, internal and external water potentials, temperature, irradiance, and carbon dioxide (CO2).
+Amphistomatous leaves have stomata on both abaxial and adaxial leaf surfaces.
+This feature is considered an adaptation to alleviate CO2 diffusion limitations in productive environments where other factors are not limiting as the diffusion path length from stomate to chloroplast is effectively halved.
+Plants can also reduce CO2 limitations through other aspects of optimal stomatal anatomy: stomatal density, distribution and patterning, and size.
+A number of studies have demonstrated that stomata are overdispersed on a single leaf surface; however, much less is known about stomatal anatomy in amphistomatous plants, especially the coordination between leaf surfaces, despite their prevelance in nature and near ubiquity among crop species.
+Here we use novel spatial statistics based on simulations and photosynthesis modelling to test hypotheses about how amphistomatous plants may optimize CO2 limitations in the model angiosperm Arabidopsis thaliana grown in different light environments.
+We find that 1) stomata are overdispersed, but not ideally dispersed, on both leaf surfaces across all light treatments; 2) abaxial and adaxial leaf surfaces are independent; and 3) the theoretical improvements to photosynthesis from abaxial-adaxial stomatal coordination are miniscule (~1%) across the range of feasible traitspace.
+However, we also find that 4) stomatal size is correlated with the mesophyll volume that it supplies with CO2, suggesting that plants may optimize CO2 diffusion limitations through alternative pathways other than ideal, uniform stomatal spacing.
+We discuss the myriad developmental, physical, and evolutionary constraits which may prohibit plants from reaching the theoretical adaptive peak of uniform stomatal spacing and inter-surface stomatal coordination.
+These findings contribute to our understanding of amphistomatous plants' exploration of the leaf anatomy adaptive landscape."
 keywords:
@@
   This work was supported by NSF Award 1929167 to CDM and NSF Award 2307341 to TNB.
-bibliography: stomata-spacing.bib
+bibliography: stomata-spacing_new.bib
 link-citations: yes
@@
 size, density, distribution, and patterning) and movement regulate gas exchange during photosynthesis, namely CO$_2$ assimilation and water loss through transpiration.
-Since waxy cuticles are mostly impermeable to CO$_2$ and H$_2$O, stomata are the primary entry points through which gas exchange occurs despite making up a small percentage of the leaf area [@lange_responses_1971].
+Since waxy cuticles are mostly impermeable to CO$_2$ and H$_2$O, stomata are the primary entry and exit points through which gas exchange occurs despite making up a small percentage of the leaf area [@lange_responses_1971].
 Stomata consist of two guard cells which open and close upon changes in turgor pressure or hormonal cues [@mcadam_linking_2016].
@@
 
-Many successful predictions about stomata and other leaf traits can be made by hypothesizing that natural selection should optimizes CO$_2$ gain per unit of water loss [@cowan_stomatal_1977; @buckley_optimal_2017; @sperry_predicting_2017].
-However, stomatal anatomy may be partially constrained by physical and developments limits on phenotypic expression [@croxdale_stomatal_2000; @harrison_influence_2020; @muir_how_2023].
-Sometimes optimization leads to similar phenotypes across many disparate species.
-For example, almost all stomata follow the one cell spacing rule to maintain proper stomatal functioning [@geisler_oriented_2000; @dow_physiological_2014]; however some species (notably in *Begonia*) appear to benefit from overlapping vapor shells caused by stomatal clustering [@yi_gan_stomatal_2010; @lehmann_effects_2015; @papanatsiou_stomatal_2017].
-Stomatal traits also vary adaptively in different environments.
-Stomatal density positively co-varies with irradiance during leaf development and negatively co-varies with CO$_2$ concentration [@gay_influence_1975; @schoch_dependence_1980; @woodward_stomatal_1987; @royer_stomatal_2001], consistent with optimality predictions.
-Stomatal size is jointly controlled by genome size, light, and stomatal density [@jordan_environmental_2015].
-Size positively co-varies with genome size [@roddy_scaling_2020] and negatively co-varies with stomatal density [@camargo_density_2011].
+Many successful predictions about stomata and other C3 leaf traits can be made by hypothesizing that natural selection should optimize CO$_2$ gain per unit of water loss for any given set of environmental parameters, including their variability [@cowan_stomatal_1977; @buckley_optimal_2017; @sperry_predicting_2017].
 Total stomatal area (size $\times$ density) is optimized for operational conductance ($g_\text{s,op}$) rather than maximum conductance ($g_\text{s,max}$) such that stomatal apertures are most responsive to changes in the environment at their operational aperture [@franks_physiological_2012; @liu_scaling_2021].
 Stomatal aperture can compensate for maladaptive stomatal densities to an extent [@bussis_stomatal_2006], but stomatal density and size ultimately determine a leaf's theoretical $g_\text{s,max}$ [@sack_developmental_2016], which is proportional to $g_\text{s,op}$ [@murray_consistent_2020].
-Additionally, low stomatal densities lead to irregular and insufficient CO$_2$ supply and reduced photosynthetic efficiency in areas far from stomata [@pieruschka_lateral_2006; @morison_lateral_2005], while high stomatal densities can reduce water use efficiency (WUE) [@bussis_stomatal_2006] and incur excessive metabolic costs [@deans_optimization_2020].
-In most species, stomata occur on the abaxial (usually lower) leaf surface; but amphistomy, the occurrence of stomata on both abaxial and adaxial leaf surfaces, is also prevalent in high light environments with constant or intermittent access to sufficient water [@mott_adaptive_1982; @jordan_using_2014; @muir_light_2018; @drake_two_2019; @muir_is_2019].
+Additionally, low stomatal densities lead to irregular and insufficient CO$_2$ supply and reduced photosynthetic efficiency in leaf areas far from stomata [@pieruschka_lateral_2006; @morison_lateral_2005], while high stomatal densities can reduce water use efficiency (WUE) [@bussis_stomatal_2006] and incur excessive metabolic costs [@deans_optimization_2020].
+Stomatal density positively co-varies with irradiance during leaf development and negatively co-varies with CO$_2$ concentration [@gay_influence_1975; @schoch_dependence_1980; @woodward_stomatal_1987; @royer_stomatal_2001], consistent with optimality predictions.
+In most species, stomata occur only on the abaxial (usually lower) leaf surface; but amphistomy, the occurrence of stomata on both abaxial and adaxial leaf surfaces, is also prevalent in high light environments with constant or intermittent access to sufficient water [@mott_adaptive_1982; @jordan_using_2014; @muir_light_2018; @drake_two_2019; @muir_is_2019].
 Amphistomy effectively halves the CO$_2$ diffusion path length and boundary layer resistance by doubling boundary layer conductance [@parkhurst_adaptive_1978; @harrison_influence_2020; @mott_amphistomy_1991].
-Historically, stomatal patterning in dicot angiosperms was thought to be random with an exclusionary distance surrounding each stomate [@sachs_developmental_1974]; however, the developmental controls of stomatal patterning are poorly understood and likely more complex than random development along the leaf surface.
-@croxdale_stomatal_2000] reviews three developmental theories which attempt to explain stomatal patterning in angiosperms: inhibition, cell lineage, and cell cycle, ultimately arguing for a cell cycle based control of stomatal patterning.
+Despite these optimality predictions, stomatal anatomy may be partially constrained by physical and developmental limits on phenotypic expression [@croxdale_stomatal_2000; @harrison_influence_2020; @muir_how_2023].
+
+A number of physical and developmental processes constrain stomatal anatomy traitspace.
+For example, almost all stomata follow the one cell spacing rule to maintain proper stomatal functioning as guard cell movement requires the rapid exchange of ions with neighboring epidermal cells (i.e.
+subsidiary cells) [@geisler_oriented_2000; @dow_physiological_2014].
+This would prevent stomata from being strongly clustered; however, some species (notably in *Begonia*) appear to benefit from the overlapping vapor shells caused by stomatal clustering in dry environments [@yi_gan_stomatal_2010; @lehmann_effects_2015; @papanatsiou_stomatal_2017].
+Historically, stomatal patterning in dicot angiosperms was thought to be random with an exclusionary distance surrounding each stomate [@sachs_developmental_1974]; however, the developmental controls of stomatal patterning are more complex than random development along the leaf surface.
+@croxdale_stomatal_2000 reviews three developmental theories which attempt to explain stomatal patterning in angiosperms: inhibition, cell lineage, and cell cycle, ultimately arguing for a cell cycle based control of stomatal patterning.
+@pillitteri_mechanisms_2012 review the short and long distance signalling pathways associated with stomatal spacing and development, which include cell to cell communication and whole plant integration to ensure the proper spacing of stomata across a single leaf surface depending on environmental ques.
+Much less is known about the development of stomata on the adaxial leaf surface in amphistomatous plants.
+Stomatal size is additionally constrained by genome size with larger genomes leading to larger minimum guard cell size [@jordan_environmental_2015].
+Despite these limitations, ecophysiological theory still predicts optimal stomatal anatomy, the details of which are discussed below.
 
@@
 Maximum photosynthetic rate ($A_\text{max}$) in $C_3$ plants is generally co-limited by biochemistry and diffusion, but modulated by light availability [@parkhurst_intercellular_1990; @manter_ci_2004; @carriqui_diffusional_2015].
-Low light decreasing CO$_2$ demand by limiting electron transport rate, leading to relatively high internal CO$_2$ concentration ($C_\text{i}$) and low $A_\text{max}$ [@kaiser_metabolic_2016].
+Low light decreases CO$_2$ demand by limiting electron transport rate, leading to relatively high internal CO$_2$ concentration ($C_\text{i}$) and low $A_\text{max}$ [@kaiser_metabolic_2016].
 In contrast, well hydrated leaves with open stomata in high light, photosynthesis is often limited by CO$_2$ supply as resistances from the boundary layer, stomatal pore, and mesophyll can result in insufficient CO$C_2$ supply at the chloroplast to maxmimize photosynthesis [@farquhar_biochemical_1980; @lehmeier_cell_2017].
@@
 
-## Stomatal distribution is nonrandom, but far from ideal
+## Stomatal distribution is non-random, but far from ideal
 
@@
 thaliana* leaves partially supports our overall hypothesis that natural selection minimizes the average distance between stomata and chloroplasts, for a given overall stomatal density.
-There are three nonmutually exclusive hypotheses for why several of our predictions were wrong.
+There are three non-mutually exclusive hypotheses for why several of our predictions were wrong.
 First, our predictions are wrong becuase they are based on overly simplistic assumptions about epidermal and mesophyll anatomy.
 Second, natural selection may be constrained by developmental processes that prevent phenotypes from reaching their adaptive optima.
-Third, the benefit of some traits may be of too little consequence to result in fitness differences large enough to be respond to selection.
+Third, the benefit of some traits may be of too little consequence to result in fitness differences large enough to respond to selection.
 We consider the plausibility of these alternative hypotheses below and present ideas for future work to test them.
 
-[this paragraph should discuss why our assumptions may be too simplistic]
 We assume an idealized leaf epidermal and mesophyll structure that is homogenous and unconstrained by other tradeoffs.
 Real leaves not only provide pathways for CO$_2$ diffusion, but must supply water, intercept light, and deter herbivores and pathogens.
-These competing interests results in nonuniform epidermal and mesophyll structure that could alter our predictions about optimal stomatal spacing.
-[explain some examples]
+These competing interests result in nonuniform epidermal and mesophyll structure that could alter our predictions about optimal stomatal spacing.
+In order to maintain consistent leaf water potential across the lamina, stomatal density must be coordinated with vein density [@fiorin_transport_2016].
+Thus, stomatal spacing may be optimized not at the interstomatal level, but at a higher level, coordinating water transport and water loss.
+We also assume a uniform mesophyll; however, in ab-adaxially differentiated leaves, this assumption is overly simplistic.
+The pallisade mesophyll is more tightly packed than the spongy mesophyll as an adaptation to intercept light efficiently, so lateral diffusion may be more limiting in the adaxial portion of the leaf.
+This may explain why adaxial leaf surfaces have consistently higher dispersion indices than abaxial surfaces across all light treatments (\autoref{fig:single-surface}).
+Future gas exchange models should include multiple parameter sets for different strata of mesophyll tissue.
 
-[this paragraph should discuss developmental constraint] No developmental pathway exists to ensure the ideal placement of stomata on the leaf. [add more explanation]
+No developmental pathway exists to ensure an idealized placement of stomata on the leaf surface.
+Rather, stomatal development is a dynamic process that must be plastic to environmental ques.
+Leaves develop based on short and long distance signalling pathways which relay information about incoming light, humidity, temperature, and surrounding stomata to developing leaf tissues [@pillitteri_mechanisms_2012].
+Our results showing an intermediate level of dispersion in stomatal spacing may be best explained by these developmental pathways which ensure the proper spacing of stomata, with an added random effect brought about by the necessity for plasticity in stomatal development (\autoref{fig:single-surface}).
+However, any nonideal randomness in stomatal spacing may offset by simultaneous and coordinated development of the IAS [@baillie_developmental_2020].
+The fact that stomata which supply a greater mesophyll volume tend to be larger suggests that plants may use coordinated development of multiple leaf tissues to negate nonideal stomatal spacing (\autoref{fig:length-area}).
 
-[this paragraph should discuss limits on natural selection] The gas exchange model demonstrate little photosynthetic gain from abaxial-adaxial stomatal coordination (\autoref{fig:model_summary}).
+In amphistomatous leaves, ideal stomatal spacing is complicated by a third dimension.
+Our gas exchange model demonstrates little photosynthetic gain from abaxial-adaxial stomatal coordination (\autoref{fig:model_summary}).
 Even though lateral diffusion may limit photosynthesis [@morison_lateral_2005], the marginal gain from optimally offsetting stomata is not sufficient to generate fitness differences relative to the strength of genetic drift (i.e.
 the drift-barrier).
-We can similarly extrapolate that an ideal, equilateral triangular stomatal spacing is only slightly better than a suboptimal overdispered pattern.
+We can similarly extrapolate that an ideal, equilateral triangular stomatal spacing is only slightly better than a suboptimal pattern.
+Any benefit garnered by ideal stomatal spacing may be additionally offset by a cost to developmental flexibility in variable environments [@pillitteri_mechanisms_2012; @baillie_developmental_2020].
 Explaining these observations as the result of weak selection is in tension with the finding that stomatal size and zone positively covary, which would suggest that small changes in lateral diffusion distance are significant.
 As described above, the positive correlation between stomatal size and zone may be explained by common developmental processes rather than as an adaptation to maximize CO$_2$ diffusion.
+Such flexibility in mature stomate size is an alternate peak in the adaptive landscape which cannot be assessed with our simplistic model assumptions.
+In any case, there is no evidence for coordinated development of both leaf surfaces, and very little theoretical benefit to photosynthesis, except in marginal circumstances which are exceptionally rare in nature.
 
-[discuss significance of our study and future directions] Our study corroborates previous studies which demonstrate that stomata are non-randomly distributed along the leaf surface as a result of developmental mechanisms such as spatially biased arrest of stomatal initials [@boetsch_arrest_1995], oriented asymmetric cell division [@geisler_oriented_2000], and cell cycle controls [@croxdale_stomatal_2000].
-We do not investigate the potential developmental pathways that influence stomatal dispersion in this study; however, they are important to consider as these pathways could limit plants from reaching the theoretical peak in the adaptive landscape: uniform stomatal dispersion.
+Our study corroborates previous studies which demonstrate that stomata are non-randomly distributed along the leaf surface as a result of developmental mechanisms such as spatially biased arrest of stomatal initials [@boetsch_arrest_1995], oriented asymmetric cell division [@geisler_oriented_2000], and cell cycle controls [@croxdale_stomatal_2000].
+We do not investigate the potential developmental pathways that influence stomatal dispersion in this study; however, they are important to consider as these pathways could limit plants from reaching a theoretical peak in the adaptive landscape: uniform stomatal dispersion.
 Instead, as this study suggests, plants may simply compensate with higher stomatal density and by fitting stomatal size to the area that they supply with CO$_2$.
-To understand why stomata are not ideally dispersed, more modelling should be done to estimate the fitness gain of stomatal dispersion.
+To understand why stomata are not ideally dispersed, more modelling (with a more complex set of assumptions including vein density and IAS structure) should be done to estimate the fitness gain of stomatal dispersion.
 Additionally, genetic manipulation studies should attempt to create mutants with clustered and ideally dispersed stomata for a comparison of their photosynthetic traits.
@@
 
-[Conclusion] Our results suggest that after optimiziing stomatal density and having developmental rules for spacing stomata relatively evenly, there may be limited gains to further optimization.
+Our results suggest that after optimizing stomatal density and having developmental rules for spacing stomata relatively evenly, there may be limited gains to further optimization.
 Therefore, developmental constraints may be necessary to make sense of some features of stomatal spacing and distribution.
+The possibility that ideal stomatal spacing is not the "tallest" fitness peak must also be explored, as stomate size is demonstrated in this study to covary with mesophyll volume supplied with CO$_2$.
+This may be especially true in highly variable environments or in large tree species with sun and shade leaves where developmental ques may change rapidly.
+The temporal component, not considered here, could also have significant implications, as CO$_2$ may only be limiting to photosynthesis during short periods when all other conditions are ideal.
+In these cases, the theoretical benefits of ideal stomatal spacing are further diminished.
+Future exploration of these competing hypotheses would require more advanced modelling, additional exploration of IAS space development and its effects on gas exchange, both real and modelled, and knowledge about how often the species of interest is CO$_2$ limited across of range of natural settings.
+Despite these additional considerations, this paper represents an important contribution to understanding the potential drivers of and limitations to stomatal anatomy in amphistomatous plants.
 

### 2023-11-10 ms/ms.tex
knitted new draft

@@
    \MyNewLabel{ADDRTXT}{CU}{%
-  Ecology and Evolutionary Biology, Universtity of Colorado, Boulder, CO
+  Ecology and Evolutionary Biology, University of Colorado, Boulder, CO
 80309%
@@
 \abstract{
-We will write abstract after Discussion and Conclusions are complete}
+Photosynthesis is co-limited by multiple factors depending on the plant
+and its environment. These include biochemical rate limitations,
+internal and external water potentials, temperature, irradiance, and
+carbon dioxide (CO2). Amphistomatous leaves have stomata on both abaxial
+and adaxial leaf surfaces. This feature is considered an adaptation to
+alleviate CO2 diffusion limitations in productive environments where
+other factors are not limiting as the diffusion path length from stomate
+to chloroplast is effectively halved. Plants can also reduce CO2
+limitations through other aspects of optimal stomatal anatomy: stomatal
+density, distribution and patterning, and size. A number of studies have
+demonstrated that stomata are overdispersed on a single leaf surface;
+however, much less is known about stomatal anatomy in amphistomatous
+plants, especially the coordination between leaf surfaces, despite their
+prevelance in nature and near ubiquity among crop species. Here we use
+novel spatial statistics based on simulations and photosynthesis
+modelling to test hypotheses about how amphistomatous plants may
+optimize CO2 limitations in the model angiosperm Arabidopsis thaliana
+grown in different light environments. We find that 1) stomata are
+overdispersed, but not ideally dispersed, on both leaf surfaces across
+all light treatments; 2) abaxial and adaxial leaf surfaces are
+independent; and 3) the theoretical improvements to photosynthesis from
+abaxial-adaxial stomatal coordination are miniscule (\textasciitilde1\%)
+across the range of feasible traitspace. However, we also find that 4)
+stomatal size is correlated with the mesophyll volume that it supplies
+with CO2, suggesting that plants may optimize CO2 diffusion limitations
+through alternative pathways other than ideal, uniform stomatal spacing.
+We discuss the myriad developmental, physical, and evolutionary
+constraits which may prohibit plants from reaching the theoretical
+adaptive peak of uniform stomatal spacing and inter-surface stomatal
+coordination. These findings contribute to our understanding of
+amphistomatous plants' exploration of the leaf anatomy adaptive
+landscape.}
 
@@
 are mostly impermeable to CO\(_2\) and H\(_2\)O, stomata are the primary
-entry points through which gas exchange occurs despite making up a small
-percentage of the leaf area \citep{lange_responses_1971}. Stomata
-consist of two guard cells which open and close upon changes in turgor
-pressure or hormonal cues \citep{mcadam_linking_2016}. The stomatal pore
-leads to an internal space known as the substomatal cavity where gases
-contact the mesophyll. Once in the mesophyll, CO\(_2\) diffuses
-throughout a network of intercellular air space (IAS) and into mesophyll
-cells where CO\(_2\) assimilation (\(A\)) occurs within the chloroplasts
-\citep{lee_diffusion_1964}. Stomatal conductance and transpiration are
-determined by numerous environmental and anatomical parameters such as
-vapor pressure deficit (VPD), irradiance, temperature, wind speed, leaf
-water potential, IAS geometry, mesophyll cell anatomy, and stomatal
-anatomy.
+entry and exit points through which gas exchange occurs despite making
+up a small percentage of the leaf area \citep{lange_responses_1971}.
+Stomata consist of two guard cells which open and close upon changes in
+turgor pressure or hormonal cues \citep{mcadam_linking_2016}. The
+stomatal pore leads to an internal space known as the substomatal cavity
+where gases contact the mesophyll. Once in the mesophyll, CO\(_2\)
+diffuses throughout a network of intercellular air space (IAS) and into
+mesophyll cells where CO\(_2\) assimilation (\(A\)) occurs within the
+chloroplasts \citep{lee_diffusion_1964}. Stomatal conductance and
+transpiration are determined by numerous environmental and anatomical
+parameters such as vapor pressure deficit (VPD), irradiance,
+temperature, wind speed, leaf water potential, IAS geometry, mesophyll
+cell anatomy, and stomatal anatomy.
 
-Many successful predictions about stomata and other leaf traits can be
-made by hypothesizing that natural selection should optimizes CO\(_2\)
-gain per unit of water loss
+Many successful predictions about stomata and other C3 leaf traits can
+be made by hypothesizing that natural selection should optimize CO\(_2\)
+gain per unit of water loss for any given set of environmental
+parameters, including their variability
 \citep{cowan_stomatal_1977, buckley_optimal_2017, sperry_predicting_2017}.
-However, stomatal anatomy may be partially constrained by physical and
-developments limits on phenotypic expression
-\citep{croxdale_stomatal_2000, harrison_influence_2020, muir_how_2023}.
-Sometimes optimization leads to similar phenotypes across many disparate
-species. For example, almost all stomata follow the one cell spacing
-rule to maintain proper stomatal functioning
-\citep{geisler_oriented_2000, dow_physiological_2014}; however some
-species (notably in \emph{Begonia}) appear to benefit from overlapping
-vapor shells caused by stomatal clustering
-\citep{yi_gan_stomatal_2010, lehmann_effects_2015, papanatsiou_stomatal_2017}.
-Stomatal traits also vary adaptively in different environments. Stomatal
-density positively co-varies with irradiance during leaf development and
-negatively co-varies with CO\(_2\) concentration
-\citep{gay_influence_1975, schoch_dependence_1980, woodward_stomatal_1987, royer_stomatal_2001},
-consistent with optimality predictions. Stomatal size is jointly
-controlled by genome size, light, and stomatal density
-\citep{jordan_environmental_2015}. Size positively co-varies with genome
-size \citep{roddy_scaling_2020} and negatively co-varies with stomatal
-density \citep{camargo_density_2011}. Total stomatal area (size
-\(\times\) density) is optimized for operational conductance
-(\(g_\text{s,op}\)) rather than maximum conductance (\(g_\text{s,max}\))
-such that stomatal apertures are most responsive to changes in the
-environment at their operational aperture
+Total stomatal area (size \(\times\) density) is optimized for
+operational conductance (\(g_\text{s,op}\)) rather than maximum
+conductance (\(g_\text{s,max}\)) such that stomatal apertures are most
+responsive to changes in the environment at their operational aperture
 \citep{franks_physiological_2012, liu_scaling_2021}. Stomatal aperture
@@
 stomatal densities lead to irregular and insufficient CO\(_2\) supply
-and reduced photosynthetic efficiency in areas far from stomata
+and reduced photosynthetic efficiency in leaf areas far from stomata
 \citep{pieruschka_lateral_2006, morison_lateral_2005}, while high
@@
 \citep{bussis_stomatal_2006} and incur excessive metabolic costs
-\citep{deans_optimization_2020}. In most species, stomata occur on the
-abaxial (usually lower) leaf surface; but amphistomy, the occurrence of
-stomata on both abaxial and adaxial leaf surfaces, is also prevalent in
-high light environments with constant or intermittent access to
-sufficient water
+\citep{deans_optimization_2020}. Stomatal density positively co-varies
+with irradiance during leaf development and negatively co-varies with
+CO\(_2\) concentration
+\citep{gay_influence_1975, schoch_dependence_1980, woodward_stomatal_1987, royer_stomatal_2001},
+consistent with optimality predictions. In most species, stomata occur
+only on the abaxial (usually lower) leaf surface; but amphistomy, the
+occurrence of stomata on both abaxial and adaxial leaf surfaces, is also
+prevalent in high light environments with constant or intermittent
+access to sufficient water
 \citep{mott_adaptive_1982, jordan_using_2014, muir_light_2018, drake_two_2019, muir_is_2019}.
@@
 \citep{parkhurst_adaptive_1978, harrison_influence_2020, mott_amphistomy_1991}.
+Despite these optimality predictions, stomatal anatomy may be partially
+constrained by physical and developmental limits on phenotypic
+expression
+\citep{croxdale_stomatal_2000, harrison_influence_2020, muir_how_2023}.
+
+A number of physical and developmental processes constrain stomatal
+anatomy traitspace. For example, almost all stomata follow the one cell
+spacing rule to maintain proper stomatal functioning as guard cell
+movement requires the rapid exchange of ions with neighboring epidermal
+cells (i.e.~subsidiary cells)
+\citep{geisler_oriented_2000, dow_physiological_2014}. This would
+prevent stomata from being strongly clustered; however, some species
+(notably in \emph{Begonia}) appear to benefit from the overlapping vapor
+shells caused by stomatal clustering in dry environments
+\citep{yi_gan_stomatal_2010, lehmann_effects_2015, papanatsiou_stomatal_2017}.
 Historically, stomatal patterning in dicot angiosperms was thought to be
@@
 \citep{sachs_developmental_1974}; however, the developmental controls of
-stomatal patterning are poorly understood and likely more complex than
-random development along the leaf surface.
-\citet{croxdale_stomatal_2000}{]} reviews three developmental theories
-which attempt to explain stomatal patterning in angiosperms: inhibition,
-cell lineage, and cell cycle, ultimately arguing for a cell cycle based
-control of stomatal patterning.
+stomatal patterning are more complex than random development along the
+leaf surface. \citet{croxdale_stomatal_2000} reviews three developmental
+theories which attempt to explain stomatal patterning in angiosperms:
+inhibition, cell lineage, and cell cycle, ultimately arguing for a cell
+cycle based control of stomatal patterning.
+\citet{pillitteri_mechanisms_2012} review the short and long distance
+signalling pathways associated with stomatal spacing and development,
+which include cell to cell communication and whole plant integration to
+ensure the proper spacing of stomata across a single leaf surface
+depending on environmental ques. Much less is known about the
+development of stomata on the adaxial leaf surface in amphistomatous
+plants. Stomatal size is additionally constrained by genome size with
+larger genomes leading to larger minimum guard cell size
+\citep{jordan_environmental_2015}. Despite these limitations,
+ecophysiological theory still predicts optimal stomatal anatomy, the
+details of which are discussed below.
 
@@
 \citep{parkhurst_intercellular_1990, manter_ci_2004, carriqui_diffusional_2015}.
-Low light decreasing CO\(_2\) demand by limiting electron transport
-rate, leading to relatively high internal CO\(_2\) concentration
+Low light decreases CO\(_2\) demand by limiting electron transport rate,
+leading to relatively high internal CO\(_2\) concentration
 (\(C_\text{i}\)) and low \(A_\text{max}\) \citep{kaiser_metabolic_2016}.
@@
 synthetic data specific to each sample image as described above. We
-tested whether light treatment affects \(\mathrm{DI}\) and stomatla
+tested whether light treatment affects \(\mathrm{DI}\) and stomatal
 density (\(D_S\)) using analysis of variance (ANOVA).
@@
 
-\hypertarget{stomatal-distribution-is-nonrandom-but-far-from-ideal}{%
-\subsection{Stomatal distribution is nonrandom, but far from
-ideal}\label{stomatal-distribution-is-nonrandom-but-far-from-ideal}}
+\hypertarget{stomatal-distribution-is-non-random-but-far-from-ideal}{%
+\subsection{Stomatal distribution is non-random, but far from
+ideal}\label{stomatal-distribution-is-non-random-but-far-from-ideal}}
 
@@
 between stomata and chloroplasts, for a given overall stomatal density.
-There are three nonmutually exclusive hypotheses for why several of our
+There are three non-mutually exclusive hypotheses for why several of our
 predictions were wrong. First, our predictions are wrong becuase they
@@
 Third, the benefit of some traits may be of too little consequence to
-result in fitness differences large enough to be respond to selection.
-We consider the plausibility of these alternative hypotheses below and
+result in fitness differences large enough to respond to selection. We
+consider the plausibility of these alternative hypotheses below and
 present ideas for future work to test them.
 
-{[}this paragraph should discuss why our assumptions may be too
-simplistic{]} We assume an idealized leaf epidermal and mesophyll
-structure that is homogenous and unconstrained by other tradeoffs. Real
-leaves not only provide pathways for CO\(_2\) diffusion, but must supply
-water, intercept light, and deter herbivores and pathogens. These
-competing interests results in nonuniform epidermal and mesophyll
-structure that could alter our predictions about optimal stomatal
-spacing. {[}explain some examples{]}
+We assume an idealized leaf epidermal and mesophyll structure that is
+homogenous and unconstrained by other tradeoffs. Real leaves not only
+provide pathways for CO\(_2\) diffusion, but must supply water,
+intercept light, and deter herbivores and pathogens. These competing
+interests result in nonuniform epidermal and mesophyll structure that
+could alter our predictions about optimal stomatal spacing. In order to
+maintain consistent leaf water potential across the lamina, stomatal
+density must be coordinated with vein density
+\citep{fiorin_transport_2016}. Thus, stomatal spacing may be optimized
+not at the interstomatal level, but at a higher level, coordinating
+water transport and water loss. We also assume a uniform mesophyll;
+however, in ab-adaxially differentiated leaves, this assumption is
+overly simplistic. The pallisade mesophyll is more tightly packed than
+the spongy mesophyll as an adaptation to intercept light efficiently, so
+lateral diffusion may be more limiting in the adaxial portion of the
+leaf. This may explain why adaxial leaf surfaces have consistently
+higher dispersion indices than abaxial surfaces across all light
+treatments (\autoref{fig:single-surface}). Future gas exchange models
+should include multiple parameter sets for different strata of mesophyll
+tissue.
 
-{[}this paragraph should discuss developmental constraint{]} No
-developmental pathway exists to ensure the ideal placement of stomata on
-the leaf. {[}add more explanation{]}
+No developmental pathway exists to ensure an idealized placement of
+stomata on the leaf surface. Rather, stomatal development is a dynamic
+process that must be plastic to environmental ques. Leaves develop based
+on short and long distance signalling pathways which relay information
+about incoming light, humidity, temperature, and surrounding stomata to
+developing leaf tissues \citep{pillitteri_mechanisms_2012}. Our results
+showing an intermediate level of dispersion in stomatal spacing may be
+best explained by these developmental pathways which ensure the proper
+spacing of stomata, with an added random effect brought about by the
+necessity for plasticity in stomatal development
+(\autoref{fig:single-surface}). However, any nonideal randomness in
+stomatal spacing may offset by simultaneous and coordinated development
+of the IAS \citep{baillie_developmental_2020}. The fact that stomata
+which supply a greater mesophyll volume tend to be larger suggests that
+plants may use coordinated development of multiple leaf tissues to
+negate nonideal stomatal spacing (\autoref{fig:length-area}).
 
-{[}this paragraph should discuss limits on natural selection{]} The gas
-exchange model demonstrate little photosynthetic gain from
-abaxial-adaxial stomatal coordination (\autoref{fig:model_summary}).
-Even though lateral diffusion may limit photosynthesis
-\citep{morison_lateral_2005}, the marginal gain from optimally
-offsetting stomata is not sufficient to generate fitness differences
-relative to the strength of genetic drift (i.e.~the drift-barrier). We
-can similarly extrapolate that an ideal, equilateral triangular stomatal
-spacing is only slightly better than a suboptimal overdispered pattern.
+In amphistomatous leaves, ideal stomatal spacing is complicated by a
+third dimension. Our gas exchange model demonstrates little
+photosynthetic gain from abaxial-adaxial stomatal coordination
+(\autoref{fig:model_summary}). Even though lateral diffusion may limit
+photosynthesis \citep{morison_lateral_2005}, the marginal gain from
+optimally offsetting stomata is not sufficient to generate fitness
+differences relative to the strength of genetic drift (i.e.~the
+drift-barrier). We can similarly extrapolate that an ideal, equilateral
+triangular stomatal spacing is only slightly better than a suboptimal
+pattern. Any benefit garnered by ideal stomatal spacing may be
+additionally offset by a cost to developmental flexibility in variable
+environments
+\citep{pillitteri_mechanisms_2012, baillie_developmental_2020}.
 Explaining these observations as the result of weak selection is in
@@
 processes rather than as an adaptation to maximize CO\(_2\) diffusion.
+Such flexibility in mature stomate size is an alternate peak in the
+adaptive landscape which cannot be assessed with our simplistic model
+assumptions. In any case, there is no evidence for coordinated
+development of both leaf surfaces, and very little theoretical benefit
+to photosynthesis, except in marginal circumstances which are
+exceptionally rare in nature.
 
-{[}discuss significance of our study and future directions{]} Our study
-corroborates previous studies which demonstrate that stomata are
-non-randomly distributed along the leaf surface as a result of
+Our study corroborates previous studies which demonstrate that stomata
+are non-randomly distributed along the leaf surface as a result of
 developmental mechanisms such as spatially biased arrest of stomatal
@@
 however, they are important to consider as these pathways could limit
-plants from reaching the theoretical peak in the adaptive landscape:
+plants from reaching a theoretical peak in the adaptive landscape:
 uniform stomatal dispersion. Instead, as this study suggests, plants may
@@
 size to the area that they supply with CO\(_2\). To understand why
-stomata are not ideally dispersed, more modelling should be done to
-estimate the fitness gain of stomatal dispersion. Additionally, genetic
-manipulation studies should attempt to create mutants with clustered and
-ideally dispersed stomata for a comparison of their photosynthetic
-traits. This could have extremely important implications for maximum
-assimilation rates in crops as most crop species are grown in high light
-where CO\(_2\) is often limiting. In drought-prone environments,
-increased stomatal dispersion may increase water use efficiency by
-reducing the number of stomata needed to achieve the same internal
-CO\(_2\) concentration, \(C_\text{i}\).
+stomata are not ideally dispersed, more modelling (with a more complex
+set of assumptions including vein density and IAS structure) should be
+done to estimate the fitness gain of stomatal dispersion. Additionally,
+genetic manipulation studies should attempt to create mutants with
+clustered and ideally dispersed stomata for a comparison of their
+photosynthetic traits. This could have extremely important implications
+for maximum assimilation rates in crops as most crop species are grown
+in high light where CO\(_2\) is often limiting. In drought-prone
+environments, increased stomatal dispersion may increase water use
+efficiency by reducing the number of stomata needed to achieve the same
+internal CO\(_2\) concentration, \(C_\text{i}\).
 
-{[}Conclusion{]} Our results suggest that after optimiziing stomatal
-density and having developmental rules for spacing stomata relatively
-evenly, there may be limited gains to further optimization. Therefore,
-developmental constraints may be necessary to make sense of some
-features of stomatal spacing and distribution.
+Our results suggest that after optimizing stomatal density and having
+developmental rules for spacing stomata relatively evenly, there may be
+limited gains to further optimization. Therefore, developmental
+constraints may be necessary to make sense of some features of stomatal
+spacing and distribution. The possibility that ideal stomatal spacing is
+not the ``tallest'' fitness peak must also be explored, as stomate size
+is demonstrated in this study to covary with mesophyll volume supplied
+with CO\(_2\). This may be especially true in highly variable
+environments or in large tree species with sun and shade leaves where
+developmental ques may change rapidly. The temporal component, not
+considered here, could also have significant implications, as CO\(_2\)
+may only be limiting to photosynthesis during short periods when all
+other conditions are ideal. In these cases, the theoretical benefits of
+ideal stomatal spacing are further diminished. Future exploration of
+these competing hypotheses would require more advanced modelling,
+additional exploration of IAS space development and its effects on gas
+exchange, both real and modelled, and knowledge about how often the
+species of interest is CO\(_2\) limited across of range of natural
+settings. Despite these additional considerations, this paper represents
+an important contribution to understanding the potential drivers of and
+limitations to stomatal anatomy in amphistomatous plants.
 
@@
 JLW and CDM conceived of the project, analyzed data, and wrote the
-manuscript. GW provided data. TNB contributed to model development and
+manuscript. GJD provided data. TNB contributed to model development and
 helped edit the manuscript.

### 2023-11-12 ms/si.qmd
added ph2d-example.pdf figure to SI

--- 2023-09-25 ms/si.qmd
@@
 title: "2-D Photosynthesis Model"
-format: docx
+format: pdf
 editor: visual
@@
 
+```{r, echo=FALSE}
+knitr::opts_knit$set(root.dir = "..")
+```
+
 We model leaf photosynthesis using a two-dimensional porous medium approximation.
-The model is solved using a finite element method (FEM) in the *R* package **deSolve** [@soetaert2010].
-Table 1 is a glossary model terms and symbols.
-Here we describe the model and associated *R* code.
+The model is solved using a finite element method (FEM) using the `steady.2d()` function in the *R* package **rootSolve** version `r packageVersion("rootSolve")` [@soetaert_practical_2009].
+Table S1 is a glossary model terms and symbols.
 
@@
 We assume that the leaf is a homogenous 2-D medium.
-In the final version, we will incorporate differences in spongy and palisade porosity, gradients in light absorption, electron transport capacity, and Rubisco concentration.
 The mesophyll is $T_\text{leaf}$ thick and the stomata are regularly spaced apart by distance $U$ on both ab- and adaxial surfaces.
@@
 This minimizes the average distance between any point in the mesophyll and its nearest stomate.
-Because of the regular spacing, we only need to model the region between a stomate on surface and the next stomate on the other surface (fig.
-1).
+Because of the regular spacing, we only need to model the region between a stomate on surface and the next stomate on the other surface (Fig.
+S1).
 The rest of the mesophyll will be the same because of symmetry.
 
-## Code
-
-I've copied the *R* code to set up the model and solve it if you want to copy and paste on your own machine. First, here are the packages you'll need.
-
-```{r, packages, message=FALSE}
+```{r, header, echo=FALSE, message=FALSE}
 
@@
 library(ggplot2)
+library(glue)
 library(readr)
+library(stringr)
 library(tidyr)
+
+source("r/functions.R")
+
+parms = read_csv("raw-data/2d-pm-parameters.csv")
+ph2d_offset = read_rds("objects/ph2d_offset.rds")
 
@@
 
-```{r, functions, echo=FALSE, message=FALSE}
-
-source("../r/functions.R")
-
-```
-
-```{r, message = FALSE, fig.cap='Figure 1: Example leaf anatomy analyzed by the 2-D FEM.', echo = FALSE, fig.height=6, fig.width=3}
+```{r, fig1, message = FALSE, fig.cap='Example leaf anatomy analyzed by the 2-D FEM.', echo = FALSE, fig.height=6, fig.width=3}
 
@@
 
-readr::read_csv("../raw-data/2d-pm-parameters.csv") |>
-  filter(!calculated) |>
+parms |>
+  # filter(!calculated) |>
   mutate(Value = glue::glue("${x}$", x = scientize(default_value))) |>
   select(Name = name, Symbol = symbol, Value, Units = units, Notes = notes) |>
-  knitr::kable()
+  knitr::kable(caption = "Table 1: glossary of model terms and mathematical symbols.")
+
 ```
 
+## Solving within gradients in CO$_2$ assimilation and concentration
+
+We extended the 1-D FEM of @earles_excess_2017 to solve a set of partial differential equations describing CO$_2$ diffusion, photosynthesis, and respiration throughout a 2-D leaf geometry.
+The diffusive flux of CO$_2$ through ab- and adaxial stoamta, intercellular airspace, and mesophyll cells was described by:
+
+$$D_\text{e} \frac{\partial^2 C_\text{ias}}{\partial x^2} = -f_\text{liq}$$ {#eq-1d_pm_flux}
+
+$$f_\text{liq} = r_\text{d} + r_\text{d} - r_\text{c}$$ {#eq-1d_pm_fliq}
+
+where
+
+$$
+D_\text{e} = \frac{\phi}{\tau} D_\text{c}
+$$ {#eq-De}
+
+is the effective diffusivity of CO$_2$ though a porous medium composed of an intercellular airspace with a porosity ($\phi$; m$^3$ m$^{-3}$) and tortuosity ($\tau$; m m$^{-1}$).
+$D_\text{c}$ is the diffusion coefficient (m s$^{-1}$) for CO$_2$ in the intercellular airspace, $C_\text{ias}$ is the \[CO$_2$\] (mol m$^{-3}$) at depth $z$ in the intercellular airspace, $f_\text{liq}$ is the volumetric rate of CO$_2$ diffusion from the intercellular airspace into the chloroplast stroma (mol m$^{-3}$ s$^{-1}$), $r_\text{c}$ is the volumetric rate of ribulose 1,5-bisphosphate (RuBP) carboxylation (mol m$^{-3}$ s$^{-1}$), $r_\text{d}$ is the volumetric respiration rate (mol m$^{-3}$ s$^{-1}$), and $r_\text{p}$ is the volumetric photorespiration rate by Rubisco (mol m$^{-3}$ s$^{-1}$).
+
+The volumetric rate of CO$_2$ diffusion from the intercellular airspace into the chloroplast stroma, $f_\text{liq}$, is defined as:
+
+$$f_\text{liq} = \frac{g_\text{liq} (C_\text{liq} - C_\text{ias})}{l_z}$$ {#eq-fliq1}
+
+where $g_\text{liq}$ is the CO$_2$ conductance from the intercellular airspace into the chloroplast stroma (m s$^{-1}$), $C_\text{liq}$ (mol m$^{-3}$) is the \[CO$_2$\] in the stroma, and $l_z$ is the finite element length through which diffusion occurs (m)." (n.b.
+$l_z$ is the same as $t_\text{elem}$ in my Table 1)
+
+In the 2-D model, we extend the flux equation to $x$ (length) and $z$ (depth) dimensions:
+
+$$D_\text{e} \nabla^2 C_\text{ias} = D_\text{e} \bigg(\frac{\partial^2 C_\text{ias}}{\partial x^2} + \frac{\partial^2 C_\text{ias}}{\partial z^2}\bigg) = -f_\text{liq}$$ {#eq-2d_pm_flux}
+
+As I worked through @earles_excess_2017 model, it did not seem like dividing $g_\text{liq}$ by $l_z$ in equation @eq-fliq1 made sense.
+The FEM is a discretization of a continuous model.
+Intuitively, it does make sense why a coarser grid (greater $l_z$) would lead to a larger \[CO$_2$\] drawdown for a given assimilation rate.
+Noting that $g_\text{liq}$ is conductance per m$^2$ of stroma, this means the length scale to divide by should be 1/(stroma area per unit bulk leaf volume).
+This is equivalent to 1/\[(stroma area per leaf area)x(leaf area per bulk leaf volume)\] = 1/\[Sc x (1/Tleaf)\] = Tleaf/Sc (for now I am assuming Sm = Sc, but we can correct this later).
+Our equation is therefore:
+
+$$f_\text{liq} = \frac{g_\text{liq} (C_\text{liq} - C_\text{ias})}{T_\text{leaf} / S_\text{m}}$$ {#eq-fliq2}
+
+I calculated assimilation and respiration the same way as @earles_excess_2017 using the standard C\$\_3\$ biochemical model.
+
+The boundary conditions are that the CO$_2$ concentration in the substomatal cavity is constant at $C_\text{stom}$. The fluxes on the left and right sides are 0 because of symmetry.
+
+## R code
+
+I've copied the *R* code to set up the model and solve it if you want to copy and paste on your own machine. I annotated the *R* code you would need to run the model and include an example result.
+
+```{r}
+
+# Variables
+vars1 = read_rds("objects/model_var.rds") |>
+  left_join(rename(parms, Variable = symbol), by = join_by(Variable))
+
+vars2 = ph2d_offset$parms[pull(vars1, r)] |>
+  as_tibble() |>
+  pivot_longer(everything(), names_to = "r") |>
+  full_join(vars1, by = join_by(r)) |>
+  mutate(
+    # convert from model unit to print unit
+    value1 = case_when(
+      r == "I_0" ~ value * 1e6,
+      r == "phi_pal" ~ value,
+      r == "T_leaf" ~ value * 1e6 - 1,
+      r == "U" ~ 2 * (value * 1e6 - 1)
+    )
+  )
+
+vars3 = vars2 |>
+  transmute(
+    s = glue("{var} = {value1}$ {Units}", var = str_remove(Variable, "\\$$"))
+  ) |>
+  pull(s) |>
+  str_c(collapse = "; ")
+
+# Guide to symbols
+symbols = parms |>
+  filter(r %in% c("C_ias", "C_liq", "I_0", "phi_pal", "T_leaf", "U")) |>
+  transmute(s = glue("{symbol} = {name}")) |> 
+  pull(s) |>
+  str_c(collapse = "; ")
+
+```
+
+extra words
+
+![Example profiles of volumetric CO$_2$ concentrations within otherwise identical amphistomatous leaves that have stomatal positions offset (top row) or aligned (bottom row) based on the 2-D porous medium model.
+When stomata are aligned, both ab- and adaxial stomata are position 0 along the $x$-axis; when stomata are offset, the adaxial stomate is positioned $U/2$ distance away.
+In this example, variables are set as: `r vars3`.
+All other parameter values are described in Table S1.
+`r symbols`.](figures/ph2d-example.pdf){fig-align="center"}
+
 ## References

### 2023-11-12 13_plot-ph2d-example.R

@@ -1,0 +1,30 @@
+source("r/header.R")
+source("r/photo_2d_pm.R")
+
+r = list(I_0 = 0.001, n_x = 101, n_z = 201, phi_pal = 0.2)
+ph2d_offset = photo_2d_pm("offset", fit_rubisco = fit_rubisco, replace = r)
+ph2d_aligned = photo_2d_pm("aligned", fit_rubisco = fit_rubisco, replace = r)
+
+# Plot results
+df_C = expand.grid(
+  z = seq_len(ph2d_offset$parms[["n_z"]]),
+  x = seq_len(ph2d_offset$parms[["n_x"]]),
+  name = c("italic(C)[ias]", "italic(C)[liq]"),
+  arrangement = c("offset", "aligned")
+) |>
+  mutate(value = c(ph2d_offset$fit$y, ph2d_aligned$fit$y))
+
+ggplot(df_C, aes(x, z, z = value)) +
+  facet_grid(arrangement ~ name, labeller = label_parsed) +
+  # geom_contour_filled() +
+  scale_x_continuous(breaks = c(0, 50, 100)) +
+  scale_y_continuous(breaks = c(0, 100, 200)) +
+  geom_contour_filled(breaks = seq(0.004, 0.015, 0.0005)) +
+  xlab(expression(distance~from~abaxial~stomate~bgroup("[", paste(mu, 'm'), "]"))) +
+  ylab(expression(distance~from~abaxial~epidermis~bgroup("[", paste(mu, 'm'), "]"))) +
+  coord_equal() +
+  guides(fill = guide_legend(expression(group('[', CO[2], ']')~mol~m^-3))) +
+  theme(axis.text.x = element_text(angle = 45, hjust = 1))
+
+ggsave("ms/figures/ph2d-example.pdf", width = 6.5, height = 4.5)
+write_rds(ph2d_offset, "objects/ph2d_offset.rds")

### 2023-11-13 ms/ms.Rmd
knitted new draft
expanding model description in si.pdf

@@
 author_contributions: |
-  JLW and CDM conceived of the project, analyzed data, and wrote the manuscript. GW provided data. TNB contributed to model development and helped edit the manuscript.
+  JLW and CDM conceived of the project, analyzed data, and wrote the manuscript. GJD provided data. TNB contributed to model development and helped edit the manuscript.
 acknowledgements: |
   This work was supported by NSF Award 1929167 to CDM and NSF Award 2307341 to TNB.
-bibliography: stomata-spacing_new.bib
+bibliography: stomata-spacing.bib
 link-citations: yes
@@
     number_sections: FALSE # Uncomment to not number sections; default TRUE
-    #number_lines: TRUE # Use `lineno` package - Default FALSE
+    number_lines: TRUE # Use `lineno` package - Default FALSE
     #number_lines_options: ["mathlines","switch"]  # Options for latex lineno package.
@@
       - booktabs # to use with knitr::kable() example below
+    #fontsize: 12pt
 header-includes:
+ - \usepackage{setspace}
+ - \doublespacing
  - \usepackage[nomarkers,tablesfirst]{endfloat} # For figures and tables at end
- - \usepackage{lineno} # For line numbering
- - \linenumbers # For line numbering
  - \usepackage{hyperref} # For autoref
@@
  - \newunicodechar{ʻ}{\okina}
-
 ---
@@
 \noindent $\mathrm{NNI}$ is calculated for each sample image as described above; $\text{median}(\mathrm{NNI_{random}})$ and $\text{median}(\mathrm{NNI_{uniform}})$ are calculated from the synthetic data specific to each sample image as described above.
-We tested whether light treatment affects $\mathrm{DI}$ and stomatla density ($D_S$) using analysis of variance (ANOVA).
+We tested whether light treatment affects $\mathrm{DI}$ and stomatal density ($D_S$) using analysis of variance (ANOVA).
 
@@
 However, such thin leaves with so few stomata are uncommon among C$_3$ plants (some CAM plants have low stomatal density [@males_stomatal_2017]).
-In other areas of parameter space, lateral diffusion limitations were small relative to those along the ab-adaxial axis (see \autoref{fig:model_example} for a representative model solution).
+In other areas of parameter space, lateral diffusion limitations were small relative to those along the ab-adaxial axis (see Fig.
+S1 for a representative model solution).
 
@@
 In these cases, the theoretical benefits of ideal stomatal spacing are further diminished.
-Future exploration of these competing hypotheses would require more advanced modelling, additional exploration of IAS space development and its effects on gas exchange, both real and modelled, and knowledge about how often the species of interest is CO$_2$ limited across of range of natural settings.
+Future exploration of these competing hypotheses would require more advanced modelling, additional exploration of IAS space development and its effects on gas exchange, both real and modeled, and knowledge about how often the species of interest is CO$_2$ limited across of range of natural settings.
 Despite these additional considerations, this paper represents an important contribution to understanding the potential drivers of and limitations to stomatal anatomy in amphistomatous plants.
@@
 # References
-
-<!-- Placeholders for supplemental figures
-
-\renewcommand\thefigure{S\arabic{figure}}    
-\renewcommand\thetable{S\arabic{table}}    
-\renewcommand\theequation{S\arabic{equation}}    
-\setcounter{figure}{0}    
-\setcounter{table}{0}    
-\setcounter{equation}{0}
-
-\begin{figure}
-  \captionsetup{labelformat=empty}
-  \caption{}
-  \label{fig:model_example}
-\end{figure}
--->

### 2023-11-13 ms/ms.tex
expanding model description in si.pdf

@@
 % line numbers
+\usepackage{lineno}
+\linenumbers
 
@@
 % Pandoc header-include feature
+\usepackage{setspace}
+\doublespacing
 \usepackage[nomarkers,tablesfirst]{endfloat}
-\usepackage{lineno}
-\linenumbers
 \usepackage{hyperref}
@@
 parameter space, lateral diffusion limitations were small relative to
-those along the ab-adaxial axis (see \autoref{fig:model_example} for a
-representative model solution).
+those along the ab-adaxial axis (see Fig. S1 for a representative model
+solution).
 
@@
 additional exploration of IAS space development and its effects on gas
-exchange, both real and modelled, and knowledge about how often the
+exchange, both real and modeled, and knowledge about how often the
 species of interest is CO\(_2\) limited across of range of natural
@@
 limitations to stomatal anatomy in amphistomatous plants.
-
-\hypertarget{references}{%
-\section{References}\label{references}}
 
@@
 
+\renewcommand\refname{References}
+
 \bibliographystyle{abbrvnat}

### 2023-11-13 ms/si.qmd
expanding model description in si.pdf

@@
 title: "2-D Photosynthesis Model"
-format: pdf
-editor: visual
-bibliography: ../ms/stomata-spacing.bib
+format:
+  pdf:
+    keep-tex: true  
+    toc: false
+    pdf-engine: xelatex
+header-includes: |
+  \usepackage{fontspec}
+  \setmainfont{Times New Roman}
+  \usepackage{fancyhdr}
+  \pagestyle{fancy}
+  % \fancyhead[C]{Triplett \textit{et al.} — \textit{American Journal of Botany} 2023 – Appendix S1}
+  \usepackage{newunicodechar,graphicx}
+  \DeclareRobustCommand{\okina}{\raisebox{\dimexpr\fontcharht\font`A-\height}{\scalebox{0.8}{`}}}
+  \newunicodechar{ʻ}{\okina}
+  \renewcommand\thefigure{S\arabic{figure}}    
+  \renewcommand\thetable{S\arabic{table}}    
+  \renewcommand\theequation{S\arabic{equation}}
+  \addtokomafont{disposition}{\rmfamily}
+  \RedeclareSectionCommand[
+    font=\normalfont\Large]{section}
+  \RedeclareSectionCommand[
+    font=\normalfont\normalsize\bfseries]{subsection}
+  \RedeclareSectionCommand[
+    font=\normalfont\normalsize\itshape]{subsubsection}
+  \RedeclareSectionCommand[
+    font=\normalfont\normalsize]{paragraph}
+  \usepackage{lscape}
+bibliography: stomata-spacing.bib
 ---
@@
 
-parms = read_csv("raw-data/2d-pm-parameters.csv")
+parms = read_csv("raw-data/2d-pm-parameters.csv", col_types = "cccdccl")
 ph2d_offset = read_rds("objects/ph2d_offset.rds")
@@
 
-```{r, fig1, message = FALSE, fig.cap='Example leaf anatomy analyzed by the 2-D FEM.', echo = FALSE, fig.height=6, fig.width=3}
+```{r, fig1, message = FALSE, fig.cap='Example leaf anatomy analyzed by the 2-D FEM.', echo = FALSE, eval = FALSE, fig.height=6, fig.width=3}
 
@@
 
-```{r, glossary, echo = FALSE, message = FALSE}
+```{r, glossary, echo = FALSE, message = FALSE, results="asis"}
+#| label: glossary
+#| tbl-cap: "Glossary of model terms and mathematical symbols."
 
@@
   # filter(!calculated) |>
-  mutate(Value = glue::glue("${x}$", x = scientize(default_value))) |>
+  mutate(Value = ifelse(
+    is.na(default_value),
+    "$-$",
+    glue::glue("${x}$", x = scientize(default_value))
+  ),
+  units = ifelse(
+    is.na(units) | units == "NA",
+    "$-$",
+    units
+  )) |>
   select(Name = name, Symbol = symbol, Value, Units = units, Notes = notes) |>
-  knitr::kable(caption = "Table 1: glossary of model terms and mathematical symbols.")
+  knitr::kable(booktabs = TRUE, escape = FALSE, longtable = TRUE)
+  # Option 1
+  # gt::gt() |>
+  # gt::fmt_markdown()
+
+  # Option 2
+  # knitr::kable(booktabs = TRUE, escape = FALSE, longtable = TRUE) |>
+  # kableExtra::kable_styling(latex_options = c("striped", "no_vertical_lines")) |>
+  # kableExtra::column_spec(1, width = "3in") |>
+  # kableExtra::column_spec(2, width = "0.5in") |>
+  # kableExtra::column_spec(3, width = "1in") |>
+  # kableExtra::column_spec(4, width = "1.5in") |>
+  # kableExtra::column_spec(5, width = "1.5in") |>
+  # kableExtra::landscape()
 
@@
 
-## Solving within gradients in CO$_2$ assimilation and concentration
+## Solving within-leaf gradients in CO$_2$ assimilation and concentration
 
 We extended the 1-D FEM of @earles_excess_2017 to solve a set of partial differential equations describing CO$_2$ diffusion, photosynthesis, and respiration throughout a 2-D leaf geometry.
-The diffusive flux of CO$_2$ through ab- and adaxial stoamta, intercellular airspace, and mesophyll cells was described by:
+The diffusive flux of CO$_2$ through ab- and adaxial stoamta, intercellular airspace, and mesophyll cells in the $x$ (length) and $z$ (depth) dimensions is:
 
-$$D_\text{e} \frac{\partial^2 C_\text{ias}}{\partial x^2} = -f_\text{liq}$$ {#eq-1d_pm_flux}
+$$D_\text{e} \nabla^2 C_\text{ias} = D_\text{e} \bigg(\frac{\partial^2 C_\text{ias}}{\partial x^2} + \frac{\partial^2 C_\text{ias}}{\partial z^2}\bigg) = -f_\text{liq}$$ {#eq-2d_pm_flux}
 
-$$f_\text{liq} = r_\text{d} + r_\text{d} - r_\text{c}$$ {#eq-1d_pm_fliq}
+$$f_\text{liq} = r_\text{d} + r_\text{p} - r_\text{c}$$ {#eq-1d_pm_fliq}
 
@@
 $$
-D_\text{e} = \frac{\phi}{\tau} D_\text{c}
+D_\text{e} = \frac{\varphi}{\tau} D_\text{c}
 $$ {#eq-De}
 
-is the effective diffusivity of CO$_2$ though a porous medium composed of an intercellular airspace with a porosity ($\phi$; m$^3$ m$^{-3}$) and tortuosity ($\tau$; m m$^{-1}$).
-$D_\text{c}$ is the diffusion coefficient (m s$^{-1}$) for CO$_2$ in the intercellular airspace, $C_\text{ias}$ is the \[CO$_2$\] (mol m$^{-3}$) at depth $z$ in the intercellular airspace, $f_\text{liq}$ is the volumetric rate of CO$_2$ diffusion from the intercellular airspace into the chloroplast stroma (mol m$^{-3}$ s$^{-1}$), $r_\text{c}$ is the volumetric rate of ribulose 1,5-bisphosphate (RuBP) carboxylation (mol m$^{-3}$ s$^{-1}$), $r_\text{d}$ is the volumetric respiration rate (mol m$^{-3}$ s$^{-1}$), and $r_\text{p}$ is the volumetric photorespiration rate by Rubisco (mol m$^{-3}$ s$^{-1}$).
+is the effective diffusivity of CO$_2$ though a porous medium composed of an intercellular airspace with a porosity ($\varphi$; m$^3$ airspace m$^{-3}$ leaf) and tortuosity ($\tau$; m m$^{-1}$).
+$D_\text{c}$ is the diffusion coefficient (m s$^{-1}$) for CO$_2$ in the intercellular airspace, $C_\text{ias}$ is the \[CO$_2$\] (mol m$^{-3}$) at horizontal positions $x$ and depth $z$ in the intercellular airspace, $f_\text{liq}$ is the volumetric rate of CO$_2$ diffusion from the intercellular airspace into the chloroplast stroma (mol m$^{-3}$ s$^{-1}$), $r_\text{c}$ is the volumetric rate of ribulose 1,5-bisphosphate (RuBP) carboxylation (mol m$^{-3}$ s$^{-1}$), $r_\text{d}$ is the volumetric respiration rate (mol m$^{-3}$ s$^{-1}$), and $r_\text{p}$ is the volumetric photorespiration rate by Rubisco (mol m$^{-3}$ s$^{-1}$).
+Following @earles_excess_2017, $r_\text{d}$ is assumed constant per stroma surface area (Table S1) and $r_\text{p}$ is a function of carboxylation ($r_\text{c}$) and $C_\text{liq}$:
 
-The volumetric rate of CO$_2$ diffusion from the intercellular airspace into the chloroplast stroma, $f_\text{liq}$, is defined as:
+$$ r_\text{p} = r_\text{c} \frac{\Gamma^*}{C_\text{liq}} $$ {#eq-rp}
 
-$$f_\text{liq} = \frac{g_\text{liq} (C_\text{liq} - C_\text{ias})}{l_z}$$ {#eq-fliq1}
+Carboxylation rate is the minimum of the Rubisco-limited ($w_\text{c}$) or RuBP-regeneration limited ($w_\text{j}$) carboxylation rate:
 
-where $g_\text{liq}$ is the CO$_2$ conductance from the intercellular airspace into the chloroplast stroma (m s$^{-1}$), $C_\text{liq}$ (mol m$^{-3}$) is the \[CO$_2$\] in the stroma, and $l_z$ is the finite element length through which diffusion occurs (m)." (n.b.
-$l_z$ is the same as $t_\text{elem}$ in my Table 1)
+$$ r_\text{c} = \text{min}(w_\text{c}, w_\text{j}) $$ {#eq-rc}
 
-In the 2-D model, we extend the flux equation to $x$ (length) and $z$ (depth) dimensions:
+where
 
-$$D_\text{e} \nabla^2 C_\text{ias} = D_\text{e} \bigg(\frac{\partial^2 C_\text{ias}}{\partial x^2} + \frac{\partial^2 C_\text{ias}}{\partial z^2}\bigg) = -f_\text{liq}$$ {#eq-2d_pm_flux}
+$$ w_\text{c} = \frac{k_\text{c} X_\text{c} C_\text{liq}}{K_\text{m} + C_\text{liq}},~\text{and}$$ {#eq-wc}
 
-As I worked through @earles_excess_2017 model, it did not seem like dividing $g_\text{liq}$ by $l_z$ in equation @eq-fliq1 made sense.
-The FEM is a discretization of a continuous model.
-Intuitively, it does make sense why a coarser grid (greater $l_z$) would lead to a larger \[CO$_2$\] drawdown for a given assimilation rate.
-Noting that $g_\text{liq}$ is conductance per m$^2$ of stroma, this means the length scale to divide by should be 1/(stroma area per unit bulk leaf volume).
-This is equivalent to 1/\[(stroma area per leaf area)x(leaf area per bulk leaf volume)\] = 1/\[Sc x (1/Tleaf)\] = Tleaf/Sc (for now I am assuming Sm = Sc, but we can correct this later).
-Our equation is therefore:
+$$ w_\text{j} = \frac{C_\text{liq} j_\text{e}}{4 C_\text{liq} + 8 \Gamma^*}.$$ {#eq-wj}
+
+$k_\text{c}$ is the catalytic rate of Rubisco (m$^{-1}$) and $K_\text{m}$ is effective Michaelis-Menten constant for Rubisco (mol m$^{-3}$).
+Following @earles_excess_2017, we assumed the relative concentration of Rubisco follows that of @nishio_carbon_1993, but scaled such that the bulk leaf Rubisco concentration integrates to $X_\text{c}$ described in Table S1.
+We estimated a continuous function describing the relative Rubisco profile as a function of leaf depth using a generalized additive model with the `gam()` function in *R* package **mgcv** version `r packageVersion("mgcv")` [@wood_generalized_2017].
+
+The effective photosynthetic e$^{-}$ transport rate ($j_\text{e}$) is the minimum of the maximum ($j_\text{max}$) and potential ($j_\infty$) photosynthetic e$^{-}$ transport rates at each position within the mesophyll:
+
+$$ j_\text{e} = \text{min}(j_\text{max}, j_\infty) $$ {#eq-je}
+
+The local $j_\text{max}$ follows the same profile as Rubisco as a function of leaf depth and is scaled so that it integrates to $J_\text{max}$ on a leaf-area basis [@earles_excess_2017].
+Potential e$^{-}$ transport is assumed proportional to the local chloroplast concentration so that it integrates to $J_\infty$ on a leaf-area basis [@earles_excess_2017], where:
+
+$$ J_\infty = I_0 \alpha \beta \phi_\text{PSII}.$$ {#eq-Jinf}
+
+Potential e$^{-}$ transport is a product of PPFD incident on the leaf surface ($I_0$, mol m$^{-2}$ s$^{-1}$), whole-leaf light absorption ($\alpha$, mol mol$^{-1}$), the fraction of light absorbed by PSII ($\beta$, mol mol$^{-1}$), and the quantum yield of PSII e$^{-}$ transport ($\phi_\text{PSII}$, mol mol$^{-1}$).
+
+WORKING HERE
+
+The local chlorophyll concentration (SYMBOL) is a function of leaf depth following Equation XX from @borsuk_spatial_2019:
+
+The volumetric rate of CO$_2$ diffusion from the intercellular airspace into the chloroplast stroma, $f_\text{liq}$, is:
 
 $$f_\text{liq} = \frac{g_\text{liq} (C_\text{liq} - C_\text{ias})}{T_\text{leaf} / S_\text{m}}$$ {#eq-fliq2}
+
+where $g_\text{liq}$ is the CO$_2$ conductance from the intercellular airspace into the chloroplast stroma (m s$^{-1}$), $C_\text{liq}$ (mol m$^{-3}$) is the \[CO$_2$\] in the stroma, and $S_\text{m}$ is leaf surface area-to-mesophyll surface area ratio.
+Noting that $g_\text{liq}$ is conductance per m$^2$ of stroma, this means the length scale to divide by should be the inverse of stroma area per unit bulk leaf volume, i.e.
+$1/[S_\text{c} (1 / T_\text{leaf})] = T_\text{leaf} / S_\text{c}$.
+For simplicity, we assume that the entire mesophyll surface area is lined with chloroplasts, hence $S_\text{m} = S_\text{c}$.
 
@@
 
-## R code
-
-I've copied the *R* code to set up the model and solve it if you want to copy and paste on your own machine. I annotated the *R* code you would need to run the model and include an example result.
-
-```{r}
+```{r, echo = FALSE}
 
@@
 ![Example profiles of volumetric CO$_2$ concentrations within otherwise identical amphistomatous leaves that have stomatal positions offset (top row) or aligned (bottom row) based on the 2-D porous medium model.
+Stomatal positions are indicated by black points at the top and bottom of panels.
 When stomata are aligned, both ab- and adaxial stomata are position 0 along the $x$-axis; when stomata are offset, the adaxial stomate is positioned $U/2$ distance away.
@@
 
+\newpage
+
 ## References

### 2023-11-13 ms/si.tex
expanding model description in si.pdf

% Options for packages loaded elsewhere
\PassOptionsToPackage{unicode}{hyperref}
\PassOptionsToPackage{hyphens}{url}
\PassOptionsToPackage{dvipsnames,svgnames,x11names}{xcolor}
%
\documentclass[
  letterpaper,
  DIV=11,
  numbers=noendperiod]{scrartcl}

\usepackage{amsmath,amssymb}
\usepackage{iftex}
\ifPDFTeX
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
  \usepackage{textcomp} % provide euro and other symbols
\else % if luatex or xetex
  \usepackage{unicode-math}
  \defaultfontfeatures{Scale=MatchLowercase}
  \defaultfontfeatures[\rmfamily]{Ligatures=TeX,Scale=1}
\fi
\usepackage{lmodern}
\ifPDFTeX\else  
    % xetex/luatex font selection
\fi
% Use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\IfFileExists{microtype.sty}{% use microtype if available
  \usepackage[]{microtype}
  \UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\makeatletter
\@ifundefined{KOMAClassName}{% if non-KOMA class
  \IfFileExists{parskip.sty}{%
    \usepackage{parskip}
  }{% else
    \setlength{\parindent}{0pt}
    \setlength{\parskip}{6pt plus 2pt minus 1pt}}
}{% if KOMA class
  \KOMAoptions{parskip=half}}
\makeatother
\usepackage{xcolor}
\setlength{\emergencystretch}{3em} % prevent overfull lines
\setcounter{secnumdepth}{-\maxdimen} % remove section numbering
% Make \paragraph and \subparagraph free-standing
\ifx\paragraph\undefined\else
  \let\oldparagraph\paragraph
  \renewcommand{\paragraph}[1]{\oldparagraph{#1}\mbox{}}
\fi
\ifx\subparagraph\undefined\else
  \let\oldsubparagraph\subparagraph
  \renewcommand{\subparagraph}[1]{\oldsubparagraph{#1}\mbox{}}
\fi

\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}\usepackage{longtable,booktabs,array}
\usepackage{calc} % for calculating minipage widths
% Correct order of tables after \paragraph or \subparagraph
\usepackage{etoolbox}
\makeatletter
\patchcmd\longtable{\par}{\if@noskipsec\mbox{}\fi\par}{}{}
\makeatother
% Allow footnotes in longtable head/foot
\IfFileExists{footnotehyper.sty}{\usepackage{footnotehyper}}{\usepackage{footnote}}
\makesavenoteenv{longtable}
\usepackage{graphicx}
\makeatletter
\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
\makeatother
% Scale images if necessary, so that they will not overflow the page
% margins by default, and it is still possible to overwrite the defaults
% using explicit options in \includegraphics[width, height, ...]{}
\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
% Set default figure placement to htbp
\makeatletter
\def\fps@figure{htbp}
\makeatother
\newlength{\cslhangindent}
\setlength{\cslhangindent}{1.5em}
\newlength{\csllabelwidth}
\setlength{\csllabelwidth}{3em}
\newlength{\cslentryspacingunit} % times entry-spacing
\setlength{\cslentryspacingunit}{\parskip}
\newenvironment{CSLReferences}[2] % #1 hanging-ident, #2 entry spacing
 {% don't indent paragraphs
  \setlength{\parindent}{0pt}
  % turn on hanging indent if param 1 is 1
  \ifodd #1
  \let\oldpar\par
  \def\par{\hangindent=\cslhangindent\oldpar}
  \fi
  % set entry spacing
  \setlength{\parskip}{#2\cslentryspacingunit}
 }%
 {}
\usepackage{calc}
\newcommand{\CSLBlock}[1]{#1\hfill\break}
\newcommand{\CSLLeftMargin}[1]{\parbox[t]{\csllabelwidth}{#1}}
\newcommand{\CSLRightInline}[1]{\parbox[t]{\linewidth - \csllabelwidth}{#1}\break}
\newcommand{\CSLIndent}[1]{\hspace{\cslhangindent}#1}

\KOMAoption{captions}{tableheading}
\usepackage{fontspec}
\setmainfont{Times New Roman}
\usepackage{fancyhdr}
\pagestyle{fancy}
% \fancyhead[C]{Triplett \textit{et al.} — \textit{American Journal of Botany} 2023 – Appendix S1}
\usepackage{newunicodechar,graphicx}
\DeclareRobustCommand{\okina}{\raisebox{\dimexpr\fontcharht\font`A-\height}{\scalebox{0.8}{`}}}
\newunicodechar{ʻ}{\okina}
\renewcommand\thefigure{S\arabic{figure}}
\renewcommand\thetable{S\arabic{table}}
\renewcommand\theequation{S\arabic{equation}}
\addtokomafont{disposition}{\rmfamily}
\RedeclareSectionCommand[
  font=\normalfont\Large]{section}
\RedeclareSectionCommand[
  font=\normalfont\normalsize\bfseries]{subsection}
\RedeclareSectionCommand[
  font=\normalfont\normalsize\itshape]{subsubsection}
\RedeclareSectionCommand[
  font=\normalfont\normalsize]{paragraph}
\usepackage{lscape}
\makeatletter
\makeatother
\makeatletter
\makeatother
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\AtBeginDocument{%
\ifdefined\contentsname
  \renewcommand*\contentsname{Table of contents}
\else
  \newcommand\contentsname{Table of contents}
\fi
\ifdefined\listfigurename
  \renewcommand*\listfigurename{List of Figures}
\else
  \newcommand\listfigurename{List of Figures}
\fi
\ifdefined\listtablename
  \renewcommand*\listtablename{List of Tables}
\else
  \newcommand\listtablename{List of Tables}
\fi
\ifdefined\figurename
  \renewcommand*\figurename{Figure}
\else
  \newcommand\figurename{Figure}
\fi
\ifdefined\tablename
  \renewcommand*\tablename{Table}
\else
  \newcommand\tablename{Table}
\fi
}
\@ifpackageloaded{float}{}{\usepackage{float}}
\floatstyle{ruled}
\@ifundefined{c@chapter}{\newfloat{codelisting}{h}{lop}}{\newfloat{codelisting}{h}{lop}[chapter]}
\floatname{codelisting}{Listing}
\newcommand*\listoflistings{\listof{codelisting}{List of Listings}}
\makeatother
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\@ifpackageloaded{subcaption}{}{\usepackage{subcaption}}
\makeatother
\makeatletter
\@ifpackageloaded{tcolorbox}{}{\usepackage[skins,breakable]{tcolorbox}}
\makeatother
\makeatletter
\@ifundefined{shadecolor}{\definecolor{shadecolor}{rgb}{.97, .97, .97}}
\makeatother
\makeatletter
\makeatother
\makeatletter
\makeatother
\ifLuaTeX
  \usepackage{selnolig}  % disable illegal ligatures
\fi
\IfFileExists{bookmark.sty}{\usepackage{bookmark}}{\usepackage{hyperref}}
\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\urlstyle{same} % disable monospaced font for URLs
\hypersetup{
  pdftitle={2-D Photosynthesis Model},
  colorlinks=true,
  linkcolor={blue},
  filecolor={Maroon},
  citecolor={Blue},
  urlcolor={Blue},
  pdfcreator={LaTeX via pandoc}}

\title{2-D Photosynthesis Model}
\author{}
\date{}

\begin{document}
\maketitle
\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[boxrule=0pt, frame hidden, enhanced, sharp corners, borderline west={3pt}{0pt}{shadecolor}, breakable, interior hidden]}{\end{tcolorbox}}\fi

We model leaf photosynthesis using a two-dimensional porous medium
approximation. The model is solved using a finite element method (FEM)
using the \texttt{steady.2d()} function in the \emph{R} package
\textbf{rootSolve} version 1.8.2.4 (Soetaert and Herman 2009). Table S1
is a glossary model terms and symbols.

\hypertarget{leaf-anatomy}{%
\subsection{Leaf anatomy}\label{leaf-anatomy}}

We assume that the leaf is a homogenous 2-D medium. The mesophyll is
\(T_\text{leaf}\) thick and the stomata are regularly spaced apart by
distance \(U\) on both ab- and adaxial surfaces. In this scenario, we
assume that the stomata on each surface are precisely offset from each
other by distance \(U/2\). This minimizes the average distance between
any point in the mesophyll and its nearest stomate. Because of the
regular spacing, we only need to model the region between a stomate on
surface and the next stomate on the other surface (Fig. S1). The rest of
the mesophyll will be the same because of symmetry.

\begin{landscape}
\begin{longtable}[]{@{}
>{\raggedright\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.29}}
>{\raggedright\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.08}}
>{\raggedright\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.1}}
>{\raggedright\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.20}}
>{\raggedright\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.33}}@{}}
\caption{Glossary of model terms and mathematical symbols.
}\tabularnewline
\toprule\noalign{}
\begin{minipage}[b]{\linewidth}\raggedright
Name
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
Symbol
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
Value
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
Units
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
Notes
\end{minipage} \\
\midrule\noalign{}
\endfirsthead
\toprule\noalign{}
\begin{minipage}[b]{\linewidth}\raggedright
Name
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
Symbol
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
Value
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
Units
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
Notes
\end{minipage} \\
\midrule\noalign{}
\endhead
\bottomrule\noalign{}
\endlastfoot
Whole-leaf light absorption & \(\alpha\) & \(0.8\) & mol mol\(^{-1}\) &
assumed \\
Chlorophyll spatial distribution coefficient & \(b_{0,\text{chl}}\) &
\(67.5\) & \(-\) & Borsuk and Brodersen (2019);
\(f_{z,\text{chl}} = b_{0,\text{chl}} + b_{1,\text{chl}} z + b_{2,\text{chl}} z ^ 2\) \\
Chlorophyll spatial distribution coefficient & \(b_{1,\text{chl}}\) &
\(41.5\) & \(-\) & Borsuk and Brodersen (2019);
\(f_{z,\text{chl}} = b_{0,\text{chl}} + b_{1,\text{chl}} z + b_{2,\text{chl}} z ^ 2\) \\
Chlorophyll spatial distribution coefficient & \(b_{2,\text{chl}}\) &
\(-29\) & \(-\) & Borsuk and Brodersen (2019);
\(f_{z,\text{chl}} = b_{0,\text{chl}} + b_{1,\text{chl}} z + b_{2,\text{chl}} z ^ 2\) \\
Fraction of light absorbed by PSII & \(\beta\) & \(0.5\) & mol
mol\(^{-1}\) & assumed \\
{[}CO\(_2\){]} in intercellular airspace & \(C_\text{ias}\) & \(-\) &
mol m\(^{-3}\) & Equation~\ref{eq-2d_pm_flux} and
Equation~\ref{eq-fliq2} \\
{[}CO\(_2\){]} in chloroplast stroma & \(C_\text{liq}\) & \(-\) & mol
m\(^{-3}\) & Equation~\ref{eq-2d_pm_flux} \\
{[}CO\(_2\){]} in substomatal cavity & \(C_\text{stom}\) &
\(1.50 \times 10^{-2}\) & mol m\(^{-3}\) leaf & assumed \\
{[}CO\(_2\){]} compensation point & \(\Gamma^*\) &
\(1.35 \times 10^{-3}\) & mol m\(^{-3}\) stroma & Caemmerer (2000) \\
Diffusivity of {[}CO\(_2\){]} in intercellular airspace & \(D_\text{c}\)
& \(1.54 \times 10^{-5}\) & m\(^{2}\) s\(^{-1}\) & assumed \\
Effective diffusivity of {[}CO\(_2\){]} in intercellular airspace &
\(D_\text{e}\) & \(-\) & m\(^{2}\) s\(^{-1}\) & Equation~\ref{eq-De} \\
Fraction of palisade mesophyll & \(f_\text{pal}\) & \(0.6\) & 1 &
\(1 = f_\text{pal} + f_\text{spg}\) \\
Fraction of spongy mesophyll & \(f_\text{spg}\) & \(0.4\) & 1 &
\(1 = f_\text{pal} + f_\text{spg}\) \\
Chlorophyll fluoresence profile along leaf depth normalized by total
fluoresence & \(F_\text{z,\text{chl}}\) & \(-\) & 1 & Borsuk and
Brodersen (2019);
\(f_{z,\text{chl}} = b_{0,\text{chl}} + b_{0,\text{chl}} z + b_{2,\text{chl}} z ^ 2\) \\
Conductance of cell wall, plasmalemma, cytosol, chloroplast envelope,
and chloroplast stroma & \(g_\text{liq}\) & \(2.50 \times 10^{-4}\) &
m\(^3\) m\(^{-2}\) stroma s\(^{-1}\) & Evans et al. (2009) \\
PPFD incident on the leaf surface & \(I_0\) & \(1.50 \times 10^{-3}\) &
mol m\(^{-2}\) s\(^{-1}\) & assumed \\
Potential photosynthetic e\(^{-}\) transport rate on a leaf area basis &
\(J_\infty\) & \(-\) & mol m\(^{-2}\) leaf s\(^{-1}\) &
Equation~\ref{eq-Jinf} \\
Maximum photosynthetic e\(^{-}\) transport rate on a leaf area basis &
\(J_\text{max}\) & \(2.75 \times 10^{-4}\) & mol m\(^{-2}\) leaf
s\(^{-1}\) & assumed \\
Effective photosynthetic e\(^{-}\) transport rate on a stroma volume
basis & \(j_\text{e}\) & \(-\) & mol m\(^{-3}\) stroma s\(^{-1}\) &
Equation~\ref{eq-je} \\
Potential photosynthetic e\(^{-}\) transport rate on a stroma volume
basis & \(j_\infty\) & \(-\) & mol m\(^{-3}\) stroma s\(^{-1}\) & see
text \\
Maximum photosynthetic e\(^{-}\) transport rate on a stroma volume basis
& \(j_\text{max}\) & \(-\) & mol m\(^{-3}\) stroma s\(^{-1}\) & see
text \\
Catalytic rate of Rubisco & \(k_\text{c}\) & \(2.84\) & m\(^{-1}\) &
Tholen and Zhu (2011) \\
Rubisco effective \(K_\text{m}\) & \(K_\text{m}\) &
\(1.87 \times 10^{-2}\) & mol m\(^{-3}\) & Caemmerer (2000) \\
Number of elements in \(x\) direction & \(n_x\) & \(100\) & \(-\) &
\(U = 2 n_x t_\text{elem}\) \\
Number of elements in \(z\) direction & \(n_z\) & \(200\) & \(-\) &
\(T_\text{leaf} = n_z t_\text{elem}\) \\
Fraction of intercellular airspace (aka porosity), palisade &
\(\varphi_\text{pal}\) & \(0.1\) & m\(^3\) airspace m\(^{-3}\) leaf &
assumed \\
Fraction of intercellular airspace (aka porosity), spongy &
\(\varphi_\text{spg}\) & \(0.3\) & m\(^3\) airspace m\(^{-3}\) leaf &
assumed \\
Quantum yield of PSII e\(^{-}\) transport & \(\phi_\text{PSII}\) &
\(0.85\) & mol mol\(^{-1}\) & assumed \\
Volumetric rate of RuBP carboxylation & \(r_\text{c}\) & \(-\) & mol
m\(^{-2}\) stroma s\(^{-1}\) & Equation~\ref{eq-rc} \\
Volumetric respiration rate & \(r_\text{d}\) & \(6.60 \times 10^{-2}\) &
mol m\(^{-2}\) stroma s\(^{-1}\) & Earles et al. (2017); Tholen and Zhu
(2011) \\
Volumetric rate of photorespiratory CO\(_2\) release & \(r_\text{p}\) &
\(-\) & mol m\(^{-2}\) stroma s\(^{-1}\) & Earles et al. (2017);
Equation~\ref{eq-rp} \\
Leaf surface area-to-mesophyll surface area ratio, palisiade &
\(S_\text{m,pal}\) & \(20\) & m\(^2\) mesophyll m\(^{-2}\) leaf &
assumed \\
Leaf surface area-to-mesophyll surface area ratio spongy &
\(S_\text{m,spg}\) & \(2\) & m\(^2\) mesophyll m\(^{-2}\) leaf &
assumed \\
Tortuosity of intercellular airspace & \(\tau\) & \(1.55\) & m
m\(^{-1}\) & Syvertsen et al. (1995) \\
Thickness of element in both \(x\) and \(z\) directions &
\(t_\text{elem}\) & \(1.00 \times 10^{-6}\) & m &
\(T_\text{leaf} = n_z t_\text{elem}\) \\
Leaf thickness & \(T_\text{leaf}\) & \(-\) & m &
\(T_\text{leaf} = n_z t_\text{elem}\) \\
Interstomatal distance & \(U\) & \(-\) & m &
\(U = n_x t_\text{elem}\) \\
Stroma volume-to-mesophyll surface area ratio & \(V_\text{strom}\) &
\(1.74 \times 10^{-6}\) & m\(^3\) stroma m\(^{-2}\) mesophyll & Earles
et al. (2017); Tholen and Zhu (2011) \\
Rubisco-limited carboxylation rate & \(w_\text{c}\) & \(-\) & mol
m\(^{-2}\) stroma s\(^{-1}\) & Equation~\ref{eq-wc} \\
RuBP regeneration-limited carboxylation rate & \(w_\text{j}\) & \(-\) &
mol m\(^{-2}\) stroma s\(^{-1}\) & Equation~\ref{eq-wj} \\
Rubisco concentration in stroma & \(X_\text{c}\) & \(2.5\) & mol
m\(^{-3}\) stroma & Tholen and Zhu (2011); Oguchi, Hikosaka, and Hirose
(2003) \\
\end{longtable}
\end{landscape}

\hypertarget{solving-within-leaf-gradients-in-co_2-assimilation-and-concentration}{%
\subsection{\texorpdfstring{Solving within-leaf gradients in CO\(_2\)
assimilation and
concentration}{Solving within-leaf gradients in CO\_2 assimilation and concentration}}\label{solving-within-leaf-gradients-in-co_2-assimilation-and-concentration}}

We extended the 1-D FEM of Earles et al. (2017) to solve a set of
partial differential equations describing CO\(_2\) diffusion,
photosynthesis, and respiration throughout a 2-D leaf geometry. The
diffusive flux of CO\(_2\) through ab- and adaxial stoamta,
intercellular airspace, and mesophyll cells in the \(x\) (length) and
\(z\) (depth) dimensions is:

\begin{equation}\protect\hypertarget{eq-2d_pm_flux}{}{D_\text{e} \nabla^2 C_\text{ias} = D_\text{e} \bigg(\frac{\partial^2 C_\text{ias}}{\partial x^2} + \frac{\partial^2 C_\text{ias}}{\partial z^2}\bigg) = -f_\text{liq}}\label{eq-2d_pm_flux}\end{equation}

\begin{equation}\protect\hypertarget{eq-1d_pm_fliq}{}{f_\text{liq} = r_\text{d} + r_\text{p} - r_\text{c}}\label{eq-1d_pm_fliq}\end{equation}

where

\begin{equation}\protect\hypertarget{eq-De}{}{
D_\text{e} = \frac{\varphi}{\tau} D_\text{c}
}\label{eq-De}\end{equation}

is the effective diffusivity of CO\(_2\) though a porous medium composed
of an intercellular airspace with a porosity (\(\varphi\); m\(^3\)
airspace m\(^{-3}\) leaf) and tortuosity (\(\tau\); m m\(^{-1}\)).
\(D_\text{c}\) is the diffusion coefficient (m s\(^{-1}\)) for CO\(_2\)
in the intercellular airspace, \(C_\text{ias}\) is the {[}CO\(_2\){]}
(mol m\(^{-3}\)) at horizontal positions \(x\) and depth \(z\) in the
intercellular airspace, \(f_\text{liq}\) is the volumetric rate of
CO\(_2\) diffusion from the intercellular airspace into the chloroplast
stroma (mol m\(^{-3}\) s\(^{-1}\)), \(r_\text{c}\) is the volumetric
rate of ribulose 1,5-bisphosphate (RuBP) carboxylation (mol m\(^{-3}\)
s\(^{-1}\)), \(r_\text{d}\) is the volumetric respiration rate (mol
m\(^{-3}\) s\(^{-1}\)), and \(r_\text{p}\) is the volumetric
photorespiration rate by Rubisco (mol m\(^{-3}\) s\(^{-1}\)). Following
Earles et al. (2017), \(r_\text{d}\) is assumed constant per stroma
surface area (Table S1) and \(r_\text{p}\) is a function of
carboxylation (\(r_\text{c}\)) and \(C_\text{liq}\):

\begin{equation}\protect\hypertarget{eq-rp}{}{ r_\text{p} = r_\text{c} \frac{\Gamma^*}{C_\text{liq}} }\label{eq-rp}\end{equation}

Carboxylation rate is the minimum of the Rubisco-limited
(\(w_\text{c}\)) or RuBP-regeneration limited (\(w_\text{j}\))
carboxylation rate:

\begin{equation}\protect\hypertarget{eq-rc}{}{ r_\text{c} = \text{min}(w_\text{c}, w_\text{j}) }\label{eq-rc}\end{equation}

where

\begin{equation}\protect\hypertarget{eq-wc}{}{ w_\text{c} = \frac{k_\text{c} X_\text{c} C_\text{liq}}{K_\text{m} + C_\text{liq}},~\text{and}}\label{eq-wc}\end{equation}

\begin{equation}\protect\hypertarget{eq-wj}{}{ w_\text{j} = \frac{C_\text{liq} j_\text{e}}{4 C_\text{liq} + 8 \Gamma^*}.}\label{eq-wj}\end{equation}

\(k_\text{c}\) is the catalytic rate of Rubisco (m\(^{-1}\)) and
\(K_\text{m}\) is effective Michaelis-Menten constant for Rubisco (mol
m\(^{-3}\)). Following Earles et al. (2017), we assumed the relative
concentration of Rubisco follows that of Nishio, Sun, and Vogelmann
(1993), but scaled such that the bulk leaf Rubisco concentration
integrates to \(X_\text{c}\) described in Table S1. We estimated a
continuous function describing the relative Rubisco profile as a
function of leaf depth using a generalized additive model with the
\texttt{gam()} function in \emph{R} package \textbf{mgcv} version 1.9.0
(Wood 2017).

The effective photosynthetic e\(^{-}\) transport rate (\(j_\text{e}\))
is the minimum of the maximum (\(j_\text{max}\)) and potential
(\(j_\infty\)) photosynthetic e\(^{-}\) transport rates at each position
within the mesophyll:

\begin{equation}\protect\hypertarget{eq-je}{}{ j_\text{e} = \text{min}(j_\text{max}, j_\infty) }\label{eq-je}\end{equation}

The local \(j_\text{max}\) follows the same profile as Rubisco as a
function of leaf depth and is scaled so that it integrates to
\(J_\text{max}\) on a leaf-area basis (Earles et al. 2017). Potential
e\(^{-}\) transport is assumed proportional to the local chloroplast
concentration so that it integrates to \(J_\infty\) on a leaf-area basis
(Earles et al. 2017), where:

\begin{equation}\protect\hypertarget{eq-Jinf}{}{ J_\infty = I_0 \alpha \beta \phi_\text{PSII}.}\label{eq-Jinf}\end{equation}

Potential e\(^{-}\) transport is a product of PPFD incident on the leaf
surface (\(I_0\), mol m\(^{-2}\) s\(^{-1}\)), whole-leaf light
absorption (\(\alpha\), mol mol\(^{-1}\)), the fraction of light
absorbed by PSII (\(\beta\), mol mol\(^{-1}\)), and the quantum yield of
PSII e\(^{-}\) transport (\(\phi_\text{PSII}\), mol mol\(^{-1}\)).

The local chlorophyll concentration (SYMBOL) is a function of leaf depth
following Equation XX from Borsuk and Brodersen (2019):

The volumetric rate of CO\(_2\) diffusion from the intercellular
airspace into the chloroplast stroma, \(f_\text{liq}\), is:

\begin{equation}\protect\hypertarget{eq-fliq2}{}{f_\text{liq} = \frac{g_\text{liq} (C_\text{liq} - C_\text{ias})}{T_\text{leaf} / S_\text{m}}}\label{eq-fliq2}\end{equation}

where \(g_\text{liq}\) is the CO\(_2\) conductance from the
intercellular airspace into the chloroplast stroma (m s\(^{-1}\)),
\(C_\text{liq}\) (mol m\(^{-3}\)) is the {[}CO\(_2\){]} in the stroma,
and \(S_\text{m}\) is leaf surface area-to-mesophyll surface area ratio.
Noting that \(g_\text{liq}\) is conductance per m\(^2\) of stroma, this
means the length scale to divide by should be the inverse of stroma area
per unit bulk leaf volume,
i.e.~\(1/[S_\text{c} (1 / T_\text{leaf})] = T_\text{leaf} / S_\text{c}\).
For simplicity, we assume that the entire mesophyll surface area is
lined with chloroplasts, hence \(S_\text{m} = S_\text{c}\).

I calculated assimilation and respiration the same way as Earles et al.
(2017) using the standard C\$\_3\$ biochemical model.

The boundary conditions are that the CO\(_2\) concentration in the
substomatal cavity is constant at \(C_\text{stom}\). The fluxes on the
left and right sides are 0 because of symmetry.

extra words

\begin{figure}

{\centering \includegraphics{figures/ph2d-example.pdf}

}

\caption{Example profiles of volumetric CO\(_2\) concentrations within
otherwise identical amphistomatous leaves that have stomatal positions
offset (top row) or aligned (bottom row) based on the 2-D porous medium
model. Stomatal positions are indicated by black points at the top and
bottom of panels. When stomata are aligned, both ab- and adaxial stomata
are position 0 along the \(x\)-axis; when stomata are offset, the
adaxial stomate is positioned \(U/2\) distance away. In this example,
variables are set as: \(I_0 = 1000\) \(\mu\)mol m\(^{-2}\) s\(^{-1}\);
\(\varphi_\text{pal} = 0.2\) m\(^3\) airspace m\(^{-3}\) leaf;
\(T_\text{leaf} = 200\) \(\mu\)m; \(U = 200\) \(\mu\)m. All other
parameter values are described in Table S1. \(C_\text{ias}\) =
{[}CO\(_2\){]} in intercellular airspace; \(C_\text{liq}\) =
{[}CO\(_2\){]} in chloroplast stroma; \(I_0\) = PPFD incident on the
leaf surface; \(\varphi_\text{pal}\) = Fraction of intercellular
airspace (aka porosity), palisade; \(T_\text{leaf}\) = Leaf thickness;
\(U\) = Interstomatal distance.}

\end{figure}

\newpage

\hypertarget{references}{%
\subsection*{References}\label{references}}
\addcontentsline{toc}{subsection}{References}

\hypertarget{refs}{}
\begin{CSLReferences}{1}{0}
\leavevmode\vadjust pre{\hypertarget{ref-borsuk_spatial_2019}{}}%
Borsuk, Aleca M., and Craig R. Brodersen. 2019. {``The Spatial
Distribution of Chlorophyll in Leaves.''} \emph{Plant Physiology} 180
(3): 1406--17. \url{https://doi.org/10.1104/pp.19.00094}.

\leavevmode\vadjust pre{\hypertarget{ref-von_caemmerer_biochemical_2000}{}}%
Caemmerer, Susanne von. 2000. \emph{Biochemical Models of Leaf
Photosynthesis}. Techniques in Plant Science 2. Collingwood: CSIRO.

\leavevmode\vadjust pre{\hypertarget{ref-earles_excess_2017}{}}%
Earles, J. Mason, Guillaume Théroux-Rancourt, Matthew E. Gilbert, Andrew
J. McElrone, and Craig R. Brodersen. 2017. {``Excess Diffuse Light
Absorption in Upper Mesophyll Limits {CO}\(_{\textrm{2}}\) Drawdown and
Depresses Photosynthesis.''} \emph{Plant Physiology} 174 (2): 1082--96.
\url{https://doi.org/10.1104/pp.17.00223}.

\leavevmode\vadjust pre{\hypertarget{ref-evans_resistances_2009}{}}%
Evans, J. R., R. Kaldenhoff, B. Genty, and I. Terashima. 2009.
{``Resistances Along the {CO}\(_{\textrm{2}}\) Diffusion Pathway Inside
Leaves.''} \emph{Journal of Experimental Botany} 60 (8): 2235--48.
\url{https://doi.org/10.1093/jxb/erp117}.

\leavevmode\vadjust pre{\hypertarget{ref-nishio_carbon_1993}{}}%
Nishio, J. N., J. Sun, and T. C. Vogelmann. 1993. {``Carbon {Fixation}
{Gradients} Across {Spinach} {Leaves} {Do} {Not} {Follow} {Internal}
{Light} {Gradients}.''} \emph{The Plant Cell}, August, 953--61.
\url{https://doi.org/10.1105/tpc.5.8.953}.

\leavevmode\vadjust pre{\hypertarget{ref-oguchi_does_2003}{}}%
Oguchi, R., K. Hikosaka, and T. Hirose. 2003. {``Does the Photosynthetic
Light-Acclimation Need Change in Leaf Anatomy?: {Chloroplast} Volume
Change in Photosynthetic Light-Acclimation.''} \emph{Plant, Cell \&
Environment} 26 (4): 505--12.
\url{https://doi.org/10.1046/j.1365-3040.2003.00981.x}.

\leavevmode\vadjust pre{\hypertarget{ref-soetaert_practical_2009}{}}%
Soetaert, Karline, and Peter M. J. Herman, eds. 2009. \emph{A
{Practical} {Guide} to {Ecological} {Modelling}}. Dordrecht: Springer
Netherlands. \url{https://doi.org/10.1007/978-1-4020-8624-3}.

\leavevmode\vadjust pre{\hypertarget{ref-syvertsen_relationship_1995}{}}%
Syvertsen, J. P., J. Lloyd, C. McConchie, P. E. Kriedemann, and G. D.
Farquhar. 1995. {``On the Relationship Between Leaf Anatomy and
{CO}\(_{\textrm{2}}\) Diffusion Through the Mesophyll of Hypostomatous
Leaves.''} \emph{Plant, Cell and Environment} 18 (2): 149--57.
\url{https://doi.org/10.1111/j.1365-3040.1995.tb00348.x}.

\leavevmode\vadjust pre{\hypertarget{ref-tholen_mechanistic_2011}{}}%
Tholen, D., and X.-G. Zhu. 2011. {``The Mechanistic Basis of Internal
Conductance: {A} Theoretical Analysis of Mesophyll Cell Photosynthesis
and {CO}\(_{\textrm{2}}\) Diffusion.''} \emph{Plant Physiology} 156 (1):
90--105. \url{https://doi.org/10.1104/pp.111.172346}.

\leavevmode\vadjust pre{\hypertarget{ref-wood_generalized_2017}{}}%
Wood, Simon N. 2017. \emph{Generalized {Additive} {Models}: {An}
{Introduction} with {R}}. 2nd ed. Chapman; Hall/CRC.
\url{https://doi.org/10.1201/9781315370279}.

\end{CSLReferences}

\end{document}

### 2023-11-13 r/13_plot-ph2d-example.R

@@ -14,12 +14,19 @@ df_C = expand.grid(
 ) |>
   mutate(value = c(ph2d_offset$fit$y, ph2d_aligned$fit$y))
 
-ggplot(df_C, aes(x, z, z = value)) +
+df_points = crossing(
+  name = c("italic(C)[ias]", "italic(C)[liq]"),
+  arrangement = c("offset", "aligned"),
+  z = c(0, r$n_z)
+) |>
+  mutate(x = 0 + (z > 0 & arrangement == "offset") * r$n_x, value = numeric(8))
+
+ggplot(df_C, aes(x, z)) +
   facet_grid(arrangement ~ name, labeller = label_parsed) +
-  # geom_contour_filled() +
-  scale_x_continuous(breaks = c(0, 50, 100)) +
+  scale_x_continuous(breaks = c(0, 50, 100), expand = expansion(mult = rep(0.1, 2))) +
   scale_y_continuous(breaks = c(0, 100, 200)) +
-  geom_contour_filled(breaks = seq(0.004, 0.015, 0.0005)) +
+  geom_contour_filled(mapping = aes(z = value), breaks = seq(0.004, 0.015, 0.0005)) +
+  geom_point( data = df_points) +
   xlab(expression(distance~from~abaxial~stomate~bgroup("[", paste(mu, 'm'), "]"))) +
   ylab(expression(distance~from~abaxial~epidermis~bgroup("[", paste(mu, 'm'), "]"))) +
   coord_equal() +

### 2023-11-13 r/14_render-si.R

@@ -1,0 +1,27 @@
+source("r/header.R")
+
+quarto::quarto_render("ms/si.qmd")
+
+si = read_lines("ms/si.tex")
+start_longtable = which(str_detect(si, "begin\\{longtable\\}"))
+end_longtable = which(str_detect(si, "end\\{longtable\\}"))
+
+# Change column widths
+colwidths = c(
+">{\\raggedright\\arraybackslash}p{(\\columnwidth - 8\\tabcolsep) * \\real{0.29}}",
+">{\\raggedright\\arraybackslash}p{(\\columnwidth - 8\\tabcolsep) * \\real{0.08}}",
+">{\\raggedright\\arraybackslash}p{(\\columnwidth - 8\\tabcolsep) * \\real{0.1}}",
+">{\\raggedright\\arraybackslash}p{(\\columnwidth - 8\\tabcolsep) * \\real{0.20}}",
+">{\\raggedright\\arraybackslash}p{(\\columnwidth - 8\\tabcolsep) * \\real{0.33}}@{}}"
+)
+
+si[start_longtable + seq(5)] <- colwidths
+
+c(si[seq(start_longtable) - 1],
+  "\\begin{landscape}",
+  si[start_longtable:end_longtable],
+  "\\end{landscape}",
+  si[(end_longtable + 1):length(si)]) |>
+  write_lines("ms/si.tex")
+
+system('cd ms; xelatex si.tex; xelatex si.tex')

### 2023-11-13 ms/si.qmd
final draft of ms/si.pdf

@@
 
-We model leaf photosynthesis using a two-dimensional porous medium approximation.
-The model is solved using a finite element method (FEM) using the `steady.2d()` function in the *R* package **rootSolve** version `r packageVersion("rootSolve")` [@soetaert_practical_2009].
+We modeled leaf photosynthesis using a two-dimensional porous medium approximation.
+The `steady.2d()` function in the *R* package **rootSolve** version `r packageVersion("rootSolve")` [@soetaert_practical_2009] solves the model using a finite element method (FEM).
+The 2-D leaf profile is $n_x$ elements long and $n_z$ elements deep with square elements of area $t_\text{elem}^2$.
+In all cases, we set $t_\text{elem} = 1~\mu \text{m}$.
 Table S1 is a glossary model terms and symbols.
@@
 The rest of the mesophyll will be the same because of symmetry.
+This allowed us to set the boundary fluxes on the left and right sides of the leaf profile to 0.
 
@@
 
-```{r, glossary, echo = FALSE, message = FALSE, results="asis"}
-#| label: glossary
-#| tbl-cap: "Glossary of model terms and mathematical symbols."
-
-parms |>
-  # filter(!calculated) |>
-  mutate(Value = ifelse(
-    is.na(default_value),
-    "$-$",
-    glue::glue("${x}$", x = scientize(default_value))
-  ),
-  units = ifelse(
-    is.na(units) | units == "NA",
-    "$-$",
-    units
-  )) |>
-  select(Name = name, Symbol = symbol, Value, Units = units, Notes = notes) |>
-  knitr::kable(booktabs = TRUE, escape = FALSE, longtable = TRUE)
-  # Option 1
-  # gt::gt() |>
-  # gt::fmt_markdown()
-
-  # Option 2
-  # knitr::kable(booktabs = TRUE, escape = FALSE, longtable = TRUE) |>
-  # kableExtra::kable_styling(latex_options = c("striped", "no_vertical_lines")) |>
-  # kableExtra::column_spec(1, width = "3in") |>
-  # kableExtra::column_spec(2, width = "0.5in") |>
-  # kableExtra::column_spec(3, width = "1in") |>
-  # kableExtra::column_spec(4, width = "1.5in") |>
-  # kableExtra::column_spec(5, width = "1.5in") |>
-  # kableExtra::landscape()
-
-```
-
 ## Solving within-leaf gradients in CO$_2$ assimilation and concentration
@@
 
-$$D_\text{e} \nabla^2 C_\text{ias} = D_\text{e} \bigg(\frac{\partial^2 C_\text{ias}}{\partial x^2} + \frac{\partial^2 C_\text{ias}}{\partial z^2}\bigg) = -f_\text{liq}$$ {#eq-2d_pm_flux}
+$$D_\text{e} \nabla^2 C_\text{ias} = D_\text{e} \bigg(\frac{\partial^2 C_\text{ias}}{\partial x^2} + \frac{\partial^2 C_\text{ias}}{\partial z^2}\bigg) = -f_\text{liq},$$ {#eq-2d_pm_flux}
+
+where
 
@@
 
-where
+and
 
@@
 is the effective diffusivity of CO$_2$ though a porous medium composed of an intercellular airspace with a porosity ($\varphi$; m$^3$ airspace m$^{-3}$ leaf) and tortuosity ($\tau$; m m$^{-1}$).
+We assume the palisade ($\varphi_\text{pal}$) is less porous than the spongy ($\varphi_\text{spg}$) mesophyll (Table S1).
 $D_\text{c}$ is the diffusion coefficient (m s$^{-1}$) for CO$_2$ in the intercellular airspace, $C_\text{ias}$ is the \[CO$_2$\] (mol m$^{-3}$) at horizontal positions $x$ and depth $z$ in the intercellular airspace, $f_\text{liq}$ is the volumetric rate of CO$_2$ diffusion from the intercellular airspace into the chloroplast stroma (mol m$^{-3}$ s$^{-1}$), $r_\text{c}$ is the volumetric rate of ribulose 1,5-bisphosphate (RuBP) carboxylation (mol m$^{-3}$ s$^{-1}$), $r_\text{d}$ is the volumetric respiration rate (mol m$^{-3}$ s$^{-1}$), and $r_\text{p}$ is the volumetric photorespiration rate by Rubisco (mol m$^{-3}$ s$^{-1}$).
@@
 
-$$ r_\text{p} = r_\text{c} \frac{\Gamma^*}{C_\text{liq}} $$ {#eq-rp}
+$$ r_\text{p} = r_\text{c} \frac{\Gamma^*}{C_\text{liq}}. $$ {#eq-rp}
 
-Carboxylation rate is the minimum of the Rubisco-limited ($w_\text{c}$) or RuBP-regeneration limited ($w_\text{j}$) carboxylation rate:
+Carboxylation rate is the minimum of the Rubisco-limited ($w_\text{c}$) or RuBP-regeneration limited ($w_\text{j}$) carboxylation rates:
 
-$$ r_\text{c} = \text{min}(w_\text{c}, w_\text{j}) $$ {#eq-rc}
+$$ r_\text{c} = \text{min}(w_\text{c}, w_\text{j}) \frac{S_\text{m}}{T_\text{leaf}} V_\text{strom}, $$ {#eq-rc}
 
@@
 
+Multiplying by $\frac{S_\text{m}}{T_\text{leaf}} V_\text{strom}$ converts carboxylation from per area to per stroma volume units.
 $k_\text{c}$ is the catalytic rate of Rubisco (m$^{-1}$) and $K_\text{m}$ is effective Michaelis-Menten constant for Rubisco (mol m$^{-3}$).
@@
 
-$$ j_\text{e} = \text{min}(j_\text{max}, j_\infty) $$ {#eq-je}
+$$ j_\text{e} = \text{min}(j_\text{max}, j_\infty). $$ {#eq-je}
 
-The local $j_\text{max}$ follows the same profile as Rubisco as a function of leaf depth and is scaled so that it integrates to $J_\text{max}$ on a leaf-area basis [@earles_excess_2017].
-Potential e$^{-}$ transport is assumed proportional to the local chloroplast concentration so that it integrates to $J_\infty$ on a leaf-area basis [@earles_excess_2017], where:
+The local $j_\text{max}$ follows the same depth profile as Rubisco and is scaled by local $S_\text{m}$ and $V_\text{strom}$ so that it integrates to $J_\text{max}$ on a leaf-area basis [@earles_excess_2017]:
 
-$$ J_\infty = I_0 \alpha \beta \phi_\text{PSII}.$$ {#eq-Jinf}
+$$J_\text{max} = \int_0^{T_\text{leaf}} j_{\text{max},i} S_{\text{m},i} V_\text{strom} dz.$$ {#eq-jmax}
+
+Potential e$^{-}$ transport follows the same depth profile as chlorophyll concentration and is scaled by local $S_\text{m}$ and $V_\text{strom}$ so that it integrates to $J_\infty$ on a leaf-area basis [@earles_excess_2017], where:
+
+$$ J_\infty = I_0 \alpha \beta \phi_\text{PSII} = \int_0^{T_\text{leaf}} j_{\infty,i} S_{\text{m},i} V_\text{strom} dz.$$ {#eq-Jinf}
 
@@
 
-WORKING HERE
+The local chlorophyll concentration is a function of depth described by a quadratic equation [@johnson_leaf_2005]:
 
-The local chlorophyll concentration (SYMBOL) is a function of leaf depth following Equation XX from @borsuk_spatial_2019:
+$$F_\text{chl} = f(z) = b_{0,\text{chl}} + b_{1,\text{chl}} z + b_{2,\text{chl}} z ^ 2.$$ {#eq-Fchl}
 
+We used generic parameters from @borsuk_spatial_2019, rescaled to relative depth on a 0-1 rather than 0-100 interval.
+
+For mass balance, the CO$_2$ assimilated via carboxylation must be supplied by diffusive flux from the intercellular airspace into the chloroplast stroma.
 The volumetric rate of CO$_2$ diffusion from the intercellular airspace into the chloroplast stroma, $f_\text{liq}$, is:
 
-$$f_\text{liq} = \frac{g_\text{liq} (C_\text{liq} - C_\text{ias})}{T_\text{leaf} / S_\text{m}}$$ {#eq-fliq2}
+$$f_\text{liq} = \frac{g_\text{liq} (C_\text{liq} - C_\text{ias})}{T_\text{leaf} / S_\text{m}},$$ {#eq-fliq2}
 
-where $g_\text{liq}$ is the CO$_2$ conductance from the intercellular airspace into the chloroplast stroma (m s$^{-1}$), $C_\text{liq}$ (mol m$^{-3}$) is the \[CO$_2$\] in the stroma, and $S_\text{m}$ is leaf surface area-to-mesophyll surface area ratio.
+where $g_\text{liq}$ is the CO$_2$ conductance from the intercellular airspace into the chloroplast stroma (m s$^{-1}$), $C_\text{liq}$ (mol m$^{-3}$) is the \[CO$_2$\] in the stroma, and $S_\text{m}$ is mesophyll surface area-to-leaf surface area ratio.
 Noting that $g_\text{liq}$ is conductance per m$^2$ of stroma, this means the length scale to divide by should be the inverse of stroma area per unit bulk leaf volume, i.e.
@@
 For simplicity, we assume that the entire mesophyll surface area is lined with chloroplasts, hence $S_\text{m} = S_\text{c}$.
+Following @earles_excess_2017, we assumed greater surface area in the palisade ($S_\text{m,pal}$) than spongy ($S_\text{m,spg}$) mesophyll.
+The fractions of palisade and spongy mesophyll are $f_\text{pal}$ and $f_\text{spg}$, respectively.
 
-I calculated assimilation and respiration the same way as @earles_excess_2017 using the standard C\$\_3\$ biochemical model.
-
-The boundary conditions are that the CO$_2$ concentration in the substomatal cavity is constant at $C_\text{stom}$. The fluxes on the left and right sides are 0 because of symmetry.
+Following @earles_excess_2017, we set boundary conditions at the stomata, $C_\text{stom}$, to be $0.85 \times$ atmospheric CO$_2$ (Table S1).
+The fluxes on the left and right sides are 0 because of symmetry; the other fluxes on the ab- and adaxial surface are assumed 0 (i.e.
+the epidermis is impermeable to CO$_2$).
 
@@
 
-extra words
-
 ![Example profiles of volumetric CO$_2$ concentrations within otherwise identical amphistomatous leaves that have stomatal positions offset (top row) or aligned (bottom row) based on the 2-D porous medium model.
@@
 
+```{r, glossary, echo = FALSE, message = FALSE, results="asis"}
+#| label: glossary
+#| tbl-cap: "Glossary of model terms and mathematical symbols."
+
+parms |>
+  # filter(!calculated) |>
+  mutate(Value = case_when(
+    is.na(default_value) & !calculated ~ "variable",
+    is.na(default_value) & calculated ~ "calculated",
+    is.numeric(default_value) ~ glue::glue("${x}$", x = scientize(default_value))
+  ),
+  units = ifelse(
+    is.na(units) | units == "NA",
+    "$-$",
+    units
+  )) |>
+  select(Name = name, Symbol = symbol, Value, Units = units, Notes = notes) |>
+  knitr::kable(booktabs = TRUE, escape = FALSE, longtable = TRUE)
+  # Option 1
+  # gt::gt() |>
+  # gt::fmt_markdown()
+
+  # Option 2
+  # knitr::kable(booktabs = TRUE, escape = FALSE, longtable = TRUE) |>
+  # kableExtra::kable_styling(latex_options = c("striped", "no_vertical_lines")) |>
+  # kableExtra::column_spec(1, width = "3in") |>
+  # kableExtra::column_spec(2, width = "0.5in") |>
+  # kableExtra::column_spec(3, width = "1in") |>
+  # kableExtra::column_spec(4, width = "1.5in") |>
+  # kableExtra::column_spec(5, width = "1.5in") |>
+  # kableExtra::landscape()
+
+```
+
 \newpage

### 2023-11-13 ms/si.tex
final draft of ms/si.pdf

@@
 \maketitle
-\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[boxrule=0pt, frame hidden, enhanced, sharp corners, borderline west={3pt}{0pt}{shadecolor}, breakable, interior hidden]}{\end{tcolorbox}}\fi
+\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[interior hidden, boxrule=0pt, sharp corners, breakable, frame hidden, enhanced, borderline west={3pt}{0pt}{shadecolor}]}{\end{tcolorbox}}\fi
 
-We model leaf photosynthesis using a two-dimensional porous medium
-approximation. The model is solved using a finite element method (FEM)
-using the \texttt{steady.2d()} function in the \emph{R} package
-\textbf{rootSolve} version 1.8.2.4 (Soetaert and Herman 2009). Table S1
-is a glossary model terms and symbols.
+We modeled leaf photosynthesis using a two-dimensional porous medium
+approximation. The \texttt{steady.2d()} function in the \emph{R} package
+\textbf{rootSolve} version 1.8.2.4 (Soetaert and Herman 2009) solves the
+model using a finite element method (FEM). The 2-D leaf profile is
+\(n_x\) elements long and \(n_z\) elements deep with square elements of
+area \(t_\text{elem}^2\). In all cases, we set
+\(t_\text{elem} = 1~\mu \text{m}\). Table S1 is a glossary model terms
+and symbols.
 
@@
 surface and the next stomate on the other surface (Fig. S1). The rest of
-the mesophyll will be the same because of symmetry.
+the mesophyll will be the same because of symmetry. This allowed us to
+set the boundary fluxes on the left and right sides of the leaf profile
+to 0.
+
+\hypertarget{solving-within-leaf-gradients-in-co_2-assimilation-and-concentration}{%
+\subsection{\texorpdfstring{Solving within-leaf gradients in CO\(_2\)
+assimilation and
+concentration}{Solving within-leaf gradients in CO\_2 assimilation and concentration}}\label{solving-within-leaf-gradients-in-co_2-assimilation-and-concentration}}
+
+We extended the 1-D FEM of Earles et al. (2017) to solve a set of
+partial differential equations describing CO\(_2\) diffusion,
+photosynthesis, and respiration throughout a 2-D leaf geometry. The
+diffusive flux of CO\(_2\) through ab- and adaxial stoamta,
+intercellular airspace, and mesophyll cells in the \(x\) (length) and
+\(z\) (depth) dimensions is:
+
+\begin{equation}\protect\hypertarget{eq-2d_pm_flux}{}{D_\text{e} \nabla^2 C_\text{ias} = D_\text{e} \bigg(\frac{\partial^2 C_\text{ias}}{\partial x^2} + \frac{\partial^2 C_\text{ias}}{\partial z^2}\bigg) = -f_\text{liq},}\label{eq-2d_pm_flux}\end{equation}
+
+where
+
+\begin{equation}\protect\hypertarget{eq-1d_pm_fliq}{}{f_\text{liq} = r_\text{d} + r_\text{p} - r_\text{c}}\label{eq-1d_pm_fliq}\end{equation}
+
+and
+
+\begin{equation}\protect\hypertarget{eq-De}{}{
+D_\text{e} = \frac{\varphi}{\tau} D_\text{c}
+}\label{eq-De}\end{equation}
+
+is the effective diffusivity of CO\(_2\) though a porous medium composed
+of an intercellular airspace with a porosity (\(\varphi\); m\(^3\)
+airspace m\(^{-3}\) leaf) and tortuosity (\(\tau\); m m\(^{-1}\)). We
+assume the palisade (\(\varphi_\text{pal}\)) is less porous than the
+spongy (\(\varphi_\text{spg}\)) mesophyll (Table S1). \(D_\text{c}\) is
+the diffusion coefficient (m s\(^{-1}\)) for CO\(_2\) in the
+intercellular airspace, \(C_\text{ias}\) is the {[}CO\(_2\){]} (mol
+m\(^{-3}\)) at horizontal positions \(x\) and depth \(z\) in the
+intercellular airspace, \(f_\text{liq}\) is the volumetric rate of
+CO\(_2\) diffusion from the intercellular airspace into the chloroplast
+stroma (mol m\(^{-3}\) s\(^{-1}\)), \(r_\text{c}\) is the volumetric
+rate of ribulose 1,5-bisphosphate (RuBP) carboxylation (mol m\(^{-3}\)
+s\(^{-1}\)), \(r_\text{d}\) is the volumetric respiration rate (mol
+m\(^{-3}\) s\(^{-1}\)), and \(r_\text{p}\) is the volumetric
+photorespiration rate by Rubisco (mol m\(^{-3}\) s\(^{-1}\)). Following
+Earles et al. (2017), \(r_\text{d}\) is assumed constant per stroma
+surface area (Table S1) and \(r_\text{p}\) is a function of
+carboxylation (\(r_\text{c}\)) and \(C_\text{liq}\):
+
+\begin{equation}\protect\hypertarget{eq-rp}{}{ r_\text{p} = r_\text{c} \frac{\Gamma^*}{C_\text{liq}}. }\label{eq-rp}\end{equation}
+
+Carboxylation rate is the minimum of the Rubisco-limited
+(\(w_\text{c}\)) or RuBP-regeneration limited (\(w_\text{j}\))
+carboxylation rates:
+
+\begin{equation}\protect\hypertarget{eq-rc}{}{ r_\text{c} = \text{min}(w_\text{c}, w_\text{j}) \frac{S_\text{m}}{T_\text{leaf}} V_\text{strom}, }\label{eq-rc}\end{equation}
+
+where
+
+\begin{equation}\protect\hypertarget{eq-wc}{}{ w_\text{c} = \frac{k_\text{c} X_\text{c} C_\text{liq}}{K_\text{m} + C_\text{liq}},~\text{and}}\label{eq-wc}\end{equation}
+
+\begin{equation}\protect\hypertarget{eq-wj}{}{ w_\text{j} = \frac{C_\text{liq} j_\text{e}}{4 C_\text{liq} + 8 \Gamma^*}.}\label{eq-wj}\end{equation}
+
+Multiplying by \(\frac{S_\text{m}}{T_\text{leaf}} V_\text{strom}\)
+converts carboxylation from per area to per stroma volume units.
+\(k_\text{c}\) is the catalytic rate of Rubisco (m\(^{-1}\)) and
+\(K_\text{m}\) is effective Michaelis-Menten constant for Rubisco (mol
+m\(^{-3}\)). Following Earles et al. (2017), we assumed the relative
+concentration of Rubisco follows that of Nishio, Sun, and Vogelmann
+(1993), but scaled such that the bulk leaf Rubisco concentration
+integrates to \(X_\text{c}\) described in Table S1. We estimated a
+continuous function describing the relative Rubisco profile as a
+function of leaf depth using a generalized additive model with the
+\texttt{gam()} function in \emph{R} package \textbf{mgcv} version 1.9.0
+(Wood 2017).
+
+The effective photosynthetic e\(^{-}\) transport rate (\(j_\text{e}\))
+is the minimum of the maximum (\(j_\text{max}\)) and potential
+(\(j_\infty\)) photosynthetic e\(^{-}\) transport rates at each position
+within the mesophyll:
+
+\begin{equation}\protect\hypertarget{eq-je}{}{ j_\text{e} = \text{min}(j_\text{max}, j_\infty). }\label{eq-je}\end{equation}
+
+The local \(j_\text{max}\) follows the same depth profile as Rubisco and
+is scaled by local \(S_\text{m}\) and \(V_\text{strom}\) so that it
+integrates to \(J_\text{max}\) on a leaf-area basis (Earles et al.
+2017):
+
+\begin{equation}\protect\hypertarget{eq-jmax}{}{J_\text{max} = \int_0^{T_\text{leaf}} j_{\text{max},i} S_{\text{m},i} V_\text{strom} dz.}\label{eq-jmax}\end{equation}
+
+Potential e\(^{-}\) transport follows the same depth profile as
+chlorophyll concentration and is scaled by local \(S_\text{m}\) and
+\(V_\text{strom}\) so that it integrates to \(J_\infty\) on a leaf-area
+basis (Earles et al. 2017), where:
+
+\begin{equation}\protect\hypertarget{eq-Jinf}{}{ J_\infty = I_0 \alpha \beta \phi_\text{PSII} = \int_0^{T_\text{leaf}} j_{\infty,i} S_{\text{m},i} V_\text{strom} dz.}\label{eq-Jinf}\end{equation}
+
+Potential e\(^{-}\) transport is a product of PPFD incident on the leaf
+surface (\(I_0\), mol m\(^{-2}\) s\(^{-1}\)), whole-leaf light
+absorption (\(\alpha\), mol mol\(^{-1}\)), the fraction of light
+absorbed by PSII (\(\beta\), mol mol\(^{-1}\)), and the quantum yield of
+PSII e\(^{-}\) transport (\(\phi_\text{PSII}\), mol mol\(^{-1}\)).
+
+The local chlorophyll concentration is a function of depth described by
+a quadratic equation (Johnson et al. 2005):
+
+\begin{equation}\protect\hypertarget{eq-Fchl}{}{F_\text{chl} = f(z) = b_{0,\text{chl}} + b_{1,\text{chl}} z + b_{2,\text{chl}} z ^ 2.}\label{eq-Fchl}\end{equation}
+
+We used generic parameters from Borsuk and Brodersen (2019), rescaled to
+relative depth on a 0-1 rather than 0-100 interval.
+
+For mass balance, the CO\(_2\) assimilated via carboxylation must be
+supplied by diffusive flux from the intercellular airspace into the
+chloroplast stroma. The volumetric rate of CO\(_2\) diffusion from the
+intercellular airspace into the chloroplast stroma, \(f_\text{liq}\),
+is:
+
+\begin{equation}\protect\hypertarget{eq-fliq2}{}{f_\text{liq} = \frac{g_\text{liq} (C_\text{liq} - C_\text{ias})}{T_\text{leaf} / S_\text{m}},}\label{eq-fliq2}\end{equation}
+
+where \(g_\text{liq}\) is the CO\(_2\) conductance from the
+intercellular airspace into the chloroplast stroma (m s\(^{-1}\)),
+\(C_\text{liq}\) (mol m\(^{-3}\)) is the {[}CO\(_2\){]} in the stroma,
+and \(S_\text{m}\) is mesophyll surface area-to-leaf surface area ratio.
+Noting that \(g_\text{liq}\) is conductance per m\(^2\) of stroma, this
+means the length scale to divide by should be the inverse of stroma area
+per unit bulk leaf volume,
+i.e.~\(1/[S_\text{c} (1 / T_\text{leaf})] = T_\text{leaf} / S_\text{c}\).
+For simplicity, we assume that the entire mesophyll surface area is
+lined with chloroplasts, hence \(S_\text{m} = S_\text{c}\). Following
+Earles et al. (2017), we assumed greater surface area in the palisade
+(\(S_\text{m,pal}\)) than spongy (\(S_\text{m,spg}\)) mesophyll. The
+fractions of palisade and spongy mesophyll are \(f_\text{pal}\) and
+\(f_\text{spg}\), respectively.
+
+Following Earles et al. (2017), we set boundary conditions at the
+stomata, \(C_\text{stom}\), to be \(0.85 \times\) atmospheric CO\(_2\)
+(Table S1). The fluxes on the left and right sides are 0 because of
+symmetry; the other fluxes on the ab- and adaxial surface are assumed 0
+(i.e.~the epidermis is impermeable to CO\(_2\)).
+
+\begin{figure}
+
+{\centering \includegraphics{figures/ph2d-example.pdf}
+
+}
+
+\caption{Example profiles of volumetric CO\(_2\) concentrations within
+otherwise identical amphistomatous leaves that have stomatal positions
+offset (top row) or aligned (bottom row) based on the 2-D porous medium
+model. Stomatal positions are indicated by black points at the top and
+bottom of panels. When stomata are aligned, both ab- and adaxial stomata
+are position 0 along the \(x\)-axis; when stomata are offset, the
+adaxial stomate is positioned \(U/2\) distance away. In this example,
+variables are set as: \(I_0 = 1000\) \(\mu\)mol m\(^{-2}\) s\(^{-1}\);
+\(\varphi_\text{pal} = 0.2\) m\(^3\) airspace m\(^{-3}\) leaf;
+\(T_\text{leaf} = 200\) \(\mu\)m; \(U = 200\) \(\mu\)m. All other
+parameter values are described in Table S1. \(C_\text{ias}\) =
+{[}CO\(_2\){]} in intercellular airspace; \(C_\text{liq}\) =
+{[}CO\(_2\){]} in chloroplast stroma; \(I_0\) = PPFD incident on the
+leaf surface; \(\varphi_\text{pal}\) = Fraction of intercellular
+airspace (aka porosity), palisade; \(T_\text{leaf}\) = Leaf thickness;
+\(U\) = Interstomatal distance.}
+
+\end{figure}
 
@@
 \(67.5\) & \(-\) & Borsuk and Brodersen (2019);
-\(f_{z,\text{chl}} = b_{0,\text{chl}} + b_{1,\text{chl}} z + b_{2,\text{chl}} z ^ 2\) \\
+Equation~\ref{eq-Fchl} \\
 Chlorophyll spatial distribution coefficient & \(b_{1,\text{chl}}\) &
 \(41.5\) & \(-\) & Borsuk and Brodersen (2019);
-\(f_{z,\text{chl}} = b_{0,\text{chl}} + b_{1,\text{chl}} z + b_{2,\text{chl}} z ^ 2\) \\
+Equation~\ref{eq-Fchl} \\
 Chlorophyll spatial distribution coefficient & \(b_{2,\text{chl}}\) &
-\(-29\) & \(-\) & Borsuk and Brodersen (2019);
-\(f_{z,\text{chl}} = b_{0,\text{chl}} + b_{1,\text{chl}} z + b_{2,\text{chl}} z ^ 2\) \\
+\(-29\) & \(-\) & Borsuk and Brodersen (2019); Equation~\ref{eq-Fchl} \\
 Fraction of light absorbed by PSII & \(\beta\) & \(0.5\) & mol
 mol\(^{-1}\) & assumed \\
-{[}CO\(_2\){]} in intercellular airspace & \(C_\text{ias}\) & \(-\) &
-mol m\(^{-3}\) & Equation~\ref{eq-2d_pm_flux} and
+{[}CO\(_2\){]} in intercellular airspace & \(C_\text{ias}\) & calculated
+& mol m\(^{-3}\) & Equation~\ref{eq-2d_pm_flux} and
 Equation~\ref{eq-fliq2} \\
-{[}CO\(_2\){]} in chloroplast stroma & \(C_\text{liq}\) & \(-\) & mol
-m\(^{-3}\) & Equation~\ref{eq-2d_pm_flux} \\
+{[}CO\(_2\){]} in chloroplast stroma & \(C_\text{liq}\) & calculated &
+mol m\(^{-3}\) & Equation~\ref{eq-2d_pm_flux} \\
 {[}CO\(_2\){]} in substomatal cavity & \(C_\text{stom}\) &
@@
 Effective diffusivity of {[}CO\(_2\){]} in intercellular airspace &
-\(D_\text{e}\) & \(-\) & m\(^{2}\) s\(^{-1}\) & Equation~\ref{eq-De} \\
+\(D_\text{e}\) & calculated & m\(^{2}\) s\(^{-1}\) &
+Equation~\ref{eq-De} \\
 Fraction of palisade mesophyll & \(f_\text{pal}\) & \(0.6\) & 1 &
@@
 Chlorophyll fluoresence profile along leaf depth normalized by total
-fluoresence & \(F_\text{z,\text{chl}}\) & \(-\) & 1 & Borsuk and
-Brodersen (2019);
-\(f_{z,\text{chl}} = b_{0,\text{chl}} + b_{0,\text{chl}} z + b_{2,\text{chl}} z ^ 2\) \\
+fluoresence & \(F_\text{chl}\) & calculated & 1 & Borsuk and Brodersen
+(2019); Equation~\ref{eq-Fchl} \\
 Conductance of cell wall, plasmalemma, cytosol, chloroplast envelope,
@@
 m\(^3\) m\(^{-2}\) stroma s\(^{-1}\) & Evans et al. (2009) \\
-PPFD incident on the leaf surface & \(I_0\) & \(1.50 \times 10^{-3}\) &
-mol m\(^{-2}\) s\(^{-1}\) & assumed \\
+PPFD incident on the leaf surface & \(I_0\) & variable & mol m\(^{-2}\)
+s\(^{-1}\) & assumed \\
 Potential photosynthetic e\(^{-}\) transport rate on a leaf area basis &
-\(J_\infty\) & \(-\) & mol m\(^{-2}\) leaf s\(^{-1}\) &
+\(J_\infty\) & calculated & mol m\(^{-2}\) leaf s\(^{-1}\) &
 Equation~\ref{eq-Jinf} \\
@@
 Effective photosynthetic e\(^{-}\) transport rate on a stroma volume
-basis & \(j_\text{e}\) & \(-\) & mol m\(^{-3}\) stroma s\(^{-1}\) &
+basis & \(j_\text{e}\) & calculated & mol m\(^{-3}\) stroma s\(^{-1}\) &
 Equation~\ref{eq-je} \\
 Potential photosynthetic e\(^{-}\) transport rate on a stroma volume
-basis & \(j_\infty\) & \(-\) & mol m\(^{-3}\) stroma s\(^{-1}\) & see
-text \\
+basis & \(j_\infty\) & calculated & mol m\(^{-3}\) stroma s\(^{-1}\) &
+Equation~\ref{eq-Jinf} \\
 Maximum photosynthetic e\(^{-}\) transport rate on a stroma volume basis
-& \(j_\text{max}\) & \(-\) & mol m\(^{-3}\) stroma s\(^{-1}\) & see
-text \\
+& \(j_\text{max}\) & calculated & mol m\(^{-3}\) stroma s\(^{-1}\) &
+Equation~\ref{eq-jmax} \\
 Catalytic rate of Rubisco & \(k_\text{c}\) & \(2.84\) & m\(^{-1}\) &
@@
 \(1.87 \times 10^{-2}\) & mol m\(^{-3}\) & Caemmerer (2000) \\
-Number of elements in \(x\) direction & \(n_x\) & \(100\) & \(-\) &
+Number of elements in \(x\) direction & \(n_x\) & calculated & \(-\) &
 \(U = 2 n_x t_\text{elem}\) \\
-Number of elements in \(z\) direction & \(n_z\) & \(200\) & \(-\) &
+Number of elements in \(z\) direction & \(n_z\) & calculated & \(-\) &
 \(T_\text{leaf} = n_z t_\text{elem}\) \\
 Fraction of intercellular airspace (aka porosity), palisade &
-\(\varphi_\text{pal}\) & \(0.1\) & m\(^3\) airspace m\(^{-3}\) leaf &
+\(\varphi_\text{pal}\) & variable & m\(^3\) airspace m\(^{-3}\) leaf &
 assumed \\
@@
 \(0.85\) & mol mol\(^{-1}\) & assumed \\
-Volumetric rate of RuBP carboxylation & \(r_\text{c}\) & \(-\) & mol
-m\(^{-2}\) stroma s\(^{-1}\) & Equation~\ref{eq-rc} \\
+Volumetric rate of RuBP carboxylation & \(r_\text{c}\) & calculated &
+mol m\(^{-2}\) stroma s\(^{-1}\) & Equation~\ref{eq-rc} \\
 Volumetric respiration rate & \(r_\text{d}\) & \(6.60 \times 10^{-2}\) &
@@
 Volumetric rate of photorespiratory CO\(_2\) release & \(r_\text{p}\) &
-\(-\) & mol m\(^{-2}\) stroma s\(^{-1}\) & Earles et al. (2017);
+calculated & mol m\(^{-2}\) stroma s\(^{-1}\) & Earles et al. (2017);
 Equation~\ref{eq-rp} \\
-Leaf surface area-to-mesophyll surface area ratio, palisiade &
+Mesophyll surface area-to-leaf surface area ratio, palisiade &
 \(S_\text{m,pal}\) & \(20\) & m\(^2\) mesophyll m\(^{-2}\) leaf &
 assumed \\
-Leaf surface area-to-mesophyll surface area ratio spongy &
+Mesophyll surface area-to-leaf surface area ratio, spongy &
 \(S_\text{m,spg}\) & \(2\) & m\(^2\) mesophyll m\(^{-2}\) leaf &
@@
 \(T_\text{leaf} = n_z t_\text{elem}\) \\
-Leaf thickness & \(T_\text{leaf}\) & \(-\) & m &
+Leaf thickness & \(T_\text{leaf}\) & variable & m &
 \(T_\text{leaf} = n_z t_\text{elem}\) \\
-Interstomatal distance & \(U\) & \(-\) & m &
+Interstomatal distance & \(U\) & variable & m &
 \(U = n_x t_\text{elem}\) \\
@@
 et al. (2017); Tholen and Zhu (2011) \\
-Rubisco-limited carboxylation rate & \(w_\text{c}\) & \(-\) & mol
+Rubisco-limited carboxylation rate & \(w_\text{c}\) & calculated & mol
 m\(^{-2}\) stroma s\(^{-1}\) & Equation~\ref{eq-wc} \\
-RuBP regeneration-limited carboxylation rate & \(w_\text{j}\) & \(-\) &
-mol m\(^{-2}\) stroma s\(^{-1}\) & Equation~\ref{eq-wj} \\
+RuBP regeneration-limited carboxylation rate & \(w_\text{j}\) &
+calculated & mol m\(^{-2}\) stroma s\(^{-1}\) & Equation~\ref{eq-wj} \\
 Rubisco concentration in stroma & \(X_\text{c}\) & \(2.5\) & mol
@@
 \end{landscape}
-
-\hypertarget{solving-within-leaf-gradients-in-co_2-assimilation-and-concentration}{%
-\subsection{\texorpdfstring{Solving within-leaf gradients in CO\(_2\)
-assimilation and
-concentration}{Solving within-leaf gradients in CO\_2 assimilation and concentration}}\label{solving-within-leaf-gradients-in-co_2-assimilation-and-concentration}}
-
-We extended the 1-D FEM of Earles et al. (2017) to solve a set of
-partial differential equations describing CO\(_2\) diffusion,
-photosynthesis, and respiration throughout a 2-D leaf geometry. The
-diffusive flux of CO\(_2\) through ab- and adaxial stoamta,
-intercellular airspace, and mesophyll cells in the \(x\) (length) and
-\(z\) (depth) dimensions is:
-
-\begin{equation}\protect\hypertarget{eq-2d_pm_flux}{}{D_\text{e} \nabla^2 C_\text{ias} = D_\text{e} \bigg(\frac{\partial^2 C_\text{ias}}{\partial x^2} + \frac{\partial^2 C_\text{ias}}{\partial z^2}\bigg) = -f_\text{liq}}\label{eq-2d_pm_flux}\end{equation}
-
-\begin{equation}\protect\hypertarget{eq-1d_pm_fliq}{}{f_\text{liq} = r_\text{d} + r_\text{p} - r_\text{c}}\label{eq-1d_pm_fliq}\end{equation}
-
-where
-
-\begin{equation}\protect\hypertarget{eq-De}{}{
-D_\text{e} = \frac{\varphi}{\tau} D_\text{c}
-}\label{eq-De}\end{equation}
-
-is the effective diffusivity of CO\(_2\) though a porous medium composed
-of an intercellular airspace with a porosity (\(\varphi\); m\(^3\)
-airspace m\(^{-3}\) leaf) and tortuosity (\(\tau\); m m\(^{-1}\)).
-\(D_\text{c}\) is the diffusion coefficient (m s\(^{-1}\)) for CO\(_2\)
-in the intercellular airspace, \(C_\text{ias}\) is the {[}CO\(_2\){]}
-(mol m\(^{-3}\)) at horizontal positions \(x\) and depth \(z\) in the
-intercellular airspace, \(f_\text{liq}\) is the volumetric rate of
-CO\(_2\) diffusion from the intercellular airspace into the chloroplast
-stroma (mol m\(^{-3}\) s\(^{-1}\)), \(r_\text{c}\) is the volumetric
-rate of ribulose 1,5-bisphosphate (RuBP) carboxylation (mol m\(^{-3}\)
-s\(^{-1}\)), \(r_\text{d}\) is the volumetric respiration rate (mol
-m\(^{-3}\) s\(^{-1}\)), and \(r_\text{p}\) is the volumetric
-photorespiration rate by Rubisco (mol m\(^{-3}\) s\(^{-1}\)). Following
-Earles et al. (2017), \(r_\text{d}\) is assumed constant per stroma
-surface area (Table S1) and \(r_\text{p}\) is a function of
-carboxylation (\(r_\text{c}\)) and \(C_\text{liq}\):
-
-\begin{equation}\protect\hypertarget{eq-rp}{}{ r_\text{p} = r_\text{c} \frac{\Gamma^*}{C_\text{liq}} }\label{eq-rp}\end{equation}
-
-Carboxylation rate is the minimum of the Rubisco-limited
-(\(w_\text{c}\)) or RuBP-regeneration limited (\(w_\text{j}\))
-carboxylation rate:
-
-\begin{equation}\protect\hypertarget{eq-rc}{}{ r_\text{c} = \text{min}(w_\text{c}, w_\text{j}) }\label{eq-rc}\end{equation}
-
-where
-
-\begin{equation}\protect\hypertarget{eq-wc}{}{ w_\text{c} = \frac{k_\text{c} X_\text{c} C_\text{liq}}{K_\text{m} + C_\text{liq}},~\text{and}}\label{eq-wc}\end{equation}
-
-\begin{equation}\protect\hypertarget{eq-wj}{}{ w_\text{j} = \frac{C_\text{liq} j_\text{e}}{4 C_\text{liq} + 8 \Gamma^*}.}\label{eq-wj}\end{equation}
-
-\(k_\text{c}\) is the catalytic rate of Rubisco (m\(^{-1}\)) and
-\(K_\text{m}\) is effective Michaelis-Menten constant for Rubisco (mol
-m\(^{-3}\)). Following Earles et al. (2017), we assumed the relative
-concentration of Rubisco follows that of Nishio, Sun, and Vogelmann
-(1993), but scaled such that the bulk leaf Rubisco concentration
-integrates to \(X_\text{c}\) described in Table S1. We estimated a
-continuous function describing the relative Rubisco profile as a
-function of leaf depth using a generalized additive model with the
-\texttt{gam()} function in \emph{R} package \textbf{mgcv} version 1.9.0
-(Wood 2017).
-
-The effective photosynthetic e\(^{-}\) transport rate (\(j_\text{e}\))
-is the minimum of the maximum (\(j_\text{max}\)) and potential
-(\(j_\infty\)) photosynthetic e\(^{-}\) transport rates at each position
-within the mesophyll:
-
-\begin{equation}\protect\hypertarget{eq-je}{}{ j_\text{e} = \text{min}(j_\text{max}, j_\infty) }\label{eq-je}\end{equation}
-
-The local \(j_\text{max}\) follows the same profile as Rubisco as a
-function of leaf depth and is scaled so that it integrates to
-\(J_\text{max}\) on a leaf-area basis (Earles et al. 2017). Potential
-e\(^{-}\) transport is assumed proportional to the local chloroplast
-concentration so that it integrates to \(J_\infty\) on a leaf-area basis
-(Earles et al. 2017), where:
-
-\begin{equation}\protect\hypertarget{eq-Jinf}{}{ J_\infty = I_0 \alpha \beta \phi_\text{PSII}.}\label{eq-Jinf}\end{equation}
-
-Potential e\(^{-}\) transport is a product of PPFD incident on the leaf
-surface (\(I_0\), mol m\(^{-2}\) s\(^{-1}\)), whole-leaf light
-absorption (\(\alpha\), mol mol\(^{-1}\)), the fraction of light
-absorbed by PSII (\(\beta\), mol mol\(^{-1}\)), and the quantum yield of
-PSII e\(^{-}\) transport (\(\phi_\text{PSII}\), mol mol\(^{-1}\)).
-
-The local chlorophyll concentration (SYMBOL) is a function of leaf depth
-following Equation XX from Borsuk and Brodersen (2019):
-
-The volumetric rate of CO\(_2\) diffusion from the intercellular
-airspace into the chloroplast stroma, \(f_\text{liq}\), is:
-
-\begin{equation}\protect\hypertarget{eq-fliq2}{}{f_\text{liq} = \frac{g_\text{liq} (C_\text{liq} - C_\text{ias})}{T_\text{leaf} / S_\text{m}}}\label{eq-fliq2}\end{equation}
-
-where \(g_\text{liq}\) is the CO\(_2\) conductance from the
-intercellular airspace into the chloroplast stroma (m s\(^{-1}\)),
-\(C_\text{liq}\) (mol m\(^{-3}\)) is the {[}CO\(_2\){]} in the stroma,
-and \(S_\text{m}\) is leaf surface area-to-mesophyll surface area ratio.
-Noting that \(g_\text{liq}\) is conductance per m\(^2\) of stroma, this
-means the length scale to divide by should be the inverse of stroma area
-per unit bulk leaf volume,
-i.e.~\(1/[S_\text{c} (1 / T_\text{leaf})] = T_\text{leaf} / S_\text{c}\).
-For simplicity, we assume that the entire mesophyll surface area is
-lined with chloroplasts, hence \(S_\text{m} = S_\text{c}\).
-
-I calculated assimilation and respiration the same way as Earles et al.
-(2017) using the standard C\$\_3\$ biochemical model.
-
-The boundary conditions are that the CO\(_2\) concentration in the
-substomatal cavity is constant at \(C_\text{stom}\). The fluxes on the
-left and right sides are 0 because of symmetry.
-
-extra words
-
-\begin{figure}
-
-{\centering \includegraphics{figures/ph2d-example.pdf}
-
-}
-
-\caption{Example profiles of volumetric CO\(_2\) concentrations within
-otherwise identical amphistomatous leaves that have stomatal positions
-offset (top row) or aligned (bottom row) based on the 2-D porous medium
-model. Stomatal positions are indicated by black points at the top and
-bottom of panels. When stomata are aligned, both ab- and adaxial stomata
-are position 0 along the \(x\)-axis; when stomata are offset, the
-adaxial stomate is positioned \(U/2\) distance away. In this example,
-variables are set as: \(I_0 = 1000\) \(\mu\)mol m\(^{-2}\) s\(^{-1}\);
-\(\varphi_\text{pal} = 0.2\) m\(^3\) airspace m\(^{-3}\) leaf;
-\(T_\text{leaf} = 200\) \(\mu\)m; \(U = 200\) \(\mu\)m. All other
-parameter values are described in Table S1. \(C_\text{ias}\) =
-{[}CO\(_2\){]} in intercellular airspace; \(C_\text{liq}\) =
-{[}CO\(_2\){]} in chloroplast stroma; \(I_0\) = PPFD incident on the
-leaf surface; \(\varphi_\text{pal}\) = Fraction of intercellular
-airspace (aka porosity), palisade; \(T_\text{leaf}\) = Leaf thickness;
-\(U\) = Interstomatal distance.}
-
-\end{figure}
 
@@
 \url{https://doi.org/10.1093/jxb/erp117}.
+
+\leavevmode\vadjust pre{\hypertarget{ref-johnson_leaf_2005}{}}%
+Johnson, Daniel M., William K. Smith, Thomas C. Vogelmann, and Craig R.
+Brodersen. 2005. {``Leaf Architecture and Direction of Incident Light
+Influence Mesophyll Fluorescence Profiles.''} \emph{American Journal of
+Botany} 92 (9): 1425--31. \url{https://doi.org/10.3732/ajb.92.9.1425}.
 

### 2023-11-13 ms/ms.Rmd
small revisions to ms before submission

@@
 ---
-title: Does stomatal patterning in amphistomatous leaves minimize the CO$_2$ diffusion path length with leaves of *Arabidopsis thaliana*?
+title: Does stomatal patterning in amphistomatous leaves minimize the CO$_2$ diffusion path length within leaves?
 date: "`r Sys.Date()`"
@@
   - "Photosynthesis is co-limited by multiple factors depending on the plant and its environment.
-These include biochemical rate limitations, internal and external water potentials, temperature, irradiance, and carbon dioxide (CO2).
+These include biochemical rate limitations, internal and external water potentials, temperature, irradiance, and carbon dioxide (CO$_2$).
 Amphistomatous leaves have stomata on both abaxial and adaxial leaf surfaces.
-This feature is considered an adaptation to alleviate CO2 diffusion limitations in productive environments where other factors are not limiting as the diffusion path length from stomate to chloroplast is effectively halved.
-Plants can also reduce CO2 limitations through other aspects of optimal stomatal anatomy: stomatal density, distribution and patterning, and size.
-A number of studies have demonstrated that stomata are overdispersed on a single leaf surface; however, much less is known about stomatal anatomy in amphistomatous plants, especially the coordination between leaf surfaces, despite their prevelance in nature and near ubiquity among crop species.
-Here we use novel spatial statistics based on simulations and photosynthesis modelling to test hypotheses about how amphistomatous plants may optimize CO2 limitations in the model angiosperm Arabidopsis thaliana grown in different light environments.
-We find that 1) stomata are overdispersed, but not ideally dispersed, on both leaf surfaces across all light treatments; 2) abaxial and adaxial leaf surfaces are independent; and 3) the theoretical improvements to photosynthesis from abaxial-adaxial stomatal coordination are miniscule (~1%) across the range of feasible traitspace.
-However, we also find that 4) stomatal size is correlated with the mesophyll volume that it supplies with CO2, suggesting that plants may optimize CO2 diffusion limitations through alternative pathways other than ideal, uniform stomatal spacing.
-We discuss the myriad developmental, physical, and evolutionary constraits which may prohibit plants from reaching the theoretical adaptive peak of uniform stomatal spacing and inter-surface stomatal coordination.
-These findings contribute to our understanding of amphistomatous plants' exploration of the leaf anatomy adaptive landscape."
+This feature is considered an adaptation to alleviate CO$_2$ diffusion limitations in productive environments where other factors are not limiting as the diffusion path length from stomate to chloroplast is effectively halved.
+Plants can also reduce CO$_2$ limitations through other aspects of optimal stomatal anatomy: stomatal density, distribution, patterning, and size.
+A number of studies have demonstrated that stomata are overdispersed on a single leaf surface; however, much less is known about stomatal anatomy in amphistomatous leaves, especially the coordination between leaf surfaces, despite their prevelance in nature and near ubiquity among crop species.
+Here we use novel spatial statistics based on simulations and photosynthesis modeling to test hypotheses about how amphistomatous plants may optimize CO$_2$ limitations in the model angiosperm *Arabidopsis thaliana* grown in different light environments.
+We find that 1) stomata are overdispersed, but not ideally dispersed, on both leaf surfaces across all light treatments; 2) abaxial and adaxial leaf surface patterning are independent; and 3) the theoretical improvements to photosynthesis from abaxial-adaxial stomatal coordination are miniscule ($\\ll 1$%) across the range of feasible parameter space.
+However, we also find that 4) stomatal size is correlated with the mesophyll volume that it supplies with CO$_2$, suggesting that plants may optimize CO$_2$ diffusion limitations through alternative pathways other than ideal, uniform stomatal spacing.
+We discuss the developmental, physical, and evolutionary constraits which may prohibit plants from reaching the theoretical adaptive peak of uniform stomatal spacing and inter-surface stomatal coordination.
+These findings contribute to our understanding of variation in the anatomy of amphistomatous leaves."
 keywords:
@@
  - \usepackage{setspace}
- - \doublespacing
+ - \onehalfspacing
  - \usepackage[nomarkers,tablesfirst]{endfloat} # For figures and tables at end
@@
 However, reducing this distance to its absolute minimum may be constrained by developmental processes, or the photosynthetic benefit may be too small to be 'seen' by natural selection (i.e.
-the selection coefficient is less than drift barrier [@sung_drift-barrier_2012]).
+the selection coefficient is less than drift barrier *sensu* @sung_drift-barrier_2012).
 
@@
 No developmental pathway exists to ensure an idealized placement of stomata on the leaf surface.
-Rather, stomatal development is a dynamic process that must be plastic to environmental ques.
+Rather, stomatal development is a dynamic process that must be plastic to environmental cues.
 Leaves develop based on short and long distance signalling pathways which relay information about incoming light, humidity, temperature, and surrounding stomata to developing leaf tissues [@pillitteri_mechanisms_2012].
@@
 As described above, the positive correlation between stomatal size and zone may be explained by common developmental processes rather than as an adaptation to maximize CO$_2$ diffusion.
-Such flexibility in mature stomate size is an alternate peak in the adaptive landscape which cannot be assessed with our simplistic model assumptions.
 In any case, there is no evidence for coordinated development of both leaf surfaces, and very little theoretical benefit to photosynthesis, except in marginal circumstances which are exceptionally rare in nature.
@@
 Instead, as this study suggests, plants may simply compensate with higher stomatal density and by fitting stomatal size to the area that they supply with CO$_2$.
-To understand why stomata are not ideally dispersed, more modelling (with a more complex set of assumptions including vein density and IAS structure) should be done to estimate the fitness gain of stomatal dispersion.
+To understand why stomata are not ideally dispersed, more modeling (with a more complex set of assumptions including vein density and IAS structure) should be done to estimate the fitness gain of stomatal dispersion.
 Additionally, genetic manipulation studies should attempt to create mutants with clustered and ideally dispersed stomata for a comparison of their photosynthetic traits.
@@
 The possibility that ideal stomatal spacing is not the "tallest" fitness peak must also be explored, as stomate size is demonstrated in this study to covary with mesophyll volume supplied with CO$_2$.
-This may be especially true in highly variable environments or in large tree species with sun and shade leaves where developmental ques may change rapidly.
+This may be especially true in highly variable environments or in large tree species with sun and shade leaves where developmental cues may change rapidly.
 The temporal component, not considered here, could also have significant implications, as CO$_2$ may only be limiting to photosynthesis during short periods when all other conditions are ideal.
 In these cases, the theoretical benefits of ideal stomatal spacing are further diminished.
-Future exploration of these competing hypotheses would require more advanced modelling, additional exploration of IAS space development and its effects on gas exchange, both real and modeled, and knowledge about how often the species of interest is CO$_2$ limited across of range of natural settings.
-Despite these additional considerations, this paper represents an important contribution to understanding the potential drivers of and limitations to stomatal anatomy in amphistomatous plants.
+Future exploration of these competing hypotheses would require more advanced modeling, additional exploration of IAS space development and its effects on gas exchange, both real and modeled, and knowledge about how often the species of interest is CO$_2$ limited across of range of natural settings.
+Despite these additional considerations, this study represents an important contribution to understanding the potential drivers of and limitations to stomatal anatomy in amphistomatous plants.
 

### 2023-11-13 ms/ms.tex
small revisions to ms before submission

@@
 
+\usepackage{newunicodechar,graphicx}
+\DeclareRobustCommand{\okina}{\raisebox{\dimexpr\fontcharht\font`A-\height}{\scalebox{0.8}{`}}}
+\newunicodechar{ʻ}{\okina}
 % one column
@@
 \hypersetup{
-  pdftitle={Does stomatal patterning in amphistomatous leaves minimize the CO\_2 diffusion path length with leaves of Arabidopsis thaliana?},
+  pdftitle={Does stomatal patterning in amphistomatous leaves minimize the CO\_2 diffusion path length within leaves?},
   pdfkeywords={amphistomy, Arabidopsis thaliana, CO\(_2\)
@@
 \usepackage{setspace}
-\doublespacing
+\onehalfspacing
 \usepackage[nomarkers,tablesfirst]{endfloat}
@@
 \usepackage{caption}
-\usepackage{newunicodechar,graphicx}
-\DeclareRobustCommand{\okina}{\raisebox{\dimexpr\fontcharht\font`A-\height}{\scalebox{0.8}{`}}}
-\newunicodechar{ʻ}{\okina}
 % Pandoc header-include feature
@@
 \title[]{Does stomatal patterning in amphistomatous leaves minimize the
-CO\(_2\) diffusion path length with leaves of \emph{Arabidopsis
-thaliana}?}
+CO\(_2\) diffusion path length within leaves?}
 
@@
 internal and external water potentials, temperature, irradiance, and
-carbon dioxide (CO2). Amphistomatous leaves have stomata on both abaxial
-and adaxial leaf surfaces. This feature is considered an adaptation to
-alleviate CO2 diffusion limitations in productive environments where
-other factors are not limiting as the diffusion path length from stomate
-to chloroplast is effectively halved. Plants can also reduce CO2
-limitations through other aspects of optimal stomatal anatomy: stomatal
-density, distribution and patterning, and size. A number of studies have
-demonstrated that stomata are overdispersed on a single leaf surface;
-however, much less is known about stomatal anatomy in amphistomatous
-plants, especially the coordination between leaf surfaces, despite their
-prevelance in nature and near ubiquity among crop species. Here we use
-novel spatial statistics based on simulations and photosynthesis
-modelling to test hypotheses about how amphistomatous plants may
-optimize CO2 limitations in the model angiosperm Arabidopsis thaliana
-grown in different light environments. We find that 1) stomata are
-overdispersed, but not ideally dispersed, on both leaf surfaces across
-all light treatments; 2) abaxial and adaxial leaf surfaces are
-independent; and 3) the theoretical improvements to photosynthesis from
-abaxial-adaxial stomatal coordination are miniscule (\textasciitilde1\%)
-across the range of feasible traitspace. However, we also find that 4)
-stomatal size is correlated with the mesophyll volume that it supplies
-with CO2, suggesting that plants may optimize CO2 diffusion limitations
+carbon dioxide (CO\(_2\)). Amphistomatous leaves have stomata on both
+abaxial and adaxial leaf surfaces. This feature is considered an
+adaptation to alleviate CO\(_2\) diffusion limitations in productive
+environments where other factors are not limiting as the diffusion path
+length from stomate to chloroplast is effectively halved. Plants can
+also reduce CO\(_2\) limitations through other aspects of optimal
+stomatal anatomy: stomatal density, distribution, patterning, and size.
+A number of studies have demonstrated that stomata are overdispersed on
+a single leaf surface; however, much less is known about stomatal
+anatomy in amphistomatous leaves, especially the coordination between
+leaf surfaces, despite their prevelance in nature and near ubiquity
+among crop species. Here we use novel spatial statistics based on
+simulations and photosynthesis modeling to test hypotheses about how
+amphistomatous plants may optimize CO\(_2\) limitations in the model
+angiosperm \emph{Arabidopsis thaliana} grown in different light
+environments. We find that 1) stomata are overdispersed, but not ideally
+dispersed, on both leaf surfaces across all light treatments; 2) abaxial
+and adaxial leaf surface patterning are independent; and 3) the
+theoretical improvements to photosynthesis from abaxial-adaxial stomatal
+coordination are miniscule (\(\ll 1\)\%) across the range of feasible
+parameter space. However, we also find that 4) stomatal size is
+correlated with the mesophyll volume that it supplies with CO\(_2\),
+suggesting that plants may optimize CO\(_2\) diffusion limitations
 through alternative pathways other than ideal, uniform stomatal spacing.
-We discuss the myriad developmental, physical, and evolutionary
-constraits which may prohibit plants from reaching the theoretical
-adaptive peak of uniform stomatal spacing and inter-surface stomatal
-coordination. These findings contribute to our understanding of
-amphistomatous plants' exploration of the leaf anatomy adaptive
-landscape.}
+We discuss the developmental, physical, and evolutionary constraits
+which may prohibit plants from reaching the theoretical adaptive peak of
+uniform stomatal spacing and inter-surface stomatal coordination. These
+findings contribute to our understanding of variation in the anatomy of
+amphistomatous leaves.}
 
@@
 to be `seen' by natural selection (i.e.~the selection coefficient is
-less than drift barrier \citep{sung_drift-barrier_2012}).
+less than drift barrier \emph{sensu} \citet{sung_drift-barrier_2012}).
 
@@
 stomata on the leaf surface. Rather, stomatal development is a dynamic
-process that must be plastic to environmental ques. Leaves develop based
+process that must be plastic to environmental cues. Leaves develop based
 on short and long distance signalling pathways which relay information
@@
 processes rather than as an adaptation to maximize CO\(_2\) diffusion.
-Such flexibility in mature stomate size is an alternate peak in the
-adaptive landscape which cannot be assessed with our simplistic model
-assumptions. In any case, there is no evidence for coordinated
-development of both leaf surfaces, and very little theoretical benefit
-to photosynthesis, except in marginal circumstances which are
-exceptionally rare in nature.
+In any case, there is no evidence for coordinated development of both
+leaf surfaces, and very little theoretical benefit to photosynthesis,
+except in marginal circumstances which are exceptionally rare in nature.
 
@@
 size to the area that they supply with CO\(_2\). To understand why
-stomata are not ideally dispersed, more modelling (with a more complex
+stomata are not ideally dispersed, more modeling (with a more complex
 set of assumptions including vein density and IAS structure) should be
@@
 environments or in large tree species with sun and shade leaves where
-developmental ques may change rapidly. The temporal component, not
+developmental cues may change rapidly. The temporal component, not
 considered here, could also have significant implications, as CO\(_2\)
@@
 ideal stomatal spacing are further diminished. Future exploration of
-these competing hypotheses would require more advanced modelling,
+these competing hypotheses would require more advanced modeling,
 additional exploration of IAS space development and its effects on gas
@@
 species of interest is CO\(_2\) limited across of range of natural
-settings. Despite these additional considerations, this paper represents
+settings. Despite these additional considerations, this study represents
 an important contribution to understanding the potential drivers of and

### 2023-11-13 ms/test.Rmd
small revisions to ms before submission

--- 2023-09-19 ms/test.Rmd
@@
 ---
-title: "test"
+title: Does stomatal patterning in amphistomatous leaves minimize the CO$_2$ diffusion path length within leaves?
+date: "`r Sys.Date()`"
+author:
+  - name: Jacob L. Watts
+    email: Jacob.Watts-1@colorado.edu
+    corresponding_author: yes # Footnote using corresponding_author
+    affiliation: 
+      - UHM
+      - CU
+  - name: Graham J. Dow
+    email: graham.dow@niab.com
+    affiliation: NIAB
+  - name: Thomas N. Buckley
+    email: tnbuckley@ucdavis.edu
+    affiliation: UCD
+  - name: Christopher D. Muir
+    email: cdmuir@wisc.edu
+    affiliation: 
+      - UHM
+      - UWM
+    # footnote: 1 # Footnote using reference
+address:
+  - code: UHM
+    address: School of Life Sciences, University of Hawaiʻi at Mānoa, Honolulu, HI 96822
+  - code: CU
+    address: Ecology and Evolutionary Biology, University of Colorado, Boulder, CO 80309
+  - code: NIAB
+    address: Department of Crop Science and Production Systems, NIAB, Cambridge, CB3 0LE, UK
+  - code: UCD
+    address: Department of Plant Sciences, University of California, Davis, CA 95616
+  - code: UWM
+    address: Department of Botany, University of Wisconsin, Madison, WI 53706
+# footnote:
+#   - code: 1
+#     text: "Current email address: \\href{mailto:cdmuir@wisc.edu}{cdmuir@wisc.edu}"
+abstract:
+  - "Photosynthesis is co-limited by multiple factors depending on the plant and its environment.
+These include biochemical rate limitations, internal and external water potentials, temperature, irradiance, and carbon dioxide (CO$_2$).
+Amphistomatous leaves have stomata on both abaxial and adaxial leaf surfaces.
+This feature is considered an adaptation to alleviate CO$_2$ diffusion limitations in productive environments where other factors are not limiting as the diffusion path length from stomate to chloroplast is effectively halved.
+Plants can also reduce CO$_2$ limitations through other aspects of optimal stomatal anatomy: stomatal density, distribution, patterning, and size.
+A number of studies have demonstrated that stomata are overdispersed on a single leaf surface; however, much less is known about stomatal anatomy in amphistomatous leaves, especially the coordination between leaf surfaces, despite their prevelance in nature and near ubiquity among crop species.
+Here we use novel spatial statistics based on simulations and photosynthesis modeling to test hypotheses about how amphistomatous plants may optimize CO$_2$ limitations in the model angiosperm *Arabidopsis thaliana* grown in different light environments.
+We find that 1) stomata are overdispersed, but not ideally dispersed, on both leaf surfaces across all light treatments; 2) abaxial and adaxial leaf surface patterning are independent; and 3) the theoretical improvements to photosynthesis from abaxial-adaxial stomatal coordination are miniscule ($\\ll 1$%) across the range of feasible parameter space.
+However, we also find that 4) stomatal size is correlated with the mesophyll volume that it supplies with CO$_2$, suggesting that plants may optimize CO$_2$ diffusion limitations through alternative pathways other than ideal, uniform stomatal spacing.
+We discuss the developmental, physical, and evolutionary constraits which may prohibit plants from reaching the theoretical adaptive peak of uniform stomatal spacing and inter-surface stomatal coordination.
+These findings contribute to our understanding of variation in the anatomy of amphistomatous leaves."
+keywords:
+  - amphistomy
+  - Arabidopsis thaliana
+  - CO$_2$ diffusion
+  - finite element method
+  - optimality
+  - photosynthesis
+  - stomata
+fontsize: 12pt
+competing_interests: |
+  The authors declare no competing interests.
+author_contributions: |
+  JLW and CDM conceived of the project, analyzed data, and wrote the manuscript. GJD provided data. TNB contributed to model development and helped edit the manuscript.
+acknowledgements: |
+  This work was supported by NSF Award 1929167 to CDM and NSF Award 2307341 to TNB.
+bibliography: stomata-spacing.bib
+link-citations: yes
 output: 
-  pdf_document: default
-  html_document: default
+  rticles::oup_article:
+    latex_engine: xelatex
+    oup_version: 1 # 1 = 2020 CTAN OUP CLS package; 0 = 2009 OUP CLS package
+    # journal: "AoB PLANTS"
+    document_style: "modern" # Can be contemporary, modern, traditional
+    papersize: "large" # Can be large, medium, small
+    #citation_package: "default" # Uncomment when using a CSL; default "natbib"
+    namedate: TRUE # Set FALSE to use numeric refs; Default FALSE
+    number_sections: FALSE # Uncomment to not number sections; default TRUE
+    number_lines: TRUE # Use `lineno` package - Default FALSE
+    #number_lines_options: ["mathlines","switch"]  # Options for latex lineno package.
+    onecolumn: TRUE # Uncomment for one column format; default FALSE
+    extra_dependencies:
+      - booktabs # to use with knitr::kable() example below
+    #fontsize: 12pt
+header-includes:
+ - \usepackage{setspace}
+ - \onehalfspacing
+ - \usepackage[nomarkers,tablesfirst]{endfloat} # For figures and tables at end
+ - \usepackage{hyperref} # For autoref
+ - \renewcommand{\figureautorefname}{Fig.}
+ - \usepackage[detect-none]{siunitx} # for \numrange{}{}
+ - \sisetup{range-phrase = \text{--}}
+ - \usepackage{caption} # for \captionsetup{}
+ - \usepackage{newunicodechar,graphicx}
+ - \DeclareRobustCommand{\okina}{\raisebox{\dimexpr\fontcharht\font`A-\height}{\scalebox{0.8}{`}}}
+ - \newunicodechar{ʻ}{\okina}
 ---
@@
 ```{r setup, include=FALSE}
-knitr::opts_chunk$set(echo = TRUE)
+knitr::opts_chunk$set(
+  echo = FALSE, 
+  message = FALSE,
+  fig.pos = 'th',
+  out.width = '100%', dpi = 300,
+  fig.env = "figure"
+)
+knitr::opts_knit$set(root.dir = "..")
 ```
 
-```{r dispersion, fig.cap = "Stomata are more dispersed than expected under the null model of uniform random position (dispersion index = 0) but far from a distribution that maximizes distance between stomata (dispersion index = 1)", echo = TRUE}
+# Introduction
 
-knitr::include_graphics("figures/dispersion.pdf")
-
-```
+Stomatal anatomy (e.g.
+size, density, distribution, and patterning) and movement regulate gas exchange during photosynthesis, namely CO$_2$ assimilation and water loss through transpiration.
+Since waxy cuticles are mostly impermeable to CO$_2$ and H$_2$O, stomata are the primary entry and exit points through which gas exchange occurs despite making up a small percentage of the leaf area [@lange_responses_1971].
+Stomata consist of two guard cells which open and close upon changes in turgor pressure or hormonal cues [@mcadam_linking_2016].
+The stomatal pore leads to an internal space known as the substomatal cavity where gases contact the mesophyll.
+Once in the mesophyll, CO$_2$ diffuses throughout a network of intercellular air space (IAS) and into mesophyll cells where CO$_2$ assimilation ($A$) occurs within the chloroplasts [@lee_diffusion_1964].
+Stomatal conductance and transpiration are determined by numerous environmental and anatomical parameters such as vapor pressure deficit (VPD), irradiance, temperature, wind speed, leaf water potential, IAS geometry, mesophyll cell anatomy, and stomatal anatomy.

### 2023-11-13 ms/test.tex
small revisions to ms before submission

--- 2023-11-13 ms/ms.tex
@@
 
-\usepackage{newunicodechar,graphicx}
-\DeclareRobustCommand{\okina}{\raisebox{\dimexpr\fontcharht\font`A-\height}{\scalebox{0.8}{`}}}
-\newunicodechar{ʻ}{\okina}
 % one column
@@
 
-Many successful predictions about stomata and other C3 leaf traits can
-be made by hypothesizing that natural selection should optimize CO\(_2\)
-gain per unit of water loss for any given set of environmental
-parameters, including their variability
-\citep{cowan_stomatal_1977, buckley_optimal_2017, sperry_predicting_2017}.
-Total stomatal area (size \(\times\) density) is optimized for
-operational conductance (\(g_\text{s,op}\)) rather than maximum
-conductance (\(g_\text{s,max}\)) such that stomatal apertures are most
-responsive to changes in the environment at their operational aperture
-\citep{franks_physiological_2012, liu_scaling_2021}. Stomatal aperture
-can compensate for maladaptive stomatal densities to an extent
-\citep{bussis_stomatal_2006}, but stomatal density and size ultimately
-determine a leaf's theoretical \(g_\text{s,max}\)
-\citep{sack_developmental_2016}, which is proportional to
-\(g_\text{s,op}\) \citep{murray_consistent_2020}. Additionally, low
-stomatal densities lead to irregular and insufficient CO\(_2\) supply
-and reduced photosynthetic efficiency in leaf areas far from stomata
-\citep{pieruschka_lateral_2006, morison_lateral_2005}, while high
-stomatal densities can reduce water use efficiency (WUE)
-\citep{bussis_stomatal_2006} and incur excessive metabolic costs
-\citep{deans_optimization_2020}. Stomatal density positively co-varies
-with irradiance during leaf development and negatively co-varies with
-CO\(_2\) concentration
-\citep{gay_influence_1975, schoch_dependence_1980, woodward_stomatal_1987, royer_stomatal_2001},
-consistent with optimality predictions. In most species, stomata occur
-only on the abaxial (usually lower) leaf surface; but amphistomy, the
-occurrence of stomata on both abaxial and adaxial leaf surfaces, is also
-prevalent in high light environments with constant or intermittent
-access to sufficient water
-\citep{mott_adaptive_1982, jordan_using_2014, muir_light_2018, drake_two_2019, muir_is_2019}.
-Amphistomy effectively halves the CO\(_2\) diffusion path length and
-boundary layer resistance by doubling boundary layer conductance
-\citep{parkhurst_adaptive_1978, harrison_influence_2020, mott_amphistomy_1991}.
-Despite these optimality predictions, stomatal anatomy may be partially
-constrained by physical and developmental limits on phenotypic
-expression
-\citep{croxdale_stomatal_2000, harrison_influence_2020, muir_how_2023}.
-
-A number of physical and developmental processes constrain stomatal
-anatomy traitspace. For example, almost all stomata follow the one cell
-spacing rule to maintain proper stomatal functioning as guard cell
-movement requires the rapid exchange of ions with neighboring epidermal
-cells (i.e.~subsidiary cells)
-\citep{geisler_oriented_2000, dow_physiological_2014}. This would
-prevent stomata from being strongly clustered; however, some species
-(notably in \emph{Begonia}) appear to benefit from the overlapping vapor
-shells caused by stomatal clustering in dry environments
-\citep{yi_gan_stomatal_2010, lehmann_effects_2015, papanatsiou_stomatal_2017}.
-Historically, stomatal patterning in dicot angiosperms was thought to be
-random with an exclusionary distance surrounding each stomate
-\citep{sachs_developmental_1974}; however, the developmental controls of
-stomatal patterning are more complex than random development along the
-leaf surface. \citet{croxdale_stomatal_2000} reviews three developmental
-theories which attempt to explain stomatal patterning in angiosperms:
-inhibition, cell lineage, and cell cycle, ultimately arguing for a cell
-cycle based control of stomatal patterning.
-\citet{pillitteri_mechanisms_2012} review the short and long distance
-signalling pathways associated with stomatal spacing and development,
-which include cell to cell communication and whole plant integration to
-ensure the proper spacing of stomata across a single leaf surface
-depending on environmental ques. Much less is known about the
-development of stomata on the adaxial leaf surface in amphistomatous
-plants. Stomatal size is additionally constrained by genome size with
-larger genomes leading to larger minimum guard cell size
-\citep{jordan_environmental_2015}. Despite these limitations,
-ecophysiological theory still predicts optimal stomatal anatomy, the
-details of which are discussed below.
-
-The patterning and spacing of stomata on the leaf affects photosynthesis
-in \(C_3\) leaves by altering the CO\(_2\) diffusion path length from
-stomata to sites of carboxylation in the mesophyll. Maximum
-photosynthetic rate (\(A_\text{max}\)) in \(C_3\) plants is generally
-co-limited by biochemistry and diffusion, but modulated by light
-availability
-\citep{parkhurst_intercellular_1990, manter_ci_2004, carriqui_diffusional_2015}.
-Low light decreases CO\(_2\) demand by limiting electron transport rate,
-leading to relatively high internal CO\(_2\) concentration
-(\(C_\text{i}\)) and low \(A_\text{max}\) \citep{kaiser_metabolic_2016}.
-In contrast, well hydrated leaves with open stomata in high light,
-photosynthesis is often limited by CO\(_2\) supply as resistances from
-the boundary layer, stomatal pore, and mesophyll can result in
-insufficient CO\(C_2\) supply at the chloroplast to maxmimize
-photosynthesis \citep{farquhar_biochemical_1980, lehmeier_cell_2017}. In
-this study, we focus primarily on how stomatal patterning affects
-diffusion, ignoring boundary layer and mesophyll resistances.
-
-To maximize CO\(_2\) supply from the stomatal pore to chloroplasts,
-stomata should be uniformly distributed in an equilateral triangular
-grid on the leaf surface so as to minimize stomatal number and CO\(_2\)
-diffusion path length \citep{parkhurst_diffusion_1994}. As the diffusion
-rate of CO\(_2\) though liquid is approximately \(10^4\times\) slower
-than CO\(_2\) diffusion through air, mesophyll resistance is generally
-thought to be primarily limited by liquid diffusion
-\citep{aalto_three-dimensional_2002, evans_resistances_2009}, but
-diffusion through the IAS has also been shown to be a rate limiting
-process because the tortuous, disjunct nature of the IAS can greatly
-increase diffusion path lengths \citep{harwood_understanding_2021}.
-Additionally, tortuosity is higher in horizontal directions (parallel to
-leaf surface) than vertical directions (perpendicular to leaf surface)
-because of the cylindrical shape and vertical arrangement of pallisade
-mesophyll cells \citep{earles_beyond_2018, harwood_understanding_2021}.
-However, the ratio of lateral to vertical diffusion rate is still
-largely unknown
-\citep{morison_lateral_2005, pieruschka_lateral_2005, pieruschka_lateral_2006}.
-Depending on the thickness of the leaf, porosity of the leaf mesophyll,
-tortuosity of the IAS, and lateral to vertical diffusion rate ratio,
-minimizing diffusion path length for CO\(_2\) via optimally distributed
-stomata may yield significant increases in CO\(_2\) supply for
-photosynthesis and higher \(A_\text{max}\).
-
-We hypothesized that natural selection will favor stomatal patterning
-and distribution to minimize the diffusion path length. In
-amphistomatous leaves, this would be accomplished by 1) a dispersed,
-equilateral triangular distribution of stomata on both abaxial and
-adaxial leaf surfaces and 2) coordinated stomatal spacing on each
-surface that offsets the position of stomata
-(\autoref{fig:ideal-amphi-grid}). Coordination between leaf surfaces is
-defined in this study as the occurrence of stomata in areas farthest
-from stomata on the opposite leaf surface. Additionally, because
-CO\(_2\) is more limiting for photosynthesis under high light, we
-hypothesize that in high light 3) there should be more stomata, and 4)
-stomata should be more uniformly distributed than in low light. Finally,
-as stomatal densities are selected for optimal operational aperture, we
-hypothesize that 5) stomatal length will be positively correlated with
-the area of the leaf surface to which it is closest. We refer to this as
-the `stomatal zone', the leaf area surrounding a focal stomate closest
-to that stomate and therefore the zone it supplies with CO\(_2\)). This
-way, each stomate can be optimally sized relative to the mesophyll
-volume it supplies.
-
-\begin{figure}[ht]
-\includegraphics[width = 4in]{figures/ideal-amphi-grid.pdf}
-\caption{Idealized amphistomatous stomatal grid with uniform stomatal patterning and perfect abaxial-adaxial coordination.}
-\label{fig:ideal-amphi-grid}
-\end{figure}
-
-To test these hypotheses, we grew the model plant \emph{Arabidopsis
-thaliana} in high, medium, and low light and measured stomatal density,
-size, and patterning on both leaf surfaces, and spatial coordination
-between them. We use Voronoi tessellation techniques to calculate
-stomatal zones. We also used a 2-D porous medium approximation of
-CO\(_2\) diffusion and photosynthesis to predict the photosynthetic
-advantage of optimal versus suboptimal coordination in stomatal
-coordination between surfaces. Specifically, we predicted that traits
-which affect diffusion path length (leaf thickness, stomatal density,
-leaf porosity, lateral-vertical diffusion rate ratio), diffusion rate
-(temperature, pressure), and CO\(_2\) demand (Rubisco concentration,
-light) would modulate the advantage of optimal stomatal arrangement
-following the relationships outlined in \autoref{tab:hypotheses}. Here,
-we integrate over reasonable parameter space to determine the
-ecophysiological context most likely to favor stomatal spatial
-coordination in amphistomatous leaves.
-
-\begin{table}[ht]
-\centering
-\begin{tabular}{ll}
-  \hline
-trait & relationship \\ 
-  \hline
-leaf thickness & + \\ 
-  interstomatal distance & + \\ 
-  leaf porosity & - \\ 
-  light & + \\ 
-   \hline
-\end{tabular}
-\caption{A summary of the hypothesized relationships between leaf traits and environmental conditions and photosynthetic advantage of stomatal spatial coordination in amphistomatous leaves.} 
-\label{tab:hypotheses}
-\end{table}
-
-\hypertarget{materials-and-methods}{%
-\section{Materials and methods}\label{materials-and-methods}}
-
-\hypertarget{data-preparation}{%
-\subsection{Data Preparation}\label{data-preparation}}
-
-Plant material, growth conditions, and three-dimensional confocal
-imaging are described in \citet{dow_disruption_2017}. Briefly, Columbia
-(Col-0) ecotype of \emph{Arabidopsis thaliana} (L.) Heynh. plants were
-grown in three different light environments: low light (PAR = 50
-\(\mu \text{mol}~\text{m}^{-2}~\text{s}^{-1}\)), medium light (100
-\(\mu \text{mol}~\text{m}^{-2}~\text{s}^{-1}\)), and high light (200
-\(\mu \text{mol}~\text{m}^{-2}~\text{s}^{-1}\)). PAR stands for
-photosynthetically active radiation. Seeds were surface-sterilized and
-stratified at 4°C for 3--5 d in 0.15\% agarose solution and then sown
-directly into Pro-Mix HP soil (Premier Horticulture; Quakerstown, PA,
-USA) and supplemented with Scott's Osmocote Classic 14-14-14 fertilizer
-(Scotts-Sierra, Marysville, OH, USA). At 10--14 d, seedlings were
-thinned so only one seedling per container remained. Plants were grown
-to maturity in growth chambers where the conditions were as follows: 16
-: 8 h, 22 : 20°C, day : night cycle. Imaging of the epidermis and
-internal leaf structures was performed using a Leica SP5 confocal
-microscope with the protocol developed by
-\citet{wuyts_high-contrast_2010} with additional modification described
-in \citet{dow_disruption_2017}. We captured 132 images in total, making
-66 abaxial-adaxial image pairs. We measured stomatal position and size
-using ImageJ \citep{schneider_nih_2012}.
-
-\hypertarget{single-surface-analyses}{%
-\subsection{Single surface analyses}\label{single-surface-analyses}}
-
-We tested whether stomata are non-randomly distributed by comparing the
-observed stomatal patterning to a random uniform pattern. For each leaf
-surface image with \(n\) stomata we generated \(10^3\) synthetic
-surfaces with \(n\) stomata uniformly randomly distributed on the
-surface. For each sample image, we compared the observed Nearest
-Neighbor Index (\(\mathrm{NNI}\)) to the null distribution of
-\(\mathrm{NNI}\) values calculated from the synthetic data set.
-\(\mathrm{NNI}\) is the ratio of observed mean distance
-(\(\overline{D}_O\)) to the expected mean distance (\(\overline{D}_E\))
-where \(\overline{D}_E\) is:
-
-\begin{equation}\label{eq:emd}
-  \overline{D}_E = \frac{0.5}{\sqrt{A_\text{leaf} / n_\text{stomata}}}.
-\end{equation}
-
-\noindent \(A_\text{leaf}\) is leaf area visible in the sampled field
-and \(n_\text{stomata}\) the number of stomata. \(\overline{D}_E\) is
-the theoretical average distance to the nearest neighbor of each stomate
-if stomata were uniformly randomly distributed
-\citep{clark_distance_1954}. \(\overline{D}_O\) calculated for each
-synthetic data set is:
-
-\begin{equation}\label{eq:omd}
-  \overline{D}_O = \frac{\sum_{i=1}^{n_\text{stomata}}d_i}{n_\text{stomata}},
-\end{equation}
-
-\noindent where \(d_i\) is the distance between \(\text{stomate}_i\) and
-its nearest neighbor. We calculated \(\mathrm{NNI}\) using the \emph{R}
-package \textbf{spatialEco} version 2.0.1 \citep{evans_spatialeco_2023}.
-The observed stomatal distribution is dispersed relative to a uniform
-random distribution if the observed \(NNI\) is greater than 95\% of the
-synthetic \(\mathrm{NNI}\) values (one-tailed test). We adjusted
-\(P\)-values to account for multiple comparisons using the
-Benjamini-Hochberg \citep{benjamini_controlling_1995} false discovery
-rate procedure implemented in the \emph{R} package \textbf{multtest}
-version 2.56.0 \citep{wong_multiple_2005}.
-
-For each sample image, we also simulated \(10^3\) synthetic data with
-\(n\) stomata ideally dispersed in an equilateral triangular grid. For
-these grids, we integrated over plausible stomatal densities and then
-conditioned on stomatal grids with exactly \(n\) stomata. The simulated
-stomatal count was drawn from a Poisson distribution with the mean
-parameter \(\lambda\) drawn from a Gamma distribution with shape \(n\)
-and scale 1 (\(\lambda \sim \Gamma(n, 1)\)). \(\Gamma(n, 1)\) is the
-posterior distribution of \(\lambda\) with a flat prior distribution.
-This allows us to integrate over uncertainty in the stomatal density
-from the sample image.
-
-We developed a dispersion index \(\mathrm{DI}\) to quantify how close
-observed stomatal distributions are to random uniform versus maximally
-dispered in an equilateral triangular grid. \(\mathrm{DI}\) varies from
-zero to one, where zero is uniformly random and one is ideally
-dispersed:
-
-\begin{equation}\label{eq:disp}
-  \mathrm{DI} = \frac{\mathrm{NNI} - \text{median}(\mathrm{NNI_{random}})}{\text{median}(\mathrm{NNI_{uniform}}) - \text{median}(\mathrm{NNI_{random}})}
-\end{equation}
-
-\noindent \(\mathrm{NNI}\) is calculated for each sample image as
-described above; \(\text{median}(\mathrm{NNI_{random}})\) and
-\(\text{median}(\mathrm{NNI_{uniform}})\) are calculated from the
-synthetic data specific to each sample image as described above. We
-tested whether light treatment affects \(\mathrm{DI}\) and stomatal
-density (\(D_S\)) using analysis of variance (ANOVA).
-
-Finally, we examined the relationship between stomatal zone area and
-stomatal length using a Bayesian linear mixed-effects model fit with the
-\emph{R} package \textbf{brms} version 2.20.4
-\citep{burkner_brms_2017, burkner_advanced_2018} and \emph{Stan} version
-2.33.1 \citep{stan_development_team_stan_2023}. Stomatal zone area was
-calculated using Voronoi tessellation
-(e.g.~\autoref{fig:tessellation-example}). The stomatal zone area,
-\(S_\text{area}\), is the region of the leaf surface whose distance to
-stomate, \(S\), is less than the distance to any other stomate, \(S\).
-Stomatal length was measured in ImageJ \citep{schneider_nih_2012}. We
-modeled fixed effects of surface, light treatment, stomatal length, and
-their 2- and 3-way interactions on \(\sqrt{S_\text{area}}\). We included
-random intercepts, random effects of surface, random slopes, and random
-surface-by-slope interactions within both plant and individual to
-account for nonindependence of stomata within the same plant or
-individual. We also modeled residual variance as a function of light
-treatment. We sampled the posterior distribution from 4 chains with 1000
-iterations each after 1000 warmup iterations. We calculated convergence
-diagnostics (\(\hat{R}\)) and effective sample sizes following
-\citet{vehtari_rank-normalization_2021}. We estimated the marginal slope
-and 95\% highest posterior density (HPD) intervals between stomatal
-length and \(\sqrt{S_\text{area}}\) using the \emph{emtrends} function
-in the \emph{R} package \textbf{emmeans} version 1.8.8
-\citep{lenth_emmeans_2023}.
-
-\begin{figure}[ht]
-\includegraphics[height = \textheight]{figures/tessellation-example.pdf}
-\caption{Examples of synthetic and real leaf surfaces.
-A) Uniform random synthetic leaf surface; B) Example of real leaf surface; C) Regularly distributed synthetic leaf surface.
-The zone defined by each stomate was calculated with voronoi tessellation and correlated with stomatal length in real leaves.}
-\label{fig:tessellation-example}
-\end{figure}
-
-\hypertarget{paired-abaxial-and-adaxial-surface-analysis}{%
-\subsection{Paired Abaxial and Adaxial Surface
-Analysis}\label{paired-abaxial-and-adaxial-surface-analysis}}
-
-To test whether the position of ab- and adaxial stomata are coordinated
-we compared the observed distribution to a null distribution where the
-positions on each surface are random. For each pair of surfaces
-(observed or synthetic) we calculated the distance squared between each
-pixel to the nearest stomatal centroid with the \emph{R} package
-\textbf{raster} version 3.6.26. We refer to this as the `nearest
-stomatal distance' or NSD. Then we calculated the pixel-wise Pearson
-correlation coefficient. If stomatal positions on each surface are
-coordinated to minimize the distance between mesophyll and the nearest
-stomate, then we expect a negative correlation. A pixel that is far from
-a stomate on one surface should be near a stomate on the other surface
-(\autoref{fig:ideal-amphi-grid}). We generated a null distribution of
-the correlation coefficient by simulating \(10^3\) synthetic data sets
-for each observed pair. For each synthetic data set, we simulated
-stomatal position using a random uniform distribution, as described
-above, matching the number of stomata on abaxial and adaxial leaf
-surfaces. Stomatal positions on each surface are coordinated if the
-correlation coefficient is greater than 95\% of the synthetic
-correlation values (one-tailed test).
-
-\hypertarget{modeling-photosynthesis}{%
-\subsection{Modeling Photosynthesis}\label{modeling-photosynthesis}}
-
-We modeled photosynthesis CO\(_2\) assimilation rate using a
-spatially-explicit two-dimensional reaction diffusion model using a
-porous medium approximation \citep{parkhurst_diffusion_1994} using the
-finite element method (FEM) following \citet{earles_excess_2017}.
-Consider a two-dimensional leaf where stomata occur on each surface in a
-regular sequence with interstomatal distance \(U\). The main outcome we
-assessed is the advantage of offsetting the position of stomata on each
-surface compared to have stomata on the same \(x\) position on each
-surface. With these assumptions, by symmetry, we only need to model two
-stomata, one abaxial and one adaxial, from \(x = 0\) to \(x = U/2\) and
-from the adaxial surface at \(y = 0\) to the abaxial surface at
-\(y = L\), the leaf thickness. We arbitrarily set the adaxial stomate at
-\(x = 0\) and toggled the abaxial stomata position between \(x = U/2\)
-(offset) or \(x = 0\) (below adaxial stomate). The `coordination
-advantage' of offset stomatal position on each surface is the
-photosynthetic rate of the leaf with offset stomata compared to that
-with stomata aligned in the same \(x\) position:
-
-\begin{equation} \label{eq:coordination_advantage}
-  \text{coordination advantage} = \frac{A_\text{offset}}{A_\text{aligned}}
-\end{equation}
-
-We modeled the coordination advantage over a range of leaf thicknesses,
-stomatal densities, photosynthetic capacities, and light environments to
-understand when offsetting stomatal position on each surface might
-deliver a significant photosynthetic advantage
-(\autoref{tab:model_var}). The complete model description is available
-in the Supporting Information.
-
-\begin{table}
-
-\caption{\label{tab:model_var}The parameter range of model variables tested for their effect on coordination advantage (\autoref{eq:coordination_advantage}) using a 2-D porous medium approximation.
-We used regularly spaced values within each range and simulated across all combinations.
-Here we converted model units to more conventional units (e.g.
-m to $\mu$m).
-$I_0$: PPFD incident on the leaf surface; $\varphi_\text{pal}$: Fraction of intercellular airspace (aka porosity), palisade; $T_\text{leaf}$: Leaf thickness; $U$: Interstomatal distance}
-\centering
-\begin{tabular}[t]{lll}
-\toprule
-Variable & Parameter range & Units\\
-\midrule
-$I_0$ & $50-1000$ & $\mu$mol m$^{-2}$ s$^{-1}$\\
-$\varphi_\text{pal}$ & $0.1-0.3$ & m$^3$ airspace m$^{-3}$ leaf\\
-$T_\text{leaf}$ & $101-501$ & $\mu$m\\
-$U$ & $17-169$ & $\mu$m\\
-\bottomrule
-\end{tabular}
-\end{table}
-
-\hypertarget{results}{%
-\section{Results}\label{results}}
-
-Stomatal density of \emph{Arabidopsis thaliana} varies among light
-treatments (ANOVA, \(F_{2,126} = 681, P = 2.88 \times 10^{-68}\))
-because the density is much greater in the high light treatment
-(\autoref{fig:density}). Density is consistently greater on abaxial leaf
-surfaces across all light treatments (ANOVA,
-\(F_{1,126} = 44.2, P = 8.21 \times 10^{-10}\); \autoref{fig:density}).
-There is no evidence for an interaction between light treatment and
-surface (ANOVA, \(F_{2,126} = 2.75 \times 10^{-2}, P = 0.973\)). Leaves
-are amphistomatous with a mean stomatal density ratio of 0.44.
-
-\begin{figure}[ht]
-\includegraphics[width = 4in]{figures/density.pdf}
-\caption{Stomatal density is higher in \textit{A.
-thaliana} plants grown under high light conditions.
-We determined statistical significance between light treatments using Tukey post-hoc tests.
-$^*~0.05 > P \ge 0.01; ^{**}~0.01 > P \ge 0.001; ^{***}~0.0001 > P \ge 0.0001; ^{***}~ P <0.0001$.}
-\label{fig:density}
-\end{figure}
-
-\hypertarget{stomatal-distribution-is-non-random-but-far-from-ideal}{%
-\subsection{Stomatal distribution is non-random, but far from
-ideal}\label{stomatal-distribution-is-non-random-but-far-from-ideal}}
-
-Many leaf surfaces (37 of 132, 28\%) are significantly overdispersed
-compared to a random uniform distribution, but none were close to an
-ideal hexagonal pattern (dispersion index = 1;
-\autoref{fig:single-surface}). Before controlling for multiple
-comparisons, 43.2\% are significantly overdispersed. The dispersion
-index differs significantly among light treatments (ANOVA,
-\(F_{2,126} = 8.55, P = 3.30 \times 10^{-4}\)) because the medium light
-treatment is significantly less than the low treatment
-(\autoref{fig:single-surface}). Dispersion index is consistently greater
-on adaxial leaf surfaces across all light treatments (ANOVA,
-\(F_{1,126} = 28.8, P = 3.67 \times 10^{-7}\);
-\autoref{fig:single-surface}). There is no evidence for an interaction
-between light treatment and surface (ANOVA,
-\(F_{2,126} = 0.577, P = 0.563\)).
-
-\begin{figure}[ht]
-\includegraphics[width=4in]{figures/single-surface.pdf}
-\caption{Stomata are more dispersed than expected under the null model of uniform random position (dispersion index = 0) but far from a distribution that maximizes distance between stomata (dispersion index = 1).
-We determined statistical significance between light treatments using Tukey post-hoc tests.
-$^*~0.05 > P \ge 0.01; ^{**}~0.01 > P \ge 0.001; ^{***}~0.0001 > P \ge 0.0001; ^{***}~ P <0.0001$.}
-\label{fig:single-surface}
-\end{figure}
-
-\hypertarget{no-evidence-for-coordinated-stomatal-position-between-surfaces}{%
-\subsection{No evidence for coordinated stomatal position between
-surfaces}\label{no-evidence-for-coordinated-stomatal-position-between-surfaces}}
-
-There is no evidence of spatial coordination between abaxial and adaxial
-leaf surfaces. The pixel-wise correlation between nearest stomatal
-distance (NSD) squared on paired abaxial and adaxial leaf surfaces is
-not significantly less than zero in any of the 66 leaves
-(\autoref{fig:dual-surface}). Before controlling for multiple
-comparisons, 3\% are significantly \emph{positively} correlated. The NSD
-correlation is not different among light treatments (ANOVA,
-\(F_{2,63} = 2.28, P = 0.111\); \autoref{fig:dual-surface}).
-
-\begin{figure}[ht]
-\includegraphics[width = 4in]{figures/dual-surface.pdf}
-\caption{Pixel-wise correlation between near stomatal distance (NSD) squared on paired abaxial and adaxial leaf surfaces.
-Dashed line indicates zero correlation.
-Weak positive correlations are not significantly different from zero after correcting for multiple comparisons.
-The correlation does not differ among light treatments.}
-\label{fig:dual-surface}
-\end{figure}
-
-\hypertarget{larger-stomata-supply-larger-mesophyll-volumes}{%
-\subsection{Larger stomata supply larger mesophyll
-volumes}\label{larger-stomata-supply-larger-mesophyll-volumes}}
-
-All parameters converged (\(\hat{R} < 1.01\)) and effective sample sizes
-were exceeded \(10^3\). Across all light treatments and leaf surfaces,
-stomatal length and stomatal area are weakly positively correlated
-(\autoref{fig:length-area}). The slope was significantly greater than
-zero for all abaxial surfaces, but not for the adaxial surface in low
-and medium light treatments. The estimated marginal slopes and 95\% HPD
-intervals for each combination of light and surface is: low light,
-abaxial surface: 1.928 {[}\numrange{0.779}{3.133}{]}; low light, adaxial
-surface: 1.745 {[}\numrange{-0.041}{3.373}{]}; medium light, abaxial
-surface: 1.085 {[}\numrange{0.328}{1.957}{]}; medium light, adaxial
-surface: 0.656 {[}\numrange{-0.399}{1.691}{]}; high light, abaxial
-surface: 0.597 {[}\numrange{0.316}{0.911}{]}; high light, adaxial
-surface: 1.269 {[}\numrange{0.831}{1.721}{]}.
-
-\begin{figure}[ht]
-\includegraphics[width = 4in]{figures/length-area.pdf}
-\caption{Stomatal length and stomatal zone area are positively correlated. Linear regression lines and 95\% confidence ribbons are from a Bayesian linear mixed-effects model.}
-\label{fig:length-area}
-\end{figure}
-
-\hypertarget{little-benefit-of-coordinated-stomatal-arrangement}{%
-\subsection{Little benefit of coordinated stomatal
-arrangement}\label{little-benefit-of-coordinated-stomatal-arrangement}}
-
-We used the finite element method (FEM) to model CO\(_2\) diffusion
-within the leaf and photosynthesis as a 2-D porous medium. Across all
-realistic parts of parameter space, the coordination advantage is much
-less than 0.01 (\autoref{fig:model_summary}). For reference, a
-log-response of ratio is 0.01 is approximately 1\%. The only exception
-was for thin leaves (\(T_\text{leaf} = 100~\mu \text{m}\)) with few
-stomata (\(U = 338~\mu \text{m}\), which corresponds to a stomatal
-density of \(\approx 10~\text{mm}^{-2}\)), where lateral diffusion is
-major constraint on CO\(_2\) supply. However, such thin leaves with so
-few stomata are uncommon among C\(_3\) plants (some CAM plants have low
-stomatal density \citep{males_stomatal_2017}). In other areas of
-parameter space, lateral diffusion limitations were small relative to
-those along the ab-adaxial axis (see Fig. S1 for a representative model
-solution).
-
-\begin{figure}[ht]
-\includegraphics[width = 5in]{figures/model_summary.pdf}
-\caption{There is little photosynthetic benefit of offsetting stomatal position each surface based on a 2-D model of photosynthesis.
-The coordination advantage (\autoref{eq:coordination_advantage}) is close to zero under nearly all of the parameter space \autoref{tab:model_var}, meaning that the photosynthetic rate of amphistomatous leaves with stomata optimally offset is nearly equal to leaves with stomata on each surface in the same position along the leaf plane.
-$I_0$: PPFD incident on the leaf surface; $\varphi_\text{pal}$: Fraction of intercellular airspace (aka porosity), palisade; $T_\text{leaf}$: Leaf thickness; $U$: Interstomatal distance.}
-\label{fig:model_summary}
-\end{figure}
-
-\hypertarget{discussion}{%
-\section{Discussion}\label{discussion}}
-
-Stomata cost resources to maintain \citep{deans_optimization_2020} and
-expose leaves to risks such as hydraulic failure
-\citep{wang_theoretical_2020} or infection by plant pathogens
-\citep{melotto_stomatal_2017}. Therefore leaves should develop enough
-stomata to adequately supply CO\(_2\) to chloroplasts, but not
-overinvest. A widespread hypothesis in plant ecophysiology is that
-natural selection optimizes traits like stomatal size, density, and
-distribution to maximize carbon gain relative to any costs in a given
-environmental context. In principle, spacing stomata to minimize the
-average distance between stomatal pores and chloroplasts within the
-mesophyll should increase carbon gain, all else being equal. However,
-reducing this distance to its absolute minimum may be constrained by
-developmental processes, or the photosynthetic benefit may be too small
-to be `seen' by natural selection (i.e.~the selection coefficient is
-less than drift barrier \emph{sensu} \citet{sung_drift-barrier_2012}).
-
-We tested five related hypotheses about stomatal spacing in
-amphistomatous leaves using the model angiosperm \emph{Arabidopsis
-thaliana} grown under different light intensities. First, we predicted
-that stomata on each surface are overdispersed relative to a random
-uniform distribution, which should increase CO\(_2\) supply. Stomata on
-each surface are overdispersed (\autoref{fig:single-surface}), but are
-not ideally dispersed in an equilateral triangular grid as would be
-optimal to minimize CO\(_2\) diffusion path length and equalize the area
-supplied by each stomate (\autoref{fig:tessellation-example}). Second,
-we predicted that an optimal amphistomatous leaf has offset stomata such
-that stomata are more likely to appear on one leaf surface if there is
-not a stomata directly opposite it on the other surface as shown in
-\autoref{fig:ideal-amphi-grid}. However, there is no evidence for
-coordination and the positions on each surface appear independent,
-regardless of light treatment (\autoref{fig:dual-surface}). Third, we
-predicted that plants respond plastically to higher light intensity by
-increasing stomatal density. \emph{Arabidopsis} plants grown under high
-light had higher stomatal density than the same genotype growns under
-low and medium light intensity (\autoref{fig:density}). However, we
-found no support for our fourth prediction that stomata would be more
-evenly dispersed at high light intensity (\autoref{fig:single-surface}).
-Finally, we predicted that within leaf variation in stomatal size would
-correlate with stomatal spacing, as larger stomata can supply larger
-volumes of adjacent mesophyll. In all three light treatments, stomatal
-size positively covaried with the stomatal zone, i.e.~adjacent region of
-mesophyll that would be supplied by that stomate
-(\autoref{fig:length-area}).
-
-Stomatal spacing on \emph{A. thaliana} leaves partially supports our
-overall hypothesis that natural selection minimizes the average distance
-between stomata and chloroplasts, for a given overall stomatal density.
-There are three non-mutually exclusive hypotheses for why several of our
-predictions were wrong. First, our predictions are wrong becuase they
-are based on overly simplistic assumptions about epidermal and mesophyll
-anatomy. Second, natural selection may be constrained by developmental
-processes that prevent phenotypes from reaching their adaptive optima.
-Third, the benefit of some traits may be of too little consequence to
-result in fitness differences large enough to respond to selection. We
-consider the plausibility of these alternative hypotheses below and
-present ideas for future work to test them.
-
-We assume an idealized leaf epidermal and mesophyll structure that is
-homogenous and unconstrained by other tradeoffs. Real leaves not only
-provide pathways for CO\(_2\) diffusion, but must supply water,
-intercept light, and deter herbivores and pathogens. These competing
-interests result in nonuniform epidermal and mesophyll structure that
-could alter our predictions about optimal stomatal spacing. In order to
-maintain consistent leaf water potential across the lamina, stomatal
-density must be coordinated with vein density
-\citep{fiorin_transport_2016}. Thus, stomatal spacing may be optimized
-not at the interstomatal level, but at a higher level, coordinating
-water transport and water loss. We also assume a uniform mesophyll;
-however, in ab-adaxially differentiated leaves, this assumption is
-overly simplistic. The pallisade mesophyll is more tightly packed than
-the spongy mesophyll as an adaptation to intercept light efficiently, so
-lateral diffusion may be more limiting in the adaxial portion of the
-leaf. This may explain why adaxial leaf surfaces have consistently
-higher dispersion indices than abaxial surfaces across all light
-treatments (\autoref{fig:single-surface}). Future gas exchange models
-should include multiple parameter sets for different strata of mesophyll
-tissue.
-
-No developmental pathway exists to ensure an idealized placement of
-stomata on the leaf surface. Rather, stomatal development is a dynamic
-process that must be plastic to environmental cues. Leaves develop based
-on short and long distance signalling pathways which relay information
-about incoming light, humidity, temperature, and surrounding stomata to
-developing leaf tissues \citep{pillitteri_mechanisms_2012}. Our results
-showing an intermediate level of dispersion in stomatal spacing may be
-best explained by these developmental pathways which ensure the proper
-spacing of stomata, with an added random effect brought about by the
-necessity for plasticity in stomatal development
-(\autoref{fig:single-surface}). However, any nonideal randomness in
-stomatal spacing may offset by simultaneous and coordinated development
-of the IAS \citep{baillie_developmental_2020}. The fact that stomata
-which supply a greater mesophyll volume tend to be larger suggests that
-plants may use coordinated development of multiple leaf tissues to
-negate nonideal stomatal spacing (\autoref{fig:length-area}).
-
-In amphistomatous leaves, ideal stomatal spacing is complicated by a
-third dimension. Our gas exchange model demonstrates little
-photosynthetic gain from abaxial-adaxial stomatal coordination
-(\autoref{fig:model_summary}). Even though lateral diffusion may limit
-photosynthesis \citep{morison_lateral_2005}, the marginal gain from
-optimally offsetting stomata is not sufficient to generate fitness
-differences relative to the strength of genetic drift (i.e.~the
-drift-barrier). We can similarly extrapolate that an ideal, equilateral
-triangular stomatal spacing is only slightly better than a suboptimal
-pattern. Any benefit garnered by ideal stomatal spacing may be
-additionally offset by a cost to developmental flexibility in variable
-environments
-\citep{pillitteri_mechanisms_2012, baillie_developmental_2020}.
-Explaining these observations as the result of weak selection is in
-tension with the finding that stomatal size and zone positively covary,
-which would suggest that small changes in lateral diffusion distance are
-significant. As described above, the positive correlation between
-stomatal size and zone may be explained by common developmental
-processes rather than as an adaptation to maximize CO\(_2\) diffusion.
-In any case, there is no evidence for coordinated development of both
-leaf surfaces, and very little theoretical benefit to photosynthesis,
-except in marginal circumstances which are exceptionally rare in nature.
-
-Our study corroborates previous studies which demonstrate that stomata
-are non-randomly distributed along the leaf surface as a result of
-developmental mechanisms such as spatially biased arrest of stomatal
-initials \citep{boetsch_arrest_1995}, oriented asymmetric cell division
-\citep{geisler_oriented_2000}, and cell cycle controls
-\citep{croxdale_stomatal_2000}. We do not investigate the potential
-developmental pathways that influence stomatal dispersion in this study;
-however, they are important to consider as these pathways could limit
-plants from reaching a theoretical peak in the adaptive landscape:
-uniform stomatal dispersion. Instead, as this study suggests, plants may
-simply compensate with higher stomatal density and by fitting stomatal
-size to the area that they supply with CO\(_2\). To understand why
-stomata are not ideally dispersed, more modeling (with a more complex
-set of assumptions including vein density and IAS structure) should be
-done to estimate the fitness gain of stomatal dispersion. Additionally,
-genetic manipulation studies should attempt to create mutants with
-clustered and ideally dispersed stomata for a comparison of their
-photosynthetic traits. This could have extremely important implications
-for maximum assimilation rates in crops as most crop species are grown
-in high light where CO\(_2\) is often limiting. In drought-prone
-environments, increased stomatal dispersion may increase water use
-efficiency by reducing the number of stomata needed to achieve the same
-internal CO\(_2\) concentration, \(C_\text{i}\).
-
-Our results suggest that after optimizing stomatal density and having
-developmental rules for spacing stomata relatively evenly, there may be
-limited gains to further optimization. Therefore, developmental
-constraints may be necessary to make sense of some features of stomatal
-spacing and distribution. The possibility that ideal stomatal spacing is
-not the ``tallest'' fitness peak must also be explored, as stomate size
-is demonstrated in this study to covary with mesophyll volume supplied
-with CO\(_2\). This may be especially true in highly variable
-environments or in large tree species with sun and shade leaves where
-developmental cues may change rapidly. The temporal component, not
-considered here, could also have significant implications, as CO\(_2\)
-may only be limiting to photosynthesis during short periods when all
-other conditions are ideal. In these cases, the theoretical benefits of
-ideal stomatal spacing are further diminished. Future exploration of
-these competing hypotheses would require more advanced modeling,
-additional exploration of IAS space development and its effects on gas
-exchange, both real and modeled, and knowledge about how often the
-species of interest is CO\(_2\) limited across of range of natural
-settings. Despite these additional considerations, this study represents
-an important contribution to understanding the potential drivers of and
-limitations to stomatal anatomy in amphistomatous plants.
-
 \section{Competing interests}
@@
 
-\renewcommand\refname{References}
-
 \bibliographystyle{abbrvnat}

### 2023-11-13 r/15_prepare-submission.R

@@ -7,7 +7,7 @@ x = which(ms %in% c(
 ))
 
 readr::write_lines(c(ms[1:6], ms[x], ms[7:(min(x) - 1)], ms[(max(x) + 1): length(ms)]), "ms/ms.tex")
-system("cd ms; xelatex ms.tex")
+system("cd ms; xelatex ms.tex; xelatex ms.tex")
 
 # Copy and rename files for submission
 file.copy("ms/ms.pdf", "aobp1/ms.pdf", overwrite = TRUE)

### 2023-11-14 r/15_prepare-submission.R
fixed reference issue in ms

@@ -6,7 +6,7 @@ x = which(ms %in% c(
   "\\newunicodechar{ʻ}{\\okina}"
 ))
 
-readr::write_lines(c(ms[1:6], ms[x], ms[7:(min(x) - 1)], ms[(max(x) + 1): length(ms)]), "ms/ms.tex")
+readr::write_lines(c(ms[1:6], ms[x], ms[7:(min(x) - 1)], ms[(max(x) + 1):length(ms)]), "ms/ms.tex")
 system("cd ms; xelatex ms.tex; xelatex ms.tex")
 
 # Copy and rename files for submission

### 2024-01-17 ms/ms.Rmd
Response to review, ms edits, and bibliography file.

@@
 Amphistomatous leaves have stomata on both abaxial and adaxial leaf surfaces.
-This feature is considered an adaptation to alleviate CO$_2$ diffusion limitations in productive environments where other factors are not limiting as the diffusion path length from stomate to chloroplast is effectively halved.
-Plants can also reduce CO$_2$ limitations through other aspects of optimal stomatal anatomy: stomatal density, distribution, patterning, and size.
-A number of studies have demonstrated that stomata are overdispersed on a single leaf surface; however, much less is known about stomatal anatomy in amphistomatous leaves, especially the coordination between leaf surfaces, despite their prevelance in nature and near ubiquity among crop species.
-Here we use novel spatial statistics based on simulations and photosynthesis modeling to test hypotheses about how amphistomatous plants may optimize CO$_2$ limitations in the model angiosperm *Arabidopsis thaliana* grown in different light environments.
-We find that 1) stomata are overdispersed, but not ideally dispersed, on both leaf surfaces across all light treatments; 2) abaxial and adaxial leaf surface patterning are independent; and 3) the theoretical improvements to photosynthesis from abaxial-adaxial stomatal coordination are miniscule ($\\ll 1$%) across the range of feasible parameter space.
+This feature is considered an adaptation to alleviate CO$_2$ diffusion limitations in productive environments as the diffusion path length from stomate to chloroplast is effectively halved in amphistomatous leaves.
+Plants may also reduce CO$_2$ limitations through other aspects of optimal stomatal anatomy: stomatal density, distribution, patterning, and size.
+A number of studies have demonstrated that stomata are overdispersed compared to a random distribution on a single leaf surface; however, despite their prevelance in nature and near ubiquity among crop species, much less is known about stomatal anatomy in amphistomatous leaves, especially the coordination between leaf surfaces.
+Here we use novel spatial statistics based on simulations and photosynthesis modeling to test hypotheses about how amphistomatous plants may optimize CO$_2$ diffusion in the model angiosperm *Arabidopsis thaliana* grown in different light environments.
+We find that 1) stomata are overdispersed, but not ideally dispersed, on both leaf surfaces across all light treatments; 2) the patterning of stomata on abaxial and adaxial leaf surfaces is independent; and 3) the theoretical improvements to photosynthesis from abaxial-adaxial stomatal coordination are miniscule ($\\ll 1$%) across the range of feasible parameter space.
 However, we also find that 4) stomatal size is correlated with the mesophyll volume that it supplies with CO$_2$, suggesting that plants may optimize CO$_2$ diffusion limitations through alternative pathways other than ideal, uniform stomatal spacing.
-We discuss the developmental, physical, and evolutionary constraits which may prohibit plants from reaching the theoretical adaptive peak of uniform stomatal spacing and inter-surface stomatal coordination.
+We discuss the developmental, physical, and evolutionary constraits which may prohibit plants from reaching this theoretical adaptive peak of uniform stomatal spacing and inter-surface stomatal coordination.
 These findings contribute to our understanding of variation in the anatomy of amphistomatous leaves."
@@
 Stomatal conductance and transpiration are determined by numerous environmental and anatomical parameters such as vapor pressure deficit (VPD), irradiance, temperature, wind speed, leaf water potential, IAS geometry, mesophyll cell anatomy, and stomatal anatomy.
+The latter of these is the focus of this study, with discussion of other interacting variables.
 
@@
 Total stomatal area (size $\times$ density) is optimized for operational conductance ($g_\text{s,op}$) rather than maximum conductance ($g_\text{s,max}$) such that stomatal apertures are most responsive to changes in the environment at their operational aperture [@franks_physiological_2012; @liu_scaling_2021].
-Stomatal aperture can compensate for maladaptive stomatal densities to an extent [@bussis_stomatal_2006], but stomatal density and size ultimately determine a leaf's theoretical $g_\text{s,max}$ [@sack_developmental_2016], which is proportional to $g_\text{s,op}$ [@murray_consistent_2020].
-Additionally, low stomatal densities lead to irregular and insufficient CO$_2$ supply and reduced photosynthetic efficiency in leaf areas far from stomata [@pieruschka_lateral_2006; @morison_lateral_2005], while high stomatal densities can reduce water use efficiency (WUE) [@bussis_stomatal_2006] and incur excessive metabolic costs [@deans_optimization_2020].
+Stomatal aperture can compensate for maladaptive stomatal densities to an extent [@bussis_stomatal_2006], but stomatal density and size ultimately determine a leaf's theoretical $g_\text{s,max}$ [@sack_developmental_2016], which is proportional to $g_\text{s,op}$ [@mcelwain_using_2016; @murray_consistent_2020].
+Additionally, low stomatal densities lead to irregular and insufficient CO$_2$ supply and reduced photosynthetic efficiency in leaf areas far from stomata [@pieruschka_lateral_2006; @morison_lateral_2005], while high stomatal densities can reduce water use efficiency (WUE) [@bussis_stomatal_2006] and incur excessive metabolic costs [@de_boer_optimal_2016; @deans_optimization_2020].
 Stomatal density positively co-varies with irradiance during leaf development and negatively co-varies with CO$_2$ concentration [@gay_influence_1975; @schoch_dependence_1980; @woodward_stomatal_1987; @royer_stomatal_2001], consistent with optimality predictions.
 In most species, stomata occur only on the abaxial (usually lower) leaf surface; but amphistomy, the occurrence of stomata on both abaxial and adaxial leaf surfaces, is also prevalent in high light environments with constant or intermittent access to sufficient water [@mott_adaptive_1982; @jordan_using_2014; @muir_light_2018; @drake_two_2019; @muir_is_2019].
-Amphistomy effectively halves the CO$_2$ diffusion path length and boundary layer resistance by doubling boundary layer conductance [@parkhurst_adaptive_1978; @harrison_influence_2020; @mott_amphistomy_1991].
+Amphistomy effectively halves the CO$_2$ diffusion path length and boundary layer resistance by doubling boundary layer conductance [@parkhurst_adaptive_1978; @mott_amphistomy_1991 @harrison_influence_2020].
+Ab- and adaxial leaf surfaces were found to function independent of one another in wheat, an important crop, with the adaxial surface demonstrating higher photosynthetic capacity [@wall_stomata_2022].
+These results highlight the utmost importance of amphistomy for some plants.
 Despite these optimality predictions, stomatal anatomy may be partially constrained by physical and developmental limits on phenotypic expression [@croxdale_stomatal_2000; @harrison_influence_2020; @muir_how_2023].
@@
 This would prevent stomata from being strongly clustered; however, some species (notably in *Begonia*) appear to benefit from the overlapping vapor shells caused by stomatal clustering in dry environments [@yi_gan_stomatal_2010; @lehmann_effects_2015; @papanatsiou_stomatal_2017].
-Historically, stomatal patterning in dicot angiosperms was thought to be random with an exclusionary distance surrounding each stomate [@sachs_developmental_1974]; however, the developmental controls of stomatal patterning are more complex than random development along the leaf surface.
+Historically, stomatal patterning in dicot angiosperms was thought to be random with an exclusionary distance surrounding each stomate [@sachs_developmental_1974]; however, the developmental controls of stomatal patterning are more complex than was first thought.
 @croxdale_stomatal_2000 reviews three developmental theories which attempt to explain stomatal patterning in angiosperms: inhibition, cell lineage, and cell cycle, ultimately arguing for a cell cycle based control of stomatal patterning.
@@
 Low light decreases CO$_2$ demand by limiting electron transport rate, leading to relatively high internal CO$_2$ concentration ($C_\text{i}$) and low $A_\text{max}$ [@kaiser_metabolic_2016].
-In contrast, well hydrated leaves with open stomata in high light, photosynthesis is often limited by CO$_2$ supply as resistances from the boundary layer, stomatal pore, and mesophyll can result in insufficient CO$C_2$ supply at the chloroplast to maxmimize photosynthesis [@farquhar_biochemical_1980; @lehmeier_cell_2017].
-In this study, we focus primarily on how stomatal patterning affects diffusion, ignoring boundary layer and mesophyll resistances.
+In contrast, well hydrated leaves with open stomata in high light, photosynthesis is often limited by CO$_2$ supply as resistances from the boundary layer, stomatal pore, sub-stomatal cavity, and mesophyll can result in insufficient CO$C_2$ supply at the chloroplast to maxmimize photosynthesis [@farquhar_biochemical_1980; @lehmeier_cell_2017].
+In this study, we focus primarily on how stomatal patterning affects diffusion.
 
+Assuming uniform mesophyll diffusion resistance in all directions (porous medium), an ideal stomatal anatomy can be predicted.
 To maximize CO$_2$ supply from the stomatal pore to chloroplasts, stomata should be uniformly distributed in an equilateral triangular grid on the leaf surface so as to minimize stomatal number and CO$_2$ diffusion path length [@parkhurst_diffusion_1994].
-As the diffusion rate of CO$_2$ though liquid is approximately $10^4\times$ slower than CO$_2$ diffusion through air, mesophyll resistance is generally thought to be primarily limited by liquid diffusion [@aalto_three-dimensional_2002; @evans_resistances_2009], but diffusion through the IAS has also been shown to be a rate limiting process because the tortuous, disjunct nature of the IAS can greatly increase diffusion path lengths [@harwood_understanding_2021].
+With this assumption, an equilateral triangular grid is ideal because it standardizes the distance between stomata (for a given stomatal density) and thereby also standardizes the mesophyll volume with which each stomate is spatially closest to, ultimately reducing redundancy associated with clustered stomata and allowing fewer stomata to supply the same leaf area (or the same number of stomata to adequately supply a larger leaf area).
+
+Such an assumption, though an oversimplification, is a powerful tool for photosynthesis modelling, and may provide insight into how real leaves diverge from this.
+In real leaves, as the diffusion rate of CO$_2$ though liquid is approximately $10^4\times$ slower than CO$_2$ diffusion through air, mesophyll resistance is generally thought to be primarily limited by liquid diffusion [@aalto_three-dimensional_2002; @evans_resistances_2009], but diffusion through the IAS has also been shown to be a rate limiting process because the tortuous, disjunct nature of the IAS can greatly increase diffusion path lengths [@harwood_understanding_2021].
 Additionally, tortuosity is higher in horizontal directions (parallel to leaf surface) than vertical directions (perpendicular to leaf surface) because of the cylindrical shape and vertical arrangement of pallisade mesophyll cells [@earles_beyond_2018; @harwood_understanding_2021].
-However, the ratio of lateral to vertical diffusion rate is still largely unknown [@morison_lateral_2005; @pieruschka_lateral_2005; @pieruschka_lateral_2006].
+However, the ratio of lateral to vertical diffusion rate is still largely unknown and may be a highly pliable trait in leaves [@morison_lateral_2005; @pieruschka_lateral_2005; @pieruschka_lateral_2006; @morison_does_2007].
 Depending on the thickness of the leaf, porosity of the leaf mesophyll, tortuosity of the IAS, and lateral to vertical diffusion rate ratio, minimizing diffusion path length for CO$_2$ via optimally distributed stomata may yield significant increases in CO$_2$ supply for photosynthesis and higher $A_\text{max}$.
+Or plants may simply coordinate the development of stomata and mesophyll IAS to reach another optimal solution which does not rely on uniformly distributed stomata [@baillie_developmental_2020].
 
-We hypothesized that natural selection will favor stomatal patterning and distribution to minimize the diffusion path length.
-In amphistomatous leaves, this would be accomplished by 1) a dispersed, equilateral triangular distribution of stomata on both abaxial and adaxial leaf surfaces and 2) coordinated stomatal spacing on each surface that offsets the position of stomata (\autoref{fig:ideal-amphi-grid}).
+We hypothesized that, in the absence of any constraint and assuming uniform mesophyll diffusion resistance, natural selection will favor stomatal patterning and distribution to minimize the diffusion path length.
+In amphistomatous leaves, this would be accomplished by 1) a uniform, equilateral triangular distribution of stomata on both abaxial and adaxial leaf surfaces and 2) coordinated stomatal spacing on each surface that offsets the position of stomata (\autoref{fig:ideal-amphi-grid}).
 Coordination between leaf surfaces is defined in this study as the occurrence of stomata in areas farthest from stomata on the opposite leaf surface.
-Additionally, because CO$_2$ is more limiting for photosynthesis under high light, we hypothesize that in high light 3) there should be more stomata, and 4) stomata should be more uniformly distributed than in low light.
-Finally, as stomatal densities are selected for optimal operational aperture, we hypothesize that 5) stomatal length will be positively correlated with the area of the leaf surface to which it is closest.
+Additionally, because CO$_2$ is more limiting for photosynthesis under high light, we hypothesize that in high light 3) there should be more stomata, and 4) stomata should be more overdispersed compared to a random distribution than in low light.
+Finally, since, in measures of whole leaves, stomatal area rather than stomatal density is optimized for operational conductance, we hypothesize that 5) stomatal length (and hence its area) will be positively correlated with the area of the leaf surface to which it is spatial closest as defined by Voronoi tessellation techniques.
 We refer to this as the 'stomatal zone', the leaf area surrounding a focal stomate closest to that stomate and therefore the zone it supplies with CO$_2$).
-This way, each stomate can be optimally sized relative to the mesophyll volume it supplies.
+This way, each stomate may be optimally sized relative to the mesophyll volume it supplies.
 
@@
 Specifically, we predicted that traits which affect diffusion path length (leaf thickness, stomatal density, leaf porosity, lateral-vertical diffusion rate ratio), diffusion rate (temperature, pressure), and CO$_2$ demand (Rubisco concentration, light) would modulate the advantage of optimal stomatal arrangement following the relationships outlined in \autoref{tab:hypotheses}.
-Here, we integrate over reasonable parameter space to determine the ecophysiological context most likely to favor stomatal spatial coordination in amphistomatous leaves.
+Here, we integrate over reasonable parameter space to determine the ecophysiological context most likely to favor stomatal coordination in amphistomatous leaves.
 
@@
 We captured 132 images in total, making 66 abaxial-adaxial image pairs.
-We measured stomatal position and size using ImageJ [@schneider_nih_2012].
+Images were square with an area of 0.386 mm$^2$.
+We measured stomatal position and length using ImageJ [@schneider_nih_2012].
+A number of synthetic leaf surface datasets were also prepared (details below) so as to be mathematically explicit when defining null distributions against which to test our hypotheses and to avoid any methodological influence on our results (e.g.
+stomatal density uncertainty, boundary effects when calculating patterning metrics: uniform random, non-random overdispersion, and uniform).
+All synthetic leaf surfaces were simulated based on the size of the real leaf images and stomatal densities matched those of real leaf images.
 
@@
 
-We tested whether stomata are non-randomly distributed by comparing the observed stomatal patterning to a random uniform pattern.
-For each leaf surface image with $n$ stomata we generated $10^3$ synthetic surfaces with $n$ stomata uniformly randomly distributed on the surface.
+We tested whether stomata are non-randomly distributed by comparing each observed, real leaf stomatal pattern to an array of synthetic random uniform patterns.
+For each observed leaf surface image with $n$ stomata we generated $10^3$ synthetic surfaces with $n$ stomata uniformly randomly distributed on the surface.
 For each sample image, we compared the observed Nearest Neighbor Index ($\mathrm{NNI}$) to the null distribution of $\mathrm{NNI}$ values calculated from the synthetic data set.
@@
 
-For each sample image, we also simulated $10^3$ synthetic data with $n$ stomata ideally dispersed in an equilateral triangular grid.
-For these grids, we integrated over plausible stomatal densities and then conditioned on stomatal grids with exactly $n$ stomata.
+For each sample image, we also simulated $10^3$ synthetic leaf surfaces with $n$ stomata ideally, uniformly dispersed in an equilateral triangular grid.
+To account for uncertainty in the stomatal density of each sample image with $n$ stomata, we integrated over plausible stomatal densities and then conditioned on synthetic leaf surfaces with exactly $n$ stomata.
 The simulated stomatal count was drawn from a Poisson distribution with the mean parameter $\lambda$ drawn from a Gamma distribution with shape $n$ and scale 1 ($\lambda \sim \Gamma(n, 1)$).
 $\Gamma(n, 1)$ is the posterior distribution of $\lambda$ with a flat prior distribution.
-This allows us to integrate over uncertainty in the stomatal density from the sample image.
+This integration was necessary to remove any artifacts of uncertainty in the true stomatal density of the sample leaves.
 
-We developed a dispersion index $\mathrm{DI}$ to quantify how close observed stomatal distributions are to random uniform versus maximally dispered in an equilateral triangular grid.
+We developed a dispersion index $\mathrm{DI}$ to quantify how close observed stomatal distributions are to random uniform versus ideally dispered in an equilateral triangular grid.
 $\mathrm{DI}$ varies from zero to one, where zero is uniformly random and one is ideally dispersed:
@@
 
-<!-- CDM: I am not sure this part is needed
-Uniformly distributed stomata on a leaf with stomatal density ($D_S$) maximize $\overline{D}_O$ in an equilateral triangle pattern with side lengths ($s$) equal to two times the incircle radius or apothem, $a$, of a regular hexagon with area ($A_\text{hex}$) equal to $A_\text{leaf}$ divided by the number of stomata ($n_\text{stomata}$) on the leaf according to Eq.
-\ref{eq:eq2}:
-
-\begin{equation}\label{eq:eq2}
-  s = \frac{\sqrt{2}} {3^{1/4} \sqrt{n_\text{stomata} / A_\text{leaf}}}
-\end{equation}
-
-As all sides of an equilateral triangle have the same length, the average nearest neighbor distance for a given $D_S$ is maximized and each stomate occupies the same hexagonal area ($A_\text{hex}$).
-As $DI$ approaches 1, the more uniform an area each of its stomata supply with CO$_2$ during photosynthesis.
- -->
-
-Finally, we examined the relationship between stomatal zone area and stomatal length using a Bayesian linear mixed-effects model fit with the *R* package **brms** version `r fit$version$brms` [@burkner_brms_2017; @burkner_advanced_2018] and *Stan* version `r fit$version$cmdstan` [@stan_development_team_stan_2023].
+Finally, we examined the relationship between stomatal zone area and stomatal length of real stomata using a Bayesian linear mixed-effects model fit with the *R* package **brms** version `r fit$version$brms` [@burkner_brms_2017; @burkner_advanced_2018] and *Stan* version `r fit$version$cmdstan` [@stan_development_team_stan_2023].
 Stomatal zone area was calculated using Voronoi tessellation (e.g.
@@
 \caption{Examples of synthetic and real leaf surfaces.
-A) Uniform random synthetic leaf surface; B) Example of real leaf surface; C) Regularly distributed synthetic leaf surface.
+A) Uniform random synthetic leaf surface; B) Example of real leaf surface; C) Uniformly distributed synthetic leaf surface.
 The zone defined by each stomate was calculated with voronoi tessellation and correlated with stomatal length in real leaves.}
@@
 To test whether the position of ab- and adaxial stomata are coordinated we compared the observed distribution to a null distribution where the positions on each surface are random.
-For each pair of surfaces (observed or synthetic) we calculated the distance squared between each pixel to the nearest stomatal centroid with the *R* package **raster** version `r packageVersion("raster")`.
+For each pair of surfaces (observed or synthetic) we calculated the distance squared between each pixel of the surface to the nearest stomatal centroid with the *R* package **raster** version `r packageVersion("raster")`.
 We refer to this as the 'nearest stomatal distance' or NSD.
@@
 For each synthetic data set, we simulated stomatal position using a random uniform distribution, as described above, matching the number of stomata on abaxial and adaxial leaf surfaces.
-Stomatal positions on each surface are coordinated if the correlation coefficient is greater than 95% of the synthetic correlation values (one-tailed test).
+Stomatal positions on each surface are coordinated if the correlation coefficient of the NSD between observed ab- and adaxial surfaces is greater than 95% of the synthetic correlation values (one-tailed test).
 
@@
 
-## Stomatal distribution is non-random, but far from ideal
+## Stomatal distribution is non-random, but far from uniform
 
-Many leaf surfaces (`r nni2[['n1']]` of `r nni2[['n1']] + nni2[['n2']]`, `r round(100 * nni2[['n1']] / (nni2[['n1']] + nni2[['n2']]), 1)`%) are significantly overdispersed compared to a random uniform distribution, but none were close to an ideal hexagonal pattern (dispersion index = 1; \autoref{fig:single-surface}).
+Many leaf surfaces (`r nni2[['n1']]` of `r nni2[['n1']] + nni2[['n2']]`, `r round(100 * nni2[['n1']] / (nni2[['n1']] + nni2[['n2']]), 1)`%) are significantly overdispersed compared to a random uniform distribution, but none were close to a uniform equilateral triangular pattern (dispersion index = 1; \autoref{fig:single-surface}).
 Before controlling for multiple comparisons, `r round(100 * nni1[['n1']] / (nni1[['n1']] + nni1[['n2']]), 1)`% are significantly overdispersed.
@@
 \includegraphics[width=4in]{figures/single-surface.pdf}
-\caption{Stomata are more dispersed than expected under the null model of uniform random position (dispersion index = 0) but far from a distribution that maximizes distance between stomata (dispersion index = 1).
+\caption{Stomata are more dispersed than expected under the null model of uniform random position (dispersion index = 0) but far from a distribution that maximizes distance between stomata (dispersion index = 1; uniform distribution).
 We determined statistical significance between light treatments using Tukey post-hoc tests.
@@
 
-All parameters converged ($\hat{R} < 1.01$) and effective sample sizes were exceeded $10^3$.
+All parameters in the Bayesian linear mixed-effects model converged ($\hat{R} < 1.01$) and effective sample sizes exceeded $10^3$.
 Across all light treatments and leaf surfaces, stomatal length and stomatal area are weakly positively correlated (\autoref{fig:length-area}).
@@
 In principle, spacing stomata to minimize the average distance between stomatal pores and chloroplasts within the mesophyll should increase carbon gain, all else being equal.
-However, reducing this distance to its absolute minimum may be constrained by developmental processes, or the photosynthetic benefit may be too small to be 'seen' by natural selection (i.e.
-the selection coefficient is less than drift barrier *sensu* @sung_drift-barrier_2012).
+However, reducing this distance to its absolute minimum may be constrained by developmental processes, the photosynthetic benefit may be too small to be 'seen' by natural selection (i.e.
+the selection coefficient is less than drift barrier *sensu* @sung_drift-barrier_2012), or alternative peaks in the adaptive landscape may be reached as the adaptive peak investigated in this study is based on oversimplistic assumptions of a porous medium mesophyll.
 
@@
 First, we predicted that stomata on each surface are overdispersed relative to a random uniform distribution, which should increase CO$_2$ supply.
-Stomata on each surface are overdispersed (\autoref{fig:single-surface}), but are not ideally dispersed in an equilateral triangular grid as would be optimal to minimize CO$_2$ diffusion path length and equalize the area supplied by each stomate (\autoref{fig:tessellation-example}).
+Stomata on each surface are overdispersed (\autoref{fig:single-surface}), but are not ideally, uniformly dispersed in an equilateral triangular grid as would be optimal to minimize CO$_2$ diffusion path length and equalize the area supplied by each stomate (\autoref{fig:tessellation-example}).
 Second, we predicted that an optimal amphistomatous leaf has offset stomata such that stomata are more likely to appear on one leaf surface if there is not a stomata directly opposite it on the other surface as shown in \autoref{fig:ideal-amphi-grid}.
@@
 Finally, we predicted that within leaf variation in stomatal size would correlate with stomatal spacing, as larger stomata can supply larger volumes of adjacent mesophyll.
-In all three light treatments, stomatal size positively covaried with the stomatal zone, i.e.
+In all three light treatments, stomatal size positively co-varied with the stomatal zone, i.e.
 adjacent region of mesophyll that would be supplied by that stomate (\autoref{fig:length-area}).
@@
 There are three non-mutually exclusive hypotheses for why several of our predictions were wrong.
-First, our predictions are wrong becuase they are based on overly simplistic assumptions about epidermal and mesophyll anatomy.
-Second, natural selection may be constrained by developmental processes that prevent phenotypes from reaching their adaptive optima.
-Third, the benefit of some traits may be of too little consequence to result in fitness differences large enough to respond to selection.
+First, our predictions are wrong because they are based on overly simplistic assumptions about epidermal and mesophyll anatomy.
+In real leaves, diffusion through the mesophyll depends on the position of each cell and resulting IAS geometry.
+If, for example, horizontal diffusion was modeled to be much slower than vertical diffusion, we likely would have seen an increased benefit of ab- adaxial stomatal coordination across a greater range of feasible trait values.
+However, further information about the three-dimensional structure of the leaves would need to be incorporated into the model to address any further questions about how mesophyll anatomy may influence stomatal anatomy.
+This represents an interesting area for further research.
+Second, spatio-temporal variation of internal conditions within leaves and between stomatal responses may make uniform, coordinated stomatal surfaces less beneficial [@lawson_nature_1998; @lawson_spatial_1999].
+This is because our model assumes a uniform leaf, the internal conditions of which are periodic and solved empirically and therefore stable.
+Any horizontal concentration gradients due to environmental heterogeneity and variable induction times for interacting leaf processes may dissolve any benefit of uniform stomatal patterning.
+Third, natural selection may be constrained by developmental processes that prevent phenotypes from reaching their adaptive optima.
+Stomatal development must be plastic to environmental ques interpreted through long distance and cell to cell signalling pathways [@pillitteri_mechanisms_2012].
+This plasticity may come with the cost of being unable to orchestrate the development of an absolutely uniform stomatal grid.
+Fourth, the benefit of some traits may be of too little consequence to result in fitness differences large enough to respond to selection.
 We consider the plausibility of these alternative hypotheses below and present ideas for future work to test them.
 
-We assume an idealized leaf epidermal and mesophyll structure that is homogenous and unconstrained by other tradeoffs.
+Our model also assumes an idealized leaf epidermal and mesophyll structure that is homogenous and unconstrained by other tradeoffs.
 Real leaves not only provide pathways for CO$_2$ diffusion, but must supply water, intercept light, and deter herbivores and pathogens.
-These competing interests result in nonuniform epidermal and mesophyll structure that could alter our predictions about optimal stomatal spacing.
+All of these competing processes also happen on different time scales, and can be observed as heterogeneity in stomatal density, aperature, and internal leaf conditions across the leaf at any given moment [@lawson_nature_1998; @lawson_spatial_1999].
+These competing interests result in non-uniform epidermal and mesophyll structure that could alter predictions about optimal stomatal spacing.
 In order to maintain consistent leaf water potential across the lamina, stomatal density must be coordinated with vein density [@fiorin_transport_2016].
@@
 This may explain why adaxial leaf surfaces have consistently higher dispersion indices than abaxial surfaces across all light treatments (\autoref{fig:single-surface}).
-Future gas exchange models should include multiple parameter sets for different strata of mesophyll tissue.
+Future gas exchange models should include multiple parameter sets for different strata of mesophyll tissue and water supply structures such as veins, though these are beyond the scope of this study.
 
@@
 Our study corroborates previous studies which demonstrate that stomata are non-randomly distributed along the leaf surface as a result of developmental mechanisms such as spatially biased arrest of stomatal initials [@boetsch_arrest_1995], oriented asymmetric cell division [@geisler_oriented_2000], and cell cycle controls [@croxdale_stomatal_2000].
-We do not investigate the potential developmental pathways that influence stomatal dispersion in this study; however, they are important to consider as these pathways could limit plants from reaching a theoretical peak in the adaptive landscape: uniform stomatal dispersion.
+We do not investigate the potential developmental pathways that influence stomatal dispersion in this study; however, they are important to consider as these pathways could limit plants from reaching a theoretical peak in the adaptive landscape: uniform stomatal patterning.
 Instead, as this study suggests, plants may simply compensate with higher stomatal density and by fitting stomatal size to the area that they supply with CO$_2$.
@@
 This may be especially true in highly variable environments or in large tree species with sun and shade leaves where developmental cues may change rapidly.
-The temporal component, not considered here, could also have significant implications, as CO$_2$ may only be limiting to photosynthesis during short periods when all other conditions are ideal.
+The temporal component, not considered here, could also have significant implications, as CO$_2$ may only be limiting to photosynthesis during short, relatively rare periods when all other conditions are ideal.
 In these cases, the theoretical benefits of ideal stomatal spacing are further diminished.

### 2024-01-21 ms/response-to-reviewers.qmd
revised ms and response to reviewers

---
format: 
  pdf:
    # cite-method: biblatex
    include-in-header: 
      - preamble.tex
header-includes: |
  \addtokomafont{disposition}{\rmfamily}
  \RedeclareSectionCommand[
    font=\normalfont\Large]{section}
  \RedeclareSectionCommand[
    font=\normalfont\normalsize\bfseries]{subsection}
  \RedeclareSectionCommand[
    font=\normalfont\normalsize\itshape]{subsubsection}
  \RedeclareSectionCommand[
    font=\normalfont\normalsize]{paragraph}
editor: visual
bibliography: stomata-spacing.bib
csl: https://www.zotero.org/styles/annals-of-botany
nocite: |
  @borsuk_structural_2022, @croxdale_stomatal_2000, @dow_disruption_2017, @earles_beyond_2018, @harwood_understanding_2021, @therouxrancourt_bias_2017 
---

**Covering Letter for a Revised Manuscript**

*Manuscript reference number: 23184*

Thank you for considering our revised manuscript. We have responded to all comments in \textcolor{blue}{blue font} below.

-- Jacob Watts, on behalf of all authors

------------------------------------------------------------------------

## Response to Associate Editor

Associate Editor Evaluations: Recommendation: Accept Revision Required

Associate Editor (Comments for the Author (Required)):

Dear authors,

My apologies for the delay in returning this manuscript, as you can understand it can be difficult to find reviewers at this time of year. We have now received two comprehensive reviews of your manuscript. Both reviewers found the science to be interesting and thought provoking. I agree, it was a very interesting read. However, reviewer 1 found it difficult to understand some of the terminology and methods and reviewer 2 has requested clarity on some of the assumptions in your modeling. Both have provided detailed comments for you to address. Could you please revise the manuscript and address their comments before resubmission to AoBP.

Kind regards, William

```{=tex}
\textcolor{blue}{Thank you so much for organizing a review of this manuscript. The reviewer's comments have been extremely useful in improving the manuscript for publication in \textit{AoBP}. We appreciate your help in getting this science in print. We summarize our changes here; details on specific comments can be found below.\\
\\
We have clarified our terminology, adding in definitions where needed and standardizing all uses of specific terms. We've clarified our methodology by being more explicit about which data are real and which are synthetic. We have added all the references and relevant discussion of them as suggested by reviewer 1. We've clarified the wording of our hypotheses following the suggestion of reviewer 1. And, as was addressed by both reviewers 1 and 2, we've clarified the limitations of our model assumptions and extensively addressed them in the introduction and discussion. To accommodate some of these changes, we've made some minor changes elsewhere in the text; however, the largest changes are all reflected in this document. We hope that, with these significant clarifications, changes, and additions of important caveats, our manuscript will be considered for publication in \textit{AoBP}. Thank you for your time and consideration.}
```

## Response to Reviewer #1

The MS explore stomatal patterning on both adaxial and abaxial leaf surfaces in relative to CO2 diffusion and photosynthesis using both observations and modelling approaches. This is a topical area at the moment and some of the findings interesting and valuable to research in this area. However several sections of the MS were difficult to follow and further clarification of several aspects is required to widen the readership.

There are several descriptions used that are not entirely clear. What exactly is meant by "over dispersal" in the context of stomatal patterning. Does this mean there are too many stomata on the leaf or their spacing is not optimal but the correct number? This is not common terminology when referring to stomatal patterning. The assessment of distribution relative to a uniform random distribution needs further explanation as it is well established that stomata are randomly distributed.

**Response:** \textcolor{blue}{We appreciate the reviewer's time to read and comment on our manuscript. They have brought up important points of clarification that must be addressed before the manuscript is ready to be published.}

\textcolor{blue}{We have clarified and standardized our terminology, adding explanations when needed. It has long been known that stomata are non-randomly distributed, tending to be more uniformly distribution than random (see Croxdale et al. (2000) for a nice review of historical evidence of stomatal patterning). In Croxdale et al. (2000), there is also confusion over terminology in the published literature, so we are not sure that there are field-accepted terms. So, we have adopted our own set of terminology which mirrors that of the field and related fields which discuss the spatial patterns of discrete objects. Specifically, see lines 111-117:}

> "We compared observed stomatal patterning to an ideal pattern (uniform equilateral triangular grid) and a null model (random uniform distribution). The terminology is unfortunately confusing because the word `uniform' is used in different ways. A uniform equilateral triangular grid means that the distance between stomata is uniform; a random uniform distribution means that a stomate has an equal probability (i.e. uniform) of occurring anyone on the leaf surface. To limit confusion, we refer to the ideal pattern as uniform (equilateral triangle grid) and the null pattern as random (random uniform). When observed stomatal patterns are more dispersed than expected under random patterning, we refer to this as over-dispersed. Note, however, that over-dispersed compared to random is still less dispersed than ideal, because the ideal pattern is maximally dispersed."

In the material and methods section it is often not clear which measurements/data are real and which are synthesized.

**Response:** \textcolor{blue}{We have edited the methods section to be more explicit about which type of data are which. We’ve also added a sentence in the beginning of the methods on lines 106-109 describing how synthetic data fit into our analyses. It reads:}

> "A number of synthetic leaf surface data sets were also simulated (details below) to generate null distributions against which to test our hypotheses and to avoid any methodological influence on our results (e.g. boundary effects when calculating stomatal patterning). All synthetic leaf surfaces were simulated based on the size of the real leaf images and stomatal densities matched those of real leaf images."

There are several references missing I am surprised not to see included in the introduction and/or discussion- see specifics below.

Line 19 - should include earlier references that examined Gopp - McElwain et al., 2016

Line 24 - should include the work of De Boar et al.

Line 59 - there are several other papers that have examined lateral flux and the impact of stomata and photosynthetic consumption, including Morison et al 2007.

The work of Wall et al., 2021 is missing and should be included.

The authors have not mentioned or considered the published references that have explored heterogeneity in stomatal distribution over the leaf surface and between surfaces (see work by Weyers et al.).

**Response:** \textcolor{blue}{We appreciate the reviewer bringing these citation gaps to our attention. We have added all the references that they mention here and the appropriate discussion.}

How do these findings for Arabidopsis translate to crops. Arabidopsis grows very close the ground and therefore boundary layers esp for the abaxial surface will be very different to key C3 crops.

**Response:** \textcolor{blue}{Translating these findings to crops is beyond the scope of our study, however, these results may prove to be relevant to crops upon further investigation. We agree that the boundary layer dynamics may differ drastically between the short \textit{A. thaliana} and those of most crops. We added:}

> "However, it would be necessary to account for many other differences between *A. thaliana* and crop leaves and canopies." (lines 295-296)

Hypothesis 3 - it is well established that high light alters stomatal density.

**Response:** \textcolor{blue}{We addressed this on lines 83-84:}

> "Hypothesis 3 is already well supported in many species [@poorter_metaanalysis_2019], but it is useful here to confirm that light treatments induced plasticity in the expected direction."

The wording of hypothesis 5 is particularly unclear. "5) stomatal length will be positively correlated with the area of the leaf surface to which it is closest". What is the area of the leaf surface to which is it closest?

**Response:** \textcolor{blue}{Thanks for bringing up this potential ambiguity in our hypothesis. It now reads (lines 81-82):}

> "Stomatal length (and hence its area) will be positively correlated with the area of the leaf surface to which it is spatial closest as defined by Voronoi tessellation techniques. We refer to this as the 'stomatal zone', the leaf area surrounding a focal stomate closest to that stomate and therefore the zone it supplies with CO$_2$). This way, each stomate may be optimally sized relative to the mesophyll volume it supplies."

The high light used in the experiment is quite low at 200 umol m-2 s-1

**Response:** \textcolor{blue}{$200~\mu \text{mol}~\text{m}^{-2}~\text{s}^{-1}$ is relatively low; however, it was high enough to drastically increase stomatal density, suggesting a change in many other leaf parameters (Dow et al. 2017). We added on lines 97-99 that:}

> "*A. thaliana* responds strongly to light levels over this range [@bailey_acclimation_2001], though natural populations in open canopies can experience $\text{PAR} > 800~\mu \text{mol}~\text{m}^{-2}~\text{s}^{-1}$ [@callahan_shade-induced_2002]."

Line 52 - provide more explanation as to why the equilateral triangular grid is the ideal distribution.

**Response:** \textcolor{blue}{On lines 54-59, we added that:}

> "An equilateral triangular grid is ideal because it maximizes the average distance between stomata, for a given stomatal density, and thereby minimizes the average distance between any point in the mesophyll to its nearest stomate. Assuming a homogenous mesophyll, this is the most efficient pattern to supply CO$_2$ to a leaf volume."

Line 94 - details of the size of the image need to be included.

**Response:** \textcolor{blue}{Thanks for catching this! We've now added the text in the appropriate spot on line 106. It now reads: ``Images were square with an area of 0.386 mm$^2$.''}

Line 95 - sample image - is this real or synthetic? Terminology used needs to be clear and consistent.

**Response:** \textcolor{blue}{As noted above in response to a previous comment, we have clarified when we are using observed and synthetic data throughout. Sample image is our standard term for real leaf surface. Synthetic leaf surface is our standard term for simulated leaf surfaces against which the real leaves were compared.}

Line 105 - please provide further clarification on why the ideal dispersal is an equilateral triangle.

**Response:** \textcolor{blue}{We believe that the clarification about why this pattern is ideal has been addressed in response to the comment on line 52 above.}

Line 106 - wording of this sentence is difficult to follow.

**Response:** \textcolor{blue}{We've changed it be more clear following your suggestion on lines 132-133: "To account for uncertainty in the stomatal density of each sample image with $n$ stomata, we integrated over plausible stomatal densities and then conditioned on synthetic leaf surfaces with exactly $n$ stomata."}

Line 129 - each pixel of what?

**Response:** \textcolor{blue}{It now reads: ``pixel of the surface'' on line 156.}

Line 157 - ideal pattern is "hexagonal"?

**Response:** \textcolor{blue}{Thanks for catching this. We’ve now standardized this term to be the same as every other mention of the uniform, equilateral triangular grid. The area of leaf surface closest to a given stomate, when stoamta are arrayed in an equilateral triangular grid, forms a hexagon (see Fig. 2C).}

Line 168 – "All parameters....exceeded 103" – please clarify what you mean by this statement.

**Response:** \textcolor{blue}{Thank you for all of your comments. We’ve clarified this statement on line 195:}

> "All parameters in the Bayesian linear mixed-effects model converged ($\hat{R} < 1.01$) and effective sample sizes exceeded $10^3$"

## Response to Reviewer #2

Dear authors,

The authors made a theoretical analysis of the optimal distribution of stomata on each leaf surface and then compared it with actual data. I only have a few comments, most of them concerning the assumptions. I think the consequences of their assumptions could be better addressed and explained with further details, especially the concept of “optimal distribution”.

Please see my comments below.

**Response:** \textcolor{blue}{We really appreciate these comments as they are most insightful and will certainly make this manuscript better and more useful for its readership. If we've interpreted the concerns correctly, the main issue with the manuscript is that the limitations brought upon by our model assumptions, particularly concerning uniform mesophyll tortuosity/diffusion resistance and its consequences for our definition of ``ideal, uniform'' stomatal patterning, are not adequately explained in the introduction and discussion. We agree that these caveats were not discussed in great enough detail and have added explanation in the revised manuscript, as described in response to comments below. Nevertheless, we believe that our approach of using an idealized model that generates quantitative, testable predictions is a useful starting point for investigation. Identifying which anatomical features deviate most from our predictions provides useful information about which assumptions we might relax in the future.}

L64-65. The impact of varying tortuosity (horizontal vs. vertical) on stomatal distribution needs clearer exploration. Specifically, how do differences in adaxial and abaxial tortuosity affect your analysis? I guess this must be in coordination with different horizontal and vertical tortuosity, as the authors mentioned: if the horizontal tortuosity is higher than the vertical tortuosity, this consideration should vary, especially if the tortuosity in the adaxial and abaxial half are different.

**Response:** \textcolor{blue}{We have addressed this and the related concept of porosity in the discussion (lines 234-242):}

> "First, our predictions must be wrong because they are based on overly simplistic assumption of a homogenous porous medium within the mesophyll. Real leaf mesophylls are spatially heterogenous and chloroplasts are distributed as discrete nodes. The intercellular air space conductance is determined by its porosity and tortuosity, both of which are heterogenous within the leaf. The palisade is typically less porous than the spongy mesophyll (e.g. Théroux-Rancourt et al. 2017), which should impact the optimal patterning on stomata on ab- versus adaxial surfaces. Tortuosity is also systematically greater in the palisade in the lateral direction parallel to the leaf plane (Harwood et al. 2021). We might predict a greater coordination advantage of offset stomata by accounting for greater lateral tortuosity, but it is likely that benefit is still very small under realistic parameter space. Quantifying the patterns of heterogeneity in porosity, tortuosity, and other factors (Earles et al. 2018) using 3D imaging (e.g. Borsuk et al. 2022) will be needed to generate more realistic hypotheses about optimal stomatal spacing."

L64-66 (hypotesises 1 and 2). Stomata distribution and tortuosity of the mesophyll also contribute to modulating transpiration, not only CO2 distribution. Did the authors account for this?

**Response:** \textcolor{blue}{We agree that incorporating transpiration is needed to better model optimal traits, but this is beyond the scope of the current study since we have not included venation and there is substantial empirical uncertainty about sites of evaporation within leaves and the prevalence of subsaturated mesophyll airspaces.}

L69-71. Clarification is needed on what is meant by "the surface closest to." Does this imply an assumption of homogeneous tortuosity, leading to a triangular distribution as the only optimisation? Doesn't this already assume that tortuosity is homogeneous and, therefore, the only optimisation possible is the triangular distribution?

**Response:** \textcolor{blue}{The reviewer is correct that we are assuming homogenous mesophyll properties (not just tortuosity). Please see a previous response to Reviewer 1 above, who raised a similar point, for how we revised the manuscript to clarify this assumption. We have also addressed the limitations of this assumption thoroughly in the text.}

L140-143 and 194-198. I apologise if I have missed something. There seems to be a presupposition of homogeneous internal resistance to CO2 diffusion or tortuosity. How does this assumption define the optimal arrangement and affect the photosynthesis modelling based solely on stomatal distribution? It seems to me that the authors already define the optimal arrangement by assuming homogeneous internal resistance to CO2 diffusion or homogeneous tortuosity. So, photosynthesis modelling for given photosynthetic parameters will be a function of the distribution of the stomata only, isn't it?

**Response:** \textcolor{blue}{We have clarified and better justified our assumptions in the Introduction (lines 54-73):}

> "Assuming uniform mesophyll diffusion resistance in all directions (homogenous porous medium), an ideal stomatal anatomy can be predicted. To maximize CO$_2$ supply from the stomatal pore to chloroplasts, stomata should be uniformly distributed in an equilateral triangular grid on the leaf surface so as to minimize stomatal number and CO$_2$ diffusion path length [@parkhurst_diffusion_1994]. An equilateral triangular grid is ideal because it maximizes the average distance between stomata, for a given stomatal density, and thereby minimizes the average distance between any point in the mesophyll to its nearest stomate. Assuming a homogenous mesophyll, this is the most efficient pattern to supply CO$_2$ to a leaf volume.

> Such an assumption, though an oversimplification, is a powerful tool for photosynthesis modelling, and may provide insight into how real leaves diverge from this. In real leaves, as the diffusion rate of CO$_2$ though liquid is approximately $10^4\times$ slower than CO$_2$ diffusion through air, mesophyll resistance is generally thought to be primarily limited by liquid diffusion [@aalto_three-dimensional_2002; @evans_resistances_2009], but diffusion through the IAS has also been shown to be a rate limiting process because the tortuous, disjunct nature of the IAS can greatly increase diffusion path lengths [@harwood_understanding_2021]. Additionally, tortuosity is higher in horizontal directions (parallel to leaf surface) than vertical directions (perpendicular to leaf surface) because of the cylindrical shape and vertical arrangement of palisade mesophyll cells [@earles_beyond_2018; @harwood_understanding_2021]. However, the ratio of lateral to vertical diffusion rate is still largely unknown and may be a highly variable trait in leaves [@morison_lateral_2005; @pieruschka_lateral_2005; @pieruschka_lateral_2006; @morison_does_2007]. Depending on the thickness of the leaf, porosity of the leaf mesophyll, tortuosity of the IAS, and lateral to vertical diffusion rate ratio, minimizing diffusion path length for CO$_2$ via optimally distributed stomata may yield significant increases in CO$_2$ supply for photosynthesis and higher $A_\text{max}$. Or plants may simply coordinate the development of stomata and mesophyll IAS to reach another optimal solution which does not rely on uniformly distributed stomata [@baillie_developmental_2020].

> We hypothesized that, in the absence of any constraint and assuming homogenous mesophyll diffusion resistance, natural selection will favor stomatal patterning and distribution to minimize the diffusion path length..."

L156-157 and 192-193. I don't see how a hexagon can be, in reality, an ideal distribution when there are other leaf structures in between and below the stomata that it seems are not being accounted for in this analysis. I don't think the authors need to account for them, but please clarify the limitations of not doing so.

**Response:** \textcolor{blue}{The reviewer is correct that we do not account for other leaf structures in this set of analyses. To address their comment, we clarified the limitations of not doing so in the Discussion and added ideas for future research on lines 253-263:}

> "We assume an idealized leaf epidermal and mesophyll structure that is homogenous and unconstrained by other tradeoffs. Real leaves not only provide pathways for CO$_2$ diffusion, but must supply water, intercept light, and deter herbivores and pathogens. All of these competing processes also happen on different time scales, and can be observed as heterogeneity in stomatal density, aperture, and internal leaf conditions across the leaf at any given moment [@lawson_nature_1998; @lawson_spatial_1999]. These competing interests result in heterogeneous epidermal and mesophyll structure that could alter predictions about optimal stomatal spacing. In order to maintain consistent leaf water potential across the lamina, stomatal density must be coordinated with vein density [@fiorin_transport_2016]. Thus, stomatal spacing may be optimized not at the interstomatal level, but at a higher level, coordinating water transport and water loss. ... Future gas exchange models should incorporate heterogenous mesophyll structure and hydraulic traits such as veins."

L205-206. Shouldn't this be stated in the introductions instead of the discussion? This allows for a more in-depth discussion in the relevant section because it seems to be the most likely option, so I expected a more detailed discussion of this issue.

**Response:** \textcolor{blue}{We'’'ve added the relevant text to the Introduction on lines 54-73 as described in a previous response to another comment. We also added to the Discussion on lines 215-218:}

> "However, reducing this distance to its absolute minimum may be constrained by developmental processes or the photosynthetic benefit may be too small to be 'seen' by natural selection (i.e. the selection coefficient is less than drift barrier *sensu* @sung_drift-barrier_2012). We also consider that our definition of optimal may be incorrect because it is based on overly simplistic assumptions about leaf mesophyll structure."

L252-253. Have alternative ideal distributions been tested, considering leaf structures like vein density, trichomes and substomatal cavity volume? Excluding these factors in estimations seems to be a significant omission. It seems problematic that other important leaf features, such as tortuosity, are not considered in the estimations but are discussed in the text. Again, I don't think the authors need to account for them, but please clarify the limitations of not doing so.

**Response:** \textcolor{blue}{To the best of our knowledge, no one has quantitatively investigated the optimal pattern with more realistic mesophyll structures. Most of the literature we are aware of predates the deeper understanding of heterogenous mesophyll structure that we have today. Our study highlights the inaccuracy of some predictions based on these simplisitic assumptions and will ophefully motivate revisiting these predictions in light of better empirical information. We added a sentence in the penultimate paragraph of the manuscript on lines 289-291 that addresses this point:}

> "To understand why stomata are not ideally dispersed, more modeling (with more realistic assumptions including vein density and IAS structure) should be done to estimate the photosynthetic properties of varying stomatal patterning."

## References

### 2024-01-21 ms/ms.Rmd
revised ms and response to reviewers

@@
 
-Many successful predictions about stomata and other C3 leaf traits can be made by hypothesizing that natural selection should optimize CO$_2$ gain per unit of water loss for any given set of environmental parameters, including their variability [@cowan_stomatal_1977; @buckley_optimal_2017; @sperry_predicting_2017].
+Many successful predictions about stomata and other C$_3$ leaf traits can be made by hypothesizing that natural selection should optimize CO$_2$ gain per unit of water loss for any given set of environmental parameters, including their variability [@cowan_stomatal_1977; @buckley_optimal_2017; @sperry_predicting_2017].
 Total stomatal area (size $\times$ density) is optimized for operational conductance ($g_\text{s,op}$) rather than maximum conductance ($g_\text{s,max}$) such that stomatal apertures are most responsive to changes in the environment at their operational aperture [@franks_physiological_2012; @liu_scaling_2021].
-Stomatal aperture can compensate for maladaptive stomatal densities to an extent [@bussis_stomatal_2006], but stomatal density and size ultimately determine a leaf's theoretical $g_\text{s,max}$ [@sack_developmental_2016], which is proportional to $g_\text{s,op}$ [@mcelwain_using_2016; @murray_consistent_2020].
+Stomatal aperture can compensate for suboptimal stomatal densities to an extent [@bussis_stomatal_2006], but stomatal density and size ultimately determine a leaf's theoretical $g_\text{s,max}$ [@sack_developmental_2016], which is proportional to $g_\text{s,op}$ under typical conditions [@mcelwain_using_2016; @murray_consistent_2020].
 Additionally, low stomatal densities lead to irregular and insufficient CO$_2$ supply and reduced photosynthetic efficiency in leaf areas far from stomata [@pieruschka_lateral_2006; @morison_lateral_2005], while high stomatal densities can reduce water use efficiency (WUE) [@bussis_stomatal_2006] and incur excessive metabolic costs [@de_boer_optimal_2016; @deans_optimization_2020].
@@
 These results highlight the utmost importance of amphistomy for some plants.
-Despite these optimality predictions, stomatal anatomy may be partially constrained by physical and developmental limits on phenotypic expression [@croxdale_stomatal_2000; @harrison_influence_2020; @muir_how_2023].
 
-A number of physical and developmental processes constrain stomatal anatomy traitspace.
+Despite the success of optimality predictions, stomatal anatomy may be partially constrained by physical and developmental limits on phenotypic expression [@croxdale_stomatal_2000; @harrison_influence_2020; @muir_how_2023].
+A number of physical and developmental processes constrain stomatal anatomy trait space.
 For example, almost all stomata follow the one cell spacing rule to maintain proper stomatal functioning as guard cell movement requires the rapid exchange of ions with neighboring epidermal cells (i.e.
@@
 This would prevent stomata from being strongly clustered; however, some species (notably in *Begonia*) appear to benefit from the overlapping vapor shells caused by stomatal clustering in dry environments [@yi_gan_stomatal_2010; @lehmann_effects_2015; @papanatsiou_stomatal_2017].
-Historically, stomatal patterning in dicot angiosperms was thought to be random with an exclusionary distance surrounding each stomate [@sachs_developmental_1974]; however, the developmental controls of stomatal patterning are more complex than was first thought.
+Historically, stomatal patterning in eudicot angiosperms was thought to be random with an exclusionary distance surrounding each stomate [@sachs_developmental_1974]; however, the developmental controls of stomatal patterning are more complex.
 @croxdale_stomatal_2000 reviews three developmental theories which attempt to explain stomatal patterning in angiosperms: inhibition, cell lineage, and cell cycle, ultimately arguing for a cell cycle based control of stomatal patterning.
@@
 Much less is known about the development of stomata on the adaxial leaf surface in amphistomatous plants.
-Stomatal size is additionally constrained by genome size with larger genomes leading to larger minimum guard cell size [@jordan_environmental_2015].
+Stomatal size is additionally constrained by genome size with larger genomes leading to larger minimum guard cell size [@jordan_environmental_2015; @roddy_scaling_2020].
 Despite these limitations, ecophysiological theory still predicts optimal stomatal anatomy, the details of which are discussed below.
 
-The patterning and spacing of stomata on the leaf affects photosynthesis in $C_3$ leaves by altering the CO$_2$ diffusion path length from stomata to sites of carboxylation in the mesophyll.
-Maximum photosynthetic rate ($A_\text{max}$) in $C_3$ plants is generally co-limited by biochemistry and diffusion, but modulated by light availability [@parkhurst_intercellular_1990; @manter_ci_2004; @carriqui_diffusional_2015].
+The patterning and spacing of stomata on the leaf affects photosynthesis in C$_3$ leaves by altering the CO$_2$ diffusion path length from stomata to sites of carboxylation in the mesophyll.
+Maximum photosynthetic rate ($A_\text{max}$) in C$_3$ plants is generally co-limited by biochemistry and diffusion, but modulated by light availability [@parkhurst_intercellular_1990; @manter_ci_2004; @carriqui_diffusional_2015].
 Low light decreases CO$_2$ demand by limiting electron transport rate, leading to relatively high internal CO$_2$ concentration ($C_\text{i}$) and low $A_\text{max}$ [@kaiser_metabolic_2016].
-In contrast, well hydrated leaves with open stomata in high light, photosynthesis is often limited by CO$_2$ supply as resistances from the boundary layer, stomatal pore, sub-stomatal cavity, and mesophyll can result in insufficient CO$C_2$ supply at the chloroplast to maxmimize photosynthesis [@farquhar_biochemical_1980; @lehmeier_cell_2017].
+In contrast, well hydrated leaves with open stomata in high light, photosynthesis is often limited by CO$_2$ supply as resistances from the boundary layer, stomatal pore, sub-stomatal cavity, and mesophyll can result in insufficient CO$C_2$ supply at the chloroplast to maximize photosynthesis [@farquhar_biochemical_1980; @lehmeier_cell_2017].
 In this study, we focus primarily on how stomatal patterning affects diffusion.
 
-Assuming uniform mesophyll diffusion resistance in all directions (porous medium), an ideal stomatal anatomy can be predicted.
+Assuming uniform mesophyll diffusion resistance in all directions (homogenous porous medium), an ideal stomatal anatomy can be predicted.
 To maximize CO$_2$ supply from the stomatal pore to chloroplasts, stomata should be uniformly distributed in an equilateral triangular grid on the leaf surface so as to minimize stomatal number and CO$_2$ diffusion path length [@parkhurst_diffusion_1994].
-With this assumption, an equilateral triangular grid is ideal because it standardizes the distance between stomata (for a given stomatal density) and thereby also standardizes the mesophyll volume with which each stomate is spatially closest to, ultimately reducing redundancy associated with clustered stomata and allowing fewer stomata to supply the same leaf area (or the same number of stomata to adequately supply a larger leaf area).
+An equilateral triangular grid is ideal because it maximizes the average distance between stomata, for a given stomatal density, and thereby minimizes the average distance between any point in the mesophyll to its nearest stomate.
+Assuming a homogenous mesophyll, this is the most efficient pattern to supply CO$_2$ to a leaf volume.
 
@@
 In real leaves, as the diffusion rate of CO$_2$ though liquid is approximately $10^4\times$ slower than CO$_2$ diffusion through air, mesophyll resistance is generally thought to be primarily limited by liquid diffusion [@aalto_three-dimensional_2002; @evans_resistances_2009], but diffusion through the IAS has also been shown to be a rate limiting process because the tortuous, disjunct nature of the IAS can greatly increase diffusion path lengths [@harwood_understanding_2021].
-Additionally, tortuosity is higher in horizontal directions (parallel to leaf surface) than vertical directions (perpendicular to leaf surface) because of the cylindrical shape and vertical arrangement of pallisade mesophyll cells [@earles_beyond_2018; @harwood_understanding_2021].
-However, the ratio of lateral to vertical diffusion rate is still largely unknown and may be a highly pliable trait in leaves [@morison_lateral_2005; @pieruschka_lateral_2005; @pieruschka_lateral_2006; @morison_does_2007].
+Additionally, tortuosity is higher in horizontal directions (parallel to leaf surface) than vertical directions (perpendicular to leaf surface) because of the cylindrical shape and vertical arrangement of palisade mesophyll cells [@earles_beyond_2018; @harwood_understanding_2021].
+However, the ratio of lateral to vertical diffusion rate is still largely unknown and may be a highly variable trait in leaves [@morison_lateral_2005; @pieruschka_lateral_2005; @pieruschka_lateral_2006; @morison_does_2007].
 Depending on the thickness of the leaf, porosity of the leaf mesophyll, tortuosity of the IAS, and lateral to vertical diffusion rate ratio, minimizing diffusion path length for CO$_2$ via optimally distributed stomata may yield significant increases in CO$_2$ supply for photosynthesis and higher $A_\text{max}$.
@@
 
-We hypothesized that, in the absence of any constraint and assuming uniform mesophyll diffusion resistance, natural selection will favor stomatal patterning and distribution to minimize the diffusion path length.
+We hypothesized that, in the absence of any constraint and assuming homogenous mesophyll diffusion resistance, natural selection will favor stomatal patterning and distribution to minimize the diffusion path length.
 In amphistomatous leaves, this would be accomplished by 1) a uniform, equilateral triangular distribution of stomata on both abaxial and adaxial leaf surfaces and 2) coordinated stomatal spacing on each surface that offsets the position of stomata (\autoref{fig:ideal-amphi-grid}).
 Coordination between leaf surfaces is defined in this study as the occurrence of stomata in areas farthest from stomata on the opposite leaf surface.
-Additionally, because CO$_2$ is more limiting for photosynthesis under high light, we hypothesize that in high light 3) there should be more stomata, and 4) stomata should be more overdispersed compared to a random distribution than in low light.
+Additionally, because CO$_2$ is more limiting for photosynthesis under high light, we hypothesize that in high light 3) there should be more stomata, and 4) stomata should be more overdispersed (closer to equilateral triangular grid) compared to a random distribution than in low light.
 Finally, since, in measures of whole leaves, stomatal area rather than stomatal density is optimized for operational conductance, we hypothesize that 5) stomatal length (and hence its area) will be positively correlated with the area of the leaf surface to which it is spatial closest as defined by Voronoi tessellation techniques.
@@
 This way, each stomate may be optimally sized relative to the mesophyll volume it supplies.
+Hypothesis 3 is already well supported in many species [@poorter_metaanalysis_2019], but it is useful here to confirm that light treatments induced plasticity in the expected direction.
 
@@
 We also used a 2-D porous medium approximation of CO$_2$ diffusion and photosynthesis to predict the photosynthetic advantage of optimal versus suboptimal coordination in stomatal coordination between surfaces.
-Specifically, we predicted that traits which affect diffusion path length (leaf thickness, stomatal density, leaf porosity, lateral-vertical diffusion rate ratio), diffusion rate (temperature, pressure), and CO$_2$ demand (Rubisco concentration, light) would modulate the advantage of optimal stomatal arrangement following the relationships outlined in \autoref{tab:hypotheses}.
+Specifically, we predicted that traits which affect diffusion path length (leaf thickness, stomatal density, leaf porosity), diffusion rate (determined by temperature, pressure), and CO$_2$ demand (Rubisco concentration, light) would modulate the advantage of optimal stomatal arrangement following the relationships outlined in \autoref{tab:hypotheses}.
 Here, we integrate over reasonable parameter space to determine the ecophysiological context most likely to favor stomatal coordination in amphistomatous leaves.
@@
 PAR stands for photosynthetically active radiation.
-Seeds were surface-sterilized and stratified at 4°C for 3–5 d in 0.15% agarose solution and then sown directly into Pro-Mix HP soil (Premier Horticulture; Quakerstown, PA, USA) and supplemented with Scott's Osmocote Classic 14-14-14 fertilizer (Scotts-Sierra, Marysville, OH, USA).
+*A.
+thaliana* responds strongly to light levels over this range [@bailey_acclimation_2001], though natural populations in open canopies can experience $\text{PAR} > 800~\mu \text{mol}~\text{m}^{-2}~\text{s}^{-1}$ [@callahan_shade-induced_2002].
+Seeds were surface-sterilized and stratified at 4 °C for 3–5 d in 0.15% agarose solution and then sown directly into Pro-Mix HP soil (Premier Horticulture; Quakerstown, PA, USA) and supplemented with Scott's Osmocote Classic 14-14-14 fertilizer (Scotts-Sierra, Marysville, OH, USA).
 At 10–14 d, seedlings were thinned so only one seedling per container remained.
 Plants were grown to maturity in growth chambers where the conditions were as follows: 16 : 8 h, 22 : 20°C, day : night cycle.
-Imaging of the epidermis and internal leaf structures was performed using a Leica SP5 confocal microscope with the protocol developed by @wuyts_high-contrast_2010 with additional modification described in @dow_disruption_2017.
+Imaging of the epidermis and internal leaf structures was performed using a Leica SP5 confocal microscope (Leica Microsystems, Wetzlar, Germany) with the protocol developed by @wuyts_high-contrast_2010 with additional modification described in @dow_disruption_2017.
 We captured 132 images in total, making 66 abaxial-adaxial image pairs.
@@
 We measured stomatal position and length using ImageJ [@schneider_nih_2012].
-A number of synthetic leaf surface datasets were also prepared (details below) so as to be mathematically explicit when defining null distributions against which to test our hypotheses and to avoid any methodological influence on our results (e.g.
-stomatal density uncertainty, boundary effects when calculating patterning metrics: uniform random, non-random overdispersion, and uniform).
+A number of synthetic leaf surface data sets were also simulated (details below) to generate null distributions against which to test our hypotheses and to avoid any methodological influence on our results (e.g.
+boundary effects when calculating stomatal patterning).
 All synthetic leaf surfaces were simulated based on the size of the real leaf images and stomatal densities matched those of real leaf images.
@@
 
-We tested whether stomata are non-randomly distributed by comparing each observed, real leaf stomatal pattern to an array of synthetic random uniform patterns.
+We compared observed stomatal patterning to an ideal pattern (uniform equilateral triangular grid) and a null model (random uniform distribution).
+The terminology is unfortunately confusing because the word `uniform' is used in different ways.
+A uniform equilateral triangular grid means that the distance between stomata is uniform; a random uniform distribution means that a stomate has an equal probability (i.e.
+uniform) of occurring anyone on the leaf surface.
+To limit confusion, we refer to the ideal pattern as uniform (equilateral triangle grid) and the null pattern as random (random uniform).
+When observed stomatal patterns are more dispersed than expected under random patterning, we refer to this as overdispersed.
+Note, however, that overdispersed compared to random is still less dispersed than ideal, because the ideal pattern is maximally dispersed.
+
+We tested whether stomata overdispersed by comparing each observed, real leaf stomatal pattern to an array of synthetic data simulated from a random distribution.
 For each observed leaf surface image with $n$ stomata we generated $10^3$ synthetic surfaces with $n$ stomata uniformly randomly distributed on the surface.
@@
 We calculated $\mathrm{NNI}$ using the *R* package **spatialEco** version `r packageVersion("spatialEco")` [@evans_spatialeco_2023].
-The observed stomatal distribution is dispersed relative to a uniform random distribution if the observed $NNI$ is greater than 95% of the synthetic $\mathrm{NNI}$ values (one-tailed test).
+The observed stomatal distribution is overdispersed relative to a random distribution if the observed $NNI$ is greater than 95% of the synthetic $\mathrm{NNI}$ values (one-tailed test).
 We adjusted $P$-values to account for multiple comparisons using the Benjamini-Hochberg [@benjamini_controlling_1995] false discovery rate procedure implemented in the *R* package **multtest** version `r packageVersion("multtest")` [@wong_multiple_2005].
@@
 
-We developed a dispersion index $\mathrm{DI}$ to quantify how close observed stomatal distributions are to random uniform versus ideally dispered in an equilateral triangular grid.
-$\mathrm{DI}$ varies from zero to one, where zero is uniformly random and one is ideally dispersed:
+We developed a dispersion index $\mathrm{DI}$ to quantify how close observed stomatal patterning is to random versus ideally patterned in an equilateral triangular grid.
+$\mathrm{DI}$ varies from zero to one, where zero is random and one is ideally patterned:
 
 \begin{equation}\label{eq:disp}
-  \mathrm{DI} = \frac{\mathrm{NNI} - \text{median}(\mathrm{NNI_{random}})}{\text{median}(\mathrm{NNI_{uniform}}) - \text{median}(\mathrm{NNI_{random}})}
+  \mathrm{DI} = \frac{\mathrm{NNI} - \text{median}(\mathrm{NNI_{random}})}{\text{median}(\mathrm{NNI_{ideal}}) - \text{median}(\mathrm{NNI_{random}})}
 \end{equation}
@@
 
-Finally, we examined the relationship between stomatal zone area and stomatal length of real stomata using a Bayesian linear mixed-effects model fit with the *R* package **brms** version `r fit$version$brms` [@burkner_brms_2017; @burkner_advanced_2018] and *Stan* version `r fit$version$cmdstan` [@stan_development_team_stan_2023].
+Finally, we examined the relationship between stomatal zone area and stomatal length using a Bayesian linear mixed-effects model fit with the *R* package **brms** version `r fit$version$brms` [@burkner_brms_2017; @burkner_advanced_2018] and *Stan* version `r fit$version$cmdstan` [@stan_development_team_stan_2023].
 Stomatal zone area was calculated using Voronoi tessellation (e.g.
@@
 To test whether the position of ab- and adaxial stomata are coordinated we compared the observed distribution to a null distribution where the positions on each surface are random.
-For each pair of surfaces (observed or synthetic) we calculated the distance squared between each pixel of the surface to the nearest stomatal centroid with the *R* package **raster** version `r packageVersion("raster")`.
+For each pair of surfaces (observed or synthetic) we calculated the distance squared between each pixel of the surface to the nearest stomatal centroid with the *R* package **raster** version `r packageVersion("raster")` [@hijmans_raster_2023].
 We refer to this as the 'nearest stomatal distance' or NSD.
@@
 We generated a null distribution of the correlation coefficient by simulating $10^3$ synthetic data sets for each observed pair.
-For each synthetic data set, we simulated stomatal position using a random uniform distribution, as described above, matching the number of stomata on abaxial and adaxial leaf surfaces.
+For each synthetic data set, we simulated stomatal position using a random uniform distribution, as described above, matching the number of stomata on abaxial and adaxial leaf surfaces to the observed data.
 Stomatal positions on each surface are coordinated if the correlation coefficient of the NSD between observed ab- and adaxial surfaces is greater than 95% of the synthetic correlation values (one-tailed test).
@@
 Consider a two-dimensional leaf where stomata occur on each surface in a regular sequence with interstomatal distance $U$.
-The main outcome we assessed is the advantage of offsetting the position of stomata on each surface compared to have stomata on the same $x$ position on each surface.
+The main outcome we assessed is the advantage of offsetting the position of stomata on each surface compared to having stomata on the same $x$ position on each surface.
 With these assumptions, by symmetry, we only need to model two stomata, one abaxial and one adaxial, from $x = 0$ to $x = U/2$ and from the adaxial surface at $y = 0$ to the abaxial surface at $y = L$, the leaf thickness.
@@
 
-## Stomatal distribution is non-random, but far from uniform
+## Stomatal patterning is non-random, but far from uniform
 
-Many leaf surfaces (`r nni2[['n1']]` of `r nni2[['n1']] + nni2[['n2']]`, `r round(100 * nni2[['n1']] / (nni2[['n1']] + nni2[['n2']]), 1)`%) are significantly overdispersed compared to a random uniform distribution, but none were close to a uniform equilateral triangular pattern (dispersion index = 1; \autoref{fig:single-surface}).
+Many leaf surfaces (`r nni2[['n1']]` of `r nni2[['n1']] + nni2[['n2']]`, `r round(100 * nni2[['n1']] / (nni2[['n1']] + nni2[['n2']]), 1)`%) are significantly overdispersed compared to a random uniform distribution, but none were close to an ideal, uniform equilateral triangular pattern (dispersion index = 1; \autoref{fig:single-surface}).
 Before controlling for multiple comparisons, `r round(100 * nni1[['n1']] / (nni1[['n1']] + nni1[['n2']]), 1)`% are significantly overdispersed.
@@
 \includegraphics[width=4in]{figures/single-surface.pdf}
-\caption{Stomata are more dispersed than expected under the null model of uniform random position (dispersion index = 0) but far from a distribution that maximizes distance between stomata (dispersion index = 1; uniform distribution).
+\caption{Stomata are more dispersed than expected under the null model of random patterning (dispersion index = 0) but far from a distribution that maximizes distance between stomata (dispersion index = 1; uniform patterning).
 We determined statistical significance between light treatments using Tukey post-hoc tests.
@@
 In principle, spacing stomata to minimize the average distance between stomatal pores and chloroplasts within the mesophyll should increase carbon gain, all else being equal.
-However, reducing this distance to its absolute minimum may be constrained by developmental processes, the photosynthetic benefit may be too small to be 'seen' by natural selection (i.e.
-the selection coefficient is less than drift barrier *sensu* @sung_drift-barrier_2012), or alternative peaks in the adaptive landscape may be reached as the adaptive peak investigated in this study is based on oversimplistic assumptions of a porous medium mesophyll.
+However, reducing this distance to its absolute minimum may be constrained by developmental processes or the photosynthetic benefit may be too small to be 'seen' by natural selection (i.e.
+the selection coefficient is less than drift barrier *sensu* @sung_drift-barrier_2012).
+We also consider that our definition of optimal may be incorrect because it is based on overly simplistic assumptions about leaf mesophyll structure.
 
 We tested five related hypotheses about stomatal spacing in amphistomatous leaves using the model angiosperm *Arabidopsis thaliana* grown under different light intensities.
-First, we predicted that stomata on each surface are overdispersed relative to a random uniform distribution, which should increase CO$_2$ supply.
-Stomata on each surface are overdispersed (\autoref{fig:single-surface}), but are not ideally, uniformly dispersed in an equilateral triangular grid as would be optimal to minimize CO$_2$ diffusion path length and equalize the area supplied by each stomate (\autoref{fig:tessellation-example}).
+First, we predicted that stomata on each surface are overdispersed relative to a random distribution, which should increase CO$_2$ supply.
+Stomata on each surface are overdispersed (\autoref{fig:single-surface}), but are not ideally, uniformly patterned in an equilateral triangular grid as would be optimal to minimize CO$_2$ diffusion path length and equalize the area supplied by each stomate (\autoref{fig:tessellation-example}).
 Second, we predicted that an optimal amphistomatous leaf has offset stomata such that stomata are more likely to appear on one leaf surface if there is not a stomata directly opposite it on the other surface as shown in \autoref{fig:ideal-amphi-grid}.
@@
 Third, we predicted that plants respond plastically to higher light intensity by increasing stomatal density.
-*Arabidopsis* plants grown under high light had higher stomatal density than the same genotype growns under low and medium light intensity (\autoref{fig:density}).
-However, we found no support for our fourth prediction that stomata would be more evenly dispersed at high light intensity (\autoref{fig:single-surface}).
+*Arabidopsis* plants grown under high light had higher stomatal density than the same genotype grown under low and medium light intensity (\autoref{fig:density}).
+However, we found no support for our fourth prediction that stomatal patterning would be overdispersed at high light intensity (\autoref{fig:single-surface}).
 Finally, we predicted that within leaf variation in stomatal size would correlate with stomatal spacing, as larger stomata can supply larger volumes of adjacent mesophyll.
@@
 There are three non-mutually exclusive hypotheses for why several of our predictions were wrong.
-First, our predictions are wrong because they are based on overly simplistic assumptions about epidermal and mesophyll anatomy.
-In real leaves, diffusion through the mesophyll depends on the position of each cell and resulting IAS geometry.
-If, for example, horizontal diffusion was modeled to be much slower than vertical diffusion, we likely would have seen an increased benefit of ab- adaxial stomatal coordination across a greater range of feasible trait values.
-However, further information about the three-dimensional structure of the leaves would need to be incorporated into the model to address any further questions about how mesophyll anatomy may influence stomatal anatomy.
-This represents an interesting area for further research.
-Second, spatio-temporal variation of internal conditions within leaves and between stomatal responses may make uniform, coordinated stomatal surfaces less beneficial [@lawson_nature_1998; @lawson_spatial_1999].
+First, our predictions must be wrong because they are based on overly simplistic assumption of a homogenous porous medium within the mesophyll.
+Real leaf mesophylls are spatially heterogenous and chloroplasts are distributed as discrete nodes.
+The intercellular air space conductance is determined by its porosity and tortuosity, both of which are heterogenous within the leaf.
+The palisade is typically less porous than the spongy mesophyll (e.g.
+@therouxrancourt_bias_2017), which should impact the optimal patterning on stomata on ab- versus adaxial surfaces.
+Tortuosity is also systematically greater in the palisade in the lateral direction parallel to the leaf plane [@harwood_understanding_2021].
+We might predict a greater coordination advantage of offset stomata by accounting for greater lateral tortuosity, but it is likely that benefit is still very small under realistic parameter space.
+Quantifying the patterns of heterogeneity in porosity, tortuosity, and other factors [@earles_beyond_2018] using 3D imaging (e.g.
+@borsuk_structural_2022) will be needed to generate more realistic hypotheses about optimal stomatal spacing.
+
+Second, spatio-temporal variation of internal conditions within leaves and between stomatal responses may make uniform, coordinated stomatal surfaces less beneficial [@weyers_heterogeneity_1997; @lawson_nature_1998; @lawson_spatial_1999].
 This is because our model assumes a uniform leaf, the internal conditions of which are periodic and solved empirically and therefore stable.
-Any horizontal concentration gradients due to environmental heterogeneity and variable induction times for interacting leaf processes may dissolve any benefit of uniform stomatal patterning.
+Any horizontal concentration gradients due to environmental heterogeneity and variable induction times for interacting leaf processes may reduce the benefit of uniform stomatal patterning.
 Third, natural selection may be constrained by developmental processes that prevent phenotypes from reaching their adaptive optima.
-Stomatal development must be plastic to environmental ques interpreted through long distance and cell to cell signalling pathways [@pillitteri_mechanisms_2012].
+Stomatal development must be plastic to environmental cues interpreted through long distance and cell-to-cell signalling pathways [@pillitteri_mechanisms_2012].
 This plasticity may come with the cost of being unable to orchestrate the development of an absolutely uniform stomatal grid.
@@
 
-Our model also assumes an idealized leaf epidermal and mesophyll structure that is homogenous and unconstrained by other tradeoffs.
+We assume an idealized leaf epidermal and mesophyll structure that is homogenous and unconstrained by other tradeoffs.
 Real leaves not only provide pathways for CO$_2$ diffusion, but must supply water, intercept light, and deter herbivores and pathogens.
-All of these competing processes also happen on different time scales, and can be observed as heterogeneity in stomatal density, aperature, and internal leaf conditions across the leaf at any given moment [@lawson_nature_1998; @lawson_spatial_1999].
-These competing interests result in non-uniform epidermal and mesophyll structure that could alter predictions about optimal stomatal spacing.
+All of these competing processes also happen on different time scales, and can be observed as heterogeneity in stomatal density, aperture, and internal leaf conditions across the leaf at any given moment [@lawson_nature_1998; @lawson_spatial_1999].
+These competing interests result in heterogeneous epidermal and mesophyll structure that could alter predictions about optimal stomatal spacing.
 In order to maintain consistent leaf water potential across the lamina, stomatal density must be coordinated with vein density [@fiorin_transport_2016].
 Thus, stomatal spacing may be optimized not at the interstomatal level, but at a higher level, coordinating water transport and water loss.
-We also assume a uniform mesophyll; however, in ab-adaxially differentiated leaves, this assumption is overly simplistic.
-The pallisade mesophyll is more tightly packed than the spongy mesophyll as an adaptation to intercept light efficiently, so lateral diffusion may be more limiting in the adaxial portion of the leaf.
+For example, the palisade mesophyll is more tightly packed than the spongy mesophyll as an adaptation to intercept light efficiently, so lateral diffusion may be more limiting in the adaxial portion of the leaf.
 This may explain why adaxial leaf surfaces have consistently higher dispersion indices than abaxial surfaces across all light treatments (\autoref{fig:single-surface}).
-Future gas exchange models should include multiple parameter sets for different strata of mesophyll tissue and water supply structures such as veins, though these are beyond the scope of this study.
+Future gas exchange models should incorporate heterogenous mesophyll structure and hydraulic traits such as veins.
 
-No developmental pathway exists to ensure an idealized placement of stomata on the leaf surface.
+We are not aware of a developmental pathway that ensures an idealized placement of stomata on the leaf surface.
 Rather, stomatal development is a dynamic process that must be plastic to environmental cues.
@@
 Our results showing an intermediate level of dispersion in stomatal spacing may be best explained by these developmental pathways which ensure the proper spacing of stomata, with an added random effect brought about by the necessity for plasticity in stomatal development (\autoref{fig:single-surface}).
-However, any nonideal randomness in stomatal spacing may offset by simultaneous and coordinated development of the IAS [@baillie_developmental_2020].
-The fact that stomata which supply a greater mesophyll volume tend to be larger suggests that plants may use coordinated development of multiple leaf tissues to negate nonideal stomatal spacing (\autoref{fig:length-area}).
+However, deviations from ideal stomatal spacing may compensated for by simultaneous and coordinated development of the IAS [@baillie_developmental_2020].
+The fact that stomata which supply a greater mesophyll volume tend to be larger suggests that plants may use coordinated development of multiple leaf anatomical features to compensate for nonideal stomatal spacing (\autoref{fig:length-area}).
 
@@
 We do not investigate the potential developmental pathways that influence stomatal dispersion in this study; however, they are important to consider as these pathways could limit plants from reaching a theoretical peak in the adaptive landscape: uniform stomatal patterning.
-Instead, as this study suggests, plants may simply compensate with higher stomatal density and by fitting stomatal size to the area that they supply with CO$_2$.
-To understand why stomata are not ideally dispersed, more modeling (with a more complex set of assumptions including vein density and IAS structure) should be done to estimate the fitness gain of stomatal dispersion.
-Additionally, genetic manipulation studies should attempt to create mutants with clustered and ideally dispersed stomata for a comparison of their photosynthetic traits.
-This could have extremely important implications for maximum assimilation rates in crops as most crop species are grown in high light where CO$_2$ is often limiting.
+Instead, as this study suggests, plants may simply compensate with higher stomatal density and by modulating stomatal size to the area that they supply with CO$_2$.
+To understand why stomata are not ideally dispersed, more modeling (with more realistic assumptions including vein density and IAS structure) should be done to estimate the photosynthetic properties of varying stomatal patterning.
+Additionally, genetic manipulation studies should attempt to create mutants with clustered and uniformly patterned stomata for a comparison of their photosynthetic traits.
+This could have important implications for maximizing assimilation rates in crops as most crop species are grown in high light where CO$_2$ is often limiting.
 In drought-prone environments, increased stomatal dispersion may increase water use efficiency by reducing the number of stomata needed to achieve the same internal CO$_2$ concentration, $C_\text{i}$.
+However, it would be necessary to account for many other differences between *A.
+thaliana* and crop leaves and canopies.
 

### 2024-01-21 ms/ms.tex
fixed reference issue in ms
revised ms and response to reviewers

@@
 adaptation to alleviate CO\(_2\) diffusion limitations in productive
-environments where other factors are not limiting as the diffusion path
-length from stomate to chloroplast is effectively halved. Plants can
-also reduce CO\(_2\) limitations through other aspects of optimal
-stomatal anatomy: stomatal density, distribution, patterning, and size.
-A number of studies have demonstrated that stomata are overdispersed on
-a single leaf surface; however, much less is known about stomatal
-anatomy in amphistomatous leaves, especially the coordination between
-leaf surfaces, despite their prevelance in nature and near ubiquity
-among crop species. Here we use novel spatial statistics based on
-simulations and photosynthesis modeling to test hypotheses about how
-amphistomatous plants may optimize CO\(_2\) limitations in the model
-angiosperm \emph{Arabidopsis thaliana} grown in different light
+environments as the diffusion path length from stomate to chloroplast is
+effectively halved in amphistomatous leaves. Plants may also reduce
+CO\(_2\) limitations through other aspects of optimal stomatal anatomy:
+stomatal density, distribution, patterning, and size. A number of
+studies have demonstrated that stomata are overdispersed compared to a
+random distribution on a single leaf surface; however, despite their
+prevelance in nature and near ubiquity among crop species, much less is
+known about stomatal anatomy in amphistomatous leaves, especially the
+coordination between leaf surfaces. Here we use novel spatial statistics
+based on simulations and photosynthesis modeling to test hypotheses
+about how amphistomatous plants may optimize CO\(_2\) diffusion in the
+model angiosperm \emph{Arabidopsis thaliana} grown in different light
 environments. We find that 1) stomata are overdispersed, but not ideally
-dispersed, on both leaf surfaces across all light treatments; 2) abaxial
-and adaxial leaf surface patterning are independent; and 3) the
-theoretical improvements to photosynthesis from abaxial-adaxial stomatal
-coordination are miniscule (\(\ll 1\)\%) across the range of feasible
-parameter space. However, we also find that 4) stomatal size is
-correlated with the mesophyll volume that it supplies with CO\(_2\),
-suggesting that plants may optimize CO\(_2\) diffusion limitations
-through alternative pathways other than ideal, uniform stomatal spacing.
-We discuss the developmental, physical, and evolutionary constraits
-which may prohibit plants from reaching the theoretical adaptive peak of
-uniform stomatal spacing and inter-surface stomatal coordination. These
-findings contribute to our understanding of variation in the anatomy of
-amphistomatous leaves.}
+dispersed, on both leaf surfaces across all light treatments; 2) the
+patterning of stomata on abaxial and adaxial leaf surfaces is
+independent; and 3) the theoretical improvements to photosynthesis from
+abaxial-adaxial stomatal coordination are miniscule (\(\ll 1\)\%) across
+the range of feasible parameter space. However, we also find that 4)
+stomatal size is correlated with the mesophyll volume that it supplies
+with CO\(_2\), suggesting that plants may optimize CO\(_2\) diffusion
+limitations through alternative pathways other than ideal, uniform
+stomatal spacing. We discuss the developmental, physical, and
+evolutionary constraits which may prohibit plants from reaching this
+theoretical adaptive peak of uniform stomatal spacing and inter-surface
+stomatal coordination. These findings contribute to our understanding of
+variation in the anatomy of amphistomatous leaves.}
 
@@
 
-\hypertarget{introduction}{%
-\section{Introduction}\label{introduction}}
+\section{Introduction}\label{introduction}
 
@@
 temperature, wind speed, leaf water potential, IAS geometry, mesophyll
-cell anatomy, and stomatal anatomy.
+cell anatomy, and stomatal anatomy. The latter of these is the focus of
+this study, with discussion of other interacting variables.
 
-Many successful predictions about stomata and other C3 leaf traits can
-be made by hypothesizing that natural selection should optimize CO\(_2\)
-gain per unit of water loss for any given set of environmental
+Many successful predictions about stomata and other C\(_3\) leaf traits
+can be made by hypothesizing that natural selection should optimize
+CO\(_2\) gain per unit of water loss for any given set of environmental
 parameters, including their variability
@@
 \citep{franks_physiological_2012, liu_scaling_2021}. Stomatal aperture
-can compensate for maladaptive stomatal densities to an extent
+can compensate for suboptimal stomatal densities to an extent
 \citep{bussis_stomatal_2006}, but stomatal density and size ultimately
@@
 \citep{sack_developmental_2016}, which is proportional to
-\(g_\text{s,op}\) \citep{murray_consistent_2020}. Additionally, low
+\(g_\text{s,op}\) under typical conditions
+\citep{mcelwain_using_2016, murray_consistent_2020}. Additionally, low
 stomatal densities lead to irregular and insufficient CO\(_2\) supply
@@
 \citep{bussis_stomatal_2006} and incur excessive metabolic costs
-\citep{deans_optimization_2020}. Stomatal density positively co-varies
-with irradiance during leaf development and negatively co-varies with
-CO\(_2\) concentration
+\citep{de_boer_optimal_2016, deans_optimization_2020}. Stomatal density
+positively co-varies with irradiance during leaf development and
+negatively co-varies with CO\(_2\) concentration
 \citep{gay_influence_1975, schoch_dependence_1980, woodward_stomatal_1987, royer_stomatal_2001},
@@
 Amphistomy effectively halves the CO\(_2\) diffusion path length and
-boundary layer resistance by doubling boundary layer conductance
-\citep{parkhurst_adaptive_1978, harrison_influence_2020, mott_amphistomy_1991}.
-Despite these optimality predictions, stomatal anatomy may be partially
-constrained by physical and developmental limits on phenotypic
+boundary layer resistance by doubling boundary layer conductance \citep[
+\citet{harrison_influence_2020}]{parkhurst_adaptive_1978, mott_amphistomy_1991}.
+Ab- and adaxial leaf surfaces were found to function independent of one
+another in wheat, an important crop, with the adaxial surface
+demonstrating higher photosynthetic capacity \citep{wall_stomata_2022}.
+These results highlight the utmost importance of amphistomy for some
+plants.
+
+Despite the success of optimality predictions, stomatal anatomy may be
+partially constrained by physical and developmental limits on phenotypic
 expression
 \citep{croxdale_stomatal_2000, harrison_influence_2020, muir_how_2023}.
-
 A number of physical and developmental processes constrain stomatal
-anatomy traitspace. For example, almost all stomata follow the one cell
+anatomy trait space. For example, almost all stomata follow the one cell
 spacing rule to maintain proper stomatal functioning as guard cell
@@
 \citep{yi_gan_stomatal_2010, lehmann_effects_2015, papanatsiou_stomatal_2017}.
-Historically, stomatal patterning in dicot angiosperms was thought to be
-random with an exclusionary distance surrounding each stomate
+Historically, stomatal patterning in eudicot angiosperms was thought to
+be random with an exclusionary distance surrounding each stomate
 \citep{sachs_developmental_1974}; however, the developmental controls of
-stomatal patterning are more complex than random development along the
-leaf surface. \citet{croxdale_stomatal_2000} reviews three developmental
-theories which attempt to explain stomatal patterning in angiosperms:
-inhibition, cell lineage, and cell cycle, ultimately arguing for a cell
-cycle based control of stomatal patterning.
-\citet{pillitteri_mechanisms_2012} review the short and long distance
-signalling pathways associated with stomatal spacing and development,
-which include cell to cell communication and whole plant integration to
-ensure the proper spacing of stomata across a single leaf surface
-depending on environmental ques. Much less is known about the
+stomatal patterning are more complex. \citet{croxdale_stomatal_2000}
+reviews three developmental theories which attempt to explain stomatal
+patterning in angiosperms: inhibition, cell lineage, and cell cycle,
+ultimately arguing for a cell cycle based control of stomatal
+patterning. \citet{pillitteri_mechanisms_2012} review the short and long
+distance signalling pathways associated with stomatal spacing and
+development, which include cell to cell communication and whole plant
+integration to ensure the proper spacing of stomata across a single leaf
+surface depending on environmental ques. Much less is known about the
 development of stomata on the adaxial leaf surface in amphistomatous
@@
 larger genomes leading to larger minimum guard cell size
-\citep{jordan_environmental_2015}. Despite these limitations,
-ecophysiological theory still predicts optimal stomatal anatomy, the
-details of which are discussed below.
+\citep{jordan_environmental_2015, roddy_scaling_2020}. Despite these
+limitations, ecophysiological theory still predicts optimal stomatal
+anatomy, the details of which are discussed below.
 
 The patterning and spacing of stomata on the leaf affects photosynthesis
-in \(C_3\) leaves by altering the CO\(_2\) diffusion path length from
+in C\(_3\) leaves by altering the CO\(_2\) diffusion path length from
 stomata to sites of carboxylation in the mesophyll. Maximum
-photosynthetic rate (\(A_\text{max}\)) in \(C_3\) plants is generally
+photosynthetic rate (\(A_\text{max}\)) in C\(_3\) plants is generally
 co-limited by biochemistry and diffusion, but modulated by light
@@
 photosynthesis is often limited by CO\(_2\) supply as resistances from
-the boundary layer, stomatal pore, and mesophyll can result in
-insufficient CO\(C_2\) supply at the chloroplast to maxmimize
-photosynthesis \citep{farquhar_biochemical_1980, lehmeier_cell_2017}. In
-this study, we focus primarily on how stomatal patterning affects
-diffusion, ignoring boundary layer and mesophyll resistances.
+the boundary layer, stomatal pore, sub-stomatal cavity, and mesophyll
+can result in insufficient CO\(C_2\) supply at the chloroplast to
+maximize photosynthesis
+\citep{farquhar_biochemical_1980, lehmeier_cell_2017}. In this study, we
+focus primarily on how stomatal patterning affects diffusion.
 
+Assuming uniform mesophyll diffusion resistance in all directions
+(homogenous porous medium), an ideal stomatal anatomy can be predicted.
 To maximize CO\(_2\) supply from the stomatal pore to chloroplasts,
@@
 grid on the leaf surface so as to minimize stomatal number and CO\(_2\)
-diffusion path length \citep{parkhurst_diffusion_1994}. As the diffusion
-rate of CO\(_2\) though liquid is approximately \(10^4\times\) slower
-than CO\(_2\) diffusion through air, mesophyll resistance is generally
-thought to be primarily limited by liquid diffusion
+diffusion path length \citep{parkhurst_diffusion_1994}. An equilateral
+triangular grid is ideal because it maximizes the average distance
+between stomata, for a given stomatal density, and thereby minimizes the
+average distance between any point in the mesophyll to its nearest
+stomate. Assuming a homogenous mesophyll, this is the most efficient
+pattern to supply CO\(_2\) to a leaf volume.
+
+Such an assumption, though an oversimplification, is a powerful tool for
+photosynthesis modelling, and may provide insight into how real leaves
+diverge from this. In real leaves, as the diffusion rate of CO\(_2\)
+though liquid is approximately \(10^4\times\) slower than CO\(_2\)
+diffusion through air, mesophyll resistance is generally thought to be
+primarily limited by liquid diffusion
 \citep{aalto_three-dimensional_2002, evans_resistances_2009}, but
@@
 leaf surface) than vertical directions (perpendicular to leaf surface)
-because of the cylindrical shape and vertical arrangement of pallisade
+because of the cylindrical shape and vertical arrangement of palisade
 mesophyll cells \citep{earles_beyond_2018, harwood_understanding_2021}.
 However, the ratio of lateral to vertical diffusion rate is still
-largely unknown
-\citep{morison_lateral_2005, pieruschka_lateral_2005, pieruschka_lateral_2006}.
+largely unknown and may be a highly variable trait in leaves
+\citep{morison_lateral_2005, pieruschka_lateral_2005, pieruschka_lateral_2006, morison_does_2007}.
 Depending on the thickness of the leaf, porosity of the leaf mesophyll,
@@
 stomata may yield significant increases in CO\(_2\) supply for
-photosynthesis and higher \(A_\text{max}\).
+photosynthesis and higher \(A_\text{max}\). Or plants may simply
+coordinate the development of stomata and mesophyll IAS to reach another
+optimal solution which does not rely on uniformly distributed stomata
+\citep{baillie_developmental_2020}.
 
-We hypothesized that natural selection will favor stomatal patterning
-and distribution to minimize the diffusion path length. In
-amphistomatous leaves, this would be accomplished by 1) a dispersed,
-equilateral triangular distribution of stomata on both abaxial and
-adaxial leaf surfaces and 2) coordinated stomatal spacing on each
+We hypothesized that, in the absence of any constraint and assuming
+homogenous mesophyll diffusion resistance, natural selection will favor
+stomatal patterning and distribution to minimize the diffusion path
+length. In amphistomatous leaves, this would be accomplished by 1) a
+uniform, equilateral triangular distribution of stomata on both abaxial
+and adaxial leaf surfaces and 2) coordinated stomatal spacing on each
 surface that offsets the position of stomata
@@
 hypothesize that in high light 3) there should be more stomata, and 4)
-stomata should be more uniformly distributed than in low light. Finally,
-as stomatal densities are selected for optimal operational aperture, we
-hypothesize that 5) stomatal length will be positively correlated with
-the area of the leaf surface to which it is closest. We refer to this as
-the `stomatal zone', the leaf area surrounding a focal stomate closest
-to that stomate and therefore the zone it supplies with CO\(_2\)). This
-way, each stomate can be optimally sized relative to the mesophyll
-volume it supplies.
+stomata should be more overdispersed (closer to equilateral triangular
+grid) compared to a random distribution than in low light. Finally,
+since, in measures of whole leaves, stomatal area rather than stomatal
+density is optimized for operational conductance, we hypothesize that 5)
+stomatal length (and hence its area) will be positively correlated with
+the area of the leaf surface to which it is spatial closest as defined
+by Voronoi tessellation techniques. We refer to this as the `stomatal
+zone', the leaf area surrounding a focal stomate closest to that stomate
+and therefore the zone it supplies with CO\(_2\)). This way, each
+stomate may be optimally sized relative to the mesophyll volume it
+supplies. Hypothesis 3 is already well supported in many species
+\citep{poorter_metaanalysis_2019}, but it is useful here to confirm that
+light treatments induced plasticity in the expected direction.
 
@@
 which affect diffusion path length (leaf thickness, stomatal density,
-leaf porosity, lateral-vertical diffusion rate ratio), diffusion rate
-(temperature, pressure), and CO\(_2\) demand (Rubisco concentration,
-light) would modulate the advantage of optimal stomatal arrangement
-following the relationships outlined in \autoref{tab:hypotheses}. Here,
-we integrate over reasonable parameter space to determine the
-ecophysiological context most likely to favor stomatal spatial
-coordination in amphistomatous leaves.
+leaf porosity), diffusion rate (determined by temperature, pressure),
+and CO\(_2\) demand (Rubisco concentration, light) would modulate the
+advantage of optimal stomatal arrangement following the relationships
+outlined in \autoref{tab:hypotheses}. Here, we integrate over reasonable
+parameter space to determine the ecophysiological context most likely to
+favor stomatal coordination in amphistomatous leaves.
 
@@
 
-\hypertarget{materials-and-methods}{%
-\section{Materials and methods}\label{materials-and-methods}}
+\section{Materials and methods}\label{materials-and-methods}
 
-\hypertarget{data-preparation}{%
-\subsection{Data Preparation}\label{data-preparation}}
+\subsection{Data Preparation}\label{data-preparation}
 
@@
 \(\mu \text{mol}~\text{m}^{-2}~\text{s}^{-1}\)). PAR stands for
-photosynthetically active radiation. Seeds were surface-sterilized and
-stratified at 4°C for 3--5 d in 0.15\% agarose solution and then sown
+photosynthetically active radiation. \emph{A. thaliana} responds
+strongly to light levels over this range
+\citep{bailey_acclimation_2001}, though natural populations in open
+canopies can experience
+\(\text{PAR} > 800~\mu \text{mol}~\text{m}^{-2}~\text{s}^{-1}\)
+\citep{callahan_shade-induced_2002}. Seeds were surface-sterilized and
+stratified at 4 °C for 3--5 d in 0.15\% agarose solution and then sown
 directly into Pro-Mix HP soil (Premier Horticulture; Quakerstown, PA,
@@
 internal leaf structures was performed using a Leica SP5 confocal
-microscope with the protocol developed by
-\citet{wuyts_high-contrast_2010} with additional modification described
-in \citet{dow_disruption_2017}. We captured 132 images in total, making
-66 abaxial-adaxial image pairs. We measured stomatal position and size
-using ImageJ \citep{schneider_nih_2012}.
+microscope (Leica Microsystems, Wetzlar, Germany) with the protocol
+developed by \citet{wuyts_high-contrast_2010} with additional
+modification described in \citet{dow_disruption_2017}. We captured 132
+images in total, making 66 abaxial-adaxial image pairs. Images were
+square with an area of 0.386 mm\(^2\). We measured stomatal position and
+length using ImageJ \citep{schneider_nih_2012}. A number of synthetic
+leaf surface data sets were also simulated (details below) to generate
+null distributions against which to test our hypotheses and to avoid any
+methodological influence on our results (e.g.~boundary effects when
+calculating stomatal patterning). All synthetic leaf surfaces were
+simulated based on the size of the real leaf images and stomatal
+densities matched those of real leaf images.
 
-\hypertarget{single-surface-analyses}{%
-\subsection{Single surface analyses}\label{single-surface-analyses}}
+\subsection{Single surface analyses}\label{single-surface-analyses}
 
-We tested whether stomata are non-randomly distributed by comparing the
-observed stomatal patterning to a random uniform pattern. For each leaf
-surface image with \(n\) stomata we generated \(10^3\) synthetic
-surfaces with \(n\) stomata uniformly randomly distributed on the
-surface. For each sample image, we compared the observed Nearest
-Neighbor Index (\(\mathrm{NNI}\)) to the null distribution of
-\(\mathrm{NNI}\) values calculated from the synthetic data set.
-\(\mathrm{NNI}\) is the ratio of observed mean distance
-(\(\overline{D}_O\)) to the expected mean distance (\(\overline{D}_E\))
-where \(\overline{D}_E\) is:
+We compared observed stomatal patterning to an ideal pattern (uniform
+equilateral triangular grid) and a null model (random uniform
+distribution). The terminology is unfortunately confusing because the
+word `uniform' is used in different ways. A uniform equilateral
+triangular grid means that the distance between stomata is uniform; a
+random uniform distribution means that a stomate has an equal
+probability (i.e.~uniform) of occurring anyone on the leaf surface. To
+limit confusion, we refer to the ideal pattern as uniform (equilateral
+triangle grid) and the null pattern as random (random uniform). When
+observed stomatal patterns are more dispersed than expected under random
+patterning, we refer to this as overdispersed. Note, however, that
+overdispersed compared to random is still less dispersed than ideal,
+because the ideal pattern is maximally dispersed.
+
+We tested whether stomata overdispersed by comparing each observed, real
+leaf stomatal pattern to an array of synthetic data simulated from a
+random distribution. For each observed leaf surface image with \(n\)
+stomata we generated \(10^3\) synthetic surfaces with \(n\) stomata
+uniformly randomly distributed on the surface. For each sample image, we
+compared the observed Nearest Neighbor Index (\(\mathrm{NNI}\)) to the
+null distribution of \(\mathrm{NNI}\) values calculated from the
+synthetic data set. \(\mathrm{NNI}\) is the ratio of observed mean
+distance (\(\overline{D}_O\)) to the expected mean distance
+(\(\overline{D}_E\)) where \(\overline{D}_E\) is:
 
@@
 its nearest neighbor. We calculated \(\mathrm{NNI}\) using the \emph{R}
-package \textbf{spatialEco} version 2.0.1 \citep{evans_spatialeco_2023}.
-The observed stomatal distribution is dispersed relative to a uniform
-random distribution if the observed \(NNI\) is greater than 95\% of the
+package \textbf{spatialEco} version 2.0.2 \citep{evans_spatialeco_2023}.
+The observed stomatal distribution is overdispersed relative to a random
+distribution if the observed \(NNI\) is greater than 95\% of the
 synthetic \(\mathrm{NNI}\) values (one-tailed test). We adjusted
@@
 
-For each sample image, we also simulated \(10^3\) synthetic data with
-\(n\) stomata ideally dispersed in an equilateral triangular grid. For
-these grids, we integrated over plausible stomatal densities and then
-conditioned on stomatal grids with exactly \(n\) stomata. The simulated
-stomatal count was drawn from a Poisson distribution with the mean
-parameter \(\lambda\) drawn from a Gamma distribution with shape \(n\)
-and scale 1 (\(\lambda \sim \Gamma(n, 1)\)). \(\Gamma(n, 1)\) is the
-posterior distribution of \(\lambda\) with a flat prior distribution.
-This allows us to integrate over uncertainty in the stomatal density
-from the sample image.
+For each sample image, we also simulated \(10^3\) synthetic leaf
+surfaces with \(n\) stomata ideally, uniformly dispersed in an
+equilateral triangular grid. To account for uncertainty in the stomatal
+density of each sample image with \(n\) stomata, we integrated over
+plausible stomatal densities and then conditioned on synthetic leaf
+surfaces with exactly \(n\) stomata. The simulated stomatal count was
+drawn from a Poisson distribution with the mean parameter \(\lambda\)
+drawn from a Gamma distribution with shape \(n\) and scale 1
+(\(\lambda \sim \Gamma(n, 1)\)). \(\Gamma(n, 1)\) is the posterior
+distribution of \(\lambda\) with a flat prior distribution. This
+integration was necessary to remove any artifacts of uncertainty in the
+true stomatal density of the sample leaves.
 
 We developed a dispersion index \(\mathrm{DI}\) to quantify how close
-observed stomatal distributions are to random uniform versus maximally
-dispered in an equilateral triangular grid. \(\mathrm{DI}\) varies from
-zero to one, where zero is uniformly random and one is ideally
-dispersed:
+observed stomatal patterning is to random versus ideally patterned in an
+equilateral triangular grid. \(\mathrm{DI}\) varies from zero to one,
+where zero is random and one is ideally patterned:
 
 \begin{equation}\label{eq:disp}
-  \mathrm{DI} = \frac{\mathrm{NNI} - \text{median}(\mathrm{NNI_{random}})}{\text{median}(\mathrm{NNI_{uniform}}) - \text{median}(\mathrm{NNI_{random}})}
+  \mathrm{DI} = \frac{\mathrm{NNI} - \text{median}(\mathrm{NNI_{random}})}{\text{median}(\mathrm{NNI_{ideal}}) - \text{median}(\mathrm{NNI_{random}})}
 \end{equation}
@@
 length and \(\sqrt{S_\text{area}}\) using the \emph{emtrends} function
-in the \emph{R} package \textbf{emmeans} version 1.8.8
+in the \emph{R} package \textbf{emmeans} version 1.9.0
 \citep{lenth_emmeans_2023}.
@@
 
-\hypertarget{paired-abaxial-and-adaxial-surface-analysis}{%
 \subsection{Paired Abaxial and Adaxial Surface
-Analysis}\label{paired-abaxial-and-adaxial-surface-analysis}}
+Analysis}\label{paired-abaxial-and-adaxial-surface-analysis}
 
@@
 (observed or synthetic) we calculated the distance squared between each
-pixel to the nearest stomatal centroid with the \emph{R} package
-\textbf{raster} version 3.6.26. We refer to this as the `nearest
-stomatal distance' or NSD. Then we calculated the pixel-wise Pearson
-correlation coefficient. If stomatal positions on each surface are
-coordinated to minimize the distance between mesophyll and the nearest
-stomate, then we expect a negative correlation. A pixel that is far from
-a stomate on one surface should be near a stomate on the other surface
-(\autoref{fig:ideal-amphi-grid}). We generated a null distribution of
-the correlation coefficient by simulating \(10^3\) synthetic data sets
-for each observed pair. For each synthetic data set, we simulated
-stomatal position using a random uniform distribution, as described
-above, matching the number of stomata on abaxial and adaxial leaf
-surfaces. Stomatal positions on each surface are coordinated if the
-correlation coefficient is greater than 95\% of the synthetic
-correlation values (one-tailed test).
+pixel of the surface to the nearest stomatal centroid with the \emph{R}
+package \textbf{raster} version 3.6.26 \citep{hijmans_raster_2023}. We
+refer to this as the `nearest stomatal distance' or NSD. Then we
+calculated the pixel-wise Pearson correlation coefficient. If stomatal
+positions on each surface are coordinated to minimize the distance
+between mesophyll and the nearest stomate, then we expect a negative
+correlation. A pixel that is far from a stomate on one surface should be
+near a stomate on the other surface (\autoref{fig:ideal-amphi-grid}). We
+generated a null distribution of the correlation coefficient by
+simulating \(10^3\) synthetic data sets for each observed pair. For each
+synthetic data set, we simulated stomatal position using a random
+uniform distribution, as described above, matching the number of stomata
+on abaxial and adaxial leaf surfaces to the observed data. Stomatal
+positions on each surface are coordinated if the correlation coefficient
+of the NSD between observed ab- and adaxial surfaces is greater than
+95\% of the synthetic correlation values (one-tailed test).
 
-\hypertarget{modeling-photosynthesis}{%
-\subsection{Modeling Photosynthesis}\label{modeling-photosynthesis}}
+\subsection{Modeling Photosynthesis}\label{modeling-photosynthesis}
 
@@
 assessed is the advantage of offsetting the position of stomata on each
-surface compared to have stomata on the same \(x\) position on each
+surface compared to having stomata on the same \(x\) position on each
 surface. With these assumptions, by symmetry, we only need to model two
@@
 
-\hypertarget{results}{%
-\section{Results}\label{results}}
+\section{Results}\label{results}
 
@@
 
-\hypertarget{stomatal-distribution-is-non-random-but-far-from-ideal}{%
-\subsection{Stomatal distribution is non-random, but far from
-ideal}\label{stomatal-distribution-is-non-random-but-far-from-ideal}}
+\subsection{Stomatal patterning is non-random, but far from
+uniform}\label{stomatal-patterning-is-non-random-but-far-from-uniform}
 
@@
 compared to a random uniform distribution, but none were close to an
-ideal hexagonal pattern (dispersion index = 1;
+ideal, uniform equilateral triangular pattern (dispersion index = 1;
 \autoref{fig:single-surface}). Before controlling for multiple
@@
 \includegraphics[width=4in]{figures/single-surface.pdf}
-\caption{Stomata are more dispersed than expected under the null model of uniform random position (dispersion index = 0) but far from a distribution that maximizes distance between stomata (dispersion index = 1).
+\caption{Stomata are more dispersed than expected under the null model of random patterning (dispersion index = 0) but far from a distribution that maximizes distance between stomata (dispersion index = 1; uniform patterning).
 We determined statistical significance between light treatments using Tukey post-hoc tests.
@@
 
-\hypertarget{no-evidence-for-coordinated-stomatal-position-between-surfaces}{%
 \subsection{No evidence for coordinated stomatal position between
-surfaces}\label{no-evidence-for-coordinated-stomatal-position-between-surfaces}}
+surfaces}\label{no-evidence-for-coordinated-stomatal-position-between-surfaces}
 
@@
 
-\hypertarget{larger-stomata-supply-larger-mesophyll-volumes}{%
 \subsection{Larger stomata supply larger mesophyll
-volumes}\label{larger-stomata-supply-larger-mesophyll-volumes}}
+volumes}\label{larger-stomata-supply-larger-mesophyll-volumes}
 
-All parameters converged (\(\hat{R} < 1.01\)) and effective sample sizes
-were exceeded \(10^3\). Across all light treatments and leaf surfaces,
-stomatal length and stomatal area are weakly positively correlated
+All parameters in the Bayesian linear mixed-effects model converged
+(\(\hat{R} < 1.01\)) and effective sample sizes exceeded \(10^3\).
+Across all light treatments and leaf surfaces, stomatal length and
+stomatal area are weakly positively correlated
 (\autoref{fig:length-area}). The slope was significantly greater than
@@
 
-\hypertarget{little-benefit-of-coordinated-stomatal-arrangement}{%
 \subsection{Little benefit of coordinated stomatal
-arrangement}\label{little-benefit-of-coordinated-stomatal-arrangement}}
+arrangement}\label{little-benefit-of-coordinated-stomatal-arrangement}
 
@@
 
-\hypertarget{discussion}{%
-\section{Discussion}\label{discussion}}
+\section{Discussion}\label{discussion}
 
@@
 reducing this distance to its absolute minimum may be constrained by
-developmental processes, or the photosynthetic benefit may be too small
+developmental processes or the photosynthetic benefit may be too small
 to be `seen' by natural selection (i.e.~the selection coefficient is
 less than drift barrier \emph{sensu} \citet{sung_drift-barrier_2012}).
+We also consider that our definition of optimal may be incorrect because
+it is based on overly simplistic assumptions about leaf mesophyll
+structure.
 
@@
 that stomata on each surface are overdispersed relative to a random
-uniform distribution, which should increase CO\(_2\) supply. Stomata on
-each surface are overdispersed (\autoref{fig:single-surface}), but are
-not ideally dispersed in an equilateral triangular grid as would be
-optimal to minimize CO\(_2\) diffusion path length and equalize the area
-supplied by each stomate (\autoref{fig:tessellation-example}). Second,
-we predicted that an optimal amphistomatous leaf has offset stomata such
-that stomata are more likely to appear on one leaf surface if there is
-not a stomata directly opposite it on the other surface as shown in
-\autoref{fig:ideal-amphi-grid}. However, there is no evidence for
-coordination and the positions on each surface appear independent,
+distribution, which should increase CO\(_2\) supply. Stomata on each
+surface are overdispersed (\autoref{fig:single-surface}), but are not
+ideally, uniformly patterned in an equilateral triangular grid as would
+be optimal to minimize CO\(_2\) diffusion path length and equalize the
+area supplied by each stomate (\autoref{fig:tessellation-example}).
+Second, we predicted that an optimal amphistomatous leaf has offset
+stomata such that stomata are more likely to appear on one leaf surface
+if there is not a stomata directly opposite it on the other surface as
+shown in \autoref{fig:ideal-amphi-grid}. However, there is no evidence
+for coordination and the positions on each surface appear independent,
 regardless of light treatment (\autoref{fig:dual-surface}). Third, we
@@
 increasing stomatal density. \emph{Arabidopsis} plants grown under high
-light had higher stomatal density than the same genotype growns under
-low and medium light intensity (\autoref{fig:density}). However, we
-found no support for our fourth prediction that stomata would be more
-evenly dispersed at high light intensity (\autoref{fig:single-surface}).
+light had higher stomatal density than the same genotype grown under low
+and medium light intensity (\autoref{fig:density}). However, we found no
+support for our fourth prediction that stomatal patterning would be
+overdispersed at high light intensity (\autoref{fig:single-surface}).
 Finally, we predicted that within leaf variation in stomatal size would
@@
 volumes of adjacent mesophyll. In all three light treatments, stomatal
-size positively covaried with the stomatal zone, i.e.~adjacent region of
-mesophyll that would be supplied by that stomate
+size positively co-varied with the stomatal zone, i.e.~adjacent region
+of mesophyll that would be supplied by that stomate
 (\autoref{fig:length-area}).
@@
 There are three non-mutually exclusive hypotheses for why several of our
-predictions were wrong. First, our predictions are wrong becuase they
-are based on overly simplistic assumptions about epidermal and mesophyll
-anatomy. Second, natural selection may be constrained by developmental
-processes that prevent phenotypes from reaching their adaptive optima.
-Third, the benefit of some traits may be of too little consequence to
-result in fitness differences large enough to respond to selection. We
-consider the plausibility of these alternative hypotheses below and
-present ideas for future work to test them.
+predictions were wrong. First, our predictions must be wrong because
+they are based on overly simplistic assumption of a homogenous porous
+medium within the mesophyll. Real leaf mesophylls are spatially
+heterogenous and chloroplasts are distributed as discrete nodes. The
+intercellular air space conductance is determined by its porosity and
+tortuosity, both of which are heterogenous within the leaf. The palisade
+is typically less porous than the spongy mesophyll (e.g.
+\citet{therouxrancourt_bias_2017}), which should impact the optimal
+patterning on stomata on ab- versus adaxial surfaces. Tortuosity is also
+systematically greater in the palisade in the lateral direction parallel
+to the leaf plane \citep{harwood_understanding_2021}. We might predict a
+greater coordination advantage of offset stomata by accounting for
+greater lateral tortuosity, but it is likely that benefit is still very
+small under realistic parameter space. Quantifying the patterns of
+heterogeneity in porosity, tortuosity, and other factors
+\citep{earles_beyond_2018} using 3D imaging (e.g.
+\citet{borsuk_structural_2022}) will be needed to generate more
+realistic hypotheses about optimal stomatal spacing.
+
+Second, spatio-temporal variation of internal conditions within leaves
+and between stomatal responses may make uniform, coordinated stomatal
+surfaces less beneficial
+\citep{weyers_heterogeneity_1997, lawson_nature_1998, lawson_spatial_1999}.
+This is because our model assumes a uniform leaf, the internal
+conditions of which are periodic and solved empirically and therefore
+stable. Any horizontal concentration gradients due to environmental
+heterogeneity and variable induction times for interacting leaf
+processes may reduce the benefit of uniform stomatal patterning. Third,
+natural selection may be constrained by developmental processes that
+prevent phenotypes from reaching their adaptive optima. Stomatal
+development must be plastic to environmental cues interpreted through
+long distance and cell-to-cell signalling pathways
+\citep{pillitteri_mechanisms_2012}. This plasticity may come with the
+cost of being unable to orchestrate the development of an absolutely
+uniform stomatal grid. Fourth, the benefit of some traits may be of too
+little consequence to result in fitness differences large enough to
+respond to selection. We consider the plausibility of these alternative
+hypotheses below and present ideas for future work to test them.
 
@@
 provide pathways for CO\(_2\) diffusion, but must supply water,
-intercept light, and deter herbivores and pathogens. These competing
-interests result in nonuniform epidermal and mesophyll structure that
-could alter our predictions about optimal stomatal spacing. In order to
+intercept light, and deter herbivores and pathogens. All of these
+competing processes also happen on different time scales, and can be
+observed as heterogeneity in stomatal density, aperture, and internal
+leaf conditions across the leaf at any given moment
+\citep{lawson_nature_1998, lawson_spatial_1999}. These competing
+interests result in heterogeneous epidermal and mesophyll structure that
+could alter predictions about optimal stomatal spacing. In order to
 maintain consistent leaf water potential across the lamina, stomatal
@@
 not at the interstomatal level, but at a higher level, coordinating
-water transport and water loss. We also assume a uniform mesophyll;
-however, in ab-adaxially differentiated leaves, this assumption is
-overly simplistic. The pallisade mesophyll is more tightly packed than
-the spongy mesophyll as an adaptation to intercept light efficiently, so
-lateral diffusion may be more limiting in the adaxial portion of the
-leaf. This may explain why adaxial leaf surfaces have consistently
-higher dispersion indices than abaxial surfaces across all light
-treatments (\autoref{fig:single-surface}). Future gas exchange models
-should include multiple parameter sets for different strata of mesophyll
-tissue.
+water transport and water loss. For example, the palisade mesophyll is
+more tightly packed than the spongy mesophyll as an adaptation to
+intercept light efficiently, so lateral diffusion may be more limiting
+in the adaxial portion of the leaf. This may explain why adaxial leaf
+surfaces have consistently higher dispersion indices than abaxial
+surfaces across all light treatments (\autoref{fig:single-surface}).
+Future gas exchange models should incorporate heterogenous mesophyll
+structure and hydraulic traits such as veins.
 
-No developmental pathway exists to ensure an idealized placement of
-stomata on the leaf surface. Rather, stomatal development is a dynamic
-process that must be plastic to environmental cues. Leaves develop based
-on short and long distance signalling pathways which relay information
-about incoming light, humidity, temperature, and surrounding stomata to
-developing leaf tissues \citep{pillitteri_mechanisms_2012}. Our results
-showing an intermediate level of dispersion in stomatal spacing may be
-best explained by these developmental pathways which ensure the proper
-spacing of stomata, with an added random effect brought about by the
-necessity for plasticity in stomatal development
-(\autoref{fig:single-surface}). However, any nonideal randomness in
-stomatal spacing may offset by simultaneous and coordinated development
+We are not aware of a developmental pathway that ensures an idealized
+placement of stomata on the leaf surface. Rather, stomatal development
+is a dynamic process that must be plastic to environmental cues. Leaves
+develop based on short and long distance signalling pathways which relay
+information about incoming light, humidity, temperature, and surrounding
+stomata to developing leaf tissues \citep{pillitteri_mechanisms_2012}.
+Our results showing an intermediate level of dispersion in stomatal
+spacing may be best explained by these developmental pathways which
+ensure the proper spacing of stomata, with an added random effect
+brought about by the necessity for plasticity in stomatal development
+(\autoref{fig:single-surface}). However, deviations from ideal stomatal
+spacing may compensated for by simultaneous and coordinated development
 of the IAS \citep{baillie_developmental_2020}. The fact that stomata
 which supply a greater mesophyll volume tend to be larger suggests that
-plants may use coordinated development of multiple leaf tissues to
-negate nonideal stomatal spacing (\autoref{fig:length-area}).
+plants may use coordinated development of multiple leaf anatomical
+features to compensate for nonideal stomatal spacing
+(\autoref{fig:length-area}).
 
@@
 plants from reaching a theoretical peak in the adaptive landscape:
-uniform stomatal dispersion. Instead, as this study suggests, plants may
-simply compensate with higher stomatal density and by fitting stomatal
-size to the area that they supply with CO\(_2\). To understand why
-stomata are not ideally dispersed, more modeling (with a more complex
-set of assumptions including vein density and IAS structure) should be
-done to estimate the fitness gain of stomatal dispersion. Additionally,
-genetic manipulation studies should attempt to create mutants with
-clustered and ideally dispersed stomata for a comparison of their
-photosynthetic traits. This could have extremely important implications
-for maximum assimilation rates in crops as most crop species are grown
-in high light where CO\(_2\) is often limiting. In drought-prone
-environments, increased stomatal dispersion may increase water use
-efficiency by reducing the number of stomata needed to achieve the same
-internal CO\(_2\) concentration, \(C_\text{i}\).
+uniform stomatal patterning. Instead, as this study suggests, plants may
+simply compensate with higher stomatal density and by modulating
+stomatal size to the area that they supply with CO\(_2\). To understand
+why stomata are not ideally dispersed, more modeling (with more
+realistic assumptions including vein density and IAS structure) should
+be done to estimate the photosynthetic properties of varying stomatal
+patterning. Additionally, genetic manipulation studies should attempt to
+create mutants with clustered and uniformly patterned stomata for a
+comparison of their photosynthetic traits. This could have important
+implications for maximizing assimilation rates in crops as most crop
+species are grown in high light where CO\(_2\) is often limiting. In
+drought-prone environments, increased stomatal dispersion may increase
+water use efficiency by reducing the number of stomata needed to achieve
+the same internal CO\(_2\) concentration, \(C_\text{i}\). However, it
+would be necessary to account for many other differences between
+\emph{A. thaliana} and crop leaves and canopies.
 
@@
 considered here, could also have significant implications, as CO\(_2\)
-may only be limiting to photosynthesis during short periods when all
-other conditions are ideal. In these cases, the theoretical benefits of
-ideal stomatal spacing are further diminished. Future exploration of
-these competing hypotheses would require more advanced modeling,
-additional exploration of IAS space development and its effects on gas
-exchange, both real and modeled, and knowledge about how often the
-species of interest is CO\(_2\) limited across of range of natural
-settings. Despite these additional considerations, this study represents
-an important contribution to understanding the potential drivers of and
-limitations to stomatal anatomy in amphistomatous plants.
+may only be limiting to photosynthesis during short, relatively rare
+periods when all other conditions are ideal. In these cases, the
+theoretical benefits of ideal stomatal spacing are further diminished.
+Future exploration of these competing hypotheses would require more
+advanced modeling, additional exploration of IAS space development and
+its effects on gas exchange, both real and modeled, and knowledge about
+how often the species of interest is CO\(_2\) limited across of range of
+natural settings. Despite these additional considerations, this study
+represents an important contribution to understanding the potential
+drivers of and limitations to stomatal anatomy in amphistomatous plants.
 

### 2024-01-22 ms/response-to-reviewers.qmd
fixed some typos in revised ms and response to reviewers

@@
 ```
-
 ## Response to Reviewer #1
@@
 The terminology is unfortunately confusing because the word `uniform' is used in different ways.
-A uniform equilateral triangular grid means that the distance between stomata is uniform; a random uniform distribution means that a stomate has an equal probability (i.e.
-uniform) of occurring anyone on the leaf surface.
-To limit confusion, we refer to the ideal pattern as uniform (equilateral triangle grid) and the null pattern as random (random uniform).
-When observed stomatal patterns are more dispersed than expected under random patterning, we refer to this as over-dispersed.
-Note, however, that over-dispersed compared to random is still less dispersed than ideal, because the ideal pattern is maximally dispersed."
+A uniform equilateral triangular grid means that the distance between stomata is uniform; a random uniform distribution means that a stomate has an equal (i.e.
+uniform) probability of occurring anywhere on the leaf surface.
+To limit confusion, we refer to the ideal pattern (equilateral triangle grid) as uniform and the null pattern (random uniform) as random.
+When observed stomatal patterns are more dispersed than expected under random patterning, we refer to this as overdispersed.
+Note, however, that overdispersed compared to random is still less dispersed than ideal, because the ideal pattern is maximally dispersed."
 
@@
 
-Line 168 – "All parameters....exceeded 103" – please clarify what you mean by this statement.
+Line 168 -- "All parameters....exceeded 103" -- please clarify what you mean by this statement.
 
@@
 I only have a few comments, most of them concerning the assumptions.
-I think the consequences of their assumptions could be better addressed and explained with further details, especially the concept of “optimal distribution”.
+I think the consequences of their assumptions could be better addressed and explained with further details, especially the concept of "optimal distribution".
 

### 2024-01-22 ms/ms.Rmd
fixed some typos in revised ms and response to reviewers

@@
     email: graham.dow@niab.com
-    affiliation: NIAB
+    affiliation: NIAB g
   - name: Thomas N. Buckley
@@
 The terminology is unfortunately confusing because the word `uniform' is used in different ways.
-A uniform equilateral triangular grid means that the distance between stomata is uniform; a random uniform distribution means that a stomate has an equal probability (i.e.
-uniform) of occurring anyone on the leaf surface.
-To limit confusion, we refer to the ideal pattern as uniform (equilateral triangle grid) and the null pattern as random (random uniform).
+A uniform equilateral triangular grid means that the distance between stomata is uniform; a random uniform distribution means that a stomate has an equal (i.e.
+uniform) probability of occurring anywhere on the leaf surface.
+To limit confusion, we refer to the ideal pattern (equilateral triangle grid) as uniform and the null pattern (random uniform) as random.
 When observed stomatal patterns are more dispersed than expected under random patterning, we refer to this as overdispersed.

### 2024-01-22 ms/ms.tex
fixed some typos in revised ms and response to reviewers

@@
 \author[%
-\addrlblbycode{NIAB}%
+\addrlblbycode{NIAB g}%
 %
@@
 
-\addrbycode{NIAB}
+\addrbycode{NIAB g}
 
@@
 
-\section{Introduction}\label{introduction}
+\hypertarget{introduction}{%
+\section{Introduction}\label{introduction}}
 
@@
 
-\section{Materials and methods}\label{materials-and-methods}
-
-\subsection{Data Preparation}\label{data-preparation}
+\hypertarget{materials-and-methods}{%
+\section{Materials and methods}\label{materials-and-methods}}
+
+\hypertarget{data-preparation}{%
+\subsection{Data Preparation}\label{data-preparation}}
 
@@
 
-\subsection{Single surface analyses}\label{single-surface-analyses}
+\hypertarget{single-surface-analyses}{%
+\subsection{Single surface analyses}\label{single-surface-analyses}}
 
@@
 random uniform distribution means that a stomate has an equal
-probability (i.e.~uniform) of occurring anyone on the leaf surface. To
-limit confusion, we refer to the ideal pattern as uniform (equilateral
-triangle grid) and the null pattern as random (random uniform). When
+(i.e.~uniform) probability of occurring anywhere on the leaf surface. To
+limit confusion, we refer to the ideal pattern (equilateral triangle
+grid) as uniform and the null pattern (random uniform) as random. When
 observed stomatal patterns are more dispersed than expected under random
@@
 rate procedure implemented in the \emph{R} package \textbf{multtest}
-version 2.56.0 \citep{wong_multiple_2005}.
+version 2.58.0 \citep{wong_multiple_2005}.
 
@@
 
+\hypertarget{paired-abaxial-and-adaxial-surface-analysis}{%
 \subsection{Paired Abaxial and Adaxial Surface
-Analysis}\label{paired-abaxial-and-adaxial-surface-analysis}
+Analysis}\label{paired-abaxial-and-adaxial-surface-analysis}}
 
@@
 
-\subsection{Modeling Photosynthesis}\label{modeling-photosynthesis}
+\hypertarget{modeling-photosynthesis}{%
+\subsection{Modeling Photosynthesis}\label{modeling-photosynthesis}}
 
@@
 
-\section{Results}\label{results}
+\hypertarget{results}{%
+\section{Results}\label{results}}
 
@@
 
+\hypertarget{stomatal-patterning-is-non-random-but-far-from-uniform}{%
 \subsection{Stomatal patterning is non-random, but far from
-uniform}\label{stomatal-patterning-is-non-random-but-far-from-uniform}
+uniform}\label{stomatal-patterning-is-non-random-but-far-from-uniform}}
 
@@
 
+\hypertarget{no-evidence-for-coordinated-stomatal-position-between-surfaces}{%
 \subsection{No evidence for coordinated stomatal position between
-surfaces}\label{no-evidence-for-coordinated-stomatal-position-between-surfaces}
+surfaces}\label{no-evidence-for-coordinated-stomatal-position-between-surfaces}}
 
@@
 
+\hypertarget{larger-stomata-supply-larger-mesophyll-volumes}{%
 \subsection{Larger stomata supply larger mesophyll
-volumes}\label{larger-stomata-supply-larger-mesophyll-volumes}
+volumes}\label{larger-stomata-supply-larger-mesophyll-volumes}}
 
@@
 
+\hypertarget{little-benefit-of-coordinated-stomatal-arrangement}{%
 \subsection{Little benefit of coordinated stomatal
-arrangement}\label{little-benefit-of-coordinated-stomatal-arrangement}
+arrangement}\label{little-benefit-of-coordinated-stomatal-arrangement}}
 
@@
 
-\section{Discussion}\label{discussion}
+\hypertarget{discussion}{%
+\section{Discussion}\label{discussion}}
 

### 2024-03-03 ms/ms.Rmd
reran simulation and analysis scripts for final ms version

--- 2024-01-21 ms/ms.Rmd
@@
 
+# Data availability
+
+Custom scripts are available on a GitHub repository (https://github.com/cdmuir/stomata-spacing) and archived on Zenodo: https://doi.org/XXXX Raw data are deposited on Dryad: https://doi.org/10.5061/dryad.44j0zpcn6
+
+# Acknowledgements
+
+The authors thank two anonymous reviewers for feedback on an earlier version of this manuscript, NSF Award 1929167 to C.D.M., and NSF Award 2307341 to T.N.B.
+This is publication #218 from the School of Life Sciences, University of Hawaiʻi at Mānoa.
+
 # References

### 2024-03-03 ms/ms.tex
reran simulation and analysis scripts for final ms version

--- 2024-01-21 ms/ms.tex
@@
 length and \(\sqrt{S_\text{area}}\) using the \emph{emtrends} function
-in the \emph{R} package \textbf{emmeans} version 1.9.0
+in the \emph{R} package \textbf{emmeans} version 1.10.0
 \citep{lenth_emmeans_2023}.
@@
 Stomatal density of \emph{Arabidopsis thaliana} varies among light
-treatments (ANOVA, \(F_{2,126} = 681, P = 2.88 \times 10^{-68}\))
+treatments (ANOVA, \(F_{2,126} = 682, P = 2.58 \times 10^{-68}\))
 because the density is much greater in the high light treatment
@@
 
-Many leaf surfaces (37 of 132, 28\%) are significantly overdispersed
+Many leaf surfaces (34 of 132, 25.8\%) are significantly overdispersed
 compared to a random uniform distribution, but none were close to an
@@
 \autoref{fig:single-surface}). Before controlling for multiple
-comparisons, 43.2\% are significantly overdispersed. The dispersion
+comparisons, 40.9\% are significantly overdispersed. The dispersion
 index differs significantly among light treatments (ANOVA,
-\(F_{2,126} = 8.55, P = 3.30 \times 10^{-4}\)) because the medium light
+\(F_{2,126} = 7.87, P = 6.02 \times 10^{-4}\)) because the medium light
 treatment is significantly less than the low treatment
@@
 on adaxial leaf surfaces across all light treatments (ANOVA,
-\(F_{1,126} = 28.8, P = 3.67 \times 10^{-7}\);
+\(F_{1,126} = 29.2, P = 3.19 \times 10^{-7}\);
 \autoref{fig:single-surface}). There is no evidence for an interaction
 between light treatment and surface (ANOVA,
-\(F_{2,126} = 0.577, P = 0.563\)).
+\(F_{2,126} = 0.594, P = 0.554\)).
 
@@
 
+\section{Data availability}\label{data-availability}
+
+Custom scripts are available on a GitHub repository
+(https://github.com/cdmuir/stomata-spacing) and archived on Zenodo:
+https://doi.org/XXXX Raw data are deposited on Dryad:
+https://doi.org/10.5061/dryad.44j0zpcn6
+
+\section{Acknowledgements}\label{acknowledgements}
+
+The authors thank two anonymous reviewers for feedback on an earlier
+version of this manuscript, NSF Award 1929167 to C.D.M., and NSF Award
+2307341 to T.N.B. This is publication \#218 from the School of Life
+Sciences, University of Hawaiʻi at Mānoa.
+
 \section{Competing interests}

### 2024-03-03 r/00_process-data.R

@@ -1,6 +1,7 @@
 source("r/header.R")
 
-stomata_position = readr::read_csv("raw-data/stomata_position.csv") |>
+stomata_position = readr::read_csv("raw-data/stomata_position.csv",
+                                   show_col_types = FALSE) |>
 
   # add light column
   dplyr::mutate(
@@ -22,7 +23,7 @@ stomata_position$image_number1 = as.character(gsub("^.{0,3}", "", stomata_positi
 # Check for exact duplicates
 stomata_position |>
   dplyr::group_by(trt_leaf_number, image_number, surface) |>
-  dplyr::summarize(is_dup = duplicated(tibble(x, y))) |>
+  dplyr::reframe(is_dup = duplicated(tibble(x, y))) |>
   dplyr::filter(is_dup)
 
 # identify near duplicates
@@ -47,8 +48,7 @@ stomata_tess = stomata_position |>
   dplyr::group_modify(~ tessellate_stomata(.x, pixels_x = 512, pixels_y = 512)) |>
   dplyr::full_join(stomata_position,
                    by = c("trt_leaf_number", "image_number", "x", "y")) |>
-  mutate(area = ifelse(test = reject, NA, area))
+  dplyr::mutate(area = ifelse(test = reject, NA, area))
 
 readr::write_rds(stomata_position, "objects/stomata_position.rds")
 readr::write_rds(stomata_tess, "objects/stomata_position_length_area.rds")
-

### 2024-03-03 r/01_simulate-single-surface.R

@@ -1,6 +1,7 @@
 source("r/header.R")
 
-stomata = readr::read_csv("raw-data/stomata_position.csv") |>
+stomata = readr::read_csv("raw-data/stomata_position.csv",
+                          show_col_types = FALSE) |>
   group_by(treatment, leaf_number, image_number, surface)
 
 pb = progress_bar$new(
@@ -16,4 +17,4 @@ synthetic_data = stomata |>
     .collate = "rows"
   )
 
-write_rds(synthetic_data, "objects/synthetic_data.rds")
+readr::write_rds(synthetic_data, "objects/synthetic_data.rds")

### 2024-03-03 r/02_get-nni-single-surface.R

@@ -1,6 +1,12 @@
 source("r/header.R")
 
-stomata = readr::read_csv("raw-data/stomata_position.csv") |>
+# remove this note after everything is rerun
+# on 2024-03-02, CDM reran this script and found three slices where stomata_n had changed from raw-data and nni_observed, so re-running with update raw-data
+# tmp = readr::read_rds( "objects/nni_observed.rds")
+# which(nni_observed[,"n_stomata"] != tmp[,"n_stomata"])
+
+stomata = readr::read_csv("raw-data/stomata_position.csv",
+                          show_col_types = FALSE) |>
   dplyr::group_by(treatment, leaf_number, image_number, surface)
 
 synthetic_data = readr::read_rds("objects/synthetic_data.rds") |>
@@ -15,8 +21,9 @@ pb <- progress_bar$new(
 )
 
 nni_synthetic <- synthetic_data |>
-  purrrlyr::by_slice(get_nni, pixels_x = 512, pixels_y = 512, progress = TRUE,
-                     .to = "NNI_synthetic")
+  purrrlyr::by_slice(get_nni, pixels_x = 512, pixels_y = 512,
+                     progress = TRUE, .to = "NNI_synthetic")
 
 readr::write_rds(nni_observed, "objects/nni_observed.rds")
 readr::write_rds(nni_synthetic, "objects/nni_synthetic.rds")
+

### 2024-03-03 r/03_test-single-surface.R

@@ -5,7 +5,7 @@ nni_observed = readr::read_rds("objects/nni_observed.rds") |>
                 NNI_observed = NNI)
 
 nni_synthetic = readr::read_rds("objects/nni_synthetic.rds") |>
-  dplyr::select(treatment, leaf_number, image_number, surface, #n_stomata,
+  dplyr::select(treatment, leaf_number, image_number, surface,
                 sim, grid, NNI_synthetic)
 
 stomata_area_summ = readr::read_rds("objects/area_summ.rds")
@@ -48,12 +48,7 @@ single_surface_results <- nni_observed |>
       NNI_observed = ..6,
       p_value = length(which(sim_ran >= ..6)) / length(sim_ran),
       dispersion = (..6 - median(sim_ran)) / (median(sim_equ - median(sim_ran)))
-    ) #|>
-      #full_join(as.tibble(stomata_area_summ), by = c("treatment", "leaf_number", "image_number", "surface"))
-      #cbind(stomata_area_summ)
-
-    # I'm trying to merge these datasets using cbind and full_join. but I really
-    # don't know why it's not working.
+    )
 
     return(ret)
 
@@ -67,6 +62,3 @@ single_surface_results <- nni_observed |>
     light = tidyr::replace_na(light, "medium"))
 
 readr::write_rds(single_surface_results, "objects/single_surface_results.rds")
-
-
-

### 2024-03-03 r/04_plot-single-surface.R

@@ -32,7 +32,7 @@ plot = ggplot(
   geom_hline(yintercept = 1, linetype = "dashed") +
   stat_pvalue_manual(
     stat.test, label = "p.adj.signif",
-    y.position = c(.5, .575, .65),
+    y.position = c(.5, 0.59375, .65),
     inherit.aes = FALSE
   ) +
   theme_pubr()

### 2024-03-03 r/05_dual-surface-correlation.R

@@ -32,4 +32,3 @@ corr_synthetic = synthetic_data |>
 # write rds files
 readr::write_rds(corr_observed, "objects/corr_observed.rds")
 readr::write_rds(corr_synthetic, "objects/corr_synthetic.rds")
-

### 2024-03-03 r/06_test-dual-surface.R

@@ -32,4 +32,3 @@ dual_surface_results <- corr_observed |>
   )
 
 readr::write_rds(dual_surface_results, "objects/dual_surface_results.rds")
-

### 2024-03-03 r/07_plot-dual-surface.R

@@ -15,7 +15,7 @@ dual_surface_results = readr::read_rds("objects/dual_surface_results.rds") |>
 
 # anova
 correlation_anova = aov(corr ~ light, data = dual_surface_results)
-write_rds(correlation_anova, "objects/correlation_anova.rds")
+readr::write_rds(correlation_anova, "objects/correlation_anova.rds")
 
 stat.test.correlation = aov(corr ~ light, data = dual_surface_results) |>
   tukey_hsd()
@@ -30,4 +30,3 @@ plot = ggplot(dual_surface_results, aes(light_reorder, corr)) +
 
 ggsave("dual-surface.pdf", plot = plot, path = "ms/figures",
        width = 3, height = 3)
-

### 2024-03-03 r/08_fit-length-area.R

@@ -2,16 +2,28 @@ source("r/header.R")
 
 stomata = readr::read_rds("objects/stomata_position_length_area.rds") |>
   dplyr::ungroup() |>
-  filter(length > 5, !reject, !is.na(area)) |>
-  mutate(sqrt_area = sqrt(area), image = str_c(trt_leaf_number, image_number), sep = "-")
+  filter(length > 5,!reject,!is.na(area)) |>
+  mutate(
+    sqrt_area = sqrt(area),
+    image = str_c(trt_leaf_number, image_number),
+    sep = "-"
+  )
 
-fit = brm(bf(
-  sqrt_area ~ surface * light * length + (surface * length | trt_leaf_number) +
-    (surface * length | image),
-  sigma ~ light
-),
-data = stomata, backend = "cmdstanr", chains = 4L, cores = 4L,
-seed = 226140189, iter = 4e3, thin = 2, silent = 0
+fit = brm(
+  bf(
+    sqrt_area ~ surface * light * length + (surface * length |
+                                              trt_leaf_number) +
+      (surface * length | image),
+    sigma ~ light
+  ),
+  data = stomata,
+  backend = "cmdstanr",
+  chains = 4L,
+  cores = 4L,
+  seed = 226140189,
+  iter = 4e3,
+  thin = 2,
+  silent = 0
 )
 
 write_rds(fit, "objects/fit.rds")

### 2024-03-03 r/functions.R

@@ -416,13 +416,27 @@ get_nni = function(
   checkmate::assert_numeric(df_stomata$y, lower = 0, upper = pixels_y)
 
   df_stomata_sp <- df_stomata |>
+    dplyr::ungroup() |>
     dplyr::select(x, y) |>
-    sp::SpatialPoints()
-
-  df_stomata_sp@bbox <- matrix(c(0, 0, pixels_x, pixels_y), ncol = 2)
+    sf::st_as_sf(coords = c("x", "y"))
 
   # perform average nearest neighbor test
-  nni = nni(df_stomata_sp, win = "extent")
+  # modified from spatialEco::nni() to use custom extent with sf package
+  w <- spatstat.geom::as.owin(c(0, 512, 0, 512))
+  x <- spatstat.geom::as.ppp(sf::st_coordinates(df_stomata_sp)[, 1:2], w)
+  A <- spatstat.geom::area.owin(w)
+  obsMeanDist <- sum(spatstat.geom::nndist(x))/x$n
+  expMeanDist <- 0.5 * sqrt(A/x$n)
+  se <- 0.26136/((x$n^2/A)^0.5)
+  nni <- obsMeanDist/expMeanDist
+  z <- (obsMeanDist - expMeanDist)/se
+  nni = list(
+    NNI = nni,
+    z.score = z,
+    p = 2 * stats::pnorm(-abs(z)),
+    expected.mean.distance = expMeanDist,
+    observed.mean.distance = obsMeanDist
+  )
   nni$n_stomata <- nrow(df_stomata)
 
   # return result

### 2024-03-03 r/header.R

@@ -45,6 +45,7 @@ library(rgdal)
 library(rlist)
 library(rootSolve)
 library(rstatix)
+library(sf)
 library(sp)
 library(spatialEco)
 library(spatstat)

### 2024-03-03 r/scratch-photo_2d_pm.R

@@ -1,213 +1,0 @@
-# I think I did this before, but calculating g_liq from Evans et al. 2009
-# Measured range of r_liq = 25-300 m^2 chloroplast s bar / mol
-# 1 s / m  = 0.025 m^2 chloroplast s bar / mol
-# SO, 1000 - 12000 s / m
-# SO g_liq = 0.001 - 8.33e-5
-
-source("r/header.R")
-source("r/photo_2d_pm.R")
-
-library(furrr)
-plan(multisession, workers = 10)
-
-library(progressr)
-handlers(global = TRUE)
-handlers("progress", "beepr")
-
-input = crossing(
-  n_x = c(50, 100),
-  n_z = c(250, 500),
-  stomatal_arrangement = c("aligned", "offset"),
-  I_0 = round(seq(0.0, 0.001, 0.0001), 4)
-) %>%
-  split(~ seq_len(nrow(.)))
-
-run_photo_2d_pm = function(input, fit_rubisco) {
-
-  p = progressor(along = input)
-  out = future_map(input, \(.x, fit_rubisco) {
-
-    r = with(.x, list(I_0 = I_0, n_x = n_x, n_z = n_z))
-    ph2d = photo_2d_pm(.x$stomatal_arrangement, fit_rubisco = fit_rubisco, replace = r)
-    ph2d$parms$stomatal_arrangement = .x$stomatal_arrangement
-    ph2d$parms$A_n = calc_2d_pm_An(ph2d)
-    p()
-    ph2d
-
-  }, fit_rubisco = fit_rubisco)
-
-  out
-
-}
-
-tmp = run_photo_2d_pm(input, fit_rubisco)
-
-tmp |>
-  map("parms") |>
-  map_dfr(\(x) as_tibble(x[c("stomatal_arrangement", "n_z", "n_x", "I_0", "A_n")])) |>
-  ggplot(aes(I_0, A_n, color = stomatal_arrangement)) +
-  facet_grid(n_z ~ n_x) +
-  geom_point()
-
-
-r = list(I_0 = 0.0003, n_x = 50, n_z = 500)
-ph2d = photo_2d_pm("aligned", replace = r)
-
-ph2d = photo_2d_pm("aligned", replace = list(I_0 = 0.001))
-
-# Plot results
-df_C = expand.grid(
-  z = seq_len(ph2d$parms[["n_z"]]),
-  x = seq_len(ph2d$parms[["n_x"]]),
[…]

### 2024-03-04 ms/ms.Rmd
merged tables and placed in external document

@@
   - amphistomy
-  - Arabidopsis thaliana
+  - $\textit{Arabidopsis thaliana}$
   - CO$_2$ diffusion
@@
 acknowledgements: |
-  This work was supported by NSF Award 1929167 to CDM and NSF Award 2307341 to TNB.
+  The authors thank two anonymous reviewers for feedback on an earlier version of this manuscript, NSF Award 1929167 to C.D.M., and NSF Award 2307341 to T.N.B.
+This is publication #218 from the School of Life Sciences, University of Hawaiʻi at Mānoa.
 bibliography: stomata-spacing.bib
@@
 We also used a 2-D porous medium approximation of CO$_2$ diffusion and photosynthesis to predict the photosynthetic advantage of optimal versus suboptimal coordination in stomatal coordination between surfaces.
-Specifically, we predicted that traits which affect diffusion path length (leaf thickness, stomatal density, leaf porosity), diffusion rate (determined by temperature, pressure), and CO$_2$ demand (Rubisco concentration, light) would modulate the advantage of optimal stomatal arrangement following the relationships outlined in \autoref{tab:hypotheses}.
+Specifically, we predicted that traits which affect diffusion path length (leaf thickness, stomatal density, leaf porosity), diffusion rate (determined by temperature, pressure), and CO$_2$ demand (Rubisco concentration, light) would modulate the advantage of optimal stomatal arrangement following the relationships outlined in Table 1.
 Here, we integrate over reasonable parameter space to determine the ecophysiological context most likely to favor stomatal coordination in amphistomatous leaves.
 
-```{r hypotheses, results="asis"}
+```{r hypotheses, results="asis",eval=FALSE}
 
@@
 
-We modeled the coordination advantage over a range of leaf thicknesses, stomatal densities, photosynthetic capacities, and light environments to understand when offsetting stomatal position on each surface might deliver a significant photosynthetic advantage (\autoref{tab:model_var}).
+We modeled the coordination advantage over a range of leaf thicknesses, stomatal densities, photosynthetic capacities, and light environments to understand when offsetting stomatal position on each surface might deliver a significant photosynthetic advantage (Table 1).
 The complete model description is available in the Supporting Information.
 
-```{r, model_var, results="asis"}
+```{r, model_var, results="asis", eval=TRUE}
 #| label: tbl-model_var
@@
 
-model_var_caption = paste0("The parameter range of model variables tested for their effect on coordination advantage (\\autoref{eq:coordination_advantage}) using a 2-D porous medium approximation.
+# model_var_caption = paste0("The parameter range of model variables tested for their effect on coordination advantage (\\autoref{eq:coordination_advantage}) using a 2-D porous medium approximation.
 We used regularly spaced values within each range and simulated across all combinations.
@@
 ", symbol_descriptions)
-
-knitr::kable(
-  model_var, 
-  caption = model_var_caption,
-  label = "model_var",
-  booktabs = TRUE, format = "latex", escape = FALSE
-)
+# 
+# knitr::kable(
+#   model_var, 
+#   caption = model_var_caption,
+#   label = "model_var",
+#   booktabs = TRUE, format = "latex", escape = FALSE
+# )
 
@@
 \caption{There is little photosynthetic benefit of offsetting stomatal position each surface based on a 2-D model of photosynthesis.
-The coordination advantage (\autoref{eq:coordination_advantage}) is close to zero under nearly all of the parameter space \autoref{tab:model_var}, meaning that the photosynthetic rate of amphistomatous leaves with stomata optimally offset is nearly equal to leaves with stomata on each surface in the same position along the leaf plane.
+The coordination advantage (\autoref{eq:coordination_advantage}) is close to zero under nearly all of the parameter space (Table 1), meaning that the photosynthetic rate of amphistomatous leaves with stomata optimally offset is nearly equal to leaves with stomata on each surface in the same position along the leaf plane.
 `r symbol_descriptions`.}

### 2024-03-04 ms/ms.tex
merged tables and placed in external document

@@
   pdftitle={Does stomatal patterning in amphistomatous leaves minimize the CO\_2 diffusion path length within leaves?},
-  pdfkeywords={amphistomy, Arabidopsis thaliana, CO\(_2\)
+  pdfkeywords={amphistomy, \(\textit{Arabidopsis thaliana}\), CO\(_2\)
 diffusion, finite element method, optimality, photosynthesis, stomata},
@@
 
-\keywords{amphistomy; Arabidopsis thaliana; CO\(_2\) diffusion; finite
-element method; optimality; photosynthesis; stomata}
+\keywords{amphistomy; \(\textit{Arabidopsis thaliana}\); CO\(_2\)
+diffusion; finite element method; optimality; photosynthesis; stomata}
 
@@
 advantage of optimal stomatal arrangement following the relationships
-outlined in \autoref{tab:hypotheses}. Here, we integrate over reasonable
-parameter space to determine the ecophysiological context most likely to
-favor stomatal coordination in amphistomatous leaves.
-
-\begin{table}[ht]
-\centering
-\begin{tabular}{ll}
-  \hline
-trait & relationship \\ 
-  \hline
-leaf thickness & + \\ 
-  interstomatal distance & + \\ 
-  leaf porosity & - \\ 
-  light & + \\ 
-   \hline
-\end{tabular}
-\caption{A summary of the hypothesized relationships between leaf traits and environmental conditions and photosynthetic advantage of stomatal spatial coordination in amphistomatous leaves.} 
-\label{tab:hypotheses}
-\end{table}
+outlined in Table 1. Here, we integrate over reasonable parameter space
+to determine the ecophysiological context most likely to favor stomatal
+coordination in amphistomatous leaves.
 
@@
 understand when offsetting stomatal position on each surface might
-deliver a significant photosynthetic advantage
-(\autoref{tab:model_var}). The complete model description is available
-in the Supporting Information.
-
-\begin{table}
-
-\caption{\label{tab:model_var}The parameter range of model variables tested for their effect on coordination advantage (\autoref{eq:coordination_advantage}) using a 2-D porous medium approximation.
-We used regularly spaced values within each range and simulated across all combinations.
-Here we converted model units to more conventional units (e.g.
-m to $\mu$m).
-$I_0$: PPFD incident on the leaf surface; $\varphi_\text{pal}$: Fraction of intercellular airspace (aka porosity), palisade; $T_\text{leaf}$: Leaf thickness; $U$: Interstomatal distance}
-\centering
-\begin{tabular}[t]{lll}
-\toprule
-Variable & Parameter range & Units\\
-\midrule
-$I_0$ & $50-1000$ & $\mu$mol m$^{-2}$ s$^{-1}$\\
-$\varphi_\text{pal}$ & $0.1-0.3$ & m$^3$ airspace m$^{-3}$ leaf\\
-$T_\text{leaf}$ & $101-501$ & $\mu$m\\
-$U$ & $17-169$ & $\mu$m\\
-\bottomrule
-\end{tabular}
-\end{table}
+deliver a significant photosynthetic advantage (Table 1). The complete
+model description is available in the Supporting Information.
 
@@
 \caption{There is little photosynthetic benefit of offsetting stomatal position each surface based on a 2-D model of photosynthesis.
-The coordination advantage (\autoref{eq:coordination_advantage}) is close to zero under nearly all of the parameter space \autoref{tab:model_var}, meaning that the photosynthetic rate of amphistomatous leaves with stomata optimally offset is nearly equal to leaves with stomata on each surface in the same position along the leaf plane.
+The coordination advantage (\autoref{eq:coordination_advantage}) is close to zero under nearly all of the parameter space (Table 1), meaning that the photosynthetic rate of amphistomatous leaves with stomata optimally offset is nearly equal to leaves with stomata on each surface in the same position along the leaf plane.
 $I_0$: PPFD incident on the leaf surface; $\varphi_\text{pal}$: Fraction of intercellular airspace (aka porosity), palisade; $T_\text{leaf}$: Leaf thickness; $U$: Interstomatal distance.}

### 2024-03-04 ms/response-to-reviewers.qmd

@@
 The terminology is unfortunately confusing because the word `uniform' is used in different ways.
-A uniform equilateral triangular grid means that the distance between stomata is uniform; a random uniform distribution means that a stomate has an equal (i.e.
-uniform) probability of occurring anywhere on the leaf surface.
+A uniform equilateral triangular grid means that the distance between stomata is uniform; a random uniform distribution means that a stomate has an equal probability (i.e.
+uniform) of occurring anywhere on the leaf surface.
 To limit confusion, we refer to the ideal pattern (equilateral triangle grid) as uniform and the null pattern (random uniform) as random.

### 2024-03-04 ms/ms.Rmd

@@
 A uniform equilateral triangular grid means that the distance between stomata is uniform; a random uniform distribution means that a stomate has an equal probability (i.e.
-uniform) of occurring anyone on the leaf surface.
-To limit confusion, we refer to the ideal pattern as uniform (equilateral triangle grid) and the null pattern as random (random uniform).
+uniform) of occurring anywhere on the leaf surface.
+To limit confusion, we refer to the ideal pattern (equilateral triangle grid) as uniform and the null pattern (random uniform) as random.
 When observed stomatal patterns are more dispersed than expected under random patterning, we refer to this as overdispersed.

### 2024-03-04 ms/ms.tex

@@
 random uniform distribution means that a stomate has an equal
-probability (i.e.~uniform) of occurring anyone on the leaf surface. To
-limit confusion, we refer to the ideal pattern as uniform (equilateral
-triangle grid) and the null pattern as random (random uniform). When
+probability (i.e.~uniform) of occurring anywhere on the leaf surface. To
+limit confusion, we refer to the ideal pattern (equilateral triangle
+grid) as uniform and the null pattern (random uniform) as random. When
 observed stomatal patterns are more dispersed than expected under random

### 2024-03-04 ms/ms.Rmd
updated funding information

@@
 acknowledgements: |
-  The authors thank two anonymous reviewers for feedback on an earlier version of this manuscript, NSF Award 1929167 to C.D.M., and NSF Award 2307341 to T.N.B.
+  The authors thank two anonymous reviewers for feedback on an earlier version of this manuscript, NSF Award 1929167 to C.D.M., and USDA NIFA Hatch project 1016439 and NSF Award 2307341 to T.N.B.
 This is publication #218 from the School of Life Sciences, University of Hawaiʻi at Mānoa.
@@
 
-The authors thank two anonymous reviewers for feedback on an earlier version of this manuscript, NSF Award 1929167 to C.D.M., and NSF Award 2307341 to T.N.B.
+The authors thank two anonymous reviewers for feedback on an earlier version of this manuscript, NSF Award 1929167 to C.D.M., and USDA NIFA Hatch project 1016439 and NSF Award 2307341 to T.N.B.
 This is publication #218 from the School of Life Sciences, University of Hawaiʻi at Mānoa.

### 2024-03-04 ms/ms.tex
updated funding information

@@
 The authors thank two anonymous reviewers for feedback on an earlier
-version of this manuscript, NSF Award 1929167 to C.D.M., and NSF Award
-2307341 to T.N.B. This is publication \#218 from the School of Life
-Sciences, University of Hawaiʻi at Mānoa.
+version of this manuscript, NSF Award 1929167 to C.D.M., and USDA NIFA
+Hatch project 1016439 and NSF Award 2307341 to T.N.B. This is
+publication \#218 from the School of Life Sciences, University of
+Hawaiʻi at Mānoa.
 

### 2024-03-05 ms/ms.Rmd
added zenodo url to ms

@@
 
-Custom scripts are available on a GitHub repository (https://github.com/cdmuir/stomata-spacing) and archived on Zenodo: https://doi.org/XXXX Raw data are deposited on Dryad: https://doi.org/10.5061/dryad.44j0zpcn6
+Custom scripts are available on a GitHub repository (https://github.com/cdmuir/stomata-spacing) and archived on Zenodo: https://doi.org/10.5281/zenodo.10775962 Raw data are deposited on Dryad: https://doi.org/10.5061/dryad.44j0zpcn6
 

### 2024-03-05 ms/ms.tex
added zenodo url to ms

@@
 (https://github.com/cdmuir/stomata-spacing) and archived on Zenodo:
-https://doi.org/XXXX Raw data are deposited on Dryad:
+https://doi.org/10.5281/zenodo.10775962 Raw data are deposited on Dryad:
 https://doi.org/10.5061/dryad.44j0zpcn6
